### 2021-03-08 doc/VerticalArrangementsSheuTaragin.tex
added additional figures. strted a new document to store figures in.

\documentclass[12pt]{article}
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{setspace}   % controls line spacing
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs

\parskip     2.0mm       % space between paragraphs

\alph{footnote}         % make title footnotes alpha-numeric

\captionsetup[figure]{labelsep=space,labelfont=bf} % remove colon from figure name

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title Page
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{The Effects of Vertical Arrangements\\in a Vertical Supply Chain\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, by the Federal Trade Commission, or by its Commissioners.  }}

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
\author{Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov. } \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission}

\date{March 2021}

\begin{document}

\pagenumbering{roman}       % Roman numerals from abstract to text
\maketitle                  % print title information
\thispagestyle{empty}       % no page number on THIS page

\begin{abstract}
\noindent
\end{abstract}

\bigbreak Keywords: bargaining models; merger simulation; vertical markets

JEL classification: L13; L40; L41; L42

\newpage                    % start a new page
\pagenumbering{arabic}      % Arabic page numbers from now on
\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.85]{../output/surplussum.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus. Each blue box depicts the effects assuming that retailers are playing a Bertrand pricing game. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:surplus}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.85]{../output/CVvertincumbBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus as the number of vertically integrrated firms present in a market change. Each blue box depicts the effects assuming that retailers are playing a Bertrand pricing game. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:CVvertincumbBW}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{../output/CVbargupBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
\label{fig:CVbargupBW}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{../output/CVbargdownBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
\label{fig:CVbargdownBW}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{../output/CVbargvertBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
\label{fig:CVbargvertBW}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{../output/CVbargbothBW.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
\label{fig:CVbargbothBW}
\end{sidewaysfigure}

\end{document}

### 2021-03-08 code/genfigures.R

@@ -531,182 +531,58 @@ pbargvert_all.bw <- pbargvert_all.bw + geom_segment(
             data=data.frame(x=6.5,y=-20,Outcome=factor( "Wholesaler" ,levels=unique(res.nests.logit$Outcome))),size=3)
 
 
-##Create plots for partial  analysis
 
 
-#filter(partdata,part %in% c("cvdiff","totalDiff")) %>% group_by(Merger,Model,part) %>%
-#dplyr::summarise(p25=quantile(diff,.25),p50=quantile(diff,.5),p75=quantile(diff,.75),p05=quantile(diff,.05),p95=quantile(diff,.95))
 
-ppartial.bw <- ggplot(filter(partdata,
-
-                             part %in% c("cvdiff" ,
-                                         #"downPSDiff",
-                                         "totalDiff"
-                                         #, "upPSDiff"
-                                         )) %>%
-                        mutate(part=factor(part,levels=c("cvdiff" ,#"downPSDiff", "upPSDiff",
-                                                         "totalDiff" ),
-                                           labels=c("Consumer",#"Retailer","Wholesaler",
-                                                    "Total"))
-                               ),
-  #aes(y=diff/mktrev.pre*100,x=part,color=Model)) +
-  aes(y=diff*100,x=part,color=Model)) +
+pbargboth_all.bw <- ggplot(filter(res.nests.logit,merger =="both" ), aes(y=Outcome_value/mktrev.pre*100,
+                                                                             #avgpricedelta/mktrev.pre*100,
+                                                                             x=factor(barg,labels=MASS::fractions(relleverage)),color=vert)) +
   #geom_boxplot(outlier.alpha = 0.1) +
   stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
-coord_cartesian(ylim=c(-100,160)) +
-  scale_y_continuous(breaks=seq(-100, 160, by=20) ) +
-  #scale_y_continuous(breaks=seq(-5, 30, by=5) ) +
+  scale_y_continuous(breaks=seq(-40,35,5))+
+  coord_cartesian(ylim=c(-40,35))+
   geom_hline(yintercept=0,linetype="dashed",color="black")+
-  theme_bw()+scale_colour_tableau('Color Blind')+ theme(
-    #axis.text.x = element_text(angle = 0, hjust = 1),
-    legend.position="bottom")+
+  geom_vline(xintercept=5,linetype="dotted")+
[…]

### 2021-03-09 doc/VerticalArrangementsSheuTaragin.tex
added linear marginal cost to incumbent vertical integration

@@
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-Each blue box (on the left in each pair) depicts the effects assuming that retailers are playing a Bertrand pricing game, and each orange box (on the right in each pair) depicts the effects assuming that retailers are playing a second score auction game.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}

### 2021-03-09 code/RunSims.R

@@ -31,7 +31,7 @@ clusterEvalQ(cl,{
              if(require(RevoUtilsMath)) setMKLthreads(1)
 })
 
-samples <- 1e3
+samples <- 1e2
 ndfirms <- 2:5
 nufirms <- ndfirms
 nvfirms <- 0:4
@@ -39,7 +39,7 @@ nestParm <- c(0)
 type=c("1st")
 merger=c("up","down", "vertical","both")
 bargparm <- seq(.1,.9,.1)
-mc=c("constant"#,"linprod", "quadprod","linquad","quadlin","lincons","conslin"#,"linfirm"
+mc=c("constant","linprod"#, "quadprod","linquad","quadlin","lincons","conslin"#,"linfirm"
      )
 relleveragePre <-  (1-bargparm)/bargparm
 

### 2021-03-09 code/genfigures.R

@@ -20,9 +20,9 @@ load(file.path(simpath,"AllSims.RData"))
 relleverage=sort(unique(res.nests$relleveragePre),decreasing = TRUE)
 
 res.nests.logit <- res.nests.long <- mutate(ungroup(res.nests.long),
-                                            Merger=factor(merger,levels=c("down","up","vertical","both"),labels=c("Downstream","Upstream","Vertical","Both")),
+                                            Merger=factor(merger,levels=c("down","up","vertical","both"),labels=c("Downstream","Upstream","Vertical","Integrated")),
                                             Cost=factor(mc,labels=c("Constant"
-                                                                    #,"Linear","Quadratic","Linear/Quadratic","Quadratic/Linear","Linear/Constant","Constant/Linear"
+                                                                    ,"Linear"#,"Quadratic","Linear/Quadratic","Quadratic/Linear","Linear/Constant","Constant/Linear"
                                                                     )
                                                         ),
                                             Cost=reorder(Cost,-Outcome_value/mktrev.pre,median,na.rm=TRUE)) %>% select(-mc)
@@ -47,7 +47,7 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 
 pvertincumb.bw <-  ggplot(res.nests.logit, aes(y=Outcome_value/mktrev.pre*100,
                                            #avgpricedelta/mktrev.pre*100,
-                                           x=vert#,color=Merger
+                                           x=vert,color=Cost
                                            )
                                            ) +
   stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
@@ -66,7 +66,7 @@ pvertincumb.bw <-  ggplot(res.nests.logit, aes(y=Outcome_value/mktrev.pre*100,
   #ylab("Avg. Downstream Price Change (%)")+
   #ylab("Share-Weighted Downstream Price Change")+
   #geom_text(data=ann_text,label="Wholesale advantage")
-  #labs(colour="Cost:")+
+  labs(colour="Cost:")+
   labs(title =  "The Distributions  of Merger Outcomes as the Number of Integrated Firms Increases",
        subtitle="Outcomes are reported as a percentage of pre-merger total expenditures.\nHorizontal mergers occur between a vertically integrated and unintegrated firm."
[…]

### 2021-03-10 code/MonteCarlo.R
full run

@@ -59,7 +59,7 @@ market<-function(nfirms.down=3, # # of downstream firms
                                 up = ids$up.firm == 1 | ids$up.firm == 2,
                                 down = ids$down.firm == 1 | ids$down.firm == 2,
                                 vertical= ids$up.firm == 1 | ids$down.firm == 1,
-                                both=ids$up.firm == 1 | ids$up.firm == 2)
+                                both=ids$up.firm == 1 | ids$up.firm == 2 | ids$up.firm == 2 | ids$up.firm == 1)
 
   if(ownerPost=="vertical"){isJoint <-  ids$up.firm == 1 & ids$down.firm == 1}
 

### 2021-03-10 code/RunSims.R

@@ -31,7 +31,7 @@ clusterEvalQ(cl,{
              if(require(RevoUtilsMath)) setMKLthreads(1)
 })
 
-samples <- 1e2
+samples <- 1e3
 ndfirms <- 2:5
 nufirms <- ndfirms
 nvfirms <- 0:4

### 2021-03-11 code/MonteCarlo.R
fixed profitability calculations and reran

@@ -61,6 +61,20 @@ market<-function(nfirms.down=3, # # of downstream firms
                                 vertical= ids$up.firm == 1 | ids$down.firm == 1,
                                 both=ids$up.firm == 1 | ids$up.firm == 2 | ids$up.firm == 2 | ids$up.firm == 1)
 
+
+
+
+  if(nfirms.vert>0){
+    isJoint <-  isParty <- switch(ownerPost,
+                                 up = isParty | ids$down.firm==2,
+                                 down = isParty | ids$up.firm==2,
+                                 vertical= isParty,
+                                 both=isParty)
+
+
+  }
+
+
   if(ownerPost=="vertical"){isJoint <-  ids$up.firm == 1 & ids$down.firm == 1}
 
   ## create nesting structures for different merger types

### 2021-03-11 code/RunSims.R

@@ -73,10 +73,14 @@ genMkts <- function(x,y,z,t,m,n, b,c, large=TRUE, bargpreset="none"){
     summary=thismkt,
     res=thismkt))}
 
+isVert <- length(thismkt$vertical)>0
+
+
 shareOutDown <- thismkt$down$shareOut
 M <- thismkt$down$M
 outMargin <- thismkt$down$outMargin
 #thismkt <- calcSlopes(thismkt,constrain="global");
+
 
 mcParmUp <- thismkt$up$mcParm
 mcParmDown <- thismkt$down$mcParm
@@ -116,15 +120,20 @@ thisres <- with(thissum,data.frame(
   upPSPost=sum(upPSPost,na.rm=TRUE),
   downPSPre=sum(downPSPre, na.rm=TRUE),
   downPSPost=sum(downPSPost,na.rm=TRUE),
-  isProfitable.up=sum( upPSPost[idUp %in% 1:2],  - upPSPre[ idUp %in% 1:2],na.rm=TRUE) >0,
-  isProfitable.down=sum( downPSPost[idDown %in% 1:2],  - downPSPre[ idDown %in% 1:2],na.rm=TRUE) >0,
+  isProfitable.up=ifelse(isVert,
+                         sum( upPSPost[idUp %in% 1:2], downPSPost[idDown %in% 2],  - upPSPre[ idUp %in% 1:2],- downPSPre[idDown %in% 2],na.rm=TRUE) >0,
+                         sum( upPSPost[idUp %in% 1:2],  - upPSPre[ idUp %in% 1:2],na.rm=TRUE) >0),
+  isProfitable.down=ifelse(isVert,
+                           sum( upPSPost[idUp %in% 2],downPSPost[idDown %in% 1:2],  -upPSPre[idUp %in% 2],- downPSPre[ idDown %in% 1:2],na.rm=TRUE) >0,
+                           sum( downPSPost[idDown %in% 1:2],  - downPSPre[ idDown %in% 1:2],na.rm=TRUE) >0),
   isProfitable.vert=sum( downPSPost[idDown==1], upPSPost[idUp==1],  - downPSPre[idDown==1] , - upPSPre[idUp==1],na.rm=TRUE) >0,
+  isProfitable.both=sum( downPSPost[idDown%in% 1:2], upPSPost[idUp%in% 1:2],  - downPSPre[idDown%in% 1:2] , - upPSPre[idUp%in% 1:2],na.rm=TRUE) >0,
[…]

### 2021-04-11 doc/VerticalArrangementsSheuTaragin.tex
added explanation for figures.

@@
 
-\date{March 2021}
+\date{\today}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Data Generating Process}
+This section provides an overview of our methodology, with additional details appearing in the Appendix.
+We simulate markets by randomly sampling shares from a Dirichlet distribution for 2, 3, 4, or 5 retailers or wholesalers, respectively.\footnote{We parametrize the Dirichlet distribution so it is equivalent to a uniform distribution.
+}  We also assume that in the pre-merger state, anywhere from 0 to 4 retailers are vertically integrated with a single wholesaler.
+We assume that vertically integrated wholesalers supply inputs to non-integrated retailers and that vertically integrated retailers purchase inputs from non-integrated wholesalers.
+The price coefficient $\alpha$ is calibrated by assuming that in the pre-merger world, there is a vertically integrated outside option available to all customers.
+The other goods are differenced relative to this option, which maintains the outside good normalization.
+The market size is set to 1.
+
+We specify values for the bargaining parameter ranging from from 0.1 (wholesalers have the advantage) to 0.9 (retailers have the advantage).
+To better understand the relative bargaining strength of these parameter values, we report our results in terms of $(1-\lambda)/\lambda$, which range from $9$ (wholesaler power is nine times greater than retailer power) to $1/9$ (retailer power is nine times greater than wholesaler power).
+The bargaining parameter is identical for all of the retailers in each simulation, unless noted otherwise.
+
+For each combination of number of retailers, number of wholesalers, and bargaining parameter, we draw 1,000 different sets of market primitives.
+This results in XX million merger simulations.
+We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding XX million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+} All XX million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+
+When simulating a horizontal merger, we assign the products produced by the two largest firms in the market to a single entity post-merger.
+Similarly, when simulating a vertical merger, we assign the products produced by the largest wholesaler and the largest retailer to a single entity post-merger.
+This assignment is purposefully skewed towards mergers that are more likely to have competitive effects and to come under agency review.
+
+Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is almost \$5, and the median average retail pre-merger price is \$12.
+Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.
+Pre-merger HHIs range between 2,029 at the $25^{th}$ percentile to 5,143 at the $75^{th}$, with a median of 2,828.
+HHIs for horizontal downstream mergers increase by 1,658 points at the median, resulting in a median post-merger HHI equal to 4,270.
+HHIs for upstream mergers increase by 1,179 points at the median, resulting in a median post-merger HHI equal to 4,243.
+HHIs for vertical mergers increase by 953 points at the median, resulting in a median post-merger HHI equal to 5,174.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
+}  Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with HHIs over 2,500.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Results Overview}
+Our overall results are depicted in Figure \ref{fig:surplussum}, which  is divided into four panels, each showing how the distribution of surplus changes for a particular set of agents (consumers, retailers, or wholesalers), as well as the net effect on the market as a whole.
+Surplus is presented as a percentage change relative to total pre-merger expenditure in the downstream market.
+
+Each panel contains four pairs of box and whisker plots, with each pair corresponding to a different type of merger.
+The blue box and whisker plots (on the left in each pair) depict outcomes assuming that marginal costs are constant, and the orange box and whisker plots (on the right in each pair) show outcomes assuming that marginal costs are linear.
+The whiskers display the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median.
+Note that negative outcome values imply agent harm, and positive values imply agent benefits.
+
+We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.
+The median change is negative, indicating harm, across all four types of mergers for both cost specifications.
+However, the distributions and magnitudes differ.
+In Bertrand markets, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
+Median harm from downstream  mergers is 4.5\% of pre-merger total expenditures, 1.3 times the magnitude of that from upstream mergers, and 3.1 times the magnitude of that from vertical mergers.
+By contrast, under second score, consumer harm in upstream mergers stochastically dominates consumer harm in downstream mergers, which in turn dominates harm in vertical mergers.
+The median harm from upstream mergers is 3.5\% of pre-merger total expenditures, 4.1 times that of downstream mergers, and 15 times that of vertical mergers.
+
+Four types of mergers are almost always harmful: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.
+That upstream horizontal mergers are net harmful is somewhat surprising, as it indicates that any EDM from the unintegrated upstream firm merging with its integrated rival is outweighed by the recapture effect.
+Likewise, the fact that vertical mergers with linear marginal costs are almost never beneficial when marginal costs are linear is surprising, as it suggests that the incentive to raise rivals' cost dominates the benefit from EDM.
+%especially given that about 40\% of vertical mergers with constant marginal costs are beneficial.
+\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}
+
+%However, the fact that downstream mergers under Bertrand are almost never beneficial and frequently quite harmful is somewhat surprising, given the potential for countervailing bargaining leverage to lower wholesale prices.
+
+Harm is less prevalent for vertical mergers and downstream mergers with constant marginal costs.
+Vertical mergers in Bertrand markets benefit consumers in about 40\% of all simulations , and downstream horizontal mergers benefit consumers in XX\% of all simulations.
+The median consumer benefit among positive outcomes from vertical mergers equals 4.2\% of pre-merger total expenditures (comparable to the median consumer harm from downstream mergers under Bertrand), and the median consumer benefit from downstream mergers in auction markets equals 1.7\% of pre-merger revenues (less than two-fifths of the median consumer harm from downstream mergers under Bertrand).
+
+Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers benefit retailers, upstream mergers frequently -- though not universally-- harm retailers.
+We also find that for all merger types, the median retailer surplus under constant marginal cost first-order stochastically dominate the retail surplus distribution under linear costs, indicating that the benefits from EDM are typically greater under constant marginal costs than linear marginal costs.
+
+% Finally, it is interesting that the harm experienced in the latter instances is quite small in relative terms: median retailer harm from upstream mergers is about one-quarter of the median consumer harm from an upstream merger.
+This, along with the narrow inter-quartile range of retailer harm (less than 1\% of pre-merger revenues) indicates that retailers are passing on the bulk of the wholesale price increase to consumers.\footnote{This phenomena may also be seen by noting that the distribution of wholesaler gain from upstream mergers is similar in magnitude to the distribution of consumer losses from upstream mergers, though less skewed.
+}
+
+As for wholesaler surplus, which appears in the third panel, the effects seen there are the mirror image of those for retailers, with net harm in almost all downstream mergers and the majority of vertical mergers, and net benefits in all upstream mergers.
+Although one would expect wholesalers to be harmed by downstream mergers, the amount of harm in Bertrand markets is about one-third  of that in second score markets.
+This difference partially explains the difference in the magnitude of consumer harm between downstream mergers with Bertrand versus auctions: namely, retailers' greater ability to extract wholesaler surplus in second score auction markets translates into more savings passed thru to consumers, which offsets the harm from less retail competition.
+
+%Whereas wholesalers benefit from vertical mergers in about 17\% of simulations, retailers benefit from vertical mergers in about 99\% of simulations.
+Moreover, the median benefit to retailers is 7.5\% of pre-merger expenditures, comparable to the median loss to wholesalers of -9\%.
+Together, these observations indicate that vertical mergers are largely a rent transfer from wholesalers to retailers and consumers, with retailers keeping most of the gain.
+
+In terms of total welfare, with the exception of second score downstream mergers, our simulated mergers are typically net harmful, with second score vertical mergers yielding the highest median total harm (-2.5\% of pre-merger expenditures), followed by Bertrand downstream mergers (-2.2\%), Bertrand vertical mergers (-1.5\%), and then upstream mergers.
+Downstream mergers under auctions are slightly beneficial at the median, with a gain of 0.5\% of pre-merger expenditures.
+About 16\% of Bertrand vertical mergers and 7\%  of second score vertical mergers are net beneficial.
+
 \begin{sidewaysfigure}
@@
 }
-\label{fig:surplus}
+\label{fig:surplussum}
 \end{sidewaysfigure}

### 2021-04-12 doc/VerticalArrangementsSheuTaragin.tex
wrote up table 1 and figure 1

@@
 \usepackage{hyperref}   % displays URLs
+%packages for kable tables
+ \usepackage{booktabs}
+ \usepackage{longtable}
+ \usepackage{array}
+ \usepackage{multirow}
+ \usepackage{wrapfig}
+ \usepackage{float}
+ \usepackage{colortbl}
+ \usepackage{pdflscape}
+ \usepackage{tabu}
+ \usepackage{threeparttable}
+ \usepackage{threeparttablex}
+ \usepackage[normalem]{ulem}
+ \usepackage{makecell}
+ \usepackage{xcolor}
 
@@
 For each combination of number of retailers, number of wholesalers, and bargaining parameter, we draw 1,000 different sets of market primitives.
-This results in XX million merger simulations.
+This results in 2.2 million merger simulations.
 We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding XX million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} All XX million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+} All 2.2 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
 
@@
 
-Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is almost \$5, and the median average retail pre-merger price is \$12.
+Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is almost \$5, and the median average retail pre-merger price is \$13.
 Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.
-Pre-merger HHIs range between 2,029 at the $25^{th}$ percentile to 5,143 at the $75^{th}$, with a median of 2,828.
-HHIs for horizontal downstream mergers increase by 1,658 points at the median, resulting in a median post-merger HHI equal to 4,270.
-HHIs for upstream mergers increase by 1,179 points at the median, resulting in a median post-merger HHI equal to 4,243.
-HHIs for vertical mergers increase by 953 points at the median, resulting in a median post-merger HHI equal to 5,174.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
-}  Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with HHIs over 2,500.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
+Pre-merger HHIs range between 2,629 at the $25^{th}$ percentile to 4,287 at the $75^{th}$, with a median of 3,688.
+HHIs for horizontal downstream mergers increase by 2,027 points at the median, resulting in a median post-merger HHI equal to 4,963.
+HHIs for upstream mergers increase by 2,040 points at the median, resulting in a median post-merger HHI equal to 5,207.
+HHIs for vertical mergers increase by 1,292 points at the median, resulting in a median post-merger HHI equal to 6,016.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
+}  HHIs for integrated mergers increase by 2,800 points at the median, resulting in a median post-merger HHI equal to 7,272.
+Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with HHIs over 2,500.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
 }
@@
 In Bertrand markets, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
-Median harm from downstream  mergers is 4.5\% of pre-merger total expenditures, 1.3 times the magnitude of that from upstream mergers, and 3.1 times the magnitude of that from vertical mergers.
-By contrast, under second score, consumer harm in upstream mergers stochastically dominates consumer harm in downstream mergers, which in turn dominates harm in vertical mergers.
-The median harm from upstream mergers is 3.5\% of pre-merger total expenditures, 4.1 times that of downstream mergers, and 15 times that of vertical mergers.
+Median harm from integrated mergers is 20\% of pre-merger total expenditures, more than 2 times the magnitude of that from downstream mergers, 3.1 times the magnitude of that from upstream mergers, and 5.4 times the magnitude of vertical mergers.
+Consumer harm under linear marginal costs first-order stochastically dominates consumer harm under constant marginal costs.
+The median harm from markets with linear marginal costs is about 6.8\% of pre-merger total expenditures, 1.2 times the magnitude of markets with constant marginal costs.
 
-Four types of mergers are almost always harmful: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.
-That upstream horizontal mergers are net harmful is somewhat surprising, as it indicates that any EDM from the unintegrated upstream firm merging with its integrated rival is outweighed by the recapture effect.
+Four types of mergers are unlikely to benefit consumers: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.
+That upstream horizontal mergers are largely net harmful is somewhat surprising, as it indicates that any EDM from the unintegrated upstream firm merging with its integrated rival is outweighed by the recapture effect.
+Similar logic applies to integrated mergers.
 Likewise, the fact that vertical mergers with linear marginal costs are almost never beneficial when marginal costs are linear is surprising, as it suggests that the incentive to raise rivals' cost dominates the benefit from EDM.
@@
 
-Harm is less prevalent for vertical mergers and downstream mergers with constant marginal costs.
-Vertical mergers in Bertrand markets benefit consumers in about 40\% of all simulations , and downstream horizontal mergers benefit consumers in XX\% of all simulations.
-The median consumer benefit among positive outcomes from vertical mergers equals 4.2\% of pre-merger total expenditures (comparable to the median consumer harm from downstream mergers under Bertrand), and the median consumer benefit from downstream mergers in auction markets equals 1.7\% of pre-merger revenues (less than two-fifths of the median consumer harm from downstream mergers under Bertrand).
+Harm is less prevalent for vertical mergers, downstream mergers and integrated mergers when firms have constant marginal costs.
+Vertical mergers when firms have constant marginal costs benefit consumers in about 33\% of all simulations , downstream horizontal mergers benefit consumers in about 13\% of all simulations, and integrated mergers benefit consumers in about 4.9\% of all simulations.
+The median consumer benefit among positive outcomes equals 5.9\% of pre-merger total expenditures from vertical mergers, 9.5\% of pre-merger total expenditures from downstream mergers, and 4.9\% of integrated mergers.
 
-Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers benefit retailers, upstream mergers frequently -- though not universally-- harm retailers.
-We also find that for all merger types, the median retailer surplus under constant marginal cost first-order stochastically dominate the retail surplus distribution under linear costs, indicating that the benefits from EDM are typically greater under constant marginal costs than linear marginal costs.
+Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 79\% of all simulation.
+Moreover, there is a partial rank-ordering across mergers: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
+We also find that  for vertical and integrated mergers, retailer surplus under constant marginal cost first-order stochastically dominates the retail surplus distribution under linear costs, again suggesting that the incentive to raise rivals' costs dominates the benefits from EDM.
+By contrast, for downstream and upstream mergers, there is no clear rank-ordering between markets with constant marginal costs and markets with linear costs.
 
@@
 
-As for wholesaler surplus, which appears in the third panel, the effects seen there are the mirror image of those for retailers, with net harm in almost all downstream mergers and the majority of vertical mergers, and net benefits in all upstream mergers.
-Although one would expect wholesalers to be harmed by downstream mergers, the amount of harm in Bertrand markets is about one-third  of that in second score markets.
-This difference partially explains the difference in the magnitude of consumer harm between downstream mergers with Bertrand versus auctions: namely, retailers' greater ability to extract wholesaler surplus in second score auction markets translates into more savings passed thru to consumers, which offsets the harm from less retail competition.
+As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the mirror image of those for retailers: upstream mergers increase wholesaler surplus in about 97\% of all simulated mergers, while the other merger types decrease wholesaler surplus from anywhere between 81\% (downstream mergers) and 68\% (integrated mergers) of all simulated mergers.
+\textbf{comparing these results to Sheu/Taragin, the effects are less stark (ie less of a transfer.} Again, we see a a partial rank-ordering across mergers: the retailer surplus distribution from upstream mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
 
@@
 
-In terms of total welfare, with the exception of second score downstream mergers, our simulated mergers are typically net harmful, with second score vertical mergers yielding the highest median total harm (-2.5\% of pre-merger expenditures), followed by Bertrand downstream mergers (-2.2\%), Bertrand vertical mergers (-1.5\%), and then upstream mergers.
-Downstream mergers under auctions are slightly beneficial at the median, with a gain of 0.5\% of pre-merger expenditures.
-About 16\% of Bertrand vertical mergers and 7\%  of second score vertical mergers are net beneficial.
+In terms of total welfare, with the exception of the approximately 16\% of vertical mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful.
+Moreover there is a complete rank ordering of mergers, with total harm from integrated mergers first-order stochastically dominating total harm from downstream mergers, which dominates total harm from upstream mergers, which dominates totals harm from vertical mergers.
+Finally, there is little difference in total harm under constant and linear costs, though total harm under linear costs continues to first-order stochastically dominate total harm under constant marginal costs for vertical and integrated mergers.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables and Figures
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\include{sumtable}
+\label{tab:simsum}
+\caption{Summary Statistics}
+\end{table}
 
@@
 \includegraphics[scale=0.85]{../output/CVvertincumbBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus as the number of vertically integrrated firms present in a market change.
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus as the number of vertically integrated firms present in a market change.
 Each blue box depicts the effects assuming that retailers are playing a Bertrand pricing game.

### 2021-04-14 doc/VerticalArrangementsSheuTaragin.tex
added some figures cutting consumer suplus by number of wholsaler and retailers. Fixed a bug in how the number of integrated firms was counted. Fixed a output bug that was producing multiple duplciated rows for each monte carlo simulation.

@@
 For each combination of number of retailers, number of wholesalers, and bargaining parameter, we draw 1,000 different sets of market primitives.
-This results in 2.2 million merger simulations.
+This results in 2.4 million merger simulations.
 We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding XX million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} All 2.2 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+} All 2.4 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
 
@@
 
-Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is almost \$5, and the median average retail pre-merger price is \$13.
+Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.49, and the median average retail pre-merger price is \$13.23.
 Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.
-Pre-merger HHIs range between 2,629 at the $25^{th}$ percentile to 4,287 at the $75^{th}$, with a median of 3,688.
-HHIs for horizontal downstream mergers increase by 2,027 points at the median, resulting in a median post-merger HHI equal to 4,963.
-HHIs for upstream mergers increase by 2,040 points at the median, resulting in a median post-merger HHI equal to 5,207.
-HHIs for vertical mergers increase by 1,292 points at the median, resulting in a median post-merger HHI equal to 6,016.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
-}  HHIs for integrated mergers increase by 2,800 points at the median, resulting in a median post-merger HHI equal to 7,272.
+Pre-merger HHIs range between 2,820 at the $25^{th}$ percentile to 5,013 at the $75^{th}$, with a median of 3,689.
+HHIs for horizontal downstream mergers increase by 2,027 points at the median, resulting in a median post-merger HHI equal to 4,962.
+HHIs for upstream mergers increase by 2,041 points at the median, resulting in a median post-merger HHI equal to 5,208.
+HHIs for vertical mergers increase by 1,292 points at the median, resulting in a median post-merger HHI equal to 6,013.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
+}  HHIs for integrated mergers increase by 2,774 points at the median, resulting in a median post-merger HHI equal to 7,324.
 Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with HHIs over 2,500.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.

### 2021-04-14 code/MonteCarlo.R

@@ -59,7 +59,8 @@ market<-function(nfirms.down=3, # # of downstream firms
                                 up = ids$up.firm == 1 | ids$up.firm == 2,
                                 down = ids$down.firm == 1 | ids$down.firm == 2,
                                 vertical= ids$up.firm == 1 | ids$down.firm == 1,
-                                both=ids$up.firm == 1 | ids$up.firm == 2 | ids$up.firm == 2 | ids$up.firm == 1)
+                                both=ids$up.firm == 1 | ids$up.firm == 2 #| ids$down.firm == 2 | ids$down.firm == 1
+                                )
 
 
 
@@ -121,11 +122,15 @@ market<-function(nfirms.down=3, # # of downstream firms
 
   ## when OwnerPost equals "vertical"
   ## simulate a vertical merger between U1 and D1
-  if(ownerPost  %in%  c("vertical") || nfirms.vert>0){
 
-    if (nfirms.vert>0) vertFirms <-2:(nfirms.vert + 1)
-    else if (ownerPost=="both") vertFirms <- c(1,vertFirms)
-    else{vertFirms=NULL}
+  vertFirms=NULL
+
+  if(ownerPost  %in%  c("vertical","both") || nfirms.vert>0){
+
+
+    if (nfirms.vert>0){vertFirms <-2:(nfirms.vert + 1)}
+    if (ownerPost=="both") {vertFirms <- vertFirms+1;vertFirms <- c(1:2,vertFirms)}
+
 
     #if(!(ownerPost  %in%  c("vertical")) && min(nfirms.down,nfirms.up) >2 ) vertFirms <- vertFirms+1
 
@@ -245,8 +250,9 @@ market<-function(nfirms.down=3, # # of downstream firms
     ## integrated wholesaler bargains with other retailers
     ownerVertPost.down[vertrowsUp, ids$down.firm == 1] <- -1
 
-    vertFirms <- c(1,vertFirms)
+
     }
+
     if(ownerPost  %in%  c("up","both")){
 
     vertrowsDown <- ids$up.firm != 1  & ids$down.firm == 2
@@ -263,6 +269,8 @@ market<-function(nfirms.down=3, # # of downstream firms
 
     ## integrated wholesaler bargains with other retailers
     ownerVertPost.down[vertrowsUp, ids$down.firm == 2] <- -1
+
+
     }
 
     if(ownerPost  %in%  c("down","both")){
[…]

### 2021-04-14 code/RunSims.R

@@ -151,7 +151,7 @@ thisres$hhipost.vert <-thisres$hhidelta.up <- thisres$hhidelta.down <- thisres$h
 return(list(
   #mkt=thismkt,
   summary=thissum,
-  res=thisres))}
+  res=unique(thisres)))}
 
 repMkts <- function(x,y,z,t,m,n,b,c,large=TRUE){replicate(samples, genMkts(x,y,z,t,m,n,b,c,large),simplify = "list")}
 
@@ -196,7 +196,10 @@ res.nests <-  lapply(mkts.nests, function(x){
 
 
 res.nests <- data.table::rbindlist(res.nests)
-res.nests <- mutate(res.nests, up=factor(up), down=factor(down), vert=factor(vert),nestParm=factor(nestParm),
+res.nests <- mutate(res.nests, up=factor(up), down=factor(down),
+                    #vert=ifelse(merger=="vertical",vert+1,vert),
+                    vert=ifelse(merger=="both",vert+2,vert),
+                    vert=factor(vert),nestParm=factor(nestParm),
                     relleveragePre = (1- barg)/barg,barg=factor(barg),
                     type= ifelse(type == "1st", "Bertrand","2nd"),type=factor(type, levels=c("Bertrand","2nd")),
                     Retailers = down)
@@ -216,11 +219,8 @@ res.nests <- ungroup(res.nests) %>% group_by(merger,up,down,vert,type,nestParm,m
          isProfitable=ifelse(merger=="up", isProfitable.up,
                              ifelse(merger=="down",isProfitable.down,
                                     ifelse(merger=="vert",isProfitable.vert,isProfitable.both))),
-         relmarginPreCut=cut(relmarginPre,quantile(relmarginPre, probs=seq(0,1,.1),na.rm=TRUE), dig.lab=1, include.lowest = TRUE),
-         cv001=quantile(cv,.05,na.rm=TRUE),
-         cv999=quantile(cv,.95,na.rm=TRUE),
-         cvtrun =ifelse(cv< cv001,cv001,cv))
-res.nests$cvtrun <- with(res.nests, ifelse(cvtrun>cv999,cv999,cvtrun))
+         relmarginPreCut=cut(relmarginPre,quantile(relmarginPre, probs=seq(0,1,.1),na.rm=TRUE), dig.lab=1, include.lowest = TRUE))
+
 
 ##eliminate markets where the hhipre is 0, isMarket==0
[…]

### 2021-04-14 code/genfigures.R

@@ -45,7 +45,7 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 # ## box and whisker results for Incumbent
 #
 
-pvertincumb.bw <-  ggplot(res.nests.logit, aes(y=Outcome_value/mktrev.pre*100,
+pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(vert!= "5" & vert !="6"), aes(y=Outcome_value/mktrev.pre*100,
                                            #avgpricedelta/mktrev.pre*100,
                                            x=vert,color=Cost
                                            )
@@ -124,7 +124,7 @@ pdown_vertincumb.bw <-  ggplot(filter(res.nests.logit,Merger=="Downstream" & Out
   #ylab("Share-Weighted Downstream Price Change")+
   #geom_text(data=ann_text,label="Wholesale advantage")
   labs(colour="Cost:")+
-  labs(title =  "The Distribution of Consumer Harm fromn Downstream Mergers as the Number of Pre-merger Integrated Firms Increases",
+  labs(title =  "The Distribution of Consumer Harm from Downstream Mergers as the Number of Pre-merger Integrated Firms Increases",
        subtitle="Outcomes are reported as a percentage of pre-merger total expenditures."
        #subtitle = "1st and 2nd score auctions yields radically different predictions for downstream mergers,\n but similar predictions for upstream mergers",
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
@@ -149,7 +149,7 @@ pvert_vertincumb.bw <-  ggplot(filter(res.nests.logit,Merger=="Vertical" & Outco
   #theme_tufte(ticks=FALSE) +
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = label_both)+
-  facet_grid(Retailers~Wholesalers,scales="free_x",labeller = label_both)+
+  facet_grid(Retailers~.,scales="free_x",labeller = label_both)+
   xlab("# Pre-merger Vertically Integrated Firms")+
   ylab("Outcome (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
@@ -405,7 +405,7 @@ pfirmsvert_retailers.bw <- ggplot(filter(ungroup(res.nests.logit),merger =="vert
[…]

### 2021-04-16 doc/VerticalArrangementsSheuTaragin.tex
added bibliography. fixed explanation of figure 1 and table 1. started explaining a simplified version of figure 2.

@@
 For each combination of number of retailers, number of wholesalers, and bargaining parameter, we draw 1,000 different sets of market primitives.
-This results in 2.4 million merger simulations.
-We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding XX million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} All 2.4 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+This results in 3.85 million merger simulations.
+We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding 2.69 million markets.\textbf{discrepancy between res.nests dimension and ALL Counts in Table 1}\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+} All 2.69 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
 
@@
 
-Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.49, and the median average retail pre-merger price is \$13.23.
+Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.5, and the median average retail pre-merger price is \$13.
 Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.
 Pre-merger HHIs range between 2,820 at the $25^{th}$ percentile to 5,013 at the $75^{th}$, with a median of 3,689.
-HHIs for horizontal downstream mergers increase by 2,027 points at the median, resulting in a median post-merger HHI equal to 4,962.
-HHIs for upstream mergers increase by 2,041 points at the median, resulting in a median post-merger HHI equal to 5,208.
-HHIs for vertical mergers increase by 1,292 points at the median, resulting in a median post-merger HHI equal to 6,013.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
-}  HHIs for integrated mergers increase by 2,774 points at the median, resulting in a median post-merger HHI equal to 7,324.
-Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with HHIs over 2,500.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
+HHIs for horizontal downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.
+HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238.
+HHIs for vertical mergers increase by 1,055 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
+}  HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280.
+Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with post-merger HHIs over 2,500 points and HHI Changes greater than 200 points.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
 }
@@
 However, the distributions and magnitudes differ.
-In Bertrand markets, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
-Median harm from integrated mergers is 20\% of pre-merger total expenditures, more than 2 times the magnitude of that from downstream mergers, 3.1 times the magnitude of that from upstream mergers, and 5.4 times the magnitude of vertical mergers.
-Consumer harm under linear marginal costs first-order stochastically dominates consumer harm under constant marginal costs.
-The median harm from markets with linear marginal costs is about 6.8\% of pre-merger total expenditures, 1.2 times the magnitude of markets with constant marginal costs.
+In particular, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
+Median consumer harm from integrated mergers is about 16\% of pre-merger total expenditures, 1.37 times the magnitude of that from downstream mergers, 2.4 times the magnitude of that from upstream mergers, and more than 3.5 times the magnitude of vertical mergers.
 
-Four types of mergers are unlikely to benefit consumers: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.
+%Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
+Mergers with constant-cost firms benefit consumers in about 15\% of all mergers, while mergers with linear-cost firms benefit consumers in only 2.3\% of mergers.
+Indeed, consumer harm under linear marginal costs first-order stochastically dominates consumer harm under constant marginal costs.
+The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.
+
+Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
+Vertical mergers when firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when firms have linear costs.
+Upstream mergers with constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs.
+Downstream mergers with constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when firms have linear costs.
+Finally, integrated  mergers benefit consumers in about 10\% of all simulations when firms have constant costs, but less than 1\% of simulations when firms have linear costs.
+
+%mergers are unlikely to benefit consumers: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.
 That upstream horizontal mergers are largely net harmful is somewhat surprising, as it indicates that any EDM from the unintegrated upstream firm merging with its integrated rival is outweighed by the recapture effect.
@@
 
-Harm is less prevalent for vertical mergers, downstream mergers and integrated mergers when firms have constant marginal costs.
-Vertical mergers when firms have constant marginal costs benefit consumers in about 33\% of all simulations , downstream horizontal mergers benefit consumers in about 13\% of all simulations, and integrated mergers benefit consumers in about 4.9\% of all simulations.
-The median consumer benefit among positive outcomes equals 5.9\% of pre-merger total expenditures from vertical mergers, 9.5\% of pre-merger total expenditures from downstream mergers, and 4.9\% of integrated mergers.
-
-Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 79\% of all simulation.
+Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 70\% of all simulations.
 Moreover, there is a partial rank-ordering across mergers: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
-We also find that  for vertical and integrated mergers, retailer surplus under constant marginal cost first-order stochastically dominates the retail surplus distribution under linear costs, again suggesting that the incentive to raise rivals' costs dominates the benefits from EDM.
-By contrast, for downstream and upstream mergers, there is no clear rank-ordering between markets with constant marginal costs and markets with linear costs.
+We also find that  for upstream, vertical, and integrated mergers, retailer surplus under constant marginal cost first-order stochastically dominates the retail surplus distribution under linear costs, again suggesting that the incentive to raise rivals' costs dominates the benefits from EDM.
+By contrast, for downstream mergers, there is no clear rank-ordering between markets with constant marginal costs and markets with linear costs.
 
@@
 
-As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the mirror image of those for retailers: upstream mergers increase wholesaler surplus in about 97\% of all simulated mergers, while the other merger types decrease wholesaler surplus from anywhere between 81\% (downstream mergers) and 68\% (integrated mergers) of all simulated mergers.
-\textbf{comparing these results to Sheu/Taragin, the effects are less stark (ie less of a transfer.} Again, we see a a partial rank-ordering across mergers: the retailer surplus distribution from upstream mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
+As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 82\% of all upstream simulated mergers, 28\% of simulated vertical mergers, 20\% of simulated downstream mergers, and 17\% of simulated integrated mergers.
+%\textbf{comparing these results to \citeN{ST2017}, the effects are less stark (i.e.
+less of a transfer.}
+However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.
 
@@
 
-In terms of total welfare, with the exception of the approximately 16\% of vertical mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful.
+In terms of total welfare, with the exception of the approximately 10\% of vertical mergers and 8.5\% of integrated mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful.
 Moreover there is a complete rank ordering of mergers, with total harm from integrated mergers first-order stochastically dominating total harm from downstream mergers, which dominates total harm from upstream mergers, which dominates totals harm from vertical mergers.
-Finally, there is little difference in total harm under constant and linear costs, though total harm under linear costs continues to first-order stochastically dominate total harm under constant marginal costs for vertical and integrated mergers.
+Finally, for each merger type, total harm under linear costs first-order stochastically dominates total harm under constant marginal costs, with the greatest differences occurring for vertical and integrated mergers.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Vertically Integrated Incumbent Firms}
+Here, we investigate how the presence of one or more incumbent vertically integrated firms affects merger outcomes when firms either have constant or linear marginal costs.
+In particular, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
+We also examine merger outcomes for these integrated mergers as the number of integrated non-merging firms increases.
+
+Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top row) and total harm (bottom row) as the number of incumbent integrated firms increases from 0 to 6 firms.
+The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2017}.
+For downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
+By contrast, for vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+Likewise, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, and all rivals are unintegrated.
+The blue box and whisker plots (left) assume that firms face constant marginal costs, while the orange box and whisker plots (right) assume that firms face linear marginal costs.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Bargaining Power}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Bibliography
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\bibliographystyle{chicago}
+\bibliography{master-bib}
+\clearpage
 
@@
 \centering
-\includegraphics[scale=0.85]{../output/surplussum.png}
+\includegraphics[scale=0.9]{../output/surplussum.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus.
-Each blue box depicts the effects assuming that retailers are playing a Bertrand pricing game.
+Each blue box depicts the effects assuming that firms face constant marginal costs, while each orange box depicts the effects assuming that firms face linear marginal costs.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
@@
 \centering
-\includegraphics[scale=0.85]{../output/CVvertincumbBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus as the number of vertically integrated firms present in a market change.
-Each blue box depicts the effects assuming that retailers are playing a Bertrand pricing game.
+\includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer and total surplus as the number of vertically integrated firms present in a market change.
+Each blue box depicts the effects assuming that firms face constant marginal costs, while each orange box depicts the effects assuming that firms face linear marginal costs.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.

### 2021-04-16 code/RunSims.R

@@ -270,7 +270,7 @@ sumtable <- data.frame(res.nests)
 
 
 sumtable <- select(sumtable,merger,down, up,vert,barg,nestParm, hhipre, hhipost, hhidelta,  avgpricepre.up, avgpricepre.down, avgpricedelta,
-                    mktElast) %>% distinct()  %>% mutate(
+                    mktElast)  %>% mutate(
                                                                                          #mktCnt=length(up),
                                                                                          up = as.numeric(as.character(up)),
                                                                                          down=as.numeric(as.character(down)),
@@ -321,7 +321,7 @@ sumtable <- spread(sumtable, quant, val) %>% select(-Max,Max) %>%
   rename(Variable=variable,Merger=merger,`50th`=p50,`25th`=p25,`75th`=p75)
 
 sink("./doc/sumtable.tex")
-print(kable(sumtable,format="latex",digits=0) %>% collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle"))
+print(kable(sumtable,format="latex",digits=0) %>% collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle"))
 sink()
 
 #witholdfreq <- with(res.barg[merger=="vertical",],table(type,nsDown,useNA = "always"))

### 2021-04-16 code/genfigures.R

@@ -32,6 +32,8 @@ res.nests.logit <- res.nests.long <- mutate(ungroup(res.nests.long),
 #partdata <- filter(partdata,up!="1" & down !="1" )%>%
 #  mutate_if(is.factor,droplevels)
 
+seq_palette <- RColorBrewer::brewer.pal(name="YlGnBu",n=8)[c(4,6,8)]
+
 boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 
   r <- quantile(x, probs,na.rm=TRUE)
@@ -45,7 +47,7 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 # ## box and whisker results for Incumbent
 #
 
-pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(vert!= "5" & vert !="6"), aes(y=Outcome_value/mktrev.pre*100,
+pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","Total")), aes(y=Outcome_value/mktrev.pre*100,
                                            #avgpricedelta/mktrev.pre*100,
                                            x=vert,color=Cost
                                            )
@@ -60,7 +62,7 @@ pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(vert!= "5" & vert !="6"), a
   #theme_tufte(ticks=FALSE) +
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
-  facet_grid(Merger~Outcome,scales="free",labeller = label_context)+
+  facet_grid(Outcome~Merger,scales="free",labeller = label_context)+
   xlab("# Pre-merger Vertically Integrated Firms")+
   ylab("Outcome (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
@@ -405,7 +407,7 @@ pfirmsvert_retailers.bw <- ggplot(filter(ungroup(res.nests.logit),merger =="vert
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
   )
 
-pbargup_all.bw <- ggplot(filter(res.nests.logit,merger =="up" & vert != "3" & vert !="4"), aes(y=Outcome_value/mktrev.pre*100,
[…]

### 2021-04-19 doc/VerticalArrangementsSheuTaragin.tex
added text on bargaining power. Tweaked figures

@@
 \subsection*{Vertically Integrated Incumbent Firms}
-Here, we investigate how the presence of one or more incumbent vertically integrated firms affects merger outcomes when firms either have constant or linear marginal costs.
-In particular, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
+Here, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
 We also examine merger outcomes for these integrated mergers as the number of integrated non-merging firms increases.
 
-Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top row) and total harm (bottom row) as the number of incumbent integrated firms increases from 0 to 6 firms.
+Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms.
 The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2017}.
@@
 Likewise, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, and all rivals are unintegrated.
-The blue box and whisker plots (left) assume that firms face constant marginal costs, while the orange box and whisker plots (right) assume that firms face linear marginal costs.
+
+Starting with downstream mergers (leftmost column), absent incumbent vertical integration (0 integrated firms), not only are downstream mergers rarely beneficial, but median consumer harm  from downstream mergers is about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
+A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is postive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues.
+Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
+Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm.
+With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 20\% of markets and shrinks median consumer harm to about 6\% of pre-merger revenues.
+By contrast, with 3 integrated 3rd parties, the inter-quartile range shrinks, and median total harm falls to about 3.7\% of pre-merger revenues.
+
+Like downstream mergers, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues.
+A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues.
+Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
+Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.
+
+Turning to vertical mergers, absent incumbent integration, about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.
+The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged.
+Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%.
+Adding a fourth integrated incumbent has a similar effect to adding a third incumbent.
+
+Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms harm consumers in about 96\% of simulated markets and lowers total surplus in more than 92\% of simulated markets.
+while adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.
 
@@
 \subsection*{Bargaining Power}
+\citeN{ST2017} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power.
+Here, we show that this result continues to hold when there are rival incumbent integrated firms.
+We also show that this relationship holds for integrated mergers as well as upstream mergers when one of the merging parties is vertically integrated.
+
+Figures \ref{fig:CVbargupBW}-\ref{fig:CVbargbothBW}  depicts box and whisker plots summarizing the consumer, retailer, wholesaler and total welfare effects for each merger as the number of incumbent integrated firms increases.
+For upstream, downstream and vertical mergers, the figures depict box and whisker plots for 0,1 and 4 incumbent vertically integrated firms.
+For downstream and upstream  mergers (Figures \ref{fig:CVbargupBW} and \ref{fig:CVbargdownBW}), the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
+By contrast, for vertical mergers ((Figures \ref{fig:CVbargvertBW}), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots for integrated mergers with 2,4, or 6 integrated firms.
+When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargbothBW} depicts a merger between two integrated firms, with all rivals are unintegrated.
+Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.
 
@@
 \includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer and total surplus as the number of vertically integrated firms present in a market change.
-Each blue box depicts the effects assuming that firms face constant marginal costs, while each orange box depicts the effects assuming that firms face linear marginal costs.
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.

### 2021-04-19 code/genfigures.R

@@ -49,7 +49,8 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 
 pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","Total")), aes(y=Outcome_value/mktrev.pre*100,
                                            #avgpricedelta/mktrev.pre*100,
-                                           x=vert,color=Cost
+                                           x=vert,#color=Cost
+                                           color=Outcome
                                            )
                                            ) +
   stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
@@ -62,13 +63,13 @@ pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","
   #theme_tufte(ticks=FALSE) +
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
-  facet_grid(Outcome~Merger,scales="free",labeller = label_context)+
+  facet_grid(~Merger,scales="free",labeller = label_context)+
   xlab("# Pre-merger Vertically Integrated Firms")+
   ylab("Outcome (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
   #ylab("Share-Weighted Downstream Price Change")+
   #geom_text(data=ann_text,label="Wholesale advantage")
-  labs(colour="Cost:")+
+  labs(colour="Outcome:")+
   labs(title =  "The Distributions  of Merger Outcomes as the Number of Integrated Firms Increases",
        subtitle="Outcomes are reported as a percentage of pre-merger total expenditures.\nHorizontal mergers occur between a vertically integrated and unintegrated firm."
        #subtitle = "1st and 2nd score auctions yields radically different predictions for downstream mergers,\n but similar predictions for upstream mergers",
@@ -407,7 +408,7 @@ pfirmsvert_retailers.bw <- ggplot(filter(ungroup(res.nests.logit),merger =="vert
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
   )
 
-pbargup_all.bw <- ggplot(filter(res.nests.logit,merger =="up" & vert != "3" & vert !="1"), aes(y=Outcome_value/mktrev.pre*100,
[…]

### 2021-04-20 doc/VerticalArrangementsSheuTaragin.tex
finished a VERY ROUGH draft of results sections, excluding what to say on costs

@@
 \author{Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov.
-} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission}
+} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission \and Margaret Loudermilk\footnote{U.S Department of Justice, margaret.loudermilk@usdoj.gov}\\  U.S Department of Justice}
 
@@
 Here, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
-We also examine merger outcomes for these integrated mergers as the number of integrated non-merging firms increases.
+We also examine merger outcomes for  mergers involving at least one integrated firm as the number of integrated non-merging rivals increases.
 
 Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms.
-The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2017}.
+The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2017}\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2017} is that these results exclude downstream markets where prices are set according to a 2nd score auction, but include markets where costs are linear as well as constant.} .
 For downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
@@
 Starting with downstream mergers (leftmost column), absent incumbent vertical integration (0 integrated firms), not only are downstream mergers rarely beneficial, but median consumer harm  from downstream mergers is about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
-A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is postive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues.
+A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues.
 Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
@@
 Here, we show that this result continues to hold when there are rival incumbent integrated firms.
-We also show that this relationship holds for integrated mergers as well as upstream mergers when one of the merging parties is vertically integrated.
+We also show that this relationship holds for integrated mergers as well as downstream and upstream mergers when one of the merging parties is vertically integrated.
 
-Figures \ref{fig:CVbargupBW}-\ref{fig:CVbargbothBW}  depicts box and whisker plots summarizing the consumer, retailer, wholesaler and total welfare effects for each merger as the number of incumbent integrated firms increases.
+%Figures \ref{fig:CVbargupBW}-\ref{fig:CVbargbothBW}  depicts box and whisker plots summarizing the consumer, retailer, wholesaler and total welfare effects for each merger as the number of incumbent integrated firms increases.
 For upstream, downstream and vertical mergers, the figures depict box and whisker plots for 0,1 and 4 incumbent vertically integrated firms.
@@
 Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.
+
+Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the number of incumbent integrated firms increases from 0 to 1 and finally 4 incumbent vertically integrated firms.
+Plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2017}.
+For downstream (left) and upstream (middle) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
+By contrast, for vertical mergers (right), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+Finally, plots with 4 integrated firms increases the number of rival incumbent integrated firms in upstream and downstream mergers to 3 and in vertical mergers to 4 .
+
+For downstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm.
+Median consumer harm decreases from about 18\% of pre-merger expenditures when retailers have relatively more bargaining power to about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm decreases from 7.5\% of pre-merger expenditures when retailers have relatively more bargaining power to 2\% when wholesalers have relatively more bargaining power.
+By contrast, mergers between an integrated and unintegrated upstream supplier (1 integrated firms), causes the box and whisker plots to rotate clockwise around equal bargaining power, strengthening the positive relationship between bargaining power and both consumer and total surplus.
+Increasing wholesaler bargaining power causes median consumer harm to fall from 25\% of pre-merger revenues to -11\% of pre-merger revenues, while median total harm falls from 12.5\% to 2\% of pre-merger revenues.
+Adding additional rival incumbents (4 integrated firms) has the reverse effect, rotating the box and whisker plots counter-clockwise and weakening the still positive relationship between bargaining power and surplus.
+Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -1.7\% of pre-merger revenues, while median total harm falls from 5.3\% to 1.9\% of pre-merger revenues.
+
+Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.
+Median consumer harm increases from about 2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power.
+By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus.
+Median consumer harm increases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power.
+Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.
+
+Unlike downstream or upstream mergers, where the presence of vertically integrated incumbent firms can affect either the magnitude or direction of the relationship between bargaining power and harm, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers.
+Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power.
+
+Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots summarizing consumer, retailer, wholesaler and total surplus  for integrated mergers with 2,4, or 6 integrated firms.
+When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargbothBW} depicts a merger between two integrated firms: all rivals are unintegrated.
+Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.
+
+In terms of the reltationship between bargaining power and harm, integrated mergers are perhaps most similar to vertical mergers.
+Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
+Median consumer harm increases from about 26\% of pre-merger expenditures when retailers have relatively more bargaining power to about -1.4\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from 16\% of pre-merger expenditures when retailers have relatively more bargaining power to -13\% when wholesalers have relatively more bargaining power.
+
+A second similarity between vertical and integrated mergers is that the presence of integrated incumbents has little impact on the strong, monotonic relationship between  bargaining power and harm.
+However, unlike vertical mergers, consumer harm from integrated mergers when there are fewer incumbent integrated firms does not first-order stochastically dominate consumer harm when there are more incumbent integrated firms.
+
+An important difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power exceeds 7/3.
+In this way, integrated mergers are more similar to downstream mergers, or upstream mergers in markets with vertically integrated incumbents.
+
+\textbf{Describe the counterintuitive result that increasing retailer bargaining power harms retailers but help wholesalers.}
+
+\subsection*{Costs}
 
@@
 
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.9]{../output/CVbargupBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
+% \label{fig:CVbargupBW}
+% \end{sidewaysfigure}
+%
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.9]{../output/CVbargdownBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
+% \label{fig:CVbargdownBW}
+% \end{sidewaysfigure}
+%
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.9]{../output/CVbargvertBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
+% \label{fig:CVbargvertBW}
+% \end{sidewaysfigure}
+
 \begin{sidewaysfigure}
 \centering
-\includegraphics[scale=0.9]{../output/CVbargupBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+\includegraphics[scale=0.9]{../output/CVbargnobothBW.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which downstream, upstream, and vertical mergers affect consumer, and total surplus as the bargaining power of wholesalers relative to retailers changes.
 The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargupBW}
+\label{fig:CVbargnobothBW}
 \end{sidewaysfigure}
@@
 \centering
-\includegraphics[scale=0.9]{../output/CVbargdownBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargdownBW}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{../output/CVbargvertBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargvertBW}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{../output/CVbargbothBW.png}
+\includegraphics[scale=0.85]{../output/CVbargbothBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.

### 2021-04-20 code/genfigures.R

@@ -47,7 +47,8 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 # ## box and whisker results for Incumbent
 #
 
-pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","Total")), aes(y=Outcome_value/mktrev.pre*100,
+pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","Total") #& Cost=="Constant"
+                                                     ), aes(y=Outcome_value/mktrev.pre*100,
                                            #avgpricedelta/mktrev.pre*100,
                                            x=vert,#color=Cost
                                            color=Outcome
@@ -408,6 +409,52 @@ pfirmsvert_retailers.bw <- ggplot(filter(ungroup(res.nests.logit),merger =="vert
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
   )
 
+
+pbargnoboth.bw <- ggplot(filter(res.nests.logit,Outcome %in% c("Consumer","Total") & merger !="both" & vert %in% c("0","1","4")), aes(y=Outcome_value/mktrev.pre*100,
+                                                                                               #avgpricedelta/mktrev.pre*100,
+                                                                                               x=factor(barg,labels=MASS::fractions(relleverage)),color=vert)) +
+  #geom_boxplot(outlier.alpha = 0.1) +
+  stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
+  #coord_cartesian(ylim=c(-60,60))+
+  scale_y_continuous(breaks=seq(-100, 100, by=10) ) +
+  geom_hline(yintercept=0,linetype="dashed",color="black")+
+  geom_vline(xintercept=5,linetype="dotted")+
+  theme_bw()+#scale_colour_tableau('Color Blind')+
+  #scale_color_brewer(palette = "PuRd",direction=-1)+
+  scale_color_manual(values = seq_palette)+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position="bottom")+
+  #scale_x_discrete(labels=rev(levels(res.nests$relleveragePre)))+
[…]

### 2021-04-20 doc/VerticalArrangements.tex
minor edits. Sent to Gloria and Peggy

--- 2021-04-20 doc/VerticalArrangementsSheuTaragin.tex
@@
 This section provides an overview of our methodology, with additional details appearing in the Appendix.
-We simulate markets by randomly sampling shares from a Dirichlet distribution for 2, 3, 4, or 5 retailers or wholesalers, respectively.\footnote{We parametrize the Dirichlet distribution so it is equivalent to a uniform distribution.
-}  We also assume that in the pre-merger state, anywhere from 0 to 4 retailers are vertically integrated with a single wholesaler.
-We assume that vertically integrated wholesalers supply inputs to non-integrated retailers and that vertically integrated retailers purchase inputs from non-integrated wholesalers.
+We consider upstream and downstream mergers between either two unintegrated firms or between one vertically integrated and one unintegrated firm.
+For vertical mergers, we only consider mergers between an unintegrated wholesaler and unintegrated retailer.
+Finally, we assume that an integrated merger occurs between two vertically integrated firms\textbf{consider adding diagonal mergers by allowing for heterogeneity in the bargaining parameter.}.
+
+We simulate markets by randomly sampling shares from a Dirichlet distribution for 2, 3, 4, or 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.
+}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7.
+We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
+We assume that vertically integrated firms are no siloed: integrated wholesalers supply inputs to non-integrated retailers and that vertically integrated retailers purchase inputs from non-integrated wholesalers.
 The price coefficient $\alpha$ is calibrated by assuming that in the pre-merger world, there is a vertically integrated outside option available to all customers.
@@
 
-For each combination of number of retailers, number of wholesalers, and bargaining parameter, we draw 1,000 different sets of market primitives.
+For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
 This results in 3.85 million merger simulations.
-We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding 2.69 million markets.\textbf{discrepancy between res.nests dimension and ALL Counts in Table 1}\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding 2.69 million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
 } All 2.69 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
@@
 Each panel contains four pairs of box and whisker plots, with each pair corresponding to a different type of merger.
-The blue box and whisker plots (on the left in each pair) depict outcomes assuming that marginal costs are constant, and the orange box and whisker plots (on the right in each pair) show outcomes assuming that marginal costs are linear.
+The blue box and whisker plots (left) depict outcomes assuming that marginal costs are constant, and the orange box and whisker plots (right) show outcomes assuming that marginal costs are linear.
 The whiskers display the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median.
@@
 
-In terms of the reltationship between bargaining power and harm, integrated mergers are perhaps most similar to vertical mergers.
+In terms of the relationship between bargaining power and harm, integrated mergers are perhaps most similar to vertical mergers.
 Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
@@
 \include{sumtable}
+\caption{Summary Statistics}
 \label{tab:simsum}
-\caption{Summary Statistics}
 \end{table}

### 2021-04-23 doc/VerticalArrangements.tex
changed orders in which mergers are displayed. Experimented with cutting figure 2 in a couple of different ways

@@
 This section provides an overview of our methodology, with additional details appearing in the Appendix.
-We consider upstream and downstream mergers between either two unintegrated firms or between one vertically integrated and one unintegrated firm.
 For vertical mergers, we only consider mergers between an unintegrated wholesaler and unintegrated retailer.
+In addition, we consider upstream and downstream mergers between either two unintegrated firms or between one vertically integrated and one unintegrated firm.
 Finally, we assume that an integrated merger occurs between two vertically integrated firms\textbf{consider adding diagonal mergers by allowing for heterogeneity in the bargaining parameter.}.
@@
 We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
-We assume that vertically integrated firms are no siloed: integrated wholesalers supply inputs to non-integrated retailers and that vertically integrated retailers purchase inputs from non-integrated wholesalers.
+We assume that vertically integrated firms are not siloed: integrated wholesalers supply inputs to non-integrated retailers and that vertically integrated retailers purchase inputs from non-integrated wholesalers.
 The price coefficient $\alpha$ is calibrated by assuming that in the pre-merger world, there is a vertically integrated outside option available to all customers.
@@
 For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-This results in 3.85 million merger simulations.
-We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding 2.69 million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} All 2.69 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+This results in 1.96 million merger simulations.
+We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding 1.37 million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+} All 1.37 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
 
@@
 
-Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.5, and the median average retail pre-merger price is \$13.
+Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.6, and the median average retail pre-merger price is \$14.
 Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.
-Pre-merger HHIs range between 2,820 at the $25^{th}$ percentile to 5,013 at the $75^{th}$, with a median of 3,689.
+Pre-merger HHIs range between 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703.
 HHIs for horizontal downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.
@@
 
-Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
+%Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
 Vertical mergers when firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when firms have linear costs.
@@
 Moreover, there is a partial rank-ordering across mergers: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
-We also find that  for upstream, vertical, and integrated mergers, retailer surplus under constant marginal cost first-order stochastically dominates the retail surplus distribution under linear costs, again suggesting that the incentive to raise rivals' costs dominates the benefits from EDM.
+
+%We also find that  for upstream, vertical, and integrated mergers, retailer surplus under constant marginal cost first-order stochastically dominates the retail surplus distribution under linear costs, again suggesting that the incentive to raise rivals' costs dominates the benefits from EDM.
 By contrast, for downstream mergers, there is no clear rank-ordering between markets with constant marginal costs and markets with linear costs.
@@
 Moreover there is a complete rank ordering of mergers, with total harm from integrated mergers first-order stochastically dominating total harm from downstream mergers, which dominates total harm from upstream mergers, which dominates totals harm from vertical mergers.
-Finally, for each merger type, total harm under linear costs first-order stochastically dominates total harm under constant marginal costs, with the greatest differences occurring for vertical and integrated mergers.
+
+%Finally, for each merger type, total harm under linear costs first-order stochastically dominates total harm under constant marginal costs, with the greatest differences occurring for vertical and integrated mergers.
 
@@
 
-Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms.
+Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms.
 The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2017}\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2017} is that these results exclude downstream markets where prices are set according to a 2nd score auction, but include markets where costs are linear as well as constant.} .
@@
 
-Starting with downstream mergers (leftmost column), absent incumbent vertical integration (0 integrated firms), not only are downstream mergers rarely beneficial, but median consumer harm  from downstream mergers is about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
+Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.
+The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged.
+Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%.
+Adding a fourth integrated incumbent has a similar effect to adding a third incumbent.
+
+downstream mergers (leftmost column), absent incumbent vertical integration , not only are downstream mergers rarely beneficial, but median consumer harm  from downstream mergers is about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
 A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues.
@@
 Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.
-
-Turning to vertical mergers, absent incumbent integration, about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.
-The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged.
-Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%.
-Adding a fourth integrated incumbent has a similar effect to adding a third incumbent.
 
@@
 
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
+}
+% \label{fig:CVvertincumbBW}
+% \end{sidewaysfigure}
+
 \begin{sidewaysfigure}
 \centering
-\includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
+\includegraphics[scale=0.9]{../output/CVvertincumb_updownBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
@@
 }
-\label{fig:CVvertincumbBW}
+\label{fig:CVvertincumbupdownBW}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\centering
+\includegraphics[scale=0.9]{../output/CVvertincumb_vertincBW.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
+}
+\label{fig:CVvertincumbvertincBW}
 \end{sidewaysfigure}

### 2021-04-23 code/RunSims.R

@@ -227,6 +227,7 @@ sink("isMarketSummary.txt")
 print(table(res.nests$isMarket,useNA="ifany"))
 print(table(res.nests$isMarket,res.nests$type,useNA="ifany"))
 print(table(res.nests$isProfitable,res.nests$merger,useNA="ifany"))
+print(table(res.nests$isProfitable,res.nests$mc,useNA="ifany"))
 print(with(res.nests[is.na(res.nests$isMarket) |! res.nests$isMarket,], table(up,down,useNA="ifany")))
 sink()
 
@@ -264,8 +265,8 @@ res.nest_all.long <- select(res.nest_all,type,down,up,merger,barg,nestParm,mc,cv
   mutate(Outcome=factor(Outcome,levels=c("Consumer","Retailer","Wholesaler","Total")))
 
 
-
-sumtable <- data.frame(res.nests)
+## only summarize constant costs
+sumtable <- data.frame(res.nests) %>% filter(mc=="constant")
 
 
 
@@ -307,18 +308,18 @@ options(scipen = 999)
 sumtable <- spread(sumtable, quant, val) %>% select(-Max,Max) %>%
   mutate(
     #set=factor(set,levels=c("firm","bargaining"),labels=c("Firm Count","Bargaining Power")),
-    merger=factor(merger, labels=c("All","Integrated","Upstream","Downstream","Vertical")),
+    merger=factor(merger, levels=c("all","vertical","up","down","both"),labels=c("All","Vertical","Upstream","Downstream","Integrated")),
     variable = factor(variable, levels=c("up","down","vert","barg","nestParm",
                                          "avgpricepre.up","avgpricepre.down","mktElast",
                                          "hhipre","hhipost","hhidelta"),
-                      labels = c("\\# Wholesalers","\\# Retailers","\\# Integrated","Bargaining Power","Nesting Parameter","Avg. Upstream Price ()","Avg. Downstream Price ()","Market Elasticity",
+                      labels = c("# Wholesalers","# Retailers","# Integrated","Bargaining Power","Nesting Parameter","Avg. Upstream Price ($)","Avg. Downstream Price ($)","Market Elasticity",
                                  "Pre-Merger HHI",
                                  "Post-Merger HHI","Delta HHI"
                       )
     ),
     Markets=as.numeric(Markets)) %>%
[…]

### 2021-04-23 code/genfigures.R

@@ -19,13 +19,16 @@ load(file.path(simpath,"AllSims.RData"))
 
 relleverage=sort(unique(res.nests$relleveragePre),decreasing = TRUE)
 
+
+## Only include constant cost sims
 res.nests.logit <- res.nests.long <- mutate(ungroup(res.nests.long),
-                                            Merger=factor(merger,levels=c("down","up","vertical","both"),labels=c("Downstream","Upstream","Vertical","Integrated")),
+                                            Merger=factor(merger,levels=c("vertical","up","down","both"),labels=c("Vertical","Upstream","Downstream","Integrated")),
                                             Cost=factor(mc,labels=c("Constant"
                                                                     ,"Linear"#,"Quadratic","Linear/Quadratic","Quadratic/Linear","Linear/Constant","Constant/Linear"
                                                                     )
                                                         ),
-                                            Cost=reorder(Cost,-Outcome_value/mktrev.pre,median,na.rm=TRUE)) %>% select(-mc)
+                                            Cost=reorder(Cost,-Outcome_value/mktrev.pre,median,na.rm=TRUE)) %>% select(-mc) %>%
+                                            filter(Cost=="Constant")
 
 #res.nests.logit <- filter(as.data.frame(res.nests.logit),nestParm == "0" & Retailers !="1" & Wholesalers !="1") %>%
 #  mutate_if(is.factor,droplevels)
@@ -50,10 +53,41 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","Total") #& Cost=="Constant"
                                                      ), aes(y=Outcome_value/mktrev.pre*100,
                                            #avgpricedelta/mktrev.pre*100,
-                                           x=vert,#color=Cost
-                                           color=Outcome
+                                           x=vert#color=Cost
+                                           #,color=Outcome
                                            )
                                            ) +
[…]

### 2021-04-23 doc/VerticalArrangements.tex
tried to add Gloria's changes without deleting my own

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{The Effects of Vertical Arrangements\\in a Vertical Supply Chain\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, by the Federal Trade Commission, or by its Commissioners.
-}}
+\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, by the Federal Trade Commission, or by its Commissioners.
+Furthermore, the views expressed here should not be purported to reflect those of the U.S.
+Department of Justice.}}
 
@@
 \author{Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov.
-} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission \and Margaret Loudermilk\footnote{U.S Department of Justice, margaret.loudermilk@usdoj.gov}\\  U.S Department of Justice}
+} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission \and Margaret Loudermilk\footnote{U.S.
+Department of Justice, margaret.loudermilk@usdoj.gov}\\  U.S Department of Justice}
 
@@
 \begin{abstract}
-\noindent
+\noindent We use a standard vertical supply-chain model to study the welfare impacts of mergers that have both horizontal and vertical aspects.
+The model features logit Bertrand competition downstream and Nash Bargaining upstream.
+We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and a previously integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and a previously integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+The breadth of options we include better captures the variety of merger structures observed in practice, compared to the typical ``horizontal vs.
+vertical'' dichotomy.
+We further extend our analysis to accommodate preexisting vertical integration by third-party firms and linear marginal costs.
 \end{abstract}
 
-\bigbreak Keywords: bargaining models; merger simulation; vertical markets
+\bigbreak Keywords: bargaining models; merger simulation; vertical markets; vertical mergers
 
@@
 \section{Introduction}
+Given the many different ways a firm may be organized, it is rare that any given merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  Indeed, as the 2020 U.S.
+Department of Justice (DOJ) and Federal Trade Commission (FTC) Vertical Merger Guidelines (henceforth, the ``VMG'') state, mergers often have both horizontal and vertical aspects.
+However, the typical theories and models used to analyze mergers rarely address these complexities.
+Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), while the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
+Yet, a given merger may combine all of these effects simultaneously, which raises the necessity of balancing their impacts in a unified framework.
+
+In this paper, we use a standard merger simulation model to assess the welfare implications of these complex mergers.
+Our model is drawn directly from \citeN{ST2020}, and uses a Bertrand logit downstream retail setup alongside a Nash Bargaining wholesale negotiation upstream.
+We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and a previously integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and a previously integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+In each of these simulations, we calculate the impacts on consumer welfare and firm profits.
+Furthermore, we examine how the presence of additional integrated rivals in the market change our results.
+Our baseline simulations assume that units are produced with constant marginal costs, but we later extend our analysis to linear upward sloping marginal costs.
+
+We then provide a case study (in progress) from an industry, solid waste management, that has seen a large number of mergers with both horizontal and vertical elements.
+Broadly speaking, the supply chain of this sector involves haulers at the downstream level that bring trash to disposal facilities at the upstream level.
+A variety of firms participate in these services, with some owning their own hauling and disposal assets, while others are active in only a segment of the market.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Theory}
+We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2020}.
+This framework allows us to study a variety of merger configurations.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Downstream and Upstream Competition}
+Assume there is a set of consumers indexed by $i$ who each choose to buy a product sold by a single retailer.
+Retailers are indexed by $r$, while the wholesalers that supply these retailers are indexed by $w$.
+Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.
+We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$, and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.
+We divide the set $\mathbb{W}$ into overlapping subsets, each labeled $\mathbb{W}^r$, to indicate which wholesalers' products are carried by which retailers.
+Similarly, we divide the set of retailers $\mathbb{R}$ into overlapping subsets, each labeled $\mathbb{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.
 
+The share of consumers that choose product $w$ sold by retailer $r$ has the logit form,
+\begin{equation}
+s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{t \in \mathbb{R}} \sum_{x \in \mathbb{W}^t} \exp(\delta_{tx} - \alpha p_{tx})},
+\label{eq: nb share}
+\end{equation}
+where $\delta_{rw}$ is a quality parameter and $\alpha$ captures sensitivity to price, denoted by $p_{rw}$.
+There is an outside good whose quality parameter and price have been normalized to zero.
+The retailer's profit is given by $\pi^r = \sum_{w \in \mathbb{W}^r} [p_{rw} - p^W_{rw} - c^R_{rw}] s_{rw} M$, where $p^W_{rw}$ is the unit fee charged by wholesaler $w$ to retailer $r$, $c^R_{rw}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.
+Downstream prices are set in Bertrand equilibrium, according to
+\begin{equation}
+\sum_{x \in \mathbb{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}
+{\partial p_{rw}} + s_{rw} = 0,
+\label{eq: nb downstream foc}
+\end{equation}
+which is the first order condition for product $w$ sold by retailer $r$.
+
+Wholesale prices are set via Nash Bargaining between retailers and wholesalers.
+We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2020}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
+}  Profits for wholesaler $w$ are given by $\pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M$, where $c^W_{rw}$ is the wholesale marginal cost when dealing with retailer $r$.
+The first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
+\begin{equation}
+\begin{split}
+&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
+&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right),
+\end{split}
+\label{eq: upstream foc}
+\end{equation}
+where $\lambda \in [0,1]$ captures the bargaining power of the retailer relative to the wholesaler.
+The $\Delta s_{tx}(\mathbb{W}^r \setminus \{w\}) \equiv s_{tx}(\mathbb{W}^r \setminus \{w\}) - s_{tx}$ is the difference in the share of good $x$ sold by retailer $t$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.
+Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.
+
+Together the series of downstream and upstream first order conditions determine market equilibrium.
+This model can be solved and calibrated as described in \citeN{ST2020}.
+In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
+We loosen this restriction to allow for linear increasing marginal cost as a robustness check.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Mergers}
+Here we describe the manner in which mergers are modeled in this framework.
+We begin by describing vertical mergers, as all other cases we examine have some vertical aspects.
+Suppose that retailer $r$ and wholesaler $w$ were to merge.
+Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $w$ is given by
+\begin{equation}
+\begin{split}
+&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{EDM effect}} \\
+&+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} = 0.
+\end{split}
+\label{eq: vmerger own ret downstream foc}
+\end{equation}
+The pricing problem balances two effects.
+On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $r$ being able to access product $w$ at marginal cost.
+This force would tend to lower the resulting price.
+On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to the merging partner wholesaler $w$.
+This force would tend to raise the resulting consumer price.
+
+Turning to input prices, when wholesaler $w$ bargains with unaffiliated retailer $s$, the first order condition becomes
+\begin{equation}
+\begin{split}
+&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
+& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
+&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
+\end{split}
+\label{eq: vmerger wh upstream foc}
+\end{equation}
+which reflects the change in the disagreement payoff coming from the merger with retailer $r$.
+Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $s$, these losses are softened due to a potential for diversion to retailer $r$, which we label the ``RRC effect.''  Furthermore, the margin on product $w$ sold by retailer $r$ is potentially higher due to EDM, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
+These impacts tend to raise the resulting input price.
+
+When the merged firm is bargaining with the unaffiliated wholesaler $v$ over what input price to pay as a retailer, the bargaining first order condition becomes
+\begin{equation}
+\begin{split}
+&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
+&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
+&\left.- \underbrace{[p_{rw} - c^W_{rw} - c^R_{rw}] \Delta s_{rw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
+\end{split}
+\label{eq: vmerger ret upstream foc}
+\end{equation}
+In this case, the merged firm has two channels for potential additional profits should it cease to trade with wholesaler $v$.
+First, if retail sales are diverted to product $w$ sold by retailer $r$, those sales will earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $v$ carried by retailer $r$ could increase sales by wholesaler $w$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.
+
+Next consider a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
+Such a combination has a vertical component and a horizontal component.
+The first order condition for setting the downstream price $p_{rw}$ becomes
+\begin{equation}
+\begin{split}
+&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{direct EDM effect}}+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} \\ &+\underbrace{\underbrace{[p_{sw} - c^W_{sw} - c^R_{sw}]\frac{\partial s_{sw}}{\partial p_{rw}}}_{\text{indirect EDM Effect}} + \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \frac{\partial s_{sx}} {\partial p_{rw}}}_{\text{downstream UPP effect}}= 0.
+\end{split}
+\label{eq: intdmerger own ret downstream foc}
+\end{equation}
+Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $r$ raises its prices.
+EDM between $w$ and $s$ can actually increase this UPP impact, because when the sales are diverted to product $w$ sold by retailer $s$, those units earn a larger margin.
+This impact is what we label the ''indirect EDM effect,'' which comes through the interaction of EDM interaction with UPP.
+On the other hand, the price of retailer $s$ for product $w$ could fall due to the direct impact of EDM, since that product has a lower realized marginal cost.
+The net impact on consumers is ambiguous.
+
+If instead a preexisting integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the resulting first order condition for product $rw$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc}.
+However,  the downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of wholesaler $v$.
+The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain product $v$ at marginal cost, which in turn would raise the value of diverted sales to that product.
+Whether these additional incentives to raise prices will dominate the direct EDM impact of the lower input cost of $v$ for retailer $r$ is an empirical question.
+When considering a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.
+
+Turning to wholesale prices, again consider a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
+When wholesaler $w$ bargains with an unaffiliated retailer $u$ we have the first order condition given by,
+\begin{equation}
+\begin{split}
+&[p^W_{uw} - c^W_{uw}]s_{uw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s,u\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})\\
+& - \overbrace{\sum_{t \in \{r,s\}}\left(\overbrace{[p_{tw}-c^W_{tw}-c^R_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^t \setminus \{w\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathbb{W}^u \setminus \{w\})\right)}^{\text{RRC effect}}=\\
+&\frac{1-\lambda}{\lambda} \left([p_{uw} - p^W_{uw} - c^R_{uw}]s_{uw} - \sum_{x \in \mathbb{W}^u \setminus \{w\}} [p_{ux} - p^W_{ux} - c^R_{ux}] \Delta s_{ux}(\mathbb{W}^u \setminus \{w\})\right).
+\end{split}
+\label{eq: intdmerger wh upstream foc}
+\end{equation}
+The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $r$, both of which may potentially recapture sales should retailer $u$ lose access to product $w$.
+The merged firm has higher bargaining leverage as a result.
+If instead the integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the profits earned by firm $rw$ should the negotiation with retailer $u$ fail are augmented with the earnings of wholesaler $v$ rather than retailer $s$.
+This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with wholesaler $w$, that retailer's contract with wholesaler $v$ remains in place.
+This assumption can be loosened.
+}    In the case of a merger between two integrated retailer/wholesaler firms, all of these effects would appear.
+
+Returning to a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $v$ to supply retailer $r$, the first order condition becomes
+\begin{equation}
+\begin{split}
+&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
+&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
+&- \underbrace{\sum_{t \in \{r,s\}}[p_{tw} - c^W_{tw} - c^R_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}
+- \underbrace{\sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^r \setminus \{v\})}_{\text{retail recapture leverage effect}}\\
+&-\underbrace{\left.\sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})\right)}_{\text{wholesale recapture leverage effect}}.
+\end{split}
+\label{eq: intdmerger ret upstream foc}
+\end{equation}
+Compared to equation \eqref{eq: vmerger ret upstream foc}, now the EDM recapture effect applies to sales of product $w$ at both affiliated retailers $r$ and $s$.
+Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $r$ lose access to product $v$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture effect.
+Both of these additional terms tend to increase the bargaining leverage of the merged firm.
+If we instead examined a merger between retailer/wholesaler $rw$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $r$ for two merged wholesalers, rather than sales through two merged retailers.
+Similarly, the retail recapture leverage effect would be replaced with an additional wholesale recapture leverage effect for the newly merged wholesaler.
+For a merger between two integrated retailer/wholesaler firms, all of these effects would enter.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-04-28 doc/VerticalArrangements.tex
incorporating new figures and updating text

@@
 As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 82\% of all upstream simulated mergers, 28\% of simulated vertical mergers, 20\% of simulated downstream mergers, and 17\% of simulated integrated mergers.
-%\textbf{comparing these results to \citeN{ST2017}, the effects are less stark (i.e.
+%\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e.
 less of a transfer.}
@@
 Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms.
-The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2017}\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2017} is that these results exclude downstream markets where prices are set according to a 2nd score auction, but include markets where costs are linear as well as constant.} .
+The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction, but include markets where costs are linear as well as constant.} .
 For downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
@@
 \subsection*{Bargaining Power}
-\citeN{ST2017} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power.
+\citeN{ST2020} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power.
 Here, we show that this result continues to hold when there are rival incumbent integrated firms.
@@
 Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the number of incumbent integrated firms increases from 0 to 1 and finally 4 incumbent vertically integrated firms.
-Plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2017}.
+Plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}.
 For downstream (left) and upstream (middle) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
@@
 
+Here, we explore the role that the constant marginal cost assumption plays in driving merger outcomes.
+In theory, constant marginal costs can increase the magnitude of EDM, as the wholesaler price reduction applies equally to all units produced by the retailer.
+By contrast, if either the integrated wholesaler or retailer face linear marginal costs, increased output from the EDM is limited by the increased marginal costs, which should limit the benefit to the parties.
+Likewise, in theory, RRC could be more profitable for the merging parties under linear than constant marginal costs, as linear marginal costs cause the integrated wholesaler's price to increase more rapidly.
+
+Because EDM is plausibly greater under constant marginal costs while RRC is also plausibly less, then one might expect that on net mergers are more beneficial under constant than linear marginal costs.
+Figure \ref{fig:surplussumcost} displays box and whisker plots that depict the net effect of these different marginal cost specifications on consumer and total welfare under six different scenarios:
+linear marginal costs for the non-merging parties products (``Party Constant", dark blue), constant marginal costs for all products (``Constant", orange), linear marginal costs for retailers (``Constant/Linear", light grey), linear marginal costs for wholesalers (``Linear/Constant", dark grey), linear marginal costs for the merging parties products (``Party Linear",light blue) and linear marginal costs for all products (``Linear'',red).
+These six scenarios are ordered from least to most consumer harm in vertical mergers.
+
+Overall, consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
+Mergers with all constant-cost firms (orange) benefit consumers in about 15\% of all mergers, while mergers with all linear-cost firms benefit consumers in only 2.3\% of mergers.
+Indeed, consumer harm under the all linear marginal costs scenario first-order stochastically dominates consumer harm under the all constant marginal costs scenario.
+The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.
+
+The biggest differences in outcomes occurs for vertical mergers.
+when all firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when all firms have linear costs (light blue).
+Upstream mergers with all constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs.
+Downstream mergers with constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when firms have linear costs.
+Finally, integrated  mergers benefit consumers in about 10\% of all simulations when firms have constant costs, but less than 1\% of simulations when firms have linear costs.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
+\begin{sidewaysfigure}
+\centering
+\includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
 }
-% \label{fig:CVvertincumbBW}
-% \end{sidewaysfigure}
+\label{fig:CVvertincumbBW}
+\end{sidewaysfigure}
 
@@
 
+\begin{sidewaysfigure}
+\centering
+\includegraphics[scale=0.9]{../output/surplussum_cost.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 6 different cost scenarios.
+Dark blue boxes (``Party Constant") assume that the merging parties face constant marginal costs while other firms face linear marginal costs.
+Orange boxes (``Constant") assume that all firms face constant marginal costs.
+Light grey boxes (``Constant/Linear") assume upstream firms have constant marginal costs, while downstream firms have linear marginal costs.
+Orange box depicts the effects assuming that firms face linear marginal costs.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
+}
+\label{fig:surplussumcost}
+\end{sidewaysfigure}
 \end{document}

### 2021-04-28 code/RunSims.R

@@ -22,7 +22,7 @@ simpath <- "./data"
 #profiler <- microbenchmark::microbenchmark(times=1,{
 #Rprof(filename = "Rprof.out", append = FALSE, interval = 2, line.profiling=TRUE)
 
-cl <- makeCluster(mc <- getOption("cl.cores", 45))
+cl <- makeCluster(mc <- getOption("cl.cores", 100))
 ## to make this reproducible
 clusterSetRNGStream(cl, 3141519)
 
@@ -38,7 +38,7 @@ nestParm <- c(0)
 type=c("1st")
 merger=c("up","down", "vertical","both")
 bargparm <- seq(.1,.9,.1)
-mc=c("constant","linprod"#, "quadprod","linquad","quadlin","lincons","conslin"#,"linfirm"
+mc=c("constant","linprod","lincons","conslin","consparty","linparty"#, "quadprod","linquad","quadlin"#,"linfirm"
      )
 relleveragePre <-  (1-bargparm)/bargparm
 
@@ -196,7 +196,8 @@ res.nests <-  lapply(mkts.nests, function(x){
 
 
 res.nests <- data.table::rbindlist(res.nests)
-res.nests <- mutate(res.nests, up=factor(up), down=factor(down),
+res.nests <- mutate(res.nests,
+                    up=factor(up), down=factor(down),
                     #vert=ifelse(merger=="vertical",vert+1,vert),
                     vert=ifelse(merger=="both",vert+2,vert),
                     vert=factor(vert),nestParm=factor(nestParm),

### 2021-04-28 doc/VerticalArrangements.tex
incroprated glorias changes

--- 2021-04-23 doc/VerticalArrangements.tex
@@
 In this paper, we use a standard merger simulation model to assess the welfare implications of these complex mergers.
-Our model is drawn directly from \citeN{ST2020}, and uses a Bertrand logit downstream retail setup alongside a Nash Bargaining wholesale negotiation upstream.
+Our model is drawn directly from \citeN{ST2020}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
 We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and a previously integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and a previously integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
 In each of these simulations, we calculate the impacts on consumer welfare and firm profits.
-Furthermore, we examine how the presence of additional integrated rivals in the market change our results.
-Our baseline simulations assume that units are produced with constant marginal costs, but we later extend our analysis to linear upward sloping marginal costs.
+Furthermore, we examine how the presence of additional integrated rivals in the market changes our results.
+Our baseline simulations assume that units are produced with constant marginal costs, but we later extend our analysis to linear upward-sloping marginal costs.
 
@@
 \subsection*{Downstream and Upstream Competition}
-Assume there is a set of consumers indexed by $i$ who each choose to buy a product sold by a single retailer.
+Assume there is a set of consumers who each choose to buy a product sold by a single retailer.
 Retailers are indexed by $r$, while the wholesalers that supply these retailers are indexed by $w$.
@@
 \end{equation}
-where $\delta_{rw}$ is a quality parameter and $\alpha$ captures sensitivity to price, denoted by $p_{rw}$.
+where $\delta_{rw}$ is a quality parameter and $\alpha$ captures sensitivity to price $p_{rw}$.
 There is an outside good whose quality parameter and price have been normalized to zero.
@@
 This force would tend to lower the resulting price.
-On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to the merging partner wholesaler $w$.
+On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to wholesaler $w$.
 This force would tend to raise the resulting consumer price.
@@
 Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $r$ raises its prices.
-EDM between $w$ and $s$ can actually increase this UPP impact, because when the sales are diverted to product $w$ sold by retailer $s$, those units earn a larger margin.
-This impact is what we label the ''indirect EDM effect,'' which comes through the interaction of EDM interaction with UPP.
-On the other hand, the price of retailer $s$ for product $w$ could fall due to the direct impact of EDM, since that product has a lower realized marginal cost.
-The net impact on consumers is ambiguous.
+EDM between $w$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $w$ sold by retailer $s$, those units earn a larger margin.
+This impact is what we label the ``indirect EDM effect,'' which comes through the interaction of EDM with UPP.
 
@@
 The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain product $v$ at marginal cost, which in turn would raise the value of diverted sales to that product.
-Whether these additional incentives to raise prices will dominate the direct EDM impact of the lower input cost of $v$ for retailer $r$ is an empirical question.
+Whether these additional incentives to raise prices will dominate the direct EDM impact is an empirical question.
 When considering a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.
@@
 The merged firm has higher bargaining leverage as a result.
-If instead the integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the profits earned by firm $rw$ should the negotiation with retailer $u$ fail are augmented with the earnings of wholesaler $v$ rather than retailer $s$.
+If instead the integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the profits earned by firm $rw$ should the negotiation with retailer $u$ fail are augmented with the earnings of wholesaler $v$ rather than of retailer $s$.
 This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with wholesaler $w$, that retailer's contract with wholesaler $v$ remains in place.

### 2021-04-29 doc/VerticalArrangements.tex
added some new figures. added additional text to explain figures. still a work in progress.

--- 2021-04-28 doc/VerticalArrangements.tex
@@
 Pre-merger HHIs range between 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703.
-HHIs for horizontal downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.
-HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238.
-HHIs for vertical mergers increase by 1,055 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
-}  HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280.
+HHIs for vertical mergers increase by 1,277 points at the median, resulting in a median post-merger HHI equal to 4,889.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
+}  HHIs for upstream mergers increase by 2,004 points at the median, resulting in a median post-merger HHI equal to 5,231.
+HHIs for horizontal downstream mergers increase by 2,054 points at the median, resulting in a median post-merger HHI equal to 4,889.
+HHIs for integrated mergers increase by 2,746 points at the median, resulting in a median post-merger HHI equal to 7,220.
 Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with post-merger HHIs over 2,500 points and HHI Changes greater than 200 points.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
@@
 
-Each panel contains four pairs of box and whisker plots, with each pair corresponding to a different type of merger.
-The blue box and whisker plots (left) depict outcomes assuming that marginal costs are constant, and the orange box and whisker plots (right) show outcomes assuming that marginal costs are linear.
+Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger.
 The whiskers display the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median.
@@
 
-%Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
-Mergers with constant-cost firms benefit consumers in about 15\% of all mergers, while mergers with linear-cost firms benefit consumers in only 2.3\% of mergers.
-Indeed, consumer harm under linear marginal costs first-order stochastically dominates consumer harm under constant marginal costs.
-The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.
-
-%Consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
-Vertical mergers when firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when firms have linear costs.
-Upstream mergers with constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs.
-Downstream mergers with constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when firms have linear costs.
-Finally, integrated  mergers benefit consumers in about 10\% of all simulations when firms have constant costs, but less than 1\% of simulations when firms have linear costs.
-
-%mergers are unlikely to benefit consumers: upstream under both constant and linear marginal costs, vertical mergers with linear marginal costs, and integrated mergers with linear marginal costs.
-That upstream horizontal mergers are largely net harmful is somewhat surprising, as it indicates that any EDM from the unintegrated upstream firm merging with its integrated rival is outweighed by the recapture effect.
-Similar logic applies to integrated mergers.
-Likewise, the fact that vertical mergers with linear marginal costs are almost never beneficial when marginal costs are linear is surprising, as it suggests that the incentive to raise rivals' cost dominates the benefit from EDM.
-%especially given that about 40\% of vertical mergers with constant marginal costs are beneficial.
 \textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}
@@
 Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 70\% of all simulations.
-Moreover, there is a partial rank-ordering across mergers: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
+Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
 
@@
 In terms of total welfare, with the exception of the approximately 10\% of vertical mergers and 8.5\% of integrated mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful.
-Moreover there is a complete rank ordering of mergers, with total harm from integrated mergers first-order stochastically dominating total harm from downstream mergers, which dominates total harm from upstream mergers, which dominates totals harm from vertical mergers.
+Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers.
+Median consumer harm from integrated mergers is about 10\% of pre-merger total expenditures, X.XX times the magnitude of that from downstream mergers, X.X times the magnitude of that from upstream mergers, and more than X.X times the magnitude of vertical mergers.
 
@@
 
-Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (blue,left) and total harm (orange, right) as the number of incumbent integrated firms increases from 0 to 6 firms.
-The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction, but include markets where costs are linear as well as constant.} .
-For downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
-By contrast, for vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
-Likewise, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, and all rivals are unintegrated.
+Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms.
+The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}.\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+By contrast,for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
+Finally, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, again with no integrated rivals.
 
@@
 Adding a fourth integrated incumbent has a similar effect to adding a third incumbent.
+Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not lessen as additional additional integrated incumbents are added.
 
-downstream mergers (leftmost column), absent incumbent vertical integration , not only are downstream mergers rarely beneficial, but median consumer harm  from downstream mergers is about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
+By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues.
+A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues.
+Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
+Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.
+In summary, additional incumbent integrated firms do not substantially impact median harm, but do substantially reduce the likelihood of observing extremely harmful mergers.
+
+Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
 A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues.
@@
 
-Like downstream mergers, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues.
-A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues.
-Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
-Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.
-
 Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms harm consumers in about 96\% of simulated markets and lowers total surplus in more than 92\% of simulated markets.
-while adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.
+While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.
 
@@
 
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{../output/CVvertincumb_updownBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.9]{../output/CVvertincumb_updownBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
 }
-\label{fig:CVvertincumbupdownBW}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{../output/CVvertincumb_vertincBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
+% \label{fig:CVvertincumbupdownBW}
+% \end{sidewaysfigure}
+%
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.9]{../output/CVvertincumb_vertincBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
 }
-\label{fig:CVvertincumbvertincBW}
-\end{sidewaysfigure}
+% \label{fig:CVvertincumbvertincBW}
+% \end{sidewaysfigure}
 

### 2021-04-29 code/RunSims.R

@@ -197,7 +197,10 @@ res.nests <-  lapply(mkts.nests, function(x){
 
 res.nests <- data.table::rbindlist(res.nests)
 res.nests <- mutate(res.nests,
-                    up=factor(up), down=factor(down),
+                    up=ifelse(merger=="both",up+2,up),
+                    down=ifelse(merger=="both",down+2,down),
+                    up=factor(up),
+                    down=factor(down),
                     #vert=ifelse(merger=="vertical",vert+1,vert),
                     vert=ifelse(merger=="both",vert+2,vert),
                     vert=factor(vert),nestParm=factor(nestParm),

### 2021-04-30 doc/VerticalArrangements.tex
added additional updates

@@
 In particular, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
-Median consumer harm from integrated mergers is about 16\% of pre-merger total expenditures, 1.37 times the magnitude of that from downstream mergers, 2.4 times the magnitude of that from upstream mergers, and more than 3.5 times the magnitude of vertical mergers.
+Median consumer harm from integrated mergers is almost 14\% of pre-merger total expenditures, 1.25 times the magnitude of that from downstream mergers, 2.3 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.
 
-\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}
+%\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}
 
@@
 
-Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 70\% of all simulations.
+Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 67\% of all simulations.
 Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
@@
 
-As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 82\% of all upstream simulated mergers, 28\% of simulated vertical mergers, 20\% of simulated downstream mergers, and 17\% of simulated integrated mergers.
+As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and abput 15\% of simulated integrated mergers.
 %\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e.
@@
 
-In terms of total welfare, with the exception of the approximately 10\% of vertical mergers and 8.5\% of integrated mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful.
+In terms of total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers with constant costs that are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial.
 Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers.
-Median consumer harm from integrated mergers is about 10\% of pre-merger total expenditures, X.XX times the magnitude of that from downstream mergers, X.X times the magnitude of that from upstream mergers, and more than X.X times the magnitude of vertical mergers.
+Median consumer harm from integrated mergers is about 9.6\% of pre-merger total expenditures, 1.5 times the magnitude of that from downstream mergers, 2.5 times the magnitude of that from upstream mergers, and more than 4.3 times the magnitude of vertical mergers.
 
@@
 Here, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
-We also examine merger outcomes for  mergers involving at least one integrated firm as the number of integrated non-merging rivals increases.
+We also examine merger outcomes for  mergers as the number of incumbent integrated non-merging rivals increases.
 
 Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms.
-The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}.\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020} and is included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merging with an unintegrated retailer, a single unintegrated 3rd party is integrated, and the remaining 3rd parties (if any) are unintegrated.
 By contrast,for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
@@
 
-Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.
-The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged.
-Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%.
-Adding a fourth integrated incumbent has a similar effect to adding a third incumbent.
-Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not lessen as additional additional integrated incumbents are added.
+Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial (i.e.
+increase total surplus).
+The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 25\% and the percentage of mergers that increase  total surplus to 16\%, but leads median harm largely unchanged.
+Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefits consumers to about 27\% but does not increase the percentage of net beneficial mergers.
+Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of netbeneficial mergers to about 11\%.
+Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.
 
 By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues.
-A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues.
-Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
+A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare increases in almost 16\% of mergers and total welfare increases in 4.3\% of mergers, but median consumer harm increases to 6.3\% and median total harm increases to 4.8\% of pre-merger revenues.
+Increasing the number of integrated 3rd parties to 1 (2 integrated firms) disproportionately shrinks the inter-quartile range-- relatively more harmful mergers are eliminated than beneficial ones-- but the presence of a single integrated 3rd party has little impact on median harm.
 Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.
@@
 Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
-A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues.
+A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare increases in about 18\% of mergers and total welfare increases in 5.7\%, but median consumer harm increases to 14\% and median total harm increases to 7.3\% of pre-merger revenues.
 Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
-Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm.
-With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 20\% of markets and shrinks median consumer harm to about 6\% of pre-merger revenues.
-By contrast, with 3 integrated 3rd parties, the inter-quartile range shrinks, and median total harm falls to about 3.7\% of pre-merger revenues.
+Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm t 5.6\%.
+With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues.
+By contrast, with 3 integrated 3rd parties, the inter-quartile range of total harm shrinks, and median total harm falls to about 3.8\% of pre-merger revenues.
+These results indicate that unlike either vertical upstream mergers, integrated downstream mergers are more harmful than unintegrated downstream mergers, and that the presence of additional integrated rival incumbents reduces harm, though slowly.
 
-Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms harm consumers in about 96\% of simulated markets and lowers total surplus in more than 92\% of simulated markets.
-While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.
+Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms benefits consumers in about 7\% of simulated markets and is net beneficial in 15\% of simulated markets.
+While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of either outcome, adding a second integrated rival (4 integrated firms) eliminates the most harmful integrated mergers while also increasing median consumer harm to 14\%  and median total harm to 10.6\%.
+The presence of additional integrated incumbents has little incremental effect on  median harm.
 
@@
 
-Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the number of incumbent integrated firms increases from 0 to 1 and finally 4 incumbent vertically integrated firms.
-Plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}.
-For downstream (left) and upstream (middle) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
-By contrast, for vertical mergers (right), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter increases from 1/9 (retailers have the advantage to 9 (wholesalers have the advantage).
+also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue)incumbent vertically integrated firms.
+As in Figure \ref{fig:CVvertincumbBW_updownBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}.
+For vertical mergers (left), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
+By contrast, for upstream (middle) and downstream (left) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
 Finally, plots with 4 integrated firms increases the number of rival incumbent integrated firms in upstream and downstream mergers to 3 and in vertical mergers to 4 .
+
+For vertical mergers, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers.
+Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power.
+Across all three incumbent integrated firm scenarios, median consumer harm decreases from about 7\% when retailers have the advantage to -24\% when wholesalers have the advantage.
+Similarly, median net harm decreases from about 2.9\% of pre-merger expenditures when retailers have relatively more bargaining power to about -14\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+
+Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.
+Median consumer harm decreases from 2.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power.
+By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus.
+Median consumer harm decreases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power.
+Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.
 
@@
 Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -1.7\% of pre-merger revenues, while median total harm falls from 5.3\% to 1.9\% of pre-merger revenues.
-
-Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.
-Median consumer harm increases from about 2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power.
-By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus.
-Median consumer harm increases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power.
-Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.
-
-Unlike downstream or upstream mergers, where the presence of vertically integrated incumbent firms can affect either the magnitude or direction of the relationship between bargaining power and harm, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers.
-Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power.
 

### 2021-04-30 doc/VerticalArrangements.tex
added additional updates

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Empirical Setting}
+
+The U.S.
+waste and recycling industry generates approximately \$80 billion in annual revenues.
+In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors.
+However, solid waste companies tend not to be homogenous in their degree of vertical integration across geographies.
+As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevent local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.
+
+The vertical supply chain in the solid waste industry is primarily comprised of waste collection operations or ``haulers'' and waste disposal facilities.
+Haulers collect municipal solid waste (MSW) from businesses and residences and must dispose of it at a lawful disposal site, predominantly landfills.
+Waste disposal (upstream) is a required input into waste collection services (downstream).
+Some haulers are vertically integrated and operate their own disposal facilities.
+Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
+Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens’ waste themselves.
+Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
+Thus, mergers that combine hauling and disposal assets may incentivize the merged entity to raise its hauling rivals’ cost of disposal in order to benefit its own collection operations.
+Whether or not the merged entity is both able and incentivized to undertake such action depends upon the extent of its market power in the local disposal market, the merging parties’ profit margins in each line of business and their intensity of hauling competition with prospective disposal customers.
+
+In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.
+GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.
+Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021.
+These three merged companies along with Waste Connections, the third largest waste management company, are estimated to contol over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
+
+Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
+Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestitutre from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
+and Plaintiff States v.
+Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}} In the DOJ complaint filed in the Republic-Santek case, vertical competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{\tiny{See U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}}
+
+Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets - in the presence of differing numbers of existing integrated competitors to analyze the wefare effects from mergers with complex vertical arrangements.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-05-02 doc/VerticalArrangements.tex
finished drafts of simulation sections as well as conclusion

@@
 
-As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and abput 15\% of simulated integrated mergers.
+As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and about 15\% of simulated integrated mergers.
 %\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e.
@@
 Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefits consumers to about 27\% but does not increase the percentage of net beneficial mergers.
-Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of netbeneficial mergers to about 11\%.
+Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of net beneficial mergers to about 11\%.
 Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.
@@
 
-Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.
-Median consumer harm decreases from 2.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power.
+Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm.
+Median consumer harm increases from 2.2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 29\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 12\% when wholesalers have relatively more bargaining power.
 By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus.
-Median consumer harm decreases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power.
-Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.
+Median consumer harm decreases from about 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power to about -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -8.4\% when wholesalers have relatively more bargaining power.
+Adding 2 additional incumbent integrated firms (4 integrated firms) mitigates some of the adverse effects of mergers while still preserving the positive relationship between bargaining power and harm.
 
@@
 Median consumer harm decreases from about 18\% of pre-merger expenditures when retailers have relatively more bargaining power to about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm decreases from 7.5\% of pre-merger expenditures when retailers have relatively more bargaining power to 2\% when wholesalers have relatively more bargaining power.
+Likewise, median total harm decreases from 7.6\% of pre-merger expenditures when retailers have relatively more bargaining power to about 2\% when wholesalers have relatively more bargaining power.
 By contrast, mergers between an integrated and unintegrated upstream supplier (1 integrated firms), causes the box and whisker plots to rotate clockwise around equal bargaining power, strengthening the positive relationship between bargaining power and both consumer and total surplus.
-Increasing wholesaler bargaining power causes median consumer harm to fall from 25\% of pre-merger revenues to -11\% of pre-merger revenues, while median total harm falls from 12.5\% to 2\% of pre-merger revenues.
+Increasing wholesaler bargaining power causes median consumer harm to fall from 25\% of pre-merger revenues to -33\% of pre-merger revenues, while median total harm falls from 12.2\% to about 1\% of pre-merger revenues.
 Adding additional rival incumbents (4 integrated firms) has the reverse effect, rotating the box and whisker plots counter-clockwise and weakening the still positive relationship between bargaining power and surplus.
-Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -1.7\% of pre-merger revenues, while median total harm falls from 5.3\% to 1.9\% of pre-merger revenues.
+Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -33\% of pre-merger revenues, while median total harm falls from 5.5\% to 3.99\% of pre-merger revenues.
 
@@
 Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
-Median consumer harm increases from about 26\% of pre-merger expenditures when retailers have relatively more bargaining power to about -1.4\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from 16\% of pre-merger expenditures when retailers have relatively more bargaining power to -13\% when wholesalers have relatively more bargaining power.
+Median consumer harm increases from about 26\% of pre-merger expenditures when retailers have relatively more bargaining power to about 3.75\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Likewise, median total harm increases from 16\% of pre-merger expenditures when retailers have relatively more bargaining power to -14.5\% when wholesalers have relatively more bargaining power.
 
@@
 In theory, constant marginal costs can increase the magnitude of EDM, as the wholesaler price reduction applies equally to all units produced by the retailer.
-By contrast, if either the integrated wholesaler or retailer face linear marginal costs, increased output from the EDM is limited by the increased marginal costs, which should limit the benefit to the parties.
-Likewise, in theory, RRC could be more profitable for the merging parties under linear than constant marginal costs, as linear marginal costs cause the integrated wholesaler's price to increase more rapidly.
+By contrast, if either the integrated wholesaler or retailer face linear marginal costs, increased output from the EDM is limited by the increased marginal costs, which in turn limits the benefit to the parties.
+Likewise, in theory, RRC could be more profitable for the merging parties under linear than constant marginal costs, as linear marginal costs cause the integrated wholesaler's prices to increase more rapidly.
 
 Because EDM is plausibly greater under constant marginal costs while RRC is also plausibly less, then one might expect that on net mergers are more beneficial under constant than linear marginal costs.
-Figure \ref{fig:surplussumcost} displays box and whisker plots that depict the net effect of these different marginal cost specifications on consumer and total welfare under six different scenarios:
-linear marginal costs for the non-merging parties products (``Party Constant", dark blue), constant marginal costs for all products (``Constant", orange), linear marginal costs for retailers (``Constant/Linear", light grey), linear marginal costs for wholesalers (``Linear/Constant", dark grey), linear marginal costs for the merging parties products (``Party Linear",light blue) and linear marginal costs for all products (``Linear'',red).
-These six scenarios are ordered from least to most consumer harm in vertical mergers.
+We explore this hypothesis in Figure \ref{fig:surplussumcost}, which displays box and whisker plots that depict the net effect of these different marginal cost specifications on consumer and total welfare under four different scenarios:
+constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).
 
-Overall, consumer harm is less prevalent for mergers where firms have constant marginal costs then when firms have linear costs.
-Mergers with all constant-cost firms (orange) benefit consumers in about 15\% of all mergers, while mergers with all linear-cost firms benefit consumers in only 2.3\% of mergers.
-Indeed, consumer harm under the all linear marginal costs scenario first-order stochastically dominates consumer harm under the all constant marginal costs scenario.
+Overall, harm is less prevalent for mergers where firms have constant marginal costs than where firms have linear costs.
+Mergers with constant-cost firms increases consumer surplus  in about 15\% of all mergers and total surplus in 10.5\% of all mergers, while mergers with linear-cost firms increase consumer surplus in only 2.3\% of mergers and total surplus in 0.8\% of all mergers.
+However, while the consumer and total harm distributions for each merger type under the all-firm linear marginal costs scenario first-order stochastically dominates their counterparts under the all-firm constant marginal costs scenario, the consumer and total harm distributions under the party linear marginal costs scenario only first-order stochastically dominates the their counterparts under the party constant marginal costs scenario for vertical and integrated mergers.
 The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.
 
-The biggest differences in outcomes occurs for vertical mergers.
-when all firms have constant marginal costs benefit consumers in about 30\% of all simulations , but only in about 2.3\% of simulations when all firms have linear costs (light blue).
+Changing the cost structure consistently yields the greatest differences in outcomes for vertical mergers.
+When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of all vertical merger simulations and total surplus in about 18\% of all vertical merger simulations, but only increase consumer surplus in about 1\% of vertical merger simulations and total surplus in about 0.9\% of vertical merger simulations when all firms have linear costs.
+Moreover, qualitatively similar results hold for vertical mergers when only the merging parties' products have constant marginal costs, compared to when only the merging parties' products have linear costs.
+
+The differences between the constant and linear cost structures is more muted for both upstream and downstream mergers.
 Upstream mergers with all constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs.
-Downstream mergers with constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when firms have linear costs.
-Finally, integrated  mergers benefit consumers in about 10\% of all simulations when firms have constant costs, but less than 1\% of simulations when firms have linear costs.
+Downstream mergers with all constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when all firms have linear costs.
+Again, qualitatively similar results hold for vertical mergers when only the merging parties' products have constant marginal costs, compared to when only the merging parties' products have linear costs.
+
+Finally, while there is a marked difference in outcomes between the constant and linear cost structures for integrated  mergers under the all-firms scenario, the differences in outcomes are more muted under the party scenario, especially for consumer surplus.
+In particular, median consumer harm when all firms face constant marginal costs is 13.6\%, compared to the median consumer harm of 17.5\% when all firms face linear marginal costs.
+By contrast, median consumer harm when only the merging parties' have constant costs is 15.1\%, compared to the median consumer harm of 16.5\% when the parties' have linear costs.
+%Integrated mergers benefit consumers in about 10\% of simulations when all firms have constant costs, but less than 1\% of simulations when all firms have linear costs.
 
@@
 In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors.
-However, solid waste companies tend not to be homogenous in their degree of vertical integration across geographies.
-As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevent local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.
+However, solid waste companies tend not to be homogeneous in their degree of vertical integration across geographies.
+As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevant local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.
 
@@
 Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021.
-These three merged companies along with Waste Connections, the third largest waste management company, are estimated to contol over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
+These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
 
 Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
-Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestitutre from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
+Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
 and Plaintiff States v.
@@
 
-Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets - in the presence of differing numbers of existing integrated competitors to analyze the wefare effects from mergers with complex vertical arrangements.
+Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets - in the presence of differing numbers of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
+
+\section{Conclusion}
+
+This paper relaxes two strong assumptions made in \citeN{ST2020}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated, and that all firms employ constant marginal cost technology.
+Relaxing these constraints is important as competition authorities are called upon to investigate mergers where either the merging parties or a third party are already vertically integrated, or where the presence of a scarce input could reduce the benefit from EDM or increase the harm from RRC.
+Here, we show using both numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained.
+We also show using simulation that relative bargaining power can serve as a useful indicia for predicting consumer and total harm from mergers where these two assumptions are relaxed.
+Finally, we intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm.
+Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2020}, such as how the presence of either two-part tariffs or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
 

### 2021-05-02 code/genfigures.R

@@ -50,8 +50,11 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 
 
 
-psummary_cost.bw <-  ggplot(filter(res.nests.all,Outcome %in% c("Consumer","Total")) %>%
-                              mutate(Cost=reorder(Cost,ifelse((Outcome=="Consumer") & (Merger=="Vertical"),-Outcome_value/mktrev.pre,NA),quantile,probs=.5,na.rm=TRUE)),
+psummary_cost.bw <-  ggplot(filter(res.nests.all,Outcome %in% c("Consumer","Total") & !Cost %in% c("Constant/Linear","Linear/Constant")) %>%
+                              mutate(isParty=factor(grepl("Party",as.character(Cost)),labels=c("All","Party")),
+                                     Cost=reorder(gsub("Party ","",as.character(Cost)),ifelse((Outcome=="Consumer") & (Merger=="Vertical"),-Outcome_value/mktrev.pre,NA),quantile,probs=.5,na.rm=TRUE),
+
+                                     ),
                             aes(y=Outcome_value/mktrev.pre*100,
                                             #avgpricedelta/mktrev.pre*100,
                                             x=Merger,color=Cost
@@ -67,7 +70,7 @@ psummary_cost.bw <-  ggplot(filter(res.nests.all,Outcome %in% c("Consumer","Tota
   #theme_tufte(ticks=FALSE) +
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
-  facet_grid(~Outcome,scales="free",labeller = "label_context")+
+  facet_grid(isParty~Outcome,scales="free",labeller = "label_context")+
   xlab("Change in Surplus")+
   ylab("Outcome (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
@@ -743,7 +746,6 @@ dev.off()
 png("output/surplussum_cost.png",width = 10, height = 7, units = "in", res=300)
 print(psummary_cost.bw)
 dev.off()
-
 png("output/surplussum.png",width = 10, height = 7, units = "in", res=300)
 png("output/CVvertincumbBW.png",width = 10, height = 7, units = "in", res=300)
 print(pvertincumb.bw)

### 2021-05-02 doc/VerticalArrangements.tex
teaked cost graph so that it only summarized all firms and party cost changes. also changed the figure desciption in text

@@
 \includegraphics[scale=0.9]{../output/surplussum_cost.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 6 different cost scenarios.
-Dark blue boxes (``Party Constant") assume that the merging parties face constant marginal costs while other firms face linear marginal costs.
-Orange boxes (``Constant") assume that all firms face constant marginal costs.
-Light grey boxes (``Constant/Linear") assume upstream firms have constant marginal costs, while downstream firms have linear marginal costs.
-Orange box depicts the effects assuming that firms face linear marginal costs.
+\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
+The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue,left) or linear marginal costs (orange, right).
+The boxes in the bottom row (``Party") eitehr assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.

### 2021-05-02 code/genfigures.R

@@ -70,7 +70,7 @@ psummary_cost.bw <-  ggplot(filter(res.nests.all,Outcome %in% c("Consumer","Tota
   #theme_tufte(ticks=FALSE) +
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
-  facet_grid(isParty~Outcome,scales="free",labeller = "label_context")+
+  facet_grid(isParty~Outcome,scales="fixed",labeller = "label_context")+
   xlab("Change in Surplus")+
   ylab("Outcome (%)")+
   #ylab("Avg. Downstream Price Change (%)")+

### 2021-05-02 doc/VerticalArrangements.tex
read through my sections and incoprated nits. Fixed some figure captions.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, by the Federal Trade Commission, or by its Commissioners.
+\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, or by the Federal Trade Commission.
 Furthermore, the views expressed here should not be purported to reflect those of the U.S.
@@
 In addition, we consider upstream and downstream mergers between either two unintegrated firms or between one vertically integrated and one unintegrated firm.
-Finally, we assume that an integrated merger occurs between two vertically integrated firms\textbf{consider adding diagonal mergers by allowing for heterogeneity in the bargaining parameter.}.
+Finally, we assume that an integrated merger occurs between two vertically integrated firms
+
+%\textbf{consider adding diagonal mergers by allowing for heterogeneity in the bargaining parameter.}.
 
 We simulate markets by randomly sampling shares from a Dirichlet distribution for 2, 3, 4, or 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.
-}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7.
+}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for these simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7.
 We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
@@
 We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.
-The median change is negative, indicating harm, across all four types of mergers for both cost specifications.
+The median change is negative, indicating harm, across all four types of mergers.
 However, the distributions and magnitudes differ.
@@
 
-In terms of total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers with constant costs that are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial.
+In terms of total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial.
 Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers.
@@
 \subsection*{Vertically Integrated Incumbent Firms}
-Here, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
+Here, we examine merger outcomes when a vertically integrated firm merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
 We also examine merger outcomes for  mergers as the number of incumbent integrated non-merging rivals increases.
 
-Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms.
-The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020} and is included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merging with an unintegrated retailer, a single unintegrated 3rd party is integrated, and the remaining 3rd parties (if any) are unintegrated.
-By contrast,for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
+Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms.
+The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020} and is included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction.
+\citeN{ST2020} also do not consider mergers between integrated firms.
+}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single unintegrated 3rd party is integrated, and the remaining 3rd parties (if any) are unintegrated.
+By contrast, for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
 Finally, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, again with no integrated rivals.
@@
 The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 25\% and the percentage of mergers that increase  total surplus to 16\%, but leads median harm largely unchanged.
-Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefits consumers to about 27\% but does not increase the percentage of net beneficial mergers.
-Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of net beneficial mergers to about 11\%.
-Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.
+Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefit consumers to about 27\% but does not increase the percentage of net beneficial mergers.
+Adding a fourth integrated incumbent further increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of net beneficial mergers to about 11\%.
+Taken together, Figure \ref{fig:CVvertincumbBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.
 
@@
 Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter increases from 1/9 (retailers have the advantage to 9 (wholesalers have the advantage).
-also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue)incumbent vertically integrated firms.
-As in Figure \ref{fig:CVvertincumbBW_updownBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}.
+also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) incumbent vertically integrated firms.
+As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}.
 For vertical mergers (left), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
@@
 Across all three incumbent integrated firm scenarios, median consumer harm decreases from about 7\% when retailers have the advantage to -24\% when wholesalers have the advantage.
-Similarly, median net harm decreases from about 2.9\% of pre-merger expenditures when retailers have relatively more bargaining power to about -14\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+Similarly, median total harm decreases from about 2.9\% of pre-merger expenditures when retailers have relatively more bargaining power to about -14\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
 
@@
 Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 12\% when wholesalers have relatively more bargaining power.
-By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus.
+By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firm), there is a positive relationship between bargaining power and both consumer and total surplus.
 Median consumer harm decreases from about 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power to about -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
@@
 
-\textbf{Describe the counterintuitive result that increasing retailer bargaining power harms retailers but help wholesalers.}
+%\textbf{Describe the counterintuitive result that increasing retailer bargaining power harms retailers but help wholesalers.}
 
@@
 Relaxing these constraints is important as competition authorities are called upon to investigate mergers where either the merging parties or a third party are already vertically integrated, or where the presence of a scarce input could reduce the benefit from EDM or increase the harm from RRC.
-Here, we show using both numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained.
+Here, we show using numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained.
 We also show using simulation that relative bargaining power can serve as a useful indicia for predicting consumer and total harm from mergers where these two assumptions are relaxed.
 Finally, we intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm.
-Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2020}, such as how the presence of either two-part tariffs or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
+Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2020}, such as: how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
 
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus.
-Each blue box depicts the effects assuming that firms face constant marginal costs, while each orange box depicts the effects assuming that firms face linear marginal costs.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
@@
 \includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (top row) and total (bottom) surplus as the number of vertically integrated firms present in a market change.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
@@
 The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue,left) or linear marginal costs (orange, right).
-The boxes in the bottom row (``Party") eitehr assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).
+The boxes in the bottom row (``Party") either assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.

### 2021-05-03 doc/VerticalArrangements.tex
incorporated gloria's 2nd round of edits

--- 2021-05-02 doc/VerticalArrangements.tex
@@
 Compared to equation \eqref{eq: vmerger ret upstream foc}, now the EDM recapture effect applies to sales of product $w$ at both affiliated retailers $r$ and $s$.
-Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $r$ lose access to product $v$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture effect.
-Both of these additional terms tend to increase the bargaining leverage of the merged firm.
+Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $r$ lose access to product $v$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture leverage effect.
+All of these additional terms tend to increase the bargaining leverage of the merged firm.
 If we instead examined a merger between retailer/wholesaler $rw$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $r$ for two merged wholesalers, rather than sales through two merged retailers.
@@
 For a merger between two integrated retailer/wholesaler firms, all of these effects would enter.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-05-05 doc/VerticalArrangements.tex
incorporated Peggys edits in response to EAG clearance (Russ)

@@
 
+\interfootnotelinepenalty=10000
+
 \parskip     2.0mm       % space between paragraphs
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, or by the Federal Trade Commission.
-Furthermore, the views expressed here should not be purported to reflect those of the U.S.
+\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
+Furthermore, the views expressed here should not be purported to reflect those of the Federal Trade Commission or the U.S.
 Department of Justice.}}
@@
 \section{Introduction}
-Given the many different ways a firm may be organized, it is rare that any given merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  Indeed, as the 2020 U.S.
+Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  Indeed, as the 2020 U.S.
 Department of Justice (DOJ) and Federal Trade Commission (FTC) Vertical Merger Guidelines (henceforth, the ``VMG'') state, mergers often have both horizontal and vertical aspects.
@@
 The U.S.
-waste and recycling industry generates approximately \$80 billion in annual revenues.
-In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors.
+waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}}} In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors.
 However, solid waste companies tend not to be homogeneous in their degree of vertical integration across geographies.
@@
 
-In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.
-GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.
-Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021.
-These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
+In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{\tiny{Competitive Impact Statement: U.S.
+and Plantiff States v.
+Waste Management, Inc.
+and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{\tiny{Waste 360, \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021.\footnote{\tiny{Competitive Impact Statement: U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}} These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
 

### 2021-05-13 doc/VerticalArrangements.tex
minor changes from peggy

--- 2021-05-03 doc/VerticalArrangements.tex
@@
 
-Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
+Concerns about vertical competitive effects were raised by the reviewing agency in the Republic-Santek transaction and by public commenters in the Waste Management-ADS transaction.
 Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.

### 2021-05-14 doc/VerticalArrangements.tex
this version cleared by FTC AND DOJ

--- 2021-05-05 doc/VerticalArrangements.tex
@@
 Department of Justice (DOJ) and Federal Trade Commission (FTC) Vertical Merger Guidelines (henceforth, the ``VMG'') state, mergers often have both horizontal and vertical aspects.
-However, the typical theories and models used to analyze mergers rarely address these complexities.
+However, the theories and models used to analyze mergers often fail to address these complexities.
 Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), while the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
@@
 \centering
-\includegraphics[scale=0.9]{../output/surplussum.png}
+\includegraphics[scale=0.9]{surplussum.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus.
@@
 \centering
-\includegraphics[scale=0.9]{../output/CVvertincumbBW.png}
+\includegraphics[scale=0.9]{CVvertincumbBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (top row) and total (bottom) surplus as the number of vertically integrated firms present in a market change.
@@
 \centering
-\includegraphics[scale=0.9]{../output/CVbargnobothBW.png}
+\includegraphics[scale=0.9]{CVbargnobothBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which downstream, upstream, and vertical mergers affect consumer, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \centering
-\includegraphics[scale=0.85]{../output/CVbargbothBW.png}
+\includegraphics[scale=0.85]{CVbargbothBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \centering
-\includegraphics[scale=0.9]{../output/surplussum_cost.png}
+\includegraphics[scale=0.9]{surplussum_cost.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.

### 2021-06-03 code/RunSims.R
added codes to simulate the effects of PARTY bargaining power changing, holding non-party bargaining power fixed. Graphed results.

@@ -30,7 +30,7 @@ clusterEvalQ(cl,{
              if(require(RevoUtilsMath)) setMKLthreads(1)
 })
 
-samples <- 1e3
+samples <- 1e3 #1e3
 ndfirms <- 2:5
 nufirms <- ndfirms
 nvfirms <- 0:4
@@ -38,13 +38,15 @@ nestParm <- c(0)
 type=c("1st")
 merger=c("up","down", "vertical","both")
 bargparm <- seq(.1,.9,.1)
+assym <- c("none","rival")
 mc=c("constant","linprod","lincons","conslin","consparty","linparty"#, "quadprod","linquad","quadlin"#,"linfirm"
      )
+
 relleveragePre <-  (1-bargparm)/bargparm
 
 
 
-genMkts <- function(x,y,z,t,m,n, b,c, large=TRUE, bargpreset="none"){
+genMkts <- function(x,y,z,t,m,n, b,c,a, large=TRUE){
 
 
 
@@ -64,7 +66,7 @@ genMkts <- function(x,y,z,t,m,n, b,c, large=TRUE, bargpreset="none"){
                                                       ownerPost = m,
                                                       M=1,cost_type=as.character(c),
                                                       largeMerger = large,
-                                                      bargpreset= bargpreset)
+                                                      bargpreset = as.character(a))
                  ,silent=TRUE)
 
 
@@ -106,6 +108,7 @@ thisres <- with(thissum,data.frame(
   nestParm=n,
   merger=m,
   barg=b,
+  preset=a,
   mc=c,
   M=M,
   outMargin=outMargin,
@@ -153,7 +156,7 @@ return(list(
   summary=thissum,
   res=unique(thisres)))}
 
-repMkts <- function(x,y,z,t,m,n,b,c,large=TRUE){replicate(samples, genMkts(x,y,z,t,m,n,b,c,large),simplify = "list")}
+repMkts <- function(x,y,z,t,m,n,b,c,a,large=TRUE){replicate(samples, genMkts(x,y,z,t,m,n,b,c,a,large),simplify = "list")}
 
 clusterExport(cl,varlist = setdiff(ls(),c("cl")))
 
@@ -170,7 +173,8 @@ res.nests <- expand.grid(up=ndfirms,
                          barg = bargparm,
                          type=type,
                          merger=merger,
-                         mc=mc)
+                         mc=mc,
[…]

### 2021-06-03 code/genfigures.R

@@ -21,7 +21,7 @@ relleverage=sort(unique(res.nests$relleveragePre),decreasing = TRUE)
 
 
 ## Only include constant cost sims
-res.nests.all  <- mutate(ungroup(res.nests.long),
+res.nests.allcost  <- mutate(ungroup(res.nests.long),
                                             Merger=factor(merger,levels=c("vertical","up","down","both"),labels=c("Vertical","Upstream","Downstream","Integrated")),
                                             Cost=factor(mc,
                                                         levels=c("constant","linprod","lincons","conslin","consparty","linparty"),
@@ -32,7 +32,20 @@ res.nests.all  <- mutate(ungroup(res.nests.long),
                                             Cost=reorder(Cost,-Outcome_value/mktrev.pre,median,na.rm=TRUE)) %>% select(-mc)
 
 
-res.nests.logit <- res.nests.long <- res.nests.all %>% filter(Cost=="Constant")
+res.nests.logit <- res.nests.long <- res.nests.allcost %>% filter(Cost=="Constant")
+
+
+res.nest_rival.allcost  <- ungroup(res.nest_all.long) %>% filter(preset=="rival") %>% mutate(
+                             Merger=factor(merger,levels=c("vertical","up","down","both"),labels=c("Vertical","Upstream","Downstream","Integrated")),
+                             Cost=factor(mc,
+                                         levels=c("constant","linprod","lincons","conslin","consparty","linparty"),
+                                         labels=c("Constant"
+                                                  ,"Linear","Linear/Constant","Constant/Linear","Party Constant","Party Linear"# "Quadratic","Linear/Quadratic","Quadratic/Linear","Linear/Constant","Constant/Linear"
+                                         )
+                             ),
+                             Cost=reorder(Cost,-Outcome_value/mktrev.pre,median,na.rm=TRUE)) %>% select(-mc)
+
+
 
 #res.nests.logit <- filter(as.data.frame(res.nests.logit),nestParm == "0" & Retailers !="1" & Wholesalers !="1") %>%
 #  mutate_if(is.factor,droplevels)
@@ -50,7 +63,7 @@ boxfun <- function(x,probs=c(.05,.25,.5,.75,.95)){
 
 
 
[…]

### 2021-06-04 code/MonteCarlo.R
reran with two new preset options 'diag0' and diag'1'

@@ -16,11 +16,12 @@ market<-function(nfirms.down=3, # # of downstream firms
                  nprods.up=nprods.down, # # of upstream products
                  nfirms.vert=0, # number of integrated firms pre-merger
                  bargparm=runif(nfirms.down*nfirms.up, .25,.75), # bargaining parameters for each upstream/downstream firm pair
-                 bargpreset=c("none","party","rival","diagonal"), # fix some bargaining parameters
+                 bargpreset=c("none","party","rival","diag1","diag0"), # fix some bargaining parameters
                  # none: do not modify bargparm
                  # party: set party bargparm to 0.5, allow others to vary
                  # rival: set non-party bargparm to 0.5, allow others to vary
-                 # diagonal: Vertical only. set party bargparm to 1 ,allow others to vary
+                 # diag1: Vertical only. set party pre-merger bargparm to 1 ,allow others to vary
+                 # diag0: Vertical only. set party pre-merger bargparm to 0 ,allow others to vary
                  mcshare.up=runif(nfirms.up*nprods.up), # share of upstream costs
                  mcshare.down=runif(nfirms.down*nprods.down), # share of wholesale price
                  ownerPost = c("up","down","vertical","both"), # simulate an upstream horizontal merger, downstream horizontal merger,
@@ -141,7 +142,8 @@ market<-function(nfirms.down=3, # # of downstream firms
                      none=bargparm,
                      party=ifelse(isParty.up |isParty.down,0.5,bargparm),
                      rival=ifelse(!(isParty.up |isParty.down),0.5,bargparm),
-                     diagonal=ifelse(isParty.up & isParty.down,1,bargparm)
+                     diag1=ifelse(isParty.up & isParty.down,1,bargparm),
+                     diag0=ifelse(isParty.up & isParty.down,0,bargparm)
   )
 
   ## when OwnerPost equals "vertical"

### 2021-10-15 presentations/FTC/BargainingSimsSlides.tex
first draft of presentation for FTC brownbag

%\documentclass[handout]{beamer}
\documentclass{beamer}
\usepackage{hyperref}   % displays URLs
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{mathtools,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{setspace}   % controls line spacing
\usepackage{caption}    % controls  captions
\usepackage{bibentry}   %cite without bibliography
%\usepackage{handoutWithNotes}
%\setbeamertemplate{note page}[default]
%\setbeameroption{show notes on second screen=right}
%\setbeameroption{show only notes}
%\insertslideintonotes{0.25}
%\pgfpagesuselayout{4 on 1}[letterpaper,border shrink=5mm]

\usetheme[subsectionpage=progressbar, numbering=fraction]{metropolis}

%\setbeamertemplate{footnote}{%
%  \hangpara{2em}{1}%
%  \makebox[2em][l]{\insertfootnotemark}\footnotesize\insertfootnotetext\par%
%}
\renewcommand\thempfootnote{\arabic{mpfootnote}} % return footnotes numbers to numerals

% shrink bibliographic text
%\renewcommand{\bibfont}{\scriptsize}
\setbeamerfont{bibliography item}{size=\footnotesize}
\setbeamerfont{bibliography entry author}{size=\footnotesize}
\setbeamerfont{bibliography entry title}{size=\footnotesize}
\setbeamerfont{bibliography entry location}{size=\footnotesize}
\setbeamerfont{bibliography entry note}{size=\footnotesize}

%\author[Sheu and Taragin]{Gloria Sheu\footnote{ The views expressed herein are entirely those of the authors and should not be purported to reflect those of the U.S. Department of Justice.}\and Charles Taragin\footnotemark[1]}
\author[Sheu et. al.]{Gloria Sheu\textsuperscript{1} \and Charles Taragin\inst{2} \and Margaret Loudermilk\inst{3}}
\institute[]{\textsuperscript{1} Federal Reserve Board of Governors \and \inst{2}Federal Trade Commission \and \inst{3}U.S. Department of Justice}
\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration}
\date{October 2021}

%\author[Sheu and Taragin]{Gloria Sheu\and Charles Taragin}

\addtobeamertemplate{title page}{\footnotesize The views expressed herein are entirely those of the authors and should not be purported to reflect those of the U.S. Department of Justice.}{}
\beamertemplatenavigationsymbolsempty
%\beamerdefaultoverlayspecification{<+->}

%\usepackage{Sweave}

\begin{document}

%\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=2em}
%\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=2em}
%\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
%\fvset{listparameters={\setlength{\topsep}{0pt}}}
%\renewenvironment{Schunk}{\vspace{\topsep}}{\vspace{\topsep}}

%\AtBeginSection[]
%{
%\begin{frame}{Table of Contents}
%\tableofcontents[currentsection]
%\end{frame}
%}

% \AtBeginSection[]{
%   \begin{frame}
%   \vfill
%   \centering
%   \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
%     \usebeamerfont{title}\insertsectionhead\par%
%   \end{beamercolorbox}
%   \vfill
%   \end{frame}
% }
%
% \def\SecT#1{\begin{frame}[plain]
% \begin{block}{}
% %\centering\huge\thesection. \insertsubsection
% \centering\huge\insertsection
% \end{block}
% \vfill\Large #1
% \end{frame}}

%\begingroup
%\setbeamertemplate{footline}{\vspace*{-0.5cm}The views expressed herein are
%    entirely those of the authors\\ and should not be purported to
%    reflect those of the U.S. Department of Justice.\par}

\begin{frame}
\titlepage
\end{frame}
%\frame{\titlepage}
%\endgroup

%\part{Introduction}
%\frame{\partpage}
\section{Introduction}

\begin{frame}
\frametitle{Mergers in Supply Chains Are Complex}

\begin{itemize}
\item literature has largely focused on mergers between unintegrated firms with unintegrated rivals. What about
\begin{itemize}
\item mergers where \textbf{one} party is already vertically integrated and the other is not?
\item mergers where \textbf{both} parties are already vertically integrated?
\item mergers where \textbf{3rd parties} are vertically integrated?
\end{itemize}
\item literature has focused on mergers between firms with \textbf{constant} marginal costs and no capacity constraints.

\begin{itemize}
\item constant marginal costs can understate raising rivals costs
\item constant marginal costs can overstate EDM
\end{itemize}

\end{itemize}
\end{frame}

% \begin{table}
% \centering
% \small
% \begin{tabular}{|r|c|c|c|c|}\hline
% Matter & Industry& Year & Supply & Demand \\\hline
% HR Block/TaxAct &Tax Prep.& 2011& Bertrand & Linear\\
% Sysco/US Foods& Food Svc. Dist.&2015 &2nd Score & Logit\\
% Anthem/Cigna&Health Ins.&2017&2nd Score & Logit\\
% Aetna/Humana&Health Ins.&2017&Bertrand & Nested Logit\\\hline
% \end{tabular}
% \caption{Recent Matters that Employed Merger Simulation}
% \end{table}
% \end{frame}

\begin{frame}[shrink=1]
\frametitle{Our Contribution}

\begin{block}{Use \citeN{ST2021} to examine merger effects}
\begin{itemize}
\item when an integrated firm merges with an unintegrated upstream firm,
\item when an integrated firm merges with an unintegrated downstream firm,
\item when two integrated firms merge,
\item as the number of integrated rivals increases,
\item when marginal costs are non-constant.
\end{itemize}
\end{block}

\begin{block}{Case Studies: Solid Waste Management Mergers (In Progress)}
\begin{itemize}
\item Waste Management/Advanced Disposal Services (2020)
\item Republic/Santek (2020)
\end{itemize}
\end{block}
\end{frame}

 % \begin{frame}[shrink=10]
 % \frametitle{The Existing Literature}
 % Our paper combines methods from two literatures
 %
 % \begin{block}{Merger Simulation}
 % \begin{itemize}
 % \item Bertrand Pricing with Differentiated Products: \citeN{WF1994};
 % \item 2nd Score Auction with Differentiated Products: \citeN{Miller2014};
 % \item Bargaining with Differentiated Products: \citeN{Grennan2013}, \citeN{GNT2015}.
 % \end{itemize}
 % \end{block}
 %
 % \begin{block}{Vertical Supply Chain}
 % \begin{itemize}
 % \item Retailer/Wholesaler Relationships: \citeN{DKVB2010};
 % \item hospital-insurer contracting: \citeN{HL2017};
 % \item video content creation/distribution: \citeN{CY2012}, \citeN{CLWY2017}.
 % \end{itemize}
 % \end{block}
 %
 % \end{frame}
 %

\begin{frame}
\frametitle{Overview}
\tableofcontents
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Overview}

\begin{itemize}
  \item We model a two-level supply chain
  \begin{enumerate}
    \item Nash bargaining upstream
    \item Bertrand differentiated products logit downstream (\citeN{WF1994})
  \end{enumerate}
  \item Downstream model is purposefully kept in standard forms to facilitate calibration
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Downstream}
\begin{frame}
\frametitle{Demand}

\begin{itemize}
  \item Assume logit for downstream Bertrand
  \item Each wholesaler $w$ in the set $\mathbb{W}$ offers a single product to retailers $r$ in the set $\mathbb{R}$,
  which in turn sell to consumers
  \item Market share of wholesaler $w$'s product purchased through retailer $r$ is
  \begin{equation*}
  s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{t \in \mathbb{R}} \sum_{x \in \mathbb{W}^t} \exp(\delta_{tx}   - \alpha p_{tx})}
  \end{equation*}
  \item $\{\delta_{rw}; \forall r \in \mathbb{R}, \forall w \in \mathbb{W}\}$ and $\alpha$ are parameters to be calibrated
\end{itemize}

\end{frame}

\note{retailer is selling a number of bundled products whose quality can vary.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Bertrand}

\begin{itemize}
  \item Assume firms set prices simultaneously in Bertrand Nash equilibrium
  \item Retailer \textcolor{cyan}{constant}  marginal cost consists of wholesale fee $p^W_{rw}$ and other costs $c^R_{rw}$
  \item Profit function for retailer $r$ is
  \begin{equation*}
  \pi^r = \sum_{w \in \mathbb{W}^r} [p_{rw} - p^W_{rw} - c^R_{rw}] s_{rw} M
  \end{equation*}
  \item Resulting first order conditions determine downstream equilibrium prices
  \begin{equation*}
  \sum_{x \in \mathbb{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} = 0
  \end{equation*}
\end{itemize}

\end{frame}

\note{constant marginal costs assume a single unit of wholesaler product is used to produce a single unit of retailer product.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Upstream}
\begin{frame}
\frametitle{Nash Bargaining}

\begin{itemize}
  \item Assume a \textcolor{cyan}{linear} wholesale price set by Nash Bargaining
  \begin{equation*}
  \max_{p^W_{rw}} (\pi^r - d^r (\mathbb{W}^r \setminus \{w\}))^{\lambda_{rw}} (\pi^w - d^w (\mathbb{R}^w \setminus \{r\}
  ))^{1 - \lambda_{rw}}
  \end{equation*}
  \item $\{\lambda_{rw}; \forall r \in \mathbb{R}, \forall w \in \mathbb{W}\}$ are parameters to be calibrated
  \item Wholesaler marginal costs  are \textcolor{cyan}{constant} and equal $c^W_{rw}$
  \item Profit function for wholesaler $w$ is
  \begin{equation*}
  \pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M
  \end{equation*}
  \item All wholesalers and retailers reach an agreement in equilibrium.
\end{itemize}

\end{frame}

\note{
\begin{enumerate}
\item $\pi^r$: retailer profits from reaching agreement with wholesaler $w$
\item $d^r$: \textbf{Bargaining Leverage}. retailer profits if not agreement is reached and retailer discontinues sale of product with $w$'s input. Some sales will be recaptured from remaining retailer profits.
\item $\lambda_{rw}$: \textbf{Bargaining Power} parameter. Possibly reflects
\begin{itemize}
\item relative bargaining ability
\item relative impatience (e.g. cost of capital)
\item often assumed to be 0.5 but we will calibrate from data
\end{itemize}
\end{enumerate}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Simplifying Assumptions}

\begin{itemize}
  \item \only<1-5>{First} \only<6>{Simplified first} order condition is given by
  \only<1-5>{
  \begin{equation*}
  \begin{split}
  &\lambda_{rw}[\pi^w - d^w (\mathbb{R}^w \setminus \{r\})]
  \left( \only<1-4>{\frac{\partial \pi^r}{\partial p^W_{rw}}} \only<5>{\textcolor{blue}{-s_{rw}M}}- \only<1-2>{\frac{\partial d^r (\mathbb{W}^r \setminus \{w\})}{\partial p^W_{rw}}} \only<3-5>{\textcolor{orange}{0}}  \right) + \\
  &(1-\lambda_{rw})[\pi^r - d^r (\mathbb{W}^r \setminus \{w\})]
  \left(\only<1-4>{\frac{\partial \pi^w}{\partial p^W_{rw}}} \only<5>{\textcolor{blue}{s_{rw}M}} - \only<1-2>{\frac{\partial d^w(\mathbb{R}^w \setminus \{r\})}{\partial p^W_{rw}}}  \only<3-5>{\textcolor{orange}{0}} \right) = 0,
  \end{split}
  \end{equation*}}
  \only<6>{
  \begin{equation*}
  \pi^w - d^w (\mathbb{R}^w\setminus \{r\}) = \frac{1-\lambda_{rw}}{\lambda_{rw}} (\pi^r - d^r (\mathbb{W}^r \setminus      \{w\}))
  \end{equation*}}
  \item Use two simplifying assumptions
  \begin{enumerate}
    \item \uncover<2-6>{\textcolor{orange}{\textit{Simultaneous negotiations}}: treat other contracts as given}
    \item \uncover<4-6>{\textcolor{blue}{\textit{Simultaneous downstream pricing}}: treat downstream prices as given}
  \end{enumerate}
\end{itemize}

\end{frame}

\note{\textbf{ simultaneous negotiations} is reasonable if other contracts cannot be easily or quickly renegotiated. \textbf{ simultaneous downstream pricing} is reasonable if wholesalers do not have much insight into how downstream prices are set OR if downstream prices change at a high frequency than upstream contracts are negotiated.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Disagreement Payoffs}

\begin{itemize}
  \item Profit for wholesaler $w$ if it does not sell to retailer $r$ is
  \begin{equation*}
  d^w (\mathbb{R}^w \setminus \{r\}) = \sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] s_{tw}(\mathbb{W}^r   \setminus \{w\}) M
  \end{equation*}
  \item Profit for retailer $r$ if it does not buy from wholesaler $w$ is
  \begin{equation*}
  d^r (\mathbb{W}^r \setminus \{w\}) = \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}
  (\mathbb{W}^r \setminus \{w\}) M
  \end{equation*}
  \item Other things equal, a larger disagreement payoff increases bargaining leverage
\end{itemize}

\end{frame}

% \subsection{Calibration}
%
% \begin{frame}
% \frametitle{Downstream Parameters}
% \small
% Suppose we observe all shares and prices and at least one downstream product margin. Then we can recover $\{\delta_{rw}; \forall r \in \mathbb{R}, \forall w \in \mathbb{W}\}$ and $\alpha$ using the following equations
%
% \begin{align*}
% \ln(s_{rw})-\ln(s_{00}) &= \delta_{rw} - \alpha p_{rw},\\
% \intertext{\textbf{and} for Bertrand}
%  m^R_{rw} &= -\frac{1}{\alpha
%   \bigl(1 - \sum\limits_{x \in \mathbb{W}^r} s_{rx}\bigr)}\\
% \intertext{\emph{or} for Second Score}
% E \left[ m^R_{rw} \middle| rw \text{ wins} \right] &= -\frac{\ln
%   \bigl(1 - \sum\limits_{x \in \mathbb{W}^r} s_{rx} \bigr)}{\alpha\bigl( \sum\limits_{x \in \mathbb{W}^r} s_{rx} \bigr)}
% \end{align*}
%
% \end{frame}
%
%
% \begin{frame}
% \frametitle{Upstream Parameters}
% Further, suppose we observe wholesale margins and prices. Then using the predicted retail margins and shares from the downstream calibration we can recover $\{\lambda_{rw}; \forall r \in \mathbb{R}, \forall w \in \mathbb{W}\}$
%
% \begin{equation*}
% \begin{split}
% &m^W_{rw}s_{rw} - \sum_{t \in \mathbb{R}^w \setminus \{r\}} m^W_{tw} \Delta s_{tw}(\mathbb{W}^r \setminus \{w\}) = \\
% &\frac{1-\lambda_{rw}}{\lambda_{rw}} \left(\widehat{m^R_{rw}}s_{rw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} \widehat{m^R_{rx}} \Delta s_{rx}(\mathbb{W}^r \setminus \{w\})\right).
% \end{split}
% \end{equation*}
%
% \end{frame}

\subsection{Merger Simulation}

% \begin{frame}
% \frametitle{Downstream Horizontal Merger: Recapture Lost Sales}
% \small
% Suppose that retailer $r$ acquires retailer $s$. Downstream, $r$ ``recaptures'' sales lost to $s$ (and vice versa)
% \begin{block}{Bertrand}
% \begin{equation*}
% \sum_{x \in \mathbb{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}
% {\partial p_{rw}} + s_{rw}  +
% \textcolor{red}{\underbrace{ \sum_{v \in \mathbb{W}^s} [p_{sv} - p^W_{sv} - c^R_{sv}] \frac{\partial s_{sv}}{\partial p_{rw}}}_\text{$\uparrow$ Price}} = 0,
% \end{equation*}
% \end{block}
% \begin{block}{Second Score Auction}
% \begin{equation*}
%  E \left[ m^R_{rw} \middle| rw \text{ wins} \right] = -\frac{\ln
%   \bigl(1 - \sum\limits_{x \in \mathbb{W}^r} s_{rx} - \textcolor{red}{\overbrace{\sum_{v \in \mathbb{W}^s} s_{sv}}^\text{$\uparrow$ Price}}\bigr)}{\alpha\bigl( \sum\limits_{x \in \mathbb{W}^r} s_{rx} +  \sum_{v \in \mathbb{W}^s} s_{sv}\bigr)}
%   \end{equation*}
% \end{block}
% \end{frame}
%
% \note{
% \begin{itemize}
% \item Bertrand: new term is UPP
% \item 2nd Score: Dominant strategy to only biud highest value alternative. Avoids cannibalization
% \end{itemize}
% }
%
% \begin{frame}
% \frametitle{Downstream Horizontal Merger: Improved Bargaining Leverage}
% Suppose that retailer $r$ acquires retailer $s$. Upstream, $r$'s bargaining leverage improves. Post-merger first order condition is given by
%   \begin{align*}
%   \pi^w - d^w (\mathbb{R}^w\setminus \{r\}) &= \frac{1-\lambda^*_{rs,w}}{\lambda^*_{rs,w}} (\pi^r - d^r (\mathbb{W}^r \setminus      \{w\}) - \textcolor{red}{\underbrace{d^s (\mathbb{W}^r \setminus      \{w\})}_\text{$\uparrow$ Price}})\\
%   \intertext{Merger can also change retailer's bargaining power}
%   \lambda^*_{rs,w} &=\begin{cases}
%        \max\{\lambda_{rw}, \lambda_{sw}\} & \text{if synergy,} \\
%        \min\{\lambda_{rw}, \lambda_{sw}\} & \text{if dis-synergy,}\\
%        \lambda_{rw}                       & \text{if no change.}\\
%     \end{cases}
%   \end{align*}
% \end{frame}
%
%
%
% \begin{frame}
% \frametitle{Upstream Horizontal Merger: Improved Bargaining Leverage}
% Suppose that wholesaler $w$ acquires wholesaler $v$. Upstream, $w$'s bargaining leverage improves. Post-merger first order condition is given by
%   \begin{align*}
%   \pi^w - d^w (\mathbb{R}^w\setminus \{r\})- \textcolor{red}{\underbrace{d^v (\mathbb{R}^w \setminus      \{w\})}_\text{$\uparrow$ Price}} &= \frac{1-\lambda^*_{r, wv}}{\lambda^*_{r, wv}} (\pi^r - d^r (\mathbb{W}^r \setminus      \{w\}) )\\
%  \intertext{Merger can also change wholesaler's bargaining power}
%   \lambda^*_{r, wv} &=\begin{cases}
%        \max\{\lambda_{rw}, \lambda_{rv}\} & \text{if synergy,} \\
%        \min\{\lambda_{rw}, \lambda_{rv}\} & \text{if dis-synergy,}\\
%        \lambda_{rw}                       & \text{if no change.}\\
%     \end{cases}
%   \end{align*}
% \end{frame}
%

\begin{frame}
\frametitle{Integrated Firm: Own Retail Prices}
Assume that retailer $r$ acquires wholesaler $w$. $r$'s price downstream is determined by

\begin{equation*}
\begin{split}
\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} &+ s_{rw}\\
\textcolor{blue}{\underbrace{[p_{rw}-c^R_{rw}-c^W_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}_\text{EDM}} &+ \textcolor{red}{\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_\text{UPP}} = 0.
\end{split}
\end{equation*}

\end{frame}

\begin{frame}
\frametitle{Integrated Firm: Wholesale Price to Rival Retailer}
Assume that retailer $r$ acquires wholesaler $w$. $w$ bargains with \emph{unaffiliated} retailer $u$, yielding a wholesale price determined by

\tiny
\begin{equation*}
\begin{split}
&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
\end{split}
\label{eq: vmerger wh upstream foc}
\end{equation*}
\end{frame}

\begin{frame}
\frametitle{Integrated Firm: Wholesale Price from Wholesaler Rival}
Assume that retailer $r$ acquires wholesaler $w$. $r$ bargains with \emph{unaffiliated} wholesaler $v$, yielding a wholesale price determined by

\tiny
\begin{equation*}
\begin{split}
&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
&\left.- \underbrace{[p_{rw} - c^W_{rw} - c^R_{rw}] \Delta s_{rw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
\end{split}
\label{eq: vmerger ret upstream foc}
\end{equation*}
\end{frame}

\section{Numerical Simulations}

\begin{frame}
\frametitle{Objective}
The purpose of these numerical simulations is to use the above model to explore how these different mergers can affect market participants under a wide array of market conditions.

\end{frame}

\begin{frame}
\frametitle{Data Generating Process (1.4 million simulated mergers)}

\begin{itemize}
\item simulate markets with either 2, 3, 4 or 5 wholesalers/retailers
\item 0 to 4 \textbf{non-siloed} integrated incumbents pre-merger
\begin{itemize}
\item up to 6 for integrated mergers
\end{itemize}
\item bargaining power parameter ranges from 0.1 to 0.9 in 0.1 increments
\item marginal costs are either constant or linear.
\item eliminate
\begin{itemize}
\item unprofitable mergers
\item mergers that do no pass the HMT
\end{itemize}
\end{itemize}

\end{frame}

\subsection{Overview }%\SecT{}

\begin{frame}
\centering
\includegraphics[width=4.5in,height=3.5in]{../../output/surplussum.png}

\end{frame}

\subsection{Integrated Firms }

\begin{frame}
\centering
\includegraphics[width=4.5in,height=3.5in]{../../output/CVvertincumbBW.png}

\end{frame}

\begin{frame}
\centering
\includegraphics[width=4.5in,height=3.5in]{../../output/CVbargnobothBW.png}

\end{frame}

\note{
\begin{enumerate}

\item $\uparrow$  wholesalers $\rightarrow$ $\downarrow$ harm at increasing rate
\item $\uparrow$ retailers  $\rightarrow$ $\uparrow$ harm at decreasing rate
\item gap between Bertrand and 2nd score $\downarrow$ as \# of wholesalers and retailers $\uparrow$

\end{enumerate}

}
\begin{frame}
\centering
\includegraphics[width=4.5in,height=3.5in]{../../output/CVbargbothBW.png}

\end{frame}

\note{
\begin{enumerate}

\item $\uparrow$ retailer bargaining power diminishes harm at a linear rate

\end{enumerate}

}

\subsection{Convex Marginal Costs}

\begin{frame}
\centering
\includegraphics[width=4.5in,height=3.5in]{../../output/surplussum_cost.png}
\end{frame}

\note{
\begin{enumerate}
\item
\end{enumerate}

}

\section{Conclusion}

\begin{frame}
\frametitle{Summary}
\begin{itemize}
\item paper relaxes two strong assumptions
\begin{itemize}
\item pre-merger, no firms are vertically integrated,
\item all firms employ constant marginal cost technology,
\end{itemize}
\item simulations with these assumptions relaxed yield markedly different outcomes from those when assumptions are maintained.
\item relative bargaining power is a useful indicia for predicting consumer and total harm in these mergers.
\end{itemize}
%\bibliographystyle{chicago}
%\bibliography{master-bib}
\end{frame}

\appendix

 \begin{frame}[allowframebreaks]
 \footnotesize
 \bibliographystyle{chicago}
 \bibliography{../../doc/master-bib}
 \end{frame}

\end{document}

### 2022-02-10 presentations/FTC/BargainingSimsSlides.tex
added data for trash sims

@@
 
-\addtobeamertemplate{title page}{\footnotesize The views expressed herein are entirely those of the authors and should not be purported to reflect those of the U.S. Department of Justice.}{}
+\addtobeamertemplate{title page}{\footnotesize The views expressed herein are entirely those of the authors and should not be purported to reflect those of the Board of Governors, Federal Trade Commission, or U.S.
+Department of Justice.}{}
 \beamertemplatenavigationsymbolsempty
@@
   \item Assume firms set prices simultaneously in Bertrand Nash equilibrium
-  \item Retailer \textcolor{cyan}{constant}  marginal cost consists of wholesale fee $p^W_{rw}$ and other costs $c^R_{rw}$
+  \item Retailer  marginal cost consists of wholesale fee $p^W_{rw}$ and other  \textcolor{cyan}{constant/linear} costs $c^R_{rw}$
   \item Profit function for retailer $r$ is
@@
   \item $\{\lambda_{rw}; \forall r \in \mathbb{R}, \forall w \in \mathbb{W}\}$ are parameters to be calibrated
-  \item Wholesaler marginal costs  are \textcolor{cyan}{constant} and equal $c^W_{rw}$
+  \item Wholesaler marginal costs  are \textcolor{cyan}{constant/linear} and equal $c^W_{rw}$
   \item Profit function for wholesaler $w$ is

### 2022-02-14 code/TrashSims.R
started writing code for Trash sims (based on Anthem/Cigna Sims).

@@ -1,0 +1,352 @@
+rm(list=ls())
+
+library(antitrust)
+library(kableExtra)
+options(knitr.kable.NA = '')
+
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+
+## Load Data:
+
+combined_data <- read.csv("./doc/trash/combined_data.csv",sep=",")
+paired_share <- read.csv("./doc/trash/paired_share.csv",sep=",")
+
+# simdata <- data.frame(
+#   #`Name`= c('ANT:ASO','CIG:ASO','AET:ASO','UT:ASO','OT:ASO'),
+#   `Pre-merger Owner`= c('Anthem','Cigna','Aetna','United','Other'),
+#   `Post-merger Owner`= c('Anthem','Anthem','Aetna','United','Other'),
+#   #`Prices ($/unit)`= c(24.4,NA,NA,NA,NA),
+#   `Quantities`= c(22.35,6.3,8.6,17.19,2.87),
+#   `Margins ($/unit)`= c(19.5,NA,NA,NA,NA),
+#   `Post-merger Cost Changes($/unit)`= c(-15.7,-4.4,0,0,0),
+#   nests=factor(c(rep(FALSE,2),rep(TRUE,3))),
+#   check.names = FALSE,
+#   stringsAsFactors = FALSE
+# ) 
+
+
+simdata <- data.frame(
+  #`Name`= c('ANT:ASO','CIG:ASO','AET:ASO','UT:ASO','OT:ASO'),
+  `Pre-merger Owner`= c('Anthem','Cigna','Aetna','United','Other'),
+  `Post-merger Owner`= c('Anthem','Anthem','Aetna','United','Other'),
+  `Prices ($/unit)`= c(4356,NA,NA,NA,NA),
+  `Quantities`= c(10.53,2.97,4.05,8.10,1.35),
+  `Margins ($/unit)`= c(239.58,NA,NA,NA,NA),
+  `Post-merger Cost Changes($/unit)`= c(-84.90,-505.05,0,0,0),
+  nests=factor(c(rep(FALSE,2),rep(TRUE,3))),
+  check.names = FALSE,
+  stringsAsFactors = FALSE
+) 
+largeFirmElast <- -.09
+
+## Run Simulation: 
+simres_noeff <- auction2nd.logit.alm(
+  ownerPre = simdata$`Pre-merger Owner`,
+  ownerPost = simdata$`Post-merger Owner`,
[…]

### 2022-02-14 doc/VerticalArrangements.tex
compiled with figures for IIOC

--- 2021-05-13 doc/VerticalArrangements.tex
@@
 
-Concerns about vertical competitive effects were raised by the reviewing agency in the Republic-Santek transaction and by public commenters in the Waste Management-ADS transaction.
+Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
 Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.

### 2022-04-28 code/genfigures.R
fixed labels, horizontal line color in surplus figure

@@ -318,8 +318,8 @@ psummary.bw <-  ggplot(res.nests.logit, aes(y=Outcome_value/mktrev.pre*100,
   stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
   #coord_cartesian(ylim=c(-60,55))+
   scale_y_continuous(breaks=seq(-100,100,5))+
-  geom_hline(yintercept=0,linetype="dashed",color="black")+
-  theme_bw()+scale_colour_tableau('Color Blind')+ theme(legend.position="bottom")+
+  geom_hline(yintercept=0,linetype="dashed",color="goldenrod")+
+  theme_bw()+scale_colour_tableau('Color Blind')+ theme(legend.position="bottom",axis.text.x = element_text(angle = 45,hjust=1))+
   #scale_x_discrete(labels=rev(levels(res.barg$relleveragePre)))+
   #scale_x_discrete(drop=FALSE,labels=ifelse(levels(res.barg$relleveragePre) %in% as.character(round(relleveragePre,1)),levels(res.barg$relleveragePre),""))+
   #theme_tufte(ticks=FALSE) +

### 2022-04-29 code/genfigures.R
changed color of horizontal intercept to goldenrod

@@ -111,7 +111,7 @@ pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","
   stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
   #coord_cartesian(ylim=c(-60,55))+
   scale_y_continuous(breaks=seq(-100,100,10))+
-  geom_hline(yintercept=0,linetype="dashed",color="black")+
+  geom_hline(yintercept=0,linetype="dashed",color="goldenrod")+
   theme_bw()+scale_colour_tableau('Color Blind')+ theme(legend.position="bottom")+
   #scale_x_discrete(labels=rev(levels(res.barg$relleveragePre)))+
   #scale_x_discrete(drop=FALSE,labels=ifelse(levels(res.barg$relleveragePre) %in% as.character(round(relleveragePre,1)),levels(res.barg$relleveragePre),""))+

### 2022-04-29 doc/VerticalArrangements.tex
incorporated Gloria's 5/28 edits, including a new figure on chatanooga trash from Peggy. removed deprecated files. still working on chatanooga sims.

--- 2021-05-14 doc/VerticalArrangements.tex
@@
 \newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
-\author{Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov.
-} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission \and Margaret Loudermilk\footnote{U.S.
-Department of Justice, margaret.loudermilk@usdoj.gov}\\  U.S Department of Justice}
+\author{Margaret Loudermilk\footnote{U.S.
+Department of Justice, margaret.loudermilk@usdoj.gov} \\  U.S Department of Justice \and Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov.
+} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission }
 
@@
 The model features logit Bertrand competition downstream and Nash Bargaining upstream.
-We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and a previously integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and a previously integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
 The breadth of options we include better captures the variety of merger structures observed in practice, compared to the typical ``horizontal vs.
@@
 In this paper, we use a standard merger simulation model to assess the welfare implications of these complex mergers.
-Our model is drawn directly from \citeN{ST2020}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
-We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and a previously integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and a previously integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
+We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
 In each of these simulations, we calculate the impacts on consumer welfare and firm profits.
@@
 \section{Theory}
-We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2020}.
+We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2021}.
 This framework allows us to study a variety of merger configurations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Downstream and Upstream Competition}
+\subsection{Downstream and Upstream Competition}
 Assume there is a set of consumers who each choose to buy a product sold by a single retailer.
@@
 Wholesale prices are set via Nash Bargaining between retailers and wholesalers.
-We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2020}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
+We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
 }  Profits for wholesaler $w$ are given by $\pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M$, where $c^W_{rw}$ is the wholesale marginal cost when dealing with retailer $r$.
@@
 Together the series of downstream and upstream first order conditions determine market equilibrium.
-This model can be solved and calibrated as described in \citeN{ST2020}.
+This model can be solved and calibrated as described in \citeN{ST2021}.
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Mergers}
+\subsection{Mergers}
 Here we describe the manner in which mergers are modeled in this framework.
-We begin by describing vertical mergers, as all other cases we examine have some vertical aspects.
+We begin by describing vertical mergers, as all cases we examine have some vertical aspects.
 Suppose that retailer $r$ and wholesaler $w$ were to merge.
@@
 
-Next consider a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
+Next consider a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
 Such a combination has a vertical component and a horizontal component.
@@
 
-If instead a preexisting integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the resulting first order condition for product $rw$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc}.
+If instead an integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the resulting first order condition for product $rw$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc}.
 However,  the downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of wholesaler $v$.
@@
 
-Returning to a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $v$ to supply retailer $r$, the first order condition becomes
+Returning to a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $v$ to supply retailer $r$, the first order condition becomes
 \begin{equation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Numerical Simulations}
+In this section, we use a similar setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and allow for the presence integrated rival firms.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Data Generating Process}
-This section provides an overview of our methodology, with additional details appearing in the Appendix.
-For vertical mergers, we only consider mergers between an unintegrated wholesaler and unintegrated retailer.
-In addition, we consider upstream and downstream mergers between either two unintegrated firms or between one vertically integrated and one unintegrated firm.
-Finally, we assume that an integrated merger occurs between two vertically integrated firms
+\subsection{Data Generating Process}
+Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.
+We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.
+Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.
+We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.
+Finally, we assume that an integrated merger is between two firms that are already vertically integrated pre-merger.
 
-%\textbf{consider adding diagonal mergers by allowing for heterogeneity in the bargaining parameter.}.
+For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.
+}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7.
+We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
+Vertically integrated firms are not siloed: integrated wholesalers supply inputs to non-integrated retailers and vertically integrated retailers purchase inputs from non-integrated wholesalers.
 
-We simulate markets by randomly sampling shares from a Dirichlet distribution for 2, 3, 4, or 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.
-}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for these simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7.
-We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
-We assume that vertically integrated firms are not siloed: integrated wholesalers supply inputs to non-integrated retailers and that vertically integrated retailers purchase inputs from non-integrated wholesalers.
-The price coefficient $\alpha$ is calibrated by assuming that in the pre-merger world, there is a vertically integrated outside option available to all customers.
-The other goods are differenced relative to this option, which maintains the outside good normalization.
-The market size is set to 1.
+Our simulations focus on mergers that are more likely to have competitive effects and to therefore come under agency scruitiny.
+For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger.
+Similarly, when simulating a vertical merger, we assign the products sold by the largest wholesaler and the largest retailer to a single firm post-merger.
 
-We specify values for the bargaining parameter ranging from from 0.1 (wholesalers have the advantage) to 0.9 (retailers have the advantage).
-To better understand the relative bargaining strength of these parameter values, we report our results in terms of $(1-\lambda)/\lambda$, which range from $9$ (wholesaler power is nine times greater than retailer power) to $1/9$ (retailer power is nine times greater than wholesaler power).
-The bargaining parameter is identical for all of the retailers in each simulation, unless noted otherwise.
+The bargaining parameter ranges from from 0.1 (where wholesalers have the advantage) to 0.9 (where retailers have the advantage).
+We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$, which ranges from $9$ (wholesaler power is nine times greater than retailer power) to $1/9$ (retailer power is nine times greater than wholesaler power).
+The bargaining parameter is identical for all of the retailers in each simulation.
+We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option.
+All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.
+We set the market size to 1.
 
 For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-This results in 1.96 million merger simulations.
-We then eliminate mergers where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, yielding 1.37 million markets.\footnote{The Hypothetical Monopolist Test requires that were a monopolist to jointly own all products in a candidate market, that firm would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} All 1.37 million markets treat as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+This results in 1.96 million possible merger simulations.
+After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 1.37 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+} Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
 
-When simulating a horizontal merger, we assign the products produced by the two largest firms in the market to a single entity post-merger.
-Similarly, when simulating a vertical merger, we assign the products produced by the largest wholesaler and the largest retailer to a single entity post-merger.
-This assignment is purposefully skewed towards mergers that are more likely to have competitive effects and to come under agency review.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Overview of Simulated Output}
 
-Table \ref{tab:simsum} provides summary statistics across our various simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to calibrate and simulate the effects of mergers in a range of competitive scenarios, including the ones described here.}  The median average wholesale pre-merger price is \$4.6, and the median average retail pre-merger price is \$14.
-Because the market size is set to 1, these average prices are equal to total pre-merger expenditures.
-Pre-merger HHIs range between 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703.
-HHIs for vertical mergers increase by 1,277 points at the median, resulting in a median post-merger HHI equal to 4,889.\footnote{We compute the post-merger HHI for vertical mergers by calculating the merged firms' market share as the sum of all the shares of downstream products that either incorporate the upstream partner's input or are sold by the downstream partner.
+Table \ref{tab:simsum} provides summary statistics for our simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}  For reference, the 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with HHIs over 2,500 points.
+The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.
+}
+
+The median average wholesale pre-merger price is \$4.60, and the median average retail pre-merger price is \$14.
+Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
+Pre-merger HHIs range from 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703.
+HHIs for vertical mergers increase by 1,277 points at the median, resulting in a median post-merger HHI equal to 4,889.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
 }  HHIs for upstream mergers increase by 2,004 points at the median, resulting in a median post-merger HHI equal to 5,231.
-HHIs for horizontal downstream mergers increase by 2,054 points at the median, resulting in a median post-merger HHI equal to 4,889.
+HHIs for downstream mergers increase by 2,054 points at the median, resulting in a median post-merger HHI equal to 4,889.
 HHIs for integrated mergers increase by 2,746 points at the median, resulting in a median post-merger HHI equal to 7,220.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Results Overview}
-Our overall results are depicted in Figure \ref{fig:surplussum}, which  is divided into four panels, each showing how the distribution of surplus changes for a particular set of agents (consumers, retailers, or wholesalers), as well as the net effect on the market as a whole.
-Surplus is presented as a percentage change relative to total pre-merger expenditure in the downstream market.
-
+Figure \ref{fig:surplussum} summarizes our results for welfare.
+Each of the four panels show how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).
+Surplus is reported as a percentage change of total pre-merger expenditure in the downstream market.
 Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger.
-The whiskers display the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median.
-Note that negative outcome values imply agent harm, and positive values imply agent benefits.
+The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median.
+Negative outcome values imply harm, and positive values imply benefits.
 
 We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.
-The median change is negative, indicating harm, across all four types of mergers.
-However, the distributions and magnitudes differ.
-In particular, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
+Across all four types of mergers, the majority of cases show harm, though the distributions and magnitudes differ.
+In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
 Median consumer harm from integrated mergers is almost 14\% of pre-merger total expenditures, 1.25 times the magnitude of that from downstream mergers, 2.3 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.
 
-%\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}
+Moving on t retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations.
+Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.
 
-%However, the fact that downstream mergers under Bertrand are almost never beneficial and frequently quite harmful is somewhat surprising, given the potential for countervailing bargaining leverage to lower wholesale prices.
-
-Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 67\% of all simulations.
-Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.
-
-%We also find that  for upstream, vertical, and integrated mergers, retailer surplus under constant marginal cost first-order stochastically dominates the retail surplus distribution under linear costs, again suggesting that the incentive to raise rivals' costs dominates the benefits from EDM.
-By contrast, for downstream mergers, there is no clear rank-ordering between markets with constant marginal costs and markets with linear costs.
-
-% Finally, it is interesting that the harm experienced in the latter instances is quite small in relative terms: median retailer harm from upstream mergers is about one-quarter of the median consumer harm from an upstream merger.
-This, along with the narrow inter-quartile range of retailer harm (less than 1\% of pre-merger revenues) indicates that retailers are passing on the bulk of the wholesale price increase to consumers.\footnote{This phenomena may also be seen by noting that the distribution of wholesaler gain from upstream mergers is similar in magnitude to the distribution of consumer losses from upstream mergers, though less skewed.
-}
-
-As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and about 15\% of simulated integrated mergers.
-%\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e.
+Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 77\% of all upstream mergers, 23\% of vertical mergers, 20\% of downstream mergers, and about 15\% of integrated mergers.
+%\textbf{comparing these results to \citeN{ST2021}, the effects are less stark (i.e.
 less of a transfer.}
@@
 
-%Whereas wholesalers benefit from vertical mergers in about 17\% of simulations, retailers benefit from vertical mergers in about 99\% of simulations.
-Moreover, the median benefit to retailers is 7.5\% of pre-merger expenditures, comparable to the median loss to wholesalers of -9\%.
-Together, these observations indicate that vertical mergers are largely a rent transfer from wholesalers to retailers and consumers, with retailers keeping most of the gain.
-
-In terms of total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial.
+As for total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial.
 Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers.
@@
 
-%Finally, for each merger type, total harm under linear costs first-order stochastically dominates total harm under constant marginal costs, with the greatest differences occurring for vertical and integrated mergers.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Mergers and Vertically Integrated Incumbent Firms}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Vertically Integrated Incumbent Firms}
-Here, we examine merger outcomes when a vertically integrated firm merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm.
-We also examine merger outcomes for  mergers as the number of incumbent integrated non-merging rivals increases.
+Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
+Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases from 0 to 6.
+The plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+By contrast, for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
+For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.
 
-Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms.
-The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020} and is included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction.
-\citeN{ST2020} also do not consider mergers between integrated firms.
-}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single unintegrated 3rd party is integrated, and the remaining 3rd parties (if any) are unintegrated.
-By contrast, for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
-Finally, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, again with no integrated rivals.
-
-Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial (i.e.
-increase total surplus).
-The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 25\% and the percentage of mergers that increase  total surplus to 16\%, but leads median harm largely unchanged.
-Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefit consumers to about 27\% but does not increase the percentage of net beneficial mergers.
-Adding a fourth integrated incumbent further increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of net beneficial mergers to about 11\%.
-Taken together, Figure \ref{fig:CVvertincumbBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.
+We begin with vertical mergers, as they are our simplest case, lacking preexisting integration at either merging firm.
+Absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.
+The addition of a single rival integrated firm (denoted as 1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm steady.
+Adding more integrated rivals tends to shift some outcomes from the lower tail to the upper tail, increasing the proportion of mergers with benefits, especially for consumers.
+However, the interquartile range and median are largely unchanged.
+Thus, the presence of rival integrated firms can limit the extent of harm, particularly consumer harm, from vertical mergers, though not dramatically.
+Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.
+It appears that these rivals can therefore somewhat limit the extent of negative impacts through RRC.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Bargaining Power}
-\citeN{ST2020} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power.
+\subsection{Mergers and Bargaining Power}
+\citeN{ST2021} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power.
 Here, we show that this result continues to hold when there are rival incumbent integrated firms.
@@
 also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) incumbent vertically integrated firms.
-As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}.
+As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2021}.
 For vertical mergers (left), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
@@
 
-\subsection*{Costs}
+\subsection{Robustness to Alternative Cost Specifications}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Empirical Setting}
+\section{Empirical Application}
 
@@
 Waste Management, Inc.
-and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{\tiny{Waste 360, \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021.\footnote{\tiny{Competitive Impact Statement: U.S.
+and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{\tiny{Waste 360, \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{\tiny{Competitive Impcat Statement: U.S.
 and State of Alabama v.
-Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}} These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
+Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
 
-Concerns about vertical competitive effects were raised by the reviewing agency in the Republic-Santek transaction and by public commenters in the Waste Management-ADS transaction.
+Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
 Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
@@
 
-Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets - in the presence of differing numbers of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
+Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
+The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
 
+\subsection{Combination of vertically integrated assets: Chattanooga}
+
+The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commerical waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers.
+The CIS notes that Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors.
+In MSW disposal, the CIS identifies only one other significant competitor and Republic and Santek as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere.
+Thus, pre-merger both parties were large, vertically integrated competitiors in the Chattanooga Area.
+
+In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, which was the parties' sole competitor in the MSW disposal market.
+Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consumation since the merger was completed before the filing of the CIS for the Republic-Santek merger.
+The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.
+
+\subsubsection{MSW Disposal Volume Data}
+\label{chatt_disposal_vol}
+
+Data on MSW disposal comes from data collected by TDEQ (Tennessee Department of Environmental Quality) on Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+This data identifies the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
+The MSW disposal market definition follows that outlined in the DOJ CIS and attributing share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
+Thus, the total market quantity is defined as all MSW volumes originating in Hamilton county.
+
+Disposal volumes are measured in tons and have been combined across landfills owned by the same firm.
+The TDEQ data identifies landfills owned or operated by Republic, Santek, Waste Connections, and three market participants each with less than 0.5\% share receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% combined.
+Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out of state landfill in GA.} These three fringe participants have been excluded from the analysis.
+After rescaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.
+
+Using this data in conjunction with information on collection market volumes, discussed in Section \ref{chatt_collection_share}, the supply relationships between the upstream disposal market participants (\textit{Republic, Santek, and Waste Connections}) and downstream collection market particpants (\textit{Republic, Santek, Waste Connections, WM-ADS, and ``Regional''}) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Santek and Waste Connections exhibit excess disposal capacity.
+However, Republic's collection volumes are estimated to exceed their ability to self-supply disposal in the Chattanooga market.
+As a result, Republic will need to puchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
+We assume that this disposal is purchased from both Santek and Waste connections according to their share of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
+We further assume that WM-ADS and Regional purchased disposal capacity from both Santek and Republic according to these shares.
+The resulting relationships between the upstream and downstream market participants are captured in Figure \ref{fig:chatt_tree}.
+
+\subsubsection{MSW Disposal Price Data}
+MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
+The measure of price reflected is the ``gate rate'', which is the posted price at the landfill, measured in \$/Ton.\footnote{Disposal prices for large customer's may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
+The price used in the analysis for each is the volume weighted average for their landfills.
+
+\subsubsection{Collection Market Share Data}
+\label{chatt_collection_share}
+
+The CIS states that in the Chattanooga Area the post-merger HHI for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
+Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
+After rescaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WM-ADS, 8.3\%, and Regional, 8.3\%.
+
+However, the market shares required for implementation of the Sheu and Targain (\textbf{add ref}) framework are expressed as upstream-downstream market participant pairs.
+Following the discussion of supply relationships in Section \ref{chatt_disposal_vol}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal-Santek collection, Waste Connection disposal-Waste Connections collection) are assigned their full collection share.
+Capacity constrained, integrated firms (Republic disposal-Republic collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal-Republic collection, Waste Connections disposal-Republic collection).
+Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.
+
+\subsubsection{Cost, Margin and  Elasticity Data}
+
+Collection and disposal margins are calculated for Republic from data on revenue by line of service and components of cost of operations in the company's 2019 annual report.
+The data is reported at the company level and is not specific to the Chattanooga Area, but revenues are reported by collection segment.
+
+Collection costs in \$/Ton are estimated from Republic and Waste Management 2019 10k financial statements.
+These costs are reported at the company level across all segments.
+The estimated share of these costs attributable to the Chattanooga market, based on the number of markets in which the companies operate, is divided by the tons disposed for each company.
+Santek, Waste Connections, and Regional do not publically produce comparable financial statements.
+Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure in this market due to their lack of internal disposal capacity.
+
+A survey of demand elasticity estimates for the collection market are present in Bel and Gradus (2016) with an average of -0.34.
+These estimates are not specific to the SCCW market, but available evidence suggests that the commerical segment should be more inelastic than other segments such that analysis using these estimates would tend to be conservative for estimating the size of merger induced price effects.\footnote{Bel, G.
+and R.
+Gradus, 2016, Effects of unit-based pricing on household waste collection demand: A meta-regression analysis, Resource and Energy Economics, 44, 169-182.}
+
+\subsubsection{Chattanooga Area Results}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
 
-This paper relaxes two strong assumptions made in \citeN{ST2020}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated, and that all firms employ constant marginal cost technology.
+This paper relaxes two strong assumptions made in \citeN{ST2021}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated, and that all firms employ constant marginal cost technology.
 Relaxing these constraints is important as competition authorities are called upon to investigate mergers where either the merging parties or a third party are already vertically integrated, or where the presence of a scarce input could reduce the benefit from EDM or increase the harm from RRC.
@@
 Finally, we intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm.
-Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2020}, such as: how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
+Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2021}, such as: how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
 
@@
 \begin{table}
-\include{sumtable}
+\small{\include{sumtable}}
 \caption{Summary Statistics}
@@
 \centering
-\includegraphics[scale=0.9]{surplussum.png}
+\includegraphics[scale=0.9]{output/surplussum.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus.
@@
 \centering
-\includegraphics[scale=0.9]{CVvertincumbBW.png}
+\includegraphics[scale=0.9]{output/CVvertincumbBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (top row) and total (bottom) surplus as the number of vertically integrated firms present in a market change.
@@
 \centering
-\includegraphics[scale=0.9]{CVbargnobothBW.png}
+\includegraphics[scale=0.9]{output/CVbargnobothBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which downstream, upstream, and vertical mergers affect consumer, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \centering
-\includegraphics[scale=0.85]{CVbargbothBW.png}
+\includegraphics[scale=0.85]{output/CVbargbothBW.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \centering
-\includegraphics[scale=0.9]{surplussum_cost.png}
+\includegraphics[scale=0.9]{output/surplussum_cost.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
@@
 \end{sidewaysfigure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=1.0]{output/chattanooga_tree.png}
+\caption{Chattanooga Area MSW Disposal and Collection Market}
+\label{fig:chatt_tree}
+\end{figure}
+
 \end{document}

### 2022-04-29 code/TrashSims.R

@@ -11,146 +11,265 @@ library(ggplot2)
 
 ## Load Data:
 
-combined_data <- read.csv("./doc/trash/combined_data.csv",sep=",")
-paired_share <- read.csv("./doc/trash/paired_share.csv",sep=",")
+library(readr)
+paired_data_4_22 <- read_csv("~/Projects/bargaining_convex/trash/paired_data_4_22.csv")
 
-# simdata <- data.frame(
-#   #`Name`= c('ANT:ASO','CIG:ASO','AET:ASO','UT:ASO','OT:ASO'),
-#   `Pre-merger Owner`= c('Anthem','Cigna','Aetna','United','Other'),
-#   `Post-merger Owner`= c('Anthem','Anthem','Aetna','United','Other'),
-#   #`Prices ($/unit)`= c(24.4,NA,NA,NA,NA),
-#   `Quantities`= c(22.35,6.3,8.6,17.19,2.87),
-#   `Margins ($/unit)`= c(19.5,NA,NA,NA,NA),
-#   `Post-merger Cost Changes($/unit)`= c(-15.7,-4.4,0,0,0),
-#   nests=factor(c(rep(FALSE,2),rep(TRUE,3))),
-#   check.names = FALSE,
-#   stringsAsFactors = FALSE
-# ) 
+disposal_price_republic <-  96 # source: 4/26 correspondence with Peggy
 
+simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
+  mutate(collection_paired_share=collection_paired_share/sum(collection_paired_share),
+         disposal_dollar_margin=ifelse(disposal_firm_name=="Republic",
+                                       disposal_percent_margin*disposal_price_republic,
+                                       disposal_dollar_margin),
+         disposal_price=ifelse(disposal_firm_name=="Republic" &
+                                 collection_firm_name=="Republic",
+                               disposal_price_republic - disposal_dollar_margin,disposal_price),
+         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.7)
+         ) %>%
+  group_by(collection_firm_name) %>%
+  mutate(Name=interaction(disposal_firm_name,collection_firm_name,sep=":",drop=TRUE),
+         firmShare=sum(collection_paired_share),
+         ownerPreUp=disposal_firm_name,
[…]

### 2022-05-10 code/MonteCarlo.R
Includes Gloria's edits from 5/6 as well as correction to upstream bargaining ownership matrix.

@@ -167,7 +167,7 @@ market<-function(nfirms.down=3, # # of downstream firms
 
     bargparmPost <- bargparm
     if(ownerPost  %in%  c("vertical")){ bargparmPost[1] <- 1 }# all bargaining power resides with the retailer in a vertical  merger
-    if(ownerPost  %in%  c("up","both")){ bargparmPost[ids$up.firm == 1 & ids$down.firm == 2] <- 1 } # post-merger, u1 gives d1 a discount
+    if(ownerPost  %in%  c("up","both")){ bargparmPost[ids$up.firm == 1 & ids$down.firm == 2] <- 1 } # post-merger, u1 gives d2 a discount
     if(ownerPost  %in%  c("down","both")){ bargparmPost[ids$up.firm == 2 & ids$down.firm == 1] <- 1 } #post-merger u2 gives d1 a discount
 
     }
@@ -227,6 +227,19 @@ market<-function(nfirms.down=3, # # of downstream firms
 
     for( v in vertFirms){
 
+      ## set integrated margin disagreement payoff to 0
+      ownerPre.up[ids$up.firm==v & ids$down.firm!=v
+                   , ids$up.firm==v & ids$down.firm==v]=0
+      ## constrain upstream integrated margin to zero
+       ownerPre.up[ids$up.firm==v & ids$down.firm==v
+                   , ids$up.firm==v & ids$down.firm!=v]=0
+
+       ## set integrated margin disagreement payoff to 0
+       ownerPost.up[ids$up.firm==v & ids$down.firm!=v
+                   , ids$up.firm==v & ids$down.firm==v]=0
+       ## constrain upstream integrated margin to zero
+       ownerPost.up[ids$up.firm==v & ids$down.firm==v
+                   , ids$up.firm==v & ids$down.firm!=v]=0
       ## integrated retailer bargains with other wholesalers
       vertrowsDown <-ids$up.firm  != v  & ids$down.firm  == v
 
@@ -261,7 +274,15 @@ market<-function(nfirms.down=3, # # of downstream firms
     #ownerBargUpVertPost <-  ownerBargUpVertPre
 
     if(ownerPost  %in%  c("vertical")){
-      vertrowsDown <- ids$up.firm != 1  & ids$down.firm == 1
+
+      ## set integrated margin disagreement payoff to 0
+      ownerPost.up[ids$up.firm==1 & ids$down.firm!=1
[…]

### 2022-05-10 code/RunSims.R

@@ -22,7 +22,7 @@ simpath <- "./data"
 #profiler <- microbenchmark::microbenchmark(times=1,{
 #Rprof(filename = "Rprof.out", append = FALSE, interval = 2, line.profiling=TRUE)
 
-cl <- makeCluster(mc <- getOption("cl.cores", 100))
+cl <- makeCluster(mc <- getOption("cl.cores", 75))
 ## to make this reproducible
 clusterSetRNGStream(cl, 3141519)
 

### 2022-05-10 code/TrashSims.R

@@ -21,13 +21,25 @@ simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
          disposal_dollar_margin=ifelse(disposal_firm_name=="Republic",
                                        disposal_percent_margin*disposal_price_republic,
                                        disposal_dollar_margin),
-         disposal_price=ifelse(disposal_firm_name=="Republic" &
-                                 collection_firm_name=="Republic",
+         # Peggy suggests assuming Santek and WasteConn have the same upstream margin as Republic
+         # and use that margin to recover costs
+         disposal_price=ifelse(disposal_firm_name=="Republic",
                                disposal_price_republic - disposal_dollar_margin,disposal_price),
-         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.7)
+         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.9),
+         disposal_dollar_margin=ifelse(disposal_firm_name %in% c("Santek","WasteConn"),
+                                       max(disposal_percent_margin,na.rm=TRUE)*disposal_price,
+                                       disposal_dollar_margin)
          ) %>%
+  group_by(disposal_firm_name) %>%
+  mutate(
+    disposal_dollar_margin=max(disposal_dollar_margin,na.rm=TRUE),
+    disposal_price=ifelse(is.na(disposal_price),max(disposal_price,na.rm=TRUE) - disposal_dollar_margin,disposal_price),
+    disposal_cost= ifelse(disposal_firm_name==collection_firm_name,disposal_price, disposal_price-disposal_dollar_margin),
+    disposal_dollar_margin=ifelse(disposal_firm_name==collection_firm_name,0,disposal_dollar_margin)) %>%
   group_by(collection_firm_name) %>%
-  mutate(Name=interaction(disposal_firm_name,collection_firm_name,sep=":",drop=TRUE),
+  mutate(
+
+         Name=interaction(disposal_firm_name,collection_firm_name,sep=":",drop=TRUE),
          firmShare=sum(collection_paired_share),
          ownerPreUp=disposal_firm_name,
[…]

### 2022-05-13 doc/TrashData.tex
finished PRELIMINARY draft of merger sims. NEED TO CHECK THESE. And replace horrible table of sim results.

\begin{table}

\caption{\label{tab:}Republic/Santek Merger Simulation Inputs}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llrrrrr}
\toprule
Disposal Firm & Collection Firm & Volume (000s) & Disposal Price & Disposal Margin (\$) & Collection Margin (\$) & Collection Price\\
\midrule
Republic & Republic & 165 & 42 & 0 & 44 & 179\\
Santek & Republic & 34 & 16 & 20 &  & 174\\
Santek & Santek & 218 & 16 & 0 &  & 134\\
Santek & WM\_ADS & 30 & 16 & 20 &  & 221\\
Santek & Regional & 30 & 16 & 20 &  & 221\\
\addlinespace
WasteConn & Republic & 19 & 11 & 14 &  & 162\\
WasteConn & WasteConn & 48 & 11 & 0 &  & 110\\
WasteConn & WM\_ADS & 17 & 11 & 14 &  & 210\\
WasteConn & Regional & 17 & 11 & 14 &  & 210\\
\bottomrule
\end{tabular}}
\end{table}

### 2022-05-13 code/TrashSims.R

@@ -25,7 +25,7 @@ simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
          # and use that margin to recover costs
          disposal_price=ifelse(disposal_firm_name=="Republic",
                                disposal_price_republic - disposal_dollar_margin,disposal_price),
-         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.9),
+         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.5),
          disposal_dollar_margin=ifelse(disposal_firm_name %in% c("Santek","WasteConn"),
                                        max(disposal_percent_margin,na.rm=TRUE)*disposal_price,
                                        disposal_dollar_margin)
@@ -59,6 +59,16 @@ ownerPreUpMat=model.matrix(~-1+ownerPreUp,data=simdata)
 ownerPreUpMat=tcrossprod(ownerPreUpMat)
 ownerPreDownMat=model.matrix(~-1+ownerPreDown,data=simdata)
 ownerPreDownMat=tcrossprod(ownerPreDownMat)
+
+for (v in vertFirms){
+  ## set integrated margin disagreement payoff to 0
+  ownerPreUpMat[simdata$ownerPreUp==v & simdata$ownerPreDown!=v
+            , simdata$ownerPreUp==v & simdata$ownerPreDown==v]=0
+  ## constrain upstream integrated margin to zero
+  ownerPreUpMat[simdata$ownerPreUp==v & simdata$ownerPreDown==v
+            , simdata$ownerPreUp==v & simdata$ownerPreDown!=v]=0
+
+}
 
 sharesNoOut <- with(simdata,collection_paired_share)
 div <- tcrossprod(1/(1-sharesNoOut),sharesNoOut)*sharesNoOut
@@ -170,7 +180,7 @@ else{ alpha <- simdata$alpha_2nd[1]}
 #minBarg_2nd <- optim(rep(.5,3),minD,method="L-BFGS-B",lower=rep(0,3),upper=rep(1,3),is2nd=TRUE)
 
 ## same bargaininag parameter for wholesalers
-minBarg_bert <- optim(rep(.5,2),minD,method="L-BFGS-B",lower=rep(0,2),upper=rep(1,2))
+minBarg_bert<- optim(rep(.5,2),minD,method="L-BFGS-B",lower=rep(0,2),upper=rep(1,2))
[…]

### 2022-05-13 doc/TrashData.tex
simdata table was misreporting prices. fixed issue.

@@
 Republic & Republic & 165 & 42 & 0 & 44 & 179\\
-Santek & Republic & 34 & 16 & 20 &  & 174\\
+Santek & Republic & 34 & 36 & 20 &  & 174\\
 Santek & Santek & 218 & 16 & 0 &  & 134\\
-Santek & WM\_ADS & 30 & 16 & 20 &  & 221\\
-Santek & Regional & 30 & 16 & 20 &  & 221\\
+Santek & WM\_ADS & 30 & 36 & 20 &  & 221\\
+Santek & Regional & 30 & 36 & 20 &  & 221\\
 \addlinespace
-WasteConn & Republic & 19 & 11 & 14 &  & 162\\
+WasteConn & Republic & 19 & 25 & 14 &  & 162\\
 WasteConn & WasteConn & 48 & 11 & 0 &  & 110\\
-WasteConn & WM\_ADS & 17 & 11 & 14 &  & 210\\
-WasteConn & Regional & 17 & 11 & 14 &  & 210\\
+WasteConn & WM\_ADS & 17 & 25 & 14 &  & 210\\
+WasteConn & Regional & 17 & 25 & 14 &  & 210\\
 \bottomrule

### 2022-05-13 code/TrashSims.R

@@ -320,7 +320,7 @@ sink("./doc/TrashData.tex")
 print(kable(select(simdata,disposal_firm_name,
                    collection_firm_name,
                    disposal_volume,
-                   disposal_cost,
+                   disposal_price,
                    disposal_dollar_margin,
                    collection_dollar_margin,
                    priceDown_repmargin_2nd
@@ -329,7 +329,7 @@ print(kable(select(simdata,disposal_firm_name,
               rename("Disposal Firm"=disposal_firm_name,
                      "Collection Firm"=collection_firm_name,
                      "Volume (000s)"=disposal_volume,
-                     "Disposal Price"=disposal_cost,
+                     "Disposal Price"=disposal_price,
                      "Disposal Margin ($)"=disposal_dollar_margin,
                      "Collection Margin ($)"=collection_dollar_margin,
                      "Collection Price"=priceDown_repmargin_2nd

### 2022-05-16 code/TrashSims.R
candidate replacement figure for table 3 (TrashSims). Have not yet removed table 3.

@@ -395,24 +395,34 @@ kable(vert_sum,format = "latex",
   #collapse_rows(1:2,row_group_label_position="stack")
 sink()
 
-# firmplot <- ggplot(data=  vert_sum
-# ,
-#        aes(x=firm,y=value, fill=level)) +
-#        facet_grid(key~Efficiency,scales = "free_y") + geom_bar(stat="identity", position=position_dodge())  +
-#   xlab("Insurer") + ylab("Equilibrium Level Changes") + geom_hline(yintercept = 0,linetype="dashed") + coord_flip()+ scale_fill_grey(start = .9, end = .1) +
-#   geom_text(aes(label=
-#                   ifelse(abs(value)>0,round(value,1),NA),
-#                    hjust=ifelse(sign(value)>0,1,0),color=ifelse(level=="Down","white","black")),
-#                   position=position_dodge(width=.9),size =3) + theme_bw() +
-#   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1))
 
-#
-# png(filename="./output/AnthemCignaSimsFirm.png" ,res = 250, units="in"
-#     ,width = 4, height = 4 )
-#
-# print(firmplot_82_33)
-#
-# dev.off()
+firmplot <- ggplot(data=  vert_sum %>% mutate(Name=interaction(Disposal,Collector,drop=TRUE,sep="/"),
+                                  Name=reorder(Name,`Post-merger`* as.numeric(Effect=="Prices") * as.numeric(Level=="Collection"))) %>%
+         mutate(Change=`Post-merger` - `Pre-merger`)
+
+                %>%
+         pivot_longer(c(`Pre-merger` ,`Post-merger`),names_to = "Type",values_to = "value") %>%
+         mutate(Type=factor(Type,levels=c("Pre-merger","Post-merger"))) %>%
+         filter(!grepl("Change",Type)),
+aes(x=Name,y=value,fill=Type,label=value)) +
+  facet_grid(~Level+Effect,scales = "free_x") + geom_bar(stat="identity",
+                                                             position=position_dodge()
+                                                             #position="stack"
+                                                         )  +
+  xlab("Disposal/Collector") + ylab("Equilibrium Levels") + #geom_hline(yintercept = 0,linetype="dashed") +
+  scale_fill_brewer(type="qual",palette = "Paired") +#scale_fill_grey(start = .9, end = .1) +
[…]

### 2022-05-25 doc/VerticalArrangements.tex
finished PRELIMINARY draft of merger sims. NEED TO CHECK THESE. And replace horrible table of sim results.
candidate replacement figure for table 3 (TrashSims). Have not yet removed table 3.
incorporates Gloria's edits for Bates White Conference.

@@
 
+\begin{center}
+Preliminary and Incomplete
+\end{center}
+
 \begin{abstract}
@@
 
-In this paper, we use a standard merger simulation model to assess the welfare implications of these complex mergers.
+In this paper, we use a merger simulation model to assess the welfare implications of these complex mergers.
 Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
@@
 
+%We find.
+
 We then provide a case study (in progress) from an industry, solid waste management, that has seen a large number of mergers with both horizontal and vertical elements.
@@
 A variety of firms participate in these services, with some owning their own hauling and disposal assets, while others are active in only a segment of the market.
+
+%Our paper is related to the literature on mergers and vertical integration.
+
+%The paper proceeds as follows.
 
@@
 & - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
-&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
+&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
 \end{split}
@@
 
+%Integrated rivals.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Numerical Simulations}
-In this section, we use a similar setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and allow for the presence integrated rival firms.
+In this section, we use a similar setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and allow for the presence of integrated rival firms.
 
@@
 We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.
-Finally, we assume that an integrated merger is between two firms that are already vertically integrated pre-merger.
+Finally, we assume that an integrated merger is between two firms that are both already vertically integrated pre-merger.
 
@@
 We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
-Vertically integrated firms are not siloed: integrated wholesalers supply inputs to non-integrated retailers and vertically integrated retailers purchase inputs from non-integrated wholesalers.
+Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and vertically integrated retailers purchase inputs from wholesalers other than their integrated partner.
 
@@
 For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-This results in 1.96 million possible merger simulations.
-After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 1.37 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.
+After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 2.7 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+} Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.
 
@@
 
-The median average wholesale pre-merger price is \$4.60, and the median average retail pre-merger price is \$14.
+The median average wholesale pre-merger price is \$4.50, and the median average retail pre-merger price is \$13.
 Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
 Pre-merger HHIs range from 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703.
-HHIs for vertical mergers increase by 1,277 points at the median, resulting in a median post-merger HHI equal to 4,889.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
-}  HHIs for upstream mergers increase by 2,004 points at the median, resulting in a median post-merger HHI equal to 5,231.
-HHIs for downstream mergers increase by 2,054 points at the median, resulting in a median post-merger HHI equal to 4,889.
-HHIs for integrated mergers increase by 2,746 points at the median, resulting in a median post-merger HHI equal to 7,220.
-Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with post-merger HHIs over 2,500 points and HHI Changes greater than 200 points.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.
-}
+HHIs for vertical mergers increase by 1,298 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
+}  HHIs for downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.
+HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238.
+HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280.
 
@@
 Each of the four panels show how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).
-Surplus is reported as a percentage change of total pre-merger expenditure in the downstream market.
+Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.
 Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger.
@@
 
-Moving on t retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations.
+Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations.
 Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.
@@
 Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 77\% of all upstream mergers, 23\% of vertical mergers, 20\% of downstream mergers, and about 15\% of integrated mergers.
-%\textbf{comparing these results to \citeN{ST2021}, the effects are less stark (i.e.
-less of a transfer.}
 However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.
@@
 Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
-Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases from 0 to 6.
+Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases.
 The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
-By contrast, for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties.
+By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties.
 For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.
 
-We begin with vertical mergers, as they are our simplest case, lacking preexisting integration at either merging firm.
-Absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.
-The addition of a single rival integrated firm (denoted as 1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm steady.
+We begin with vertical mergers, as they are our simplest case, because there is never any preexisting integration at either merging firm.
+Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.
+The addition of a single rival integrated firm (1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm basically the same.
 Adding more integrated rivals tends to shift some outcomes from the lower tail to the upper tail, increasing the proportion of mergers with benefits, especially for consumers.
-However, the interquartile range and median are largely unchanged.
-Thus, the presence of rival integrated firms can limit the extent of harm, particularly consumer harm, from vertical mergers, though not dramatically.
+However, the inter-quartile range and median are largely unchanged.
+Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers, though not dramatically.
 Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.
-It appears that these rivals can therefore somewhat limit the extent of negative impacts through RRC.
+It appears that these rivals can therefore somewhat limit the extent of negative impacts that would otherwise occur through RRC.
 
-By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues.
-A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare increases in almost 16\% of mergers and total welfare increases in 4.3\% of mergers, but median consumer harm increases to 6.3\% and median total harm increases to 4.8\% of pre-merger revenues.
-Increasing the number of integrated 3rd parties to 1 (2 integrated firms) disproportionately shrinks the inter-quartile range-- relatively more harmful mergers are eliminated than beneficial ones-- but the presence of a single integrated 3rd party has little impact on median harm.
-Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged.
-In summary, additional incumbent integrated firms do not substantially impact median harm, but do substantially reduce the likelihood of observing extremely harmful mergers.
+Moving to the left-most panel of Figure \ref{fig:CVvertincumbBW}, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues.
+\citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
+Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 18\% of mergers, and total welfare increases in 5.7\%, but median consumer harm grows to 14\% of pre-merger revenues, and median total harm grows to 7.3\%.
+Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.
+On net, we find that the harms dominate in most instances.
+Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat decreases the range of outcomes and moves the distribution towards less harm.
+With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues.
+The inter-quartile range of total harm narrows, and median total harm falls to about 3.8\% of pre-merger revenues.
+As we saw with vertical mergers, it appears that the presence of third party integrated rivals can limit harm, though here the effect is a bit stronger.
 
-Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues.
-A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare increases in about 18\% of mergers and total welfare increases in 5.7\%, but median consumer harm increases to 14\% and median total harm increases to 7.3\% of pre-merger revenues.
-Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm.
-Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm t 5.6\%.
-With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues.
-By contrast, with 3 integrated 3rd parties, the inter-quartile range of total harm shrinks, and median total harm falls to about 3.8\% of pre-merger revenues.
-These results indicate that unlike either vertical upstream mergers, integrated downstream mergers are more harmful than unintegrated downstream mergers, and that the presence of additional integrated rival incumbents reduces harm, though slowly.
+Next we examine the welfare impacts of upstream mergers.
+Like with downstream mergers, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
+This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.
+Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, but median harm is largely unchanged.
+Consumer welfare increases in almost 16\% of mergers and total welfare increases in about 4.3\% of mergers, but median consumer harm and total harm are roughly steady at approximately 6.3\% and 4.8\% of pre-merger revenues, respectively.
+The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.
+Just as we saw with downstream mergers, here we find that harm dominates in most simulations.
+Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- and slightly shrinks median harm.
+Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.
 
-Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms benefits consumers in about 7\% of simulated markets and is net beneficial in 15\% of simulated markets.
-While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of either outcome, adding a second integrated rival (4 integrated firms) eliminates the most harmful integrated mergers while also increasing median consumer harm to 14\%  and median total harm to 10.6\%.
-The presence of additional integrated incumbents has little incremental effect on  median harm.
+We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW}.
+Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 7\% of simulated markets and are net beneficial in 15\% of simulated markets.
+Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.
+While adding one additional rival incumbent integrated firm (3 integrated firms in the figure) has little effect on the distribution of either consumer or total welfare, adding a second integrated rival (4 integrated firms in the figure) eliminates the most harmful integrated mergers by shortening the lower whisker.
+Adding further integrated incumbents has little incremental effect on the distribution of harm.
+
+To summarize, our analysis of Figure \ref{fig:CVvertincumbBW} yields some overall conclusions.
+First, we find that mergers where one or both of the merging firms are already integrated (and no rivals are integrated) are only beneficial in a small proportion of instances.
+Second, the presence of integrated rivals can decrease the likelihood of observing the most harmful mergers, though the strength of that effect can vary widely across types of mergers.
 
@@
 \subsection{Mergers and Bargaining Power}
-\citeN{ST2021} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power.
-Here, we show that this result continues to hold when there are rival incumbent integrated firms.
-We also show that this relationship holds for integrated mergers as well as downstream and upstream mergers when one of the merging parties is vertically integrated.
+\citeN{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power.
+Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.
+We also show that the result persists for all types of mergers when there are rival non-merging integrated firms.
 
-%Figures \ref{fig:CVbargupBW}-\ref{fig:CVbargbothBW}  depicts box and whisker plots summarizing the consumer, retailer, wholesaler and total welfare effects for each merger as the number of incumbent integrated firms increases.
-For upstream, downstream and vertical mergers, the figures depict box and whisker plots for 0,1 and 4 incumbent vertically integrated firms.
-For downstream and upstream  mergers (Figures \ref{fig:CVbargupBW} and \ref{fig:CVbargdownBW}), the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
-By contrast, for vertical mergers ((Figures \ref{fig:CVbargvertBW}), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
-Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots for integrated mergers with 2,4, or 6 integrated firms.
-When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargbothBW} depicts a merger between two integrated firms, with all rivals are unintegrated.
-Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.
+Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top panels) and total (bottom panels) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
+Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
+As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}.
+For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
+By contrast, for upstream and downstream mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated firm when all third parties are unintegrated.
+In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.
+Figure \ref{fig:CVbargbothBW} provides analogous results for integrated mergers.
+In integrated mergers, when the number of  incumbent integrated firms equals 2, this depicts a merger between two integrated firms with all rivals are unintegrated.
+Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market, respectively.
 
-Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter increases from 1/9 (retailers have the advantage to 9 (wholesalers have the advantage).
-also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) incumbent vertically integrated firms.
-As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2021}.
-For vertical mergers (left), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated.
-By contrast, for upstream (middle) and downstream (left) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated.
-Finally, plots with 4 integrated firms increases the number of rival incumbent integrated firms in upstream and downstream mergers to 3 and in vertical mergers to 4 .
+We begin by examining vertical mergers.
+We find that the presence of integrated incumbents has little impact on the relationship between relative bargaining power and welfare.
+Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit welfare when wholesalers have relatively more power, and harm welfare when retailers have relatively more power.
+Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.
+This channel becomes less relevant as retailers gain more power.
+The impact on consumers is roughly zero when wholesalers and retailers have equal power.
+Furthermore, we find that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates harm when there are more incumbent integrated firms, although the magnitude of these differences is small.
+Across all three integrated incumbent scenarios, median consumer harm moves from about -24\% of pre-merger expenditures when wholesalers have the advantage to 7\% when retailers have the advantage.
+Similarly, median total harm goes from from about -14\% of pre-merger expenditures when wholesales have the advantage to 2.9\% when retailers have the advantage.
 
-For vertical mergers, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers.
-Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power.
-Across all three incumbent integrated firm scenarios, median consumer harm decreases from about 7\% when retailers have the advantage to -24\% when wholesalers have the advantage.
-Similarly, median total harm decreases from about 2.9\% of pre-merger expenditures when retailers have relatively more bargaining power to about -14\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
+For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.
+Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargnobothBW} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW}.
+Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Likewise, median total harm moves from about 2\% of pre-merger expenditures to 7.6\%.
+In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.
+Thus, there is little to counteract harms from decreased downstream competition.
 
-Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm.
-Median consumer harm increases from 2.2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 29\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 12\% when wholesalers have relatively more bargaining power.
-By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firm), there is a positive relationship between bargaining power and both consumer and total surplus.
-Median consumer harm decreases from about 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power to about -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -8.4\% when wholesalers have relatively more bargaining power.
-Adding 2 additional incumbent integrated firms (4 integrated firms) mitigates some of the adverse effects of mergers while still preserving the positive relationship between bargaining power and harm.
+Once we allow one of the merging firms to be integrated (denoted by 1 integrated firm in the figure), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.
+The box and whisker plots rotate clockwise around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of harm.
+Median consumer harm goes from approximately -33\% of pre-merger revenues when wholesalers have the advantage to 25\% of pre-merger revenues when retailers have the advantage.
+Median total harm goes from about 1\% to 12.2\% of pre-merger revenues.
+The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
+Compared to vertical mergers, where impacts to consumers were roughly neutral at equal bargaining power, here neutrality happens at 7/3, when wholesalers have a bit more than twice the power of retailers.
+It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.
+This is intuitive, as here the downstream merger also causes an additional lessening in horizontal competition.
+Once we add rival integrated firms (see 4 integrated firms in the figure) the box and whisker plots rotate counter-clockwise, weakening but not erasing the relationship between bargaining power and harm.
+The range of outcomes also shrinks.
+Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -33\% of pre-merger revenues to 14\% of pre-merger revenues, while median total harm goes from approximately 3.99\% to 5.5\% of pre-merger revenues.
 
-For downstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm.
-Median consumer harm decreases from about 18\% of pre-merger expenditures when retailers have relatively more bargaining power to about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm decreases from 7.6\% of pre-merger expenditures when retailers have relatively more bargaining power to about 2\% when wholesalers have relatively more bargaining power.
-By contrast, mergers between an integrated and unintegrated upstream supplier (1 integrated firms), causes the box and whisker plots to rotate clockwise around equal bargaining power, strengthening the positive relationship between bargaining power and both consumer and total surplus.
-Increasing wholesaler bargaining power causes median consumer harm to fall from 25\% of pre-merger revenues to -33\% of pre-merger revenues, while median total harm falls from 12.2\% to about 1\% of pre-merger revenues.
-Adding additional rival incumbents (4 integrated firms) has the reverse effect, rotating the box and whisker plots counter-clockwise and weakening the still positive relationship between bargaining power and surplus.
-Here, increasing wholesaler bargaining power causes median consumer harm to fall from 14\% of pre-merger revenues to -33\% of pre-merger revenues, while median total harm falls from 5.5\% to 3.99\% of pre-merger revenues.
+We turn next to upstream mergers in Figure \ref{fig:CVbargnobothBW}.
+Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.
+Median consumer harm moves from about 29\% of pre-merger expenditures when wholesalers are more powerful to 2.2\% of pre-merger expenditures when retailers are more powerful.
+Likewise, median total harm goes from approximately 12\% to 0.7\% of pre-merger expenditures.
+By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figure), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
+This is intuitive, as now all three types of mergers create some form of vertical integration.
+Median consumer harm goes from roughly -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Median total harm increases from about -8.4\% to 3.5\% of pre-merger expenditures.
+Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.
 
-Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots summarizing consumer, retailer, wholesaler and total surplus  for integrated mergers with 2,4, or 6 integrated firms.
+Figure \ref{fig:CVbargbothBW} displays results for integrated mergers.
+In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers.
+Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
+Median consumer harm moves from about 3.75\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Likewise, median total harm moves from about -14.5\% when wholesalers have relatively more bargaining power to 16\% of pre-merger expenditures when retailers have relatively more bargaining power.
+A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 4 or 6 integrated firms in the figure) has little impact on the relationship between bargaining power and harm.
+However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 or more.
+The crossing at zero harm is shifted to the left.
+
+%Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots summarizing consumer, retailer, wholesaler and total surplus  for integrated mergers with 2,4, or 6 integrated firms.
 When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargbothBW} depicts a merger between two integrated firms: all rivals are unintegrated.
@@
 
-In terms of the relationship between bargaining power and harm, integrated mergers are perhaps most similar to vertical mergers.
-Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
-Median consumer harm increases from about 26\% of pre-merger expenditures when retailers have relatively more bargaining power to about 3.75\% of pre-merger expenditures when wholesalers have relatively more bargaining power.
-Likewise, median total harm increases from 16\% of pre-merger expenditures when retailers have relatively more bargaining power to -14.5\% when wholesalers have relatively more bargaining power.
-
-A second similarity between vertical and integrated mergers is that the presence of integrated incumbents has little impact on the strong, monotonic relationship between  bargaining power and harm.
-However, unlike vertical mergers, consumer harm from integrated mergers when there are fewer incumbent integrated firms does not first-order stochastically dominate consumer harm when there are more incumbent integrated firms.
-
-An important difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power exceeds 7/3.
-In this way, integrated mergers are more similar to downstream mergers, or upstream mergers in markets with vertically integrated incumbents.
-
 %\textbf{Describe the counterintuitive result that increasing retailer bargaining power harms retailers but help wholesalers.}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Robustness to Alternative Cost Specifications}
 
-Here, we explore the role that the constant marginal cost assumption plays in driving merger outcomes.
-In theory, constant marginal costs can increase the magnitude of EDM, as the wholesaler price reduction applies equally to all units produced by the retailer.
-By contrast, if either the integrated wholesaler or retailer face linear marginal costs, increased output from the EDM is limited by the increased marginal costs, which in turn limits the benefit to the parties.
-Likewise, in theory, RRC could be more profitable for the merging parties under linear than constant marginal costs, as linear marginal costs cause the integrated wholesaler's prices to increase more rapidly.
+Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allow costs to increase linearly.
+Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).
 
-Because EDM is plausibly greater under constant marginal costs while RRC is also plausibly less, then one might expect that on net mergers are more beneficial under constant than linear marginal costs.
-We explore this hypothesis in Figure \ref{fig:surplussumcost}, which displays box and whisker plots that depict the net effect of these different marginal cost specifications on consumer and total welfare under four different scenarios:
-constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).
+In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs.
+Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause the integrated wholesaler's prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
+Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.
 
-Overall, harm is less prevalent for mergers where firms have constant marginal costs than where firms have linear costs.
-Mergers with constant-cost firms increases consumer surplus  in about 15\% of all mergers and total surplus in 10.5\% of all mergers, while mergers with linear-cost firms increase consumer surplus in only 2.3\% of mergers and total surplus in 0.8\% of all mergers.
+Our simulations confirm this hypothesis.
+See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.
+Overall, harm is greater for mergers with linear costs.
+The difference is largest for vertical mergers.
+When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of simulations and total surplus in about 18\% of simulations, but only increase consumer surplus in about 1\% of simulations and total surplus in about 0.9\% of simulations when all firms have linear costs.
+
+The bottom panel of Figure \ref{fig:surplussumcost} compares a situation where only the merging firms have constant marginal costs to one where only the merging firms have linear marginal costs.
+We see that the distributions in these bottom panels are largely the same as those in the top panels.
+Thus, it appears that the key driver of our results is whether or not the merging parties have constant marginal costs.
+When the merging firms' marginal costs are constant, the distribution of outcomes moves towards less harm.
+
+%Mergers with constant-cost firms increases consumer surplus in about 15\% of all mergers and total surplus in 10.5\% of all mergers, while mergers with linear-cost firms increase consumer surplus in only 2.3\% of mergers and total surplus in 0.8\% of all mergers.
 However, while the consumer and total harm distributions for each merger type under the all-firm linear marginal costs scenario first-order stochastically dominates their counterparts under the all-firm constant marginal costs scenario, the consumer and total harm distributions under the party linear marginal costs scenario only first-order stochastically dominates the their counterparts under the party constant marginal costs scenario for vertical and integrated mergers.
@@
 
-Changing the cost structure consistently yields the greatest differences in outcomes for vertical mergers.
+%Changing the cost structure consistently yields the greatest differences in outcomes for vertical mergers.
 When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of all vertical merger simulations and total surplus in about 18\% of all vertical merger simulations, but only increase consumer surplus in about 1\% of vertical merger simulations and total surplus in about 0.9\% of vertical merger simulations when all firms have linear costs.
@@
 
-The differences between the constant and linear cost structures is more muted for both upstream and downstream mergers.
+%The differences between the constant and linear cost structures is more muted for both upstream and downstream mergers.
 Upstream mergers with all constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs.
@@
 
-Finally, while there is a marked difference in outcomes between the constant and linear cost structures for integrated  mergers under the all-firms scenario, the differences in outcomes are more muted under the party scenario, especially for consumer surplus.
+%Finally, while there is a marked difference in outcomes between the constant and linear cost structures for integrated  mergers under the all-firms scenario, the differences in outcomes are more muted under the party scenario, especially for consumer surplus.
 In particular, median consumer harm when all firms face constant marginal costs is 13.6\%, compared to the median consumer harm of 17.5\% when all firms face linear marginal costs.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Empirical Application}
+\section{Empirical Application (In Progress)}
 
@@
 
-\subsubsection{Chattanooga Area Results}
+%\subsubsection{Chattanooga Area Results}
+%In progress.
 
@@
 
-This paper relaxes two strong assumptions made in \citeN{ST2021}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated, and that all firms employ constant marginal cost technology.
-Relaxing these constraints is important as competition authorities are called upon to investigate mergers where either the merging parties or a third party are already vertically integrated, or where the presence of a scarce input could reduce the benefit from EDM or increase the harm from RRC.
+This paper relaxes an assumption made in \citeN{ST2021}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated.
+Relaxing this constraint is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are already vertically integrated.
 Here, we show using numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained.
-We also show using simulation that relative bargaining power can serve as a useful indicia for predicting consumer and total harm from mergers where these two assumptions are relaxed.
-Finally, we intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm.
-Our hope is that future research will focus on some of the other strong assumptions made in \citeN{ST2021}, such as: how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
+We intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm.
+Our hope is that future research will focus on some of the other assumptions made in \citeN{ST2021}, such as how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
 
@@
 \end{table}
+
+\small{\include{TrashData}}
+\small{\include{TrashSims}}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=1.0]{output/TrashSimsFirm.png}
+\caption{Republic/Santek Simulation Results}
+\label{fig:TrashSimsFirms}
+\end{figure}
 \end{document}

### 2022-05-27 doc/VerticalArrangements.tex
fixed bug in code that was erroenously eliminated vertical markets. Fixed bug in code that was not properly zeoring out elements for upstream integrated products. Added figures depicting profitibility across bargaining parameters and number of integrated firms. added figure comparing outcomes across models.

@@
 For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 2.7 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
 } Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.
@@
 Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
-Pre-merger HHIs range from 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703.
+Pre-merger HHIs range from 2,923 at the $25^{th}$ percentile to 4,966 at the $75^{th}$, with a median of 3,742.
 HHIs for vertical mergers increase by 1,298 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.

### 2022-05-27 code/MonteCarlo.R

@@ -227,24 +227,26 @@ market<-function(nfirms.down=3, # # of downstream firms
 
     for( v in vertFirms){
 
-      ## set integrated margin disagreement payoff to 0
-      ownerPre.up[ids$up.firm==v & ids$down.firm!=v
-                   , ids$up.firm==v & ids$down.firm==v]=0
-      ## constrain upstream integrated margin to zero
-       ownerPre.up[ids$up.firm==v & ids$down.firm==v
-                   , ids$up.firm==v & ids$down.firm!=v]=0
 
-       ## set integrated margin disagreement payoff to 0
-       ownerPost.up[ids$up.firm==v & ids$down.firm!=v
-                   , ids$up.firm==v & ids$down.firm==v]=0
-       ## constrain upstream integrated margin to zero
-       ownerPost.up[ids$up.firm==v & ids$down.firm==v
-                   , ids$up.firm==v & ids$down.firm!=v]=0
       ## integrated retailer bargains with other wholesalers
       vertrowsDown <-ids$up.firm  != v  & ids$down.firm  == v
 
       ownerPre.up[vertrowsDown, ids$up.firm == v] <- -(1-bargparm[vertrowsDown])/bargparm[vertrowsDown]
       ownerPost.up[vertrowsDown, ids$up.firm == v] <- -(1-bargparmPost[vertrowsDown])/bargparmPost[vertrowsDown]
+
+      ## set integrated margin disagreement payoff to 0
+      ownerPre.up[!(ids$up.firm==v & ids$down.firm==v)
+                  , ids$up.firm==v & ids$down.firm==v]=0
+      ## constrain upstream integrated margin to zero
+      ownerPre.up[ids$up.firm==v & ids$down.firm==v
+                  , !(ids$up.firm==v & ids$down.firm==v)]=0
+
+      ## set integrated margin disagreement payoff to 0
+      ownerPost.up[!(ids$up.firm==v & ids$down.firm==v)
+                   , ids$up.firm==v & ids$down.firm==v]=0
+      ## constrain upstream integrated margin to zero
+      ownerPost.up[ids$up.firm==v & ids$down.firm==v
+                   ,!(ids$up.firm==v & ids$down.firm==v)]=0
 
       #ownerBargUpVertPre[vertrowsDown, ids$up.firm  == v] <- -(1-bargparm[vertrowsDown])/bargparm[vertrowsDown]
     }
[…]

### 2022-05-27 code/RunSims.R

@@ -38,7 +38,8 @@ nestParm <- c(0)
 type=c("1st")
 merger=c("up","down", "vertical","both")
 bargparm <- seq(.1,.9,.1)
-assym <- c("none","diag1","diag0")
+assym <- c("none","diag1","diag0"
+           )
 mc=c("constant","linprod","lincons","conslin","consparty","linparty"#, "quadprod","linquad","quadlin"#,"linfirm"
      )
 
@@ -83,6 +84,8 @@ M <- thismkt$down$M
 outMargin <- thismkt$down$outMargin
 #thismkt <- calcSlopes(thismkt,constrain="global");
 
+isNegMCUp <- any(thismkt$up$mc<0)
+isNegMCDown <- any(thismkt$ip$mc<0)
 
 mcParmUp <- thismkt$up$mcParm
 mcParmDown <- thismkt$down$mcParm
@@ -143,6 +146,8 @@ thisres <- with(thissum,data.frame(
   # avgPartyCostParmDown.vert = weighted.mean(mcParmDown[idDown ==1],shares.pre[idDown == 1]),
   #avgpricedelta = sum(thissum$downPricePost*shares.post,na.rm=TRUE) + outMargin*(1-sumshares.post ) - sum(thissum$downPricePre*shares.pre/sumshares.pre,na.rm=TRUE) - outMargin*(1-sumshares.pre ),
   avgpricedelta = sum(thissum$downPricePost*shares.post/sumshares.post,na.rm=TRUE) - sum(thissum$downPricePre*shares.pre/sumshares.pre,na.rm=TRUE) ,
+  isNegMCUp=isNegMCUp,
+  isNegMCDown=isNegMCDown,
   hhidelta.down = 2 * prod(tapply(shares.pre/sumshares.pre * 100,idDown, sum,na.rm=TRUE)[1:2]),
   hhipre.down = sum(tapply(shares.pre/sumshares.pre * 100,idDown, sum)^2,na.rm=TRUE),
   hhidelta.up = 2 * prod(tapply(shares.pre/sumshares.pre * 100,idUp, sum,na.rm=TRUE)[1:2]),
@@ -233,7 +238,7 @@ res.nests <- ungroup(res.nests) %>% group_by(merger,up,down,vert,type,preset,nes
          avgpricedelta=avgpricedelta,
          isProfitable=ifelse(merger=="up", isProfitable.up,
                              ifelse(merger=="down",isProfitable.down,
[…]

### 2022-05-27 code/TrashSims.R

@@ -60,6 +60,8 @@ ownerPreUpMat=tcrossprod(ownerPreUpMat)
 ownerPreDownMat=model.matrix(~-1+ownerPreDown,data=simdata)
 ownerPreDownMat=tcrossprod(ownerPreDownMat)
 
+vertFirms <- intersect(simdata$ownerPreUp,simdata$ownerPreDown)
+
 for (v in vertFirms){
   ## set integrated margin disagreement payoff to 0
   ownerPreUpMat[simdata$ownerPreUp==v & simdata$ownerPreDown!=v
@@ -74,7 +76,7 @@ sharesNoOut <- with(simdata,collection_paired_share)
 div <- tcrossprod(1/(1-sharesNoOut),sharesNoOut)*sharesNoOut
 diag(div) <- -sharesNoOut
 
-vertFirms <- intersect(simdata$ownerPreUp,simdata$ownerPreDown)
+
 
 
 minD <- function(barg,is2nd=FALSE,margins=FALSE){
@@ -112,6 +114,12 @@ else{ alpha <- simdata$alpha_2nd[1]}
     ownerPreUpMat[vertrows, simdata$ownerPreUp == v] <- -(1-b[vertrows])/b[vertrows]
   }
 
+  ## set integrated margin disagreement payoff to 0,
+  ## constrain upstream integrated margin to zero
+
+  for(n in which(simdata$ownerPreUp==simdata$ownerPreDown)){
+    ownerPreUpMat[n,-n] <- ownerPreUpMat[-n,n] <- 0
+  }
 
   ownerBargDownVert  <-  ownerPreDownMat  * (1-b)/b
 
@@ -231,6 +239,20 @@ simres_noeff <- with(simdata,
   labels = as.character(Name)
 ))
 
+simres_noeff_up <- with(simdata,
+                     logit(
+                       ownerPre = ownerPreUp,
+                       ownerPost = ownerPostUp,
+                       prices = priceDown_repmargin_bert,
+                       shares = collection_paired_share,
+                       margins = marginsDown_bert_prop,
+                       mcDelta = rep(0,nrow(simdata)),
+                       insideSize = sum(disposal_volume),
+                       labels = as.character(Name)
+                     ))
+
+
+
 
 simres_noeff_2nd <- with(simdata,
                      auction2nd.logit(
@@ -246,15 +268,13 @@ simres_noeff_2nd <- with(simdata,
 
 
 
-simres_vert <- NULL
-
[…]

### 2022-05-27 code/genfigures.R

@@ -956,3 +956,10 @@ dev.off()
 png("output/CVbargdiagonalBW.png",width = 10, height = 7, units = "in", res=300)
 print(pbargdiagonal.bw)
 dev.off()
+
+png("output/pprofits.png",width = 7, height = 7, units = "in", res=300)
+print(pprofits)
+dev.off()
+png("output/pprofits_barg.png",width = 10, height = 7, units = "in", res=300)
+print(pprofits_barg)
+dev.off()

### 2022-05-27 doc/VerticalArrangements.tex
incorproated peggy's edits

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Theory}
+\section{Theory\label{sec:theory}}
 We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2021}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Empirical Application (In Progress)}
+\section*{Empirical Application}
 
@@
 Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
-Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens’ waste themselves.
+Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens' waste themselves.
 Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
-Thus, mergers that combine hauling and disposal assets may incentivize the merged entity to raise its hauling rivals’ cost of disposal in order to benefit its own collection operations.
-Whether or not the merged entity is both able and incentivized to undertake such action depends upon the extent of its market power in the local disposal market, the merging parties’ profit margins in each line of business and their intensity of hauling competition with prospective disposal customers.
+Thus, mergers that combine hauling and disposal assets may incentivize the merged entity to raise its hauling rivals' cost of disposal in order to benefit its own collection operations.
+Whether or not the merged entity is both able and incentivized to undertake such action depends upon the extent of its market power in the local disposal market, the merging parties’ profit margins in each line of business, and their intensity of hauling competition with prospective disposal customers.
 
-In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{\tiny{Competitive Impact Statement: U.S.
+In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{Competitive Impact Statement: U.S.
 and Plantiff States v.
 Waste Management, Inc.
-and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{\tiny{Waste 360, \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{\tiny{Competitive Impcat Statement: U.S.
+and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impcat Statement: U.S.
 and State of Alabama v.
-Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
+Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{Waste Business Journal,\tiny{ \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
 
-Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
+%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
 Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
 and Plaintiff States v.
-Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}} In the DOJ complaint filed in the Republic-Santek case, vertical competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{\tiny{See U.S.
+Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
+For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anticompetitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
+In addition, vertical anticompetitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S.
 and State of Alabama v.
 Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}}
-
-Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
+%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
 The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
+The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
 
@@
 The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commerical waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers.
-The CIS notes that Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors.
-In MSW disposal, the CIS identifies only one other significant competitor and Republic and Santek as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere.
+The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors.
+In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere.
 Thus, pre-merger both parties were large, vertically integrated competitiors in the Chattanooga Area.
 
-In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, which was the parties' sole competitor in the MSW disposal market.
+In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, and was the parties' sole competitor in the MSW disposal market.
 Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
@@
 
-Data on MSW disposal comes from data collected by TDEQ (Tennessee Department of Environmental Quality) on Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+Data on MSW disposal comes from data collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
 This data identifies the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
-The MSW disposal market definition follows that outlined in the DOJ CIS and attributing share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
-Thus, the total market quantity is defined as all MSW volumes originating in Hamilton county.
+The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
+Thus, the total market quantity is defined as all MSW volumes originating in Hamilton County, Tennessee where Chattanooga is located.
 
 Disposal volumes are measured in tons and have been combined across landfills owned by the same firm.
-The TDEQ data identifies landfills owned or operated by Republic, Santek, Waste Connections, and three market participants each with less than 0.5\% share receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% combined.
+The TDEQ data identifies landfills owned or operated by Republic, Santek, Waste Connections, and three other market participants (each with less than 0.5\% share) receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% combined.
 Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out of state landfill in GA.} These three fringe participants have been excluded from the analysis.
@@
 
-However, the market shares required for implementation of the Sheu and Targain (\textbf{add ref}) framework are expressed as upstream-downstream market participant pairs.
+However, the market shares required for implementation of the Sheu-Targain (\textit{\textbf{add ref}}) framework are expressed as upstream-downstream market participant pairs.
 Following the discussion of supply relationships in Section \ref{chatt_disposal_vol}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal-Santek collection, Waste Connection disposal-Waste Connections collection) are assigned their full collection share.
@@
 
-%\subsubsection{Chattanooga Area Results}
-%In progress.
+\subsubsection{Chattanooga Area Merger Simulation Results}
+
+Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of harm to consumers and total producer benefits of approximately \$13.4 million.
+Thus, on net the model predicts the merger would be harmful despite the presence of significant efficiencies from EDM that reduce prices upstream.
+These aggregate effects obscure the complicated competitive effects between these interdependent firms though.
+These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
+
+Disposal prices for pre-merger integrated pairs are unchanged post-merger.
+However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unitegrated pairs).
+The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections prices to its downstream rivals are estimated to increase 22\%.
+For the merged firm the price increase is driven by both the effects of increased bargaining leverage and upward pricing pressure, whereas, Waste Connections' price increase is driven only by the change in bargaining leverage.
+These large price increases are offset in the aggregate, volume-weighted estimate by the large decrease in the disposal price for the integrating pair post-merger due to the EDM effect, and the cost to Republic for volume disposed with Santek decreases by 60\%.
+
+Turning to the downstream market, all collection prices increase post-merger.
+The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall.
+The pre-merger integrated Republic pair's price increases 21\% from the effect of upward pricing pressure alone and Santek's increases by 23\% from a combination of UPP and increased bargaining leverage.
+The newly integrated Santek-Republic pair's price is estimated to increase by 9\%, demonstating that the upstream efficiencies are not fully passed-through to consumers.
+Pre-merger integrated rival, Waste Connections, collection price increases by approximately 5\% and share increases by 6 percentage points.
+Unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and share increases by 2 percentage points each.
+
+\subsubsection{Merger Simulation of Collection Market Only}
+
+We compare these results to those from a merger simulation considering only a merger in the collection market under logit demand, treating each upstream-downstream pair as if it were an independent firm, as shown in Figure \ref{fig:TrashSimsCompare}.
+Ignoring these vertical relationships results in an estimated 9.4\% increase in collection prices with a 15\% increase in the merging parties' price, consumer harm of \$9.6 million and a producer benefit of \$7.1 million.
+While net harm is nearly the same under both models at about \$2.5 million, estimated consumer harm is 67\% higher in the model including vertical relationships and is offset by markedly larger producer benefits.
+Further, the collection-only model overestimates diversion to the unintegrated rivals and as a result underestimates the market price effect due to UPP from the merger.
+(See also Table \ref{tab:trash_merged}.)
+
+The Republic-Santek merger was ultimately settled with the DOJ through a divestiture.
+According to the final judgement, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station in the Chattanooga area.\footnote{Final Judgment: U.S and Plaintiff States v.
+Republic Services, Inc.
+and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anticompetitive effects of the merger.
 
@@
 \small{\include{TrashSims}}
+
+\begin{table}
+\centering
+\small{\include{trash_merged_output}}
+\caption{Predicted Market Shares for Integrated Model vs. Collection Only Model}
+\label{tab:trash_merged}
+\end{table}
 
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{output/TrashSimsCompare.png}
+\caption{Merger Simulation Results Comparison}
+\label{fig:TrashSimsCompare}
+\end{figure}
+
 \end{document}

### 2022-05-27 code/TrashSims.R

@@ -459,7 +459,12 @@ aes(x=Name,y=value,fill=Type,label=value)) +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1))
 
 
+png(filename="./output/TrashSimsFirm.png" ,res = 250, units="in"
+    ,width = 6, height = 6 )
 
+print(firmplot)
+
+dev.off()
 
 
 compareplot <- ggplot(data=  compare %>% mutate(Name=interaction(Disposal,Collector,drop=TRUE,sep="/"),
@@ -481,10 +486,10 @@ compareplot <- ggplot(data=  compare %>% mutate(Name=interaction(Disposal,Collec
 
 
 
- png(filename="./output/TrashSimsFirm.png" ,res = 250, units="in"
+ png(filename="./output/TrashSimsCompare.png" ,res = 250, units="in"
      ,width = 6, height = 6 )
 
- print(firmplot)
+ print(compareplot)
 
  dev.off()
 

### 2022-05-31 doc/VerticalArrangements.tex
Finished updated simulation numbers to reflect bugfixes in code.

@@
 
-Our simulations focus on mergers that are more likely to have competitive effects and to therefore come under agency scruitiny.
+Our simulations focus on mergers that are more likely to have competitive effects and to therefore come under agency scrutiny.
 For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger.
@@
 
-The median average wholesale pre-merger price is \$4.50, and the median average retail pre-merger price is \$13.
+The median average wholesale pre-merger price is \$5.20, and the median average retail pre-merger price is \$14.
 Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
 Pre-merger HHIs range from 2,923 at the $25^{th}$ percentile to 4,966 at the $75^{th}$, with a median of 3,742.
-HHIs for vertical mergers increase by 1,298 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
-}  HHIs for downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.
-HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238.
-HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280.
+HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,046.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
+}  HHIs for downstream mergers increase by 2,044 points at the median, resulting in a median post-merger HHI equal to 4,916.
+HHIs for upstream mergers increase by 2,023 points at the median, resulting in a median post-merger HHI equal to 4,916.
+HHIs for integrated mergers increase by 2,822 points at the median, resulting in a median post-merger HHI equal to 7,135.
 
@@
 We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.
-Across all four types of mergers, the majority of cases show harm, though the distributions and magnitudes differ.
+Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.
 In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
-Median consumer harm from integrated mergers is almost 14\% of pre-merger total expenditures, 1.25 times the magnitude of that from downstream mergers, 2.3 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.
+Median consumer harm from integrated mergers is 14\% of pre-merger total expenditures, 1.45 times the magnitude of that from downstream mergers, and 3.1 times the magnitude of that from upstream mergers.
 
-Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations.
+Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas vertical mergers, downstream mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 60\% of all simulations.
 Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.
 
-Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 77\% of all upstream mergers, 23\% of vertical mergers, 20\% of downstream mergers, and about 15\% of integrated mergers.
-However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.
+Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 67\% of all upstream mergers, 11\% of vertical mergers, 18\% of downstream mergers, and about 17\% of integrated mergers.
+The rank-ordering across mergers is also the reverse of the retailer surplus distribution.
 
-As for total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial.
-Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers.
-Median consumer harm from integrated mergers is about 9.6\% of pre-merger total expenditures, 1.5 times the magnitude of that from downstream mergers, 2.5 times the magnitude of that from upstream mergers, and more than 4.3 times the magnitude of vertical mergers.
+As for total welfare, approximately 29\% of vertical mergers and 16\% of integrated mergers are beneficial, whereas only 7\% of upstream mergers and 3\% of downstream mergers are beneficial.
+Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream  mergers, which dominates total harm from vertical mergers.
+Median consumer harm from integrated mergers is about 9.5\% of pre-merger total expenditures, 1.6 times the magnitude of that from downstream mergers, 2.6 times the magnitude of that from upstream mergers, and more than 4.6 times the magnitude of vertical mergers.
 
@@
 Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
-Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases.
+Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top row) and total harm (bottom row) as the number of incumbent integrated firms increases.
 The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
@@
 
-We begin with vertical mergers, as they are our simplest case, because there is never any preexisting integration at either merging firm.
-Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.
-The addition of a single rival integrated firm (1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm basically the same.
-Adding more integrated rivals tends to shift some outcomes from the lower tail to the upper tail, increasing the proportion of mergers with benefits, especially for consumers.
-However, the inter-quartile range and median are largely unchanged.
-Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers, though not dramatically.
+We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
+Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.
+The addition of a single rival integrated firm (1 integrated firm in the figure) increases median consumer harm from 0.2\% to to 2\% while widening the distribution of outcomes slightly.
+By contrast, additional integrating rivals has little effect on median total harm.
+Adding more integrated rivals tends to increase the proportion of mergers that benefit consumers, while also increasing the proportion of mergers that reduce total surplus.
+Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers.
 Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.
@@
 
-Moving to the left-most panel of Figure \ref{fig:CVvertincumbBW}, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues.
+Moving to the second panel from the left in Figure \ref{fig:CVvertincumbBW}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
+This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.
+Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, and median harm changes modestly.
+Consumer welfare increases in almost 30\% of mergers and total welfare increases in about 10\% of mergers, but median consumer harm and total harm are roughly steady at approximately 4\% pre-merger revenues.
+The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.
+Just as we saw with downstream mergers, here we find that harm dominates in most simulations.
+Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- while only slightly changing median harm.
+Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.
+
+Next we examine the welfare impacts of downstream mergers.
+As with upstream mergers,  downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues.
 \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
-Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 18\% of mergers, and total welfare increases in 5.7\%, but median consumer harm grows to 14\% of pre-merger revenues, and median total harm grows to 7.3\%.
+Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 26\% of mergers, and total welfare increases in about 6\%, but median consumer harm grows to about 11\% of pre-merger revenues, and median total harm grows to about 7\%.
 Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.
@@
 Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat decreases the range of outcomes and moves the distribution towards less harm.
-With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues.
-The inter-quartile range of total harm narrows, and median total harm falls to about 3.8\% of pre-merger revenues.
+With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 34\% of markets and shrinks median consumer harm to about 5\% of pre-merger revenues.
+The inter-quartile range of total harm narrows, and median total harm falls to about 4\% of pre-merger revenues.
 As we saw with vertical mergers, it appears that the presence of third party integrated rivals can limit harm, though here the effect is a bit stronger.
 
-Next we examine the welfare impacts of upstream mergers.
-Like with downstream mergers, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
-This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.
-Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, but median harm is largely unchanged.
-Consumer welfare increases in almost 16\% of mergers and total welfare increases in about 4.3\% of mergers, but median consumer harm and total harm are roughly steady at approximately 6.3\% and 4.8\% of pre-merger revenues, respectively.
-The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.
-Just as we saw with downstream mergers, here we find that harm dominates in most simulations.
-Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- and slightly shrinks median harm.
-Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.
-
 We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW}.
-Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 7\% of simulated markets and are net beneficial in 15\% of simulated markets.
+Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 8\% of simulated markets and are net beneficial in 15\% of simulated markets.
 Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.
 While adding one additional rival incumbent integrated firm (3 integrated firms in the figure) has little effect on the distribution of either consumer or total welfare, adding a second integrated rival (4 integrated firms in the figure) eliminates the most harmful integrated mergers by shortening the lower whisker.
-Adding further integrated incumbents has little incremental effect on the distribution of harm.
+Adding further integrated incumbents has modest incremental effects on the distribution of harm.
 
@@
 
+We turn next to upstream mergers in Figure \ref{fig:CVbargnobothBW}.
+Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.
+Median consumer harm moves from about 28\% of pre-merger expenditures when wholesalers are more powerful to 2\% of pre-merger expenditures when retailers are more powerful.
+Likewise, median total harm goes from approximately 12\% to less than 1\% of pre-merger expenditures.
+By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figure), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
+This is intuitive, as now all three types of mergers create some form of vertical integration.
+Median consumer harm goes from roughly -23\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Median total harm increases from about -20\% to more than 3\% of pre-merger expenditures.
+Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.
+
 For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.
@@
 Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Likewise, median total harm moves from about 2\% of pre-merger expenditures to 7.6\%.
+Likewise, median total harm moves from about 2\% of pre-merger expenditures to about 8\%.
 In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.
@@
 The box and whisker plots rotate clockwise around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of harm.
-Median consumer harm goes from approximately -33\% of pre-merger revenues when wholesalers have the advantage to 25\% of pre-merger revenues when retailers have the advantage.
-Median total harm goes from about 1\% to 12.2\% of pre-merger revenues.
+Median consumer harm goes from approximately -33\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.
+Median total harm goes from about 2\% to almost 12\% of pre-merger revenues.
 The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
-Compared to vertical mergers, where impacts to consumers were roughly neutral at equal bargaining power, here neutrality happens at 7/3, when wholesalers have a bit more than twice the power of retailers.
+Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here neutrality happens between  3/2 and 7/3, when wholesalers have a bit more than 1.5 times the power of retailers.
 It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.
@@
 The range of outcomes also shrinks.
-Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -33\% of pre-merger revenues to 14\% of pre-merger revenues, while median total harm goes from approximately 3.99\% to 5.5\% of pre-merger revenues.
-
-We turn next to upstream mergers in Figure \ref{fig:CVbargnobothBW}.
-Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.
-Median consumer harm moves from about 29\% of pre-merger expenditures when wholesalers are more powerful to 2.2\% of pre-merger expenditures when retailers are more powerful.
-Likewise, median total harm goes from approximately 12\% to 0.7\% of pre-merger expenditures.
-By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figure), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
-This is intuitive, as now all three types of mergers create some form of vertical integration.
-Median consumer harm goes from roughly -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Median total harm increases from about -8.4\% to 3.5\% of pre-merger expenditures.
-Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.
+Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -35\% of pre-merger revenues to 13\% of pre-merger revenues, while median total harm goes from almost 3\% to 5\% of pre-merger revenues.
 
@@
 Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
-Median consumer harm moves from about 3.75\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Likewise, median total harm moves from about -14.5\% when wholesalers have relatively more bargaining power to 16\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Median consumer harm moves from less than -2\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Likewise, median total harm moves from about -20\% when wholesalers have relatively more bargaining power to 15\% of pre-merger expenditures when retailers have relatively more bargaining power.
 A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 4 or 6 integrated firms in the figure) has little impact on the relationship between bargaining power and harm.
@@
 The difference is largest for vertical mergers.
-When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of simulations and total surplus in about 18\% of simulations, but only increase consumer surplus in about 1\% of simulations and total surplus in about 0.9\% of simulations when all firms have linear costs.
+When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 55\% of simulations and total surplus in about 29\% of simulations, but only increase consumer surplus in about 36\% of simulations and total surplus in about 2\% of simulations when all firms have linear costs.
 
@@
 In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{Competitive Impact Statement: U.S.
-and Plantiff States v.
+and Plaintiff States v.
 Waste Management, Inc.
-and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impcat Statement: U.S.
+and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impact Statement: U.S.
 and State of Alabama v.
@@
 Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
-For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anticompetitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
-In addition, vertical anticompetitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S.
+For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
+In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S.
 and State of Alabama v.
@@
 
-The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commerical waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers.
-The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors.
+The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commercial waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers.
+The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
 In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere.
-Thus, pre-merger both parties were large, vertically integrated competitiors in the Chattanooga Area.
+Thus, pre-merger both parties were large, vertically integrated competitors in the Chattanooga Area.
 
@@
 Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
-Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consumation since the merger was completed before the filing of the CIS for the Republic-Santek merger.
+Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consummation since the merger was completed before the filing of the CIS for the Republic-Santek merger.
 The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.
@@
 Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out of state landfill in GA.} These three fringe participants have been excluded from the analysis.
-After rescaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.
+After re-scaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.
 
-Using this data in conjunction with information on collection market volumes, discussed in Section \ref{chatt_collection_share}, the supply relationships between the upstream disposal market participants (\textit{Republic, Santek, and Waste Connections}) and downstream collection market particpants (\textit{Republic, Santek, Waste Connections, WM-ADS, and ``Regional''}) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Using this data in conjunction with information on collection market volumes, discussed in Section \ref{chatt_collection_share}, the supply relationships between the upstream disposal market participants (\textit{Republic, Santek, and Waste Connections}) and downstream collection market participants (\textit{Republic, Santek, Waste Connections, WM-ADS, and ``Regional''}) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
 Santek and Waste Connections exhibit excess disposal capacity.
 However, Republic's collection volumes are estimated to exceed their ability to self-supply disposal in the Chattanooga market.
-As a result, Republic will need to puchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
+As a result, Republic will need to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
 We assume that this disposal is purchased from both Santek and Waste connections according to their share of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
@@
 Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
-After rescaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WM-ADS, 8.3\%, and Regional, 8.3\%.
+After re-scaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WM-ADS, 8.3\%, and Regional, 8.3\%.
 
-However, the market shares required for implementation of the Sheu-Targain (\textit{\textbf{add ref}}) framework are expressed as upstream-downstream market participant pairs.
+However, the market shares required for implementation of the Sheu-Taragin (\textit{\textbf{add ref}}) framework are expressed as upstream-downstream market participant pairs.
 Following the discussion of supply relationships in Section \ref{chatt_disposal_vol}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal-Santek collection, Waste Connection disposal-Waste Connections collection) are assigned their full collection share.
@@
 The estimated share of these costs attributable to the Chattanooga market, based on the number of markets in which the companies operate, is divided by the tons disposed for each company.
-Santek, Waste Connections, and Regional do not publically produce comparable financial statements.
+Santek, Waste Connections, and Regional do not publicly produce comparable financial statements.
 Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure in this market due to their lack of internal disposal capacity.
@@
 A survey of demand elasticity estimates for the collection market are present in Bel and Gradus (2016) with an average of -0.34.
-These estimates are not specific to the SCCW market, but available evidence suggests that the commerical segment should be more inelastic than other segments such that analysis using these estimates would tend to be conservative for estimating the size of merger induced price effects.\footnote{Bel, G.
+These estimates are not specific to the SCCW market, but available evidence suggests that the commercial segment should be more inelastic than other segments such that analysis using these estimates would tend to be conservative for estimating the size of merger induced price effects.\footnote{Bel, G.
 and R.
@@
 Disposal prices for pre-merger integrated pairs are unchanged post-merger.
-However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unitegrated pairs).
+However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs).
 The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections prices to its downstream rivals are estimated to increase 22\%.
@@
 The pre-merger integrated Republic pair's price increases 21\% from the effect of upward pricing pressure alone and Santek's increases by 23\% from a combination of UPP and increased bargaining leverage.
-The newly integrated Santek-Republic pair's price is estimated to increase by 9\%, demonstating that the upstream efficiencies are not fully passed-through to consumers.
+The newly integrated Santek-Republic pair's price is estimated to increase by 9\%, demonstrating that the upstream efficiencies are not fully passed-through to consumers.
 Pre-merger integrated rival, Waste Connections, collection price increases by approximately 5\% and share increases by 6 percentage points.
@@
 Republic Services, Inc.
-and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anticompetitive effects of the merger.
+and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
 

### 2022-05-31 code/genfigures.R

@@ -793,7 +793,7 @@ pbargvert_all.bw <- pbargvert_all.bw + geom_segment(
 
 
 
-pbargboth_all.bw <- ggplot(filter(res.nests.logit,Merger =="Integrated" & vert != "5" & vert !="3"), aes(y=Outcome_value/mktrev.pre*100,
+pbargboth_all.bw <- ggplot(filter(res.nests.logit,Merger =="Integrated"  & !Outcome %in% c("Retailer","Wholesaler") & vert != "5" & vert !="3"), aes(y=Outcome_value/mktrev.pre*100,
                                                                              #avgpricedelta/mktrev.pre*100,
                                                                              x=factor(barg,labels=MASS::fractions(relleverage)),color=vert)) +
   #geom_boxplot(outlier.alpha = 0.1) +
@@ -828,11 +828,11 @@ pbargboth_all.bw <- ggplot(filter(res.nests.logit,Merger =="Integrated" & vert !
 pbargboth_all.bw <- pbargboth_all.bw + geom_segment(
   aes(x=x,xend=xend,y=y,yend=y),color="black",arrow=arrow(length=unit(0.3,"cm"),ends="last",type="closed"),size=1, show.legend = FALSE,
   data=data.frame(x=5.1,y=-25,xend=7.5
-                  ,Outcome=factor( "Wholesaler" ,levels=unique(res.nests.logit$Outcome)))) +
+                  ,Outcome=factor( "Total" ,levels=unique(res.nests.logit$Outcome)))) +
   geom_text(aes(x=x,y=y),color="black",label="equal power",angle=90,
-            data=data.frame(x=4.7,y=15,Outcome=factor( "Wholesaler" ,levels=unique(res.nests.logit$Outcome))),size=3.5) +
+            data=data.frame(x=4.7,y=15,Outcome=factor( "Total" ,levels=unique(res.nests.logit$Outcome))),size=3.5) +
   geom_text(aes(x=x,y=y),color="black",label="more\n retailer\n power",
-            data=data.frame(x=6.5,y=-20,Outcome=factor( "Wholesaler" ,levels=unique(res.nests.logit$Outcome))),size=3)
+            data=data.frame(x=6.5,y=-20,Outcome=factor( "Total" ,levels=unique(res.nests.logit$Outcome))),size=3)
 
 
[…]

### 2022-05-31 doc/VerticalArrangements.tex
fixed issue with how compareplot was being outputted. Minor edits to intro and conclusion.

@@
 
-We then provide a case study (in progress) from an industry, solid waste management, that has seen a large number of mergers with both horizontal and vertical elements.
+We then employ this model to investigate the 2021 merger of two solid waste management firms, Republic and Santek, in one geographic market: Chattanooga, Tennessee.
+The solid waste management industry has seen a large number of mergers with both horizontal and vertical elements.
 Broadly speaking, the supply chain of this sector involves haulers at the downstream level that bring trash to disposal facilities at the upstream level.
 A variety of firms participate in these services, with some owning their own hauling and disposal assets, while others are active in only a segment of the market.
+We find that ignoring vertical integration substantially understates harm to consumers in Chattanooga.
 
@@
 Here, we show using numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained.
-We intend to use two recent trash mergers to investigate the importance of these assumptions in predicting merger harm.
+This is further illustrated by the results from our simulations in the Republic/Santek merger, where we find that ignoring existing vertical relationships drastically understates harm to Chattanooga consumers.
 Our hope is that future research will focus on some of the other assumptions made in \citeN{ST2021}, such as how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.

### 2022-05-31 code/TrashSims.R

@@ -459,12 +459,7 @@ aes(x=Name,y=value,fill=Type,label=value)) +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1))
 
 
-png(filename="./output/TrashSimsFirm.png" ,res = 250, units="in"
-    ,width = 6, height = 6 )
 
-print(firmplot)
-
-dev.off()
 
 
 compareplot <- ggplot(data=  compare %>% mutate(Name=interaction(Disposal,Collector,drop=TRUE,sep="/"),
@@ -486,11 +481,18 @@ compareplot <- ggplot(data=  compare %>% mutate(Name=interaction(Disposal,Collec
 
 
 
+ png(filename="./output/TrashSimsFirm.png" ,res = 250, units="in"
+     ,width = 6, height = 6 )
+
+ print(firmplot)
+
+ dev.off()
+
+
  png(filename="./output/TrashSimsCompare.png" ,res = 250, units="in"
      ,width = 6, height = 6 )
 
  print(compareplot)
 
  dev.off()
-
 

### 2022-06-01 doc/VerticalArrangements.tex
separated consumer and total welfare into 2 figures for Figs 2 and 3. Added Integrated mergers to Figure 3. dropped figure 4 from paper. Added all the new figures to the paper,but have not modified text.

@@
 Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
-Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top row) and total harm (bottom row) as the number of incumbent integrated firms increases.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+Figures \ref{fig:CVvertincumbBW_consumer} \ref{fig:CVvertincumbBW_total} depict box and whisker plots summarizing consumer harm and total harm as the number of incumbent integrated firms increases.
+The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figures \ref{fig:CVvertincumbBW_consumer} \ref{fig:CVvertincumbBW_total} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
 By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties.
@@
 We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
-Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.
+Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW_consumer}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.
 The addition of a single rival integrated firm (1 integrated firm in the figure) increases median consumer harm from 0.2\% to to 2\% while widening the distribution of outcomes slightly.
@@
 
-Moving to the second panel from the left in Figure \ref{fig:CVvertincumbBW}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
+Moving to the second panel from the left in Figure \ref{fig:CVvertincumbBW_consumer}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
 This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.
@@
 
-We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW}.
+We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW_consumer}.
 Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 8\% of simulated markets and are net beneficial in 15\% of simulated markets.
@@
 
-To summarize, our analysis of Figure \ref{fig:CVvertincumbBW} yields some overall conclusions.
+To summarize, our analysis of Figure \ref{fig:CVvertincumbBW_consumer} yields some overall conclusions.
 First, we find that mergers where one or both of the merging firms are already integrated (and no rivals are integrated) are only beneficial in a small proportion of instances.
@@
 
-Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top panels) and total (bottom panels) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
+Figure \ref{fig:CVbargincumbent_consumer}  depicts box and whisker plots summarizing the consumer (top panels) and total (bottom panels) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
 Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
-As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}.
+As in Figure \ref{fig:CVvertincumbBW_consumer}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}.
 For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
@@
 In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.
-Figure \ref{fig:CVbargbothBW} provides analogous results for integrated mergers.
+Figure \ref{fig:CVbargincumbent_consumer} provides analogous results for integrated mergers.
 In integrated mergers, when the number of  incumbent integrated firms equals 2, this depicts a merger between two integrated firms with all rivals are unintegrated.
@@
 
-We turn next to upstream mergers in Figure \ref{fig:CVbargnobothBW}.
+We turn next to upstream mergers in Figure \ref{fig:CVbargincumbent_consumer}.
 Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.
@@
 For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.
-Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargnobothBW} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW}.
+Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargincumbent_consumer} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW_consumer}.
 Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power.
@@
 
-Figure \ref{fig:CVbargbothBW} displays results for integrated mergers.
+Figure \ref{fig:CVbargincumbent_consumer} displays results for integrated mergers.
 In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers.
@@
 
-%Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots summarizing consumer, retailer, wholesaler and total surplus  for integrated mergers with 2,4, or 6 integrated firms.
-When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargbothBW} depicts a merger between two integrated firms: all rivals are unintegrated.
+%Finally, Figure \ref{fig:CVbargincumbent_consumer} depicts box and whisker plots summarizing consumer, retailer, wholesaler and total surplus  for integrated mergers with 2,4, or 6 integrated firms.
+When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargincumbent_consumer} depicts a merger between two integrated firms: all rivals are unintegrated.
 Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.
@@
 \centering
-\includegraphics[scale=0.9]{output/CVvertincumbBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (top row) and total (bottom) surplus as the number of vertically integrated firms present in a market change.
+\includegraphics[scale=0.9]{output/CVvertincumbBW_consumer.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market change.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
 }
-\label{fig:CVvertincumbBW}
+\label{fig:CVvertincumbBW_consumer}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\centering
+\includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market change.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
+}
+\label{fig:CVvertincumbBW_total}
 \end{sidewaysfigure}
@@
 \centering
-\includegraphics[scale=0.9]{output/CVbargnobothBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which downstream, upstream, and vertical mergers affect consumer, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which bertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
 The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargnobothBW}
+\label{fig:CVbargincumbent_consumer}
 \end{sidewaysfigure}
@@
 \centering
-\includegraphics[scale=0.85]{output/CVbargbothBW.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which bertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
 The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargbothBW}
+\label{fig:CVbargincumbent_total}
 \end{sidewaysfigure}
+
+% \begin{sidewaysfigure}
+% \centering
+% \includegraphics[scale=0.85]{output/CVbargbothBW.png}
+% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
+% \label{fig:CVbargbothBW}
+% \end{sidewaysfigure}
 

### 2022-06-01 code/genfigures.R

@@ -131,6 +131,71 @@ pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
   )
 
+pvertincumb_c.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer") #& Cost=="Constant"
+) %>% mutate(vert=as.numeric(as.character(vert)),
+        vert=ifelse(merger=="both",vert-2,vert),
+        vert=factor(vert)), aes(y=Outcome_value/mktrev.pre*100,
+       #avgpricedelta/mktrev.pre*100,
+       x=vert#color=Cost
+       #,color=Outcome
+)
+) +
+  stat_summary(fun.data=boxfun, geom="boxplot",position="dodge")+
+  #coord_cartesian(ylim=c(-60,55))+
+  scale_y_continuous(breaks=seq(-100,100,10))+
+  geom_hline(yintercept=0,linetype="dashed",color="goldenrod")+
+  theme_bw()+scale_colour_tableau('Color Blind')+ theme(legend.position="bottom")+
+  #scale_x_discrete(labels=rev(levels(res.barg$relleveragePre)))+
+  #scale_x_discrete(drop=FALSE,labels=ifelse(levels(res.barg$relleveragePre) %in% as.character(round(relleveragePre,1)),levels(res.barg$relleveragePre),""))+
+  #theme_tufte(ticks=FALSE) +
+  #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
+  #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
+  facet_grid(~Merger,scales="free",labeller = label_context)+
+  xlab("# Incumbent Vertically Integrated Firms")+
+  ylab("Consumer Surplus (%)")+
+  #ylab("Avg. Downstream Price Change (%)")+
+  #ylab("Share-Weighted Downstream Price Change")+
+  #geom_text(data=ann_text,label="Wholesale advantage")
+  labs(colour="Outcome:")+
+  labs(title =  "The Distribution of Consumer Surplus Changes as the Number of Integrated Firms Increases",
+       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures.\nHorizontal mergers occur between a vertically integrated and unintegrated firm."
[…]

### 2022-06-16 code/MonteCarlo.R
fixed bug is nested logit code that created too many nesting parameters when there was only a single nest. This bug affects the flat logit case. Fixing this fixed the issue Gloria raised with regards to so many upstream mergers that yileded consumer benefit.

@@ -28,7 +28,7 @@ market<-function(nfirms.down=3, # # of downstream firms
                  # or a vertical merger
                  nests, # specify nesting structure. default is merging parties in same nest
                  nestParm = 0, # single nesting parameter. Default is flat Logit
-                 isNestedTogether=TRUE, # specify if parties are in the same nest (TRUE) or different nests (FALSE)
+                 isNestedTogether=c("none","same","different"), # specify if parties are in the same nest ("same") or different nests ("different")
                  outMargin=runif(1,1,1e2),
                  shareOutDown=runif(1,.01,.1), # outside good margin
                  shareParm=rep(1,nfirms.down*nprods.down*nfirms.up*nprods.up), # parameter for dirichlet (share) distribution (default is uniform)
@@ -39,6 +39,7 @@ market<-function(nfirms.down=3, # # of downstream firms
   ownerPost <- match.arg(ownerPost)
   supply.down <- match.arg(supply.down)
   bargpreset <- match.arg(bargpreset)
+  isNestedTogether <- match.arg(isNestedTogether)
 
 
   if(any(is.na(nestParm) | nestParm >=1 | nestParm<0)){stop("nestParm must be >=0 and <1")}
@@ -103,32 +104,34 @@ market<-function(nfirms.down=3, # # of downstream firms
   #if(ownerPost=="vertical"){isJoint <-  ids$up.firm == 1 & ids$down.firm == 1}
 
   ## create nesting structures for different merger types
-  if( missing(nests)){
-    if(isNestedTogether){
+if(missing(nests) && max(nestParm,na.rm=TRUE)>0){
+    if(isNestedTogether =="same"){
       if(ownerPost %in% c("up","both")) nests <- ids$up.firm == 1 | ids$up.firm == 2
       else if(ownerPost == "down")      nests <- ids$down.firm == 1 | ids$down.firm == 2
       else if (ownerPost == "vertical") nests <- ids$up.firm == 1 | ids$down.firm == 1
     }
 
-    else{
+    else if(isNestedTogether =="different"){
       if(ownerPost %in% c("up","both")) nests <- ids$up.firm == 1 | ids$up.firm == 3
       else if(ownerPost == "down")      nests <- ids$down.firm == 1 | ids$down.firm == 3
[…]

### 2022-06-16 doc/VerticalArrangements.tex
removed nests summary from summary table. started fixing text in paper.

@@
 
-The median average wholesale pre-merger price is \$5.20, and the median average retail pre-merger price is \$14.
+The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.
 Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
-Pre-merger HHIs range from 2,923 at the $25^{th}$ percentile to 4,966 at the $75^{th}$, with a median of 3,742.
-HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,046.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
-}  HHIs for downstream mergers increase by 2,044 points at the median, resulting in a median post-merger HHI equal to 4,916.
-HHIs for upstream mergers increase by 2,023 points at the median, resulting in a median post-merger HHI equal to 4,916.
-HHIs for integrated mergers increase by 2,822 points at the median, resulting in a median post-merger HHI equal to 7,135.
+Pre-merger HHIs range from 2,911 at the $25^{th}$ percentile to 4,980 at the $75^{th}$, with a median of 3,747.
+HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,045.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
+}  HHIs for downstream mergers increase by 2,126 points at the median, resulting in a median post-merger HHI equal to 5,149.
+HHIs for upstream mergers increase by 2,069 points at the median, resulting in a median post-merger HHI equal to 4,967.
+HHIs for integrated mergers increase by 2,801 points at the median, resulting in a median post-merger HHI equal to 7,135.
 
@@
 Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.
-In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.
-Median consumer harm from integrated mergers is 14\% of pre-merger total expenditures, 1.45 times the magnitude of that from downstream mergers, and 3.1 times the magnitude of that from upstream mergers.
+In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integretaed mergers first-order stochastically dominates consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers.
+However, while consumer harm is typically greater under integrated mergers than downstream mergers, it is not first-order stochastically dominant.
+Median consumer harm from integrated mergers is about 14\% of pre-merger total expenditures, 1.3 times the magnitude of that from downstream mergers, and 3 times the magnitude of that from upstream mergers.
 

### 2022-06-16 code/RunSims.R

@@ -331,10 +331,11 @@ sumtable <- spread(sumtable, quant, val) %>% select(-Max,Max) %>%
   mutate(
     #set=factor(set,levels=c("firm","bargaining"),labels=c("Firm Count","Bargaining Power")),
     merger=factor(merger, levels=c("all","vertical","up","down","both"),labels=c("All","Vertical","Upstream","Downstream","Integrated")),
-    variable = factor(variable, levels=c("up","down","vert","barg","nestParm",
+    variable = factor(variable, levels=c("up","down","vert","barg",#"nestParm",
                                          "avgpricepre.up","avgpricepre.down","mktElast",
                                          "hhipre","hhipost","hhidelta"),
-                      labels = c("# Wholesalers","# Retailers","# Integrated","Bargaining Power","Nesting Parameter","Avg. Upstream Price ($)","Avg. Downstream Price ($)","Market Elasticity",
+                      labels = c("# Wholesalers","# Retailers","# Integrated","Bargaining Power",#"Nesting Parameter",
+                                 "Avg. Upstream Price ($)","Avg. Downstream Price ($)","Market Elasticity",
                                  "Pre-Merger HHI",
                                  "Post-Merger HHI","Delta HHI"
                       )

### 2022-07-21 doc/TrashData.tex
Changed Santek:Santek marginal costs so they equal the the calibrated marginal costs of other Santek products.

@@
 Republic & Republic & 165 & 42 & 0 & 44 & 179\\
-Santek & Republic & 34 & 36 & 20 &  & 174\\
+Santek & Republic & 34 & 36 & 20 &  & 173\\
 Santek & Santek & 218 & 16 & 0 &  & 134\\

### 2022-07-21 code/TrashSims.R

@@ -17,7 +17,11 @@ paired_data_4_22 <- read_csv("~/Projects/bargaining_convex/trash/paired_data_4_2
 disposal_price_republic <-  96 # source: 4/26 correspondence with Peggy
 
 simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
-  mutate(collection_paired_share=collection_paired_share/sum(collection_paired_share),
+  #group_by(disposal_firm_name) %>%
+  #mutate(disposal_volume=ifelse(disposal_firm_name != collection_firm_name,mean(disposal_volume[disposal_firm_name != collection_firm_name]),disposal_volume)) %>%
+  ungroup() %>%
+  mutate(#collection_paired_share=collection_paired_share/sum(collection_paired_share),
+         collection_paired_share=disposal_volume/sum(disposal_volume),
          disposal_dollar_margin=ifelse(disposal_firm_name=="Republic",
                                        disposal_percent_margin*disposal_price_republic,
                                        disposal_dollar_margin),
@@ -25,7 +29,7 @@ simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
          # and use that margin to recover costs
          disposal_price=ifelse(disposal_firm_name=="Republic",
                                disposal_price_republic - disposal_dollar_margin,disposal_price),
-         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.5),
+         barg=ifelse(disposal_firm_name==collection_firm_name,1,0.6),
          disposal_dollar_margin=ifelse(disposal_firm_name %in% c("Santek","WasteConn"),
                                        max(disposal_percent_margin,na.rm=TRUE)*disposal_price,
                                        disposal_dollar_margin)
@@ -35,7 +39,9 @@ simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
     disposal_dollar_margin=max(disposal_dollar_margin,na.rm=TRUE),
     disposal_price=ifelse(is.na(disposal_price),max(disposal_price,na.rm=TRUE) - disposal_dollar_margin,disposal_price),
[…]

### 2022-07-21 doc/VerticalArrangements.tex
Ran simulations with Dirichlet parameter 2.5. No meaningful difference, asside from lower market elasticity, so I am reverting back to Dirichlet 1 (uniform).

@@
 Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.
-In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integretaed mergers first-order stochastically dominates consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers.
+In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integrated mergers first-order stochastically dominates consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers.
 However, while consumer harm is typically greater under integrated mergers than downstream mergers, it is not first-order stochastically dominant.
-Median consumer harm from integrated mergers is about 14\% of pre-merger total expenditures, 1.3 times the magnitude of that from downstream mergers, and 3 times the magnitude of that from upstream mergers.
+Median consumer harm from integrated mergers is about 14\% of pre-merger expenditures, 1.3 times the magnitude of downstream mergers, and 3 times the magnitude from upstream mergers.
 
-Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas vertical mergers, downstream mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 60\% of all simulations.
-Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.
+Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas vertical mergers, downstream mergers, and integrated mergers almost always benefit retailers, upstream mergers harm retailers in about 60\% of all simulations.
+Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, while the retailer surplus distribution from vertical mergers stochastically dominates the distribution from upstream mergers.
+The median gain to retailers from integrated mergers is about 13\% of pre-merger expenditures,  1.5 times the magnitude of downstream mergers, and 1.1 times the magnitude of that from vertical mergers.
 
-Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 67\% of all upstream mergers, 11\% of vertical mergers, 18\% of downstream mergers, and about 17\% of integrated mergers.
-The rank-ordering across mergers is also the reverse of the retailer surplus distribution.
+Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 64\% of all upstream mergers, 11\% of vertical mergers, and about 16\% of downstream  and integrated mergers.
+Here, wholesaler surplus under integrated, downstream and upstream mergers each stochastically dominate the surplus from vertical mergers, but not one another.
+The rank ordering  of median harm, however, is the reverse of that from
 
-As for total welfare, approximately 29\% of vertical mergers and 16\% of integrated mergers are beneficial, whereas only 7\% of upstream mergers and 3\% of downstream mergers are beneficial.
+As for total welfare, approximately 29\% of vertical mergers and 17\% of integrated mergers are beneficial, whereas only 9\% of upstream mergers and 2\% of downstream mergers are beneficial.
 Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream  mergers, which dominates total harm from vertical mergers.
-Median consumer harm from integrated mergers is about 9.5\% of pre-merger total expenditures, 1.6 times the magnitude of that from downstream mergers, 2.6 times the magnitude of that from upstream mergers, and more than 4.6 times the magnitude of vertical mergers.
+Median consumer harm from integrated mergers is about 9\% of pre-merger total expenditures, 1.4 times the magnitude of that from downstream mergers, 2.6 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.
 
@@
 Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
-Figures \ref{fig:CVvertincumbBW_consumer} \ref{fig:CVvertincumbBW_total} depict box and whisker plots summarizing consumer harm and total harm as the number of incumbent integrated firms increases.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figures \ref{fig:CVvertincumbBW_consumer} \ref{fig:CVvertincumbBW_total} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total} depict box and whisker plots summarizing consumer harm and total harm as the number of incumbent integrated firms increases.
+The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figures \ref{fig:CVvertincumbBW_consumer},  \ref{fig:CVvertincumbBW_total} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
 By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties.

### 2022-07-21 code/MonteCarlo.R

@@ -31,7 +31,7 @@ market<-function(nfirms.down=3, # # of downstream firms
                  isNestedTogether=c("none","same","different"), # specify if parties are in the same nest ("same") or different nests ("different")
                  outMargin=runif(1,1,1e2),
                  shareOutDown=runif(1,.01,.1), # outside good margin
-                 shareParm=rep(1,nfirms.down*nprods.down*nfirms.up*nprods.up), # parameter for dirichlet (share) distribution (default is uniform)
+                 shareParm=rep(2.5,nfirms.down*nprods.down*nfirms.up*nprods.up), # parameter for dirichlet (share) distribution (default is uniform)
                  largeMerger=FALSE, # force the firms with the largest market shares to merge,
                  ...){
 

### 2022-07-25 doc/VerticalArrangements.tex
Reran figures using Dirichelt parameter of 1. Incorporated new numbers into paper. Minor other changes to paper.

@@
 Furthermore, the views expressed here should not be purported to reflect those of the Federal Trade Commission or the U.S.
-Department of Justice.}}
+Department of Justice.
+This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}
 
@@
 Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
-Pre-merger HHIs range from 2,911 at the $25^{th}$ percentile to 4,980 at the $75^{th}$, with a median of 3,747.
-HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,045.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
-}  HHIs for downstream mergers increase by 2,126 points at the median, resulting in a median post-merger HHI equal to 5,149.
-HHIs for upstream mergers increase by 2,069 points at the median, resulting in a median post-merger HHI equal to 4,967.
-HHIs for integrated mergers increase by 2,801 points at the median, resulting in a median post-merger HHI equal to 7,135.
+Pre-merger HHIs range from 2,911 at the $25^{th}$ percentile to 4,981 at the $75^{th}$, with a median of 3,747.
+HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,048.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
+}  HHIs for downstream mergers increase by 2,128 points at the median, resulting in a median post-merger HHI equal to 5,153.
+HHIs for upstream mergers increase by 2,070 points at the median, resulting in a median post-merger HHI equal to 4,968.
+HHIs for integrated mergers increase by 2,803 points at the median, resulting in a median post-merger HHI equal to 6,998.
 
@@
 Here, wholesaler surplus under integrated, downstream and upstream mergers each stochastically dominate the surplus from vertical mergers, but not one another.
-The rank ordering  of median harm, however, is the reverse of that from
 
@@
 We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
-Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW_consumer}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.
-The addition of a single rival integrated firm (1 integrated firm in the figure) increases median consumer harm from 0.2\% to to 2\% while widening the distribution of outcomes slightly.
+Absent incumbent integration by rivals (denoted by 0 integrated firms in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.
+The addition of a single rival integrated firm (1 integrated firm in the figure) decreases median consumer harm from about 0\% of pre-merger revenues to to 2\% while widening the distribution of outcomes slightly.
 By contrast, additional integrating rivals has little effect on median total harm.
@@
 
-Moving to the second panel from the left in Figure \ref{fig:CVvertincumbBW_consumer}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
+Moving to the second panel from the left in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
 This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.
-Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, and median harm changes modestly.
-Consumer welfare increases in almost 30\% of mergers and total welfare increases in about 10\% of mergers, but median consumer harm and total harm are roughly steady at approximately 4\% pre-merger revenues.
+Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, and median harm decreases modestly.
+Consumer welfare increases in almost 30\% of mergers and total welfare increases in about 14\% of mergers, but median consumer harm and total harm are roughly steady at approximately 4\% of pre-merger revenues.
 The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.
@@
 Next we examine the welfare impacts of downstream mergers.
-As with upstream mergers,  downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues.
+As with upstream mergers,  downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 6\% of pre-merger revenues.
 \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
-Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 26\% of mergers, and total welfare increases in about 6\%, but median consumer harm grows to about 11\% of pre-merger revenues, and median total harm grows to about 7\%.
+Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, and total welfare increases in about 2\%, but median consumer harm grows to about 12\% of pre-merger revenues, and median total harm grows to about 8\%.
 Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.
@@
 
-We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW_consumer}.
-Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 8\% of simulated markets and are net beneficial in 15\% of simulated markets.
+We finish by examining integrated mergers in the right-most panel of Figures \ref{fig:CVvertincumbBW_consumer} and\ref{fig:CVvertincumbBW_total}.
+Absent the presence of rival incumbent integrated firms (0 integrated firms in the figures), mergers between two integrated firms benefits consumers in about 6\% of simulated markets and are net beneficial in about 17\% of simulated markets.
 Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.
-While adding one additional rival incumbent integrated firm (3 integrated firms in the figure) has little effect on the distribution of either consumer or total welfare, adding a second integrated rival (4 integrated firms in the figure) eliminates the most harmful integrated mergers by shortening the lower whisker.
+Adding one additional rival incumbent integrated firm (1 integrated firms in the figures) lengthens the upper whisker, increasing the percentage of beneficial mergers, but has little effect on either median consumer or total welfare.
+Likewise, adding a second integrated rival (2 integrated firms in the figures) lengthens the upper whisker while also eliminating the most harmful integrated mergers and slightly increasing median consumer and total harms.
 Adding further integrated incumbents has modest incremental effects on the distribution of harm.
 
-To summarize, our analysis of Figure \ref{fig:CVvertincumbBW_consumer} yields some overall conclusions.
+To summarize, our analysis of Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total} yield some overall conclusions.
 First, we find that mergers where one or both of the merging firms are already integrated (and no rivals are integrated) are only beneficial in a small proportion of instances.
 Second, the presence of integrated rivals can decrease the likelihood of observing the most harmful mergers, though the strength of that effect can vary widely across types of mergers.
+Third, with the exception of downstream mergers,additional integrated firms often tend to increase the incidence of with the exception of integrated mergers, additional integrated firms do not substantively affects median consumer or total surplus, but can greatly effect outcome distribution skewness.
 
@@
 
-Figure \ref{fig:CVbargincumbent_consumer}  depicts box and whisker plots summarizing the consumer (top panels) and total (bottom panels) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
+Figures \ref{fig:CVbargincumbent_consumer}  and \ref{fig:CVbargincumbent_total}  depict box and whisker plots summarizing the consumer and total welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
 Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
-As in Figure \ref{fig:CVvertincumbBW_consumer}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}.
+As in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}.
 For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
@@
 In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.
-Figure \ref{fig:CVbargincumbent_consumer} provides analogous results for integrated mergers.
-In integrated mergers, when the number of  incumbent integrated firms equals 2, this depicts a merger between two integrated firms with all rivals are unintegrated.
-Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market, respectively.
+Finally, in integrated mergers, when the number of  incumbent integrated firms equals 0, this depicts a merger between two integrated firms with all rivals are unintegrated.
+Plots with either 1 or 4 integrated firms depict mergers between integrated incumbents when there are either 1 or 4 rival integrated firms in the market, respectively.
 
@@
 The impact on consumers is roughly zero when wholesalers and retailers have equal power.
-Furthermore, we find that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates harm when there are more incumbent integrated firms, although the magnitude of these differences is small.
+Furthermore, we find that consumer harm when there are 0 incumbent integrated firms first-order stochastically dominates harm when there are 4  incumbent integrated firms, with the difference in magnitudes decreasing as retailer bargaining power increases.
 Across all three integrated incumbent scenarios, median consumer harm moves from about -24\% of pre-merger expenditures when wholesalers have the advantage to 7\% when retailers have the advantage.
@@
 
-We turn next to upstream mergers in Figure \ref{fig:CVbargincumbent_consumer}.
-Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.
-Median consumer harm moves from about 28\% of pre-merger expenditures when wholesalers are more powerful to 2\% of pre-merger expenditures when retailers are more powerful.
-Likewise, median total harm goes from approximately 12\% to less than 1\% of pre-merger expenditures.
-By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figure), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
-This is intuitive, as now all three types of mergers create some form of vertical integration.
-Median consumer harm goes from roughly -23\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Median total harm increases from about -20\% to more than 3\% of pre-merger expenditures.
+We turn next to upstream mergers in Figures \ref{fig:CVbargincumbent_consumer} and   \ref{fig:CVbargincumbent_total}.
+Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is unlike that of downstream and vertical mergers: median harm first increases and then decreases as wholesalers have relatively more bargaining power, with the largest depicted median harm-- 12\% for consumers, 18\% on net-- occurring when relative bargaining power 7/3.
+Increasing bargaining power beyond 7/3 evidently diminishes harm because pre-merger, wholesalers have already expropriated much of the retailers' surplus.
+By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figures), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
+This is intuitive, as now all three types of mergers create some form of EDM that is to varying degrees passed through to consumers.
+Median consumer harm goes from roughly -21\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Median total harm increases from about -24\% to more than 3\% of pre-merger expenditures.
 Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.
@@
 For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.
-Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargincumbent_consumer} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW_consumer}.
-Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Likewise, median total harm moves from about 2\% of pre-merger expenditures to about 8\%.
+Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figures \ref{fig:CVbargincumbent_consumer} and \ref{fig:CVbargincumbent_total} feature a decrease in welfare, which mirrors our findings in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}.
+Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 19\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Likewise, median total harm moves from about 3\% of pre-merger expenditures to about 8\%.
 In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.
@@
 Once we allow one of the merging firms to be integrated (denoted by 1 integrated firm in the figure), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.
-The box and whisker plots rotate clockwise around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of harm.
-Median consumer harm goes from approximately -33\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.
-Median total harm goes from about 2\% to almost 12\% of pre-merger revenues.
+The box and whisker plots in Figure \ref{fig:CVbargincumbent_consumer} fold towards one another around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of consumer harm.
+Median consumer harm goes from approximately -28\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.
+Median total harm goes from almost 5\% to about 11\% of pre-merger revenues.
 The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
@@
 
-Figure \ref{fig:CVbargincumbent_consumer} displays results for integrated mergers.
+The last column in Figures \ref{fig:CVbargincumbent_consumer} and   \ref{fig:CVbargincumbent_total} display results for integrated mergers.
 In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers.
 Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
-Median consumer harm moves from less than -2\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Likewise, median total harm moves from about -20\% when wholesalers have relatively more bargaining power to 15\% of pre-merger expenditures when retailers have relatively more bargaining power.
-A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 4 or 6 integrated firms in the figure) has little impact on the relationship between bargaining power and harm.
+Median consumer harm moves from about 0\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
+Likewise, median total harm moves from about -19\% when wholesalers have relatively more bargaining power to 15\% of pre-merger expenditures when retailers have relatively more bargaining power.
+A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 1 or 4 integrated firms in the figure) has modest impact on the relationship between bargaining power and harm.
 However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 or more.
@@
 \includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which bertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
+\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
 The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
@@
 \includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which bertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
+\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
 The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.

### 2022-07-25 code/MonteCarlo.R

@@ -31,7 +31,7 @@ market<-function(nfirms.down=3, # # of downstream firms
                  isNestedTogether=c("none","same","different"), # specify if parties are in the same nest ("same") or different nests ("different")
                  outMargin=runif(1,1,1e2),
                  shareOutDown=runif(1,.01,.1), # outside good margin
-                 shareParm=rep(2.5,nfirms.down*nprods.down*nfirms.up*nprods.up), # parameter for dirichlet (share) distribution (default is uniform)
+                 shareParm=rep(1,nfirms.down*nprods.down*nfirms.up*nprods.up), # parameter for dirichlet (share) distribution (default is uniform)
                  largeMerger=FALSE, # force the firms with the largest market shares to merge,
                  ...){
 

### 2022-07-25 code/genfigures.R

@@ -119,7 +119,7 @@ pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
   facet_grid(Outcome~Merger,scales="free",labeller = label_context)+
-  xlab("# Incumbent Vertically Integrated Firms")+
+  xlab("# Vertically Integrated Firms")+
   ylab("Outcome (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
   #ylab("Share-Weighted Downstream Price Change")+
@@ -151,7 +151,7 @@ pvertincumb_c.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer"
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
   facet_grid(~Merger,scales="free",labeller = label_context)+
-  xlab("# Incumbent Vertically Integrated Firms")+
+  xlab("# Vertically Integrated Firms")+
   ylab("Consumer Surplus (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
   #ylab("Share-Weighted Downstream Price Change")+
@@ -183,7 +183,7 @@ pvertincumb_t.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Total") #
   #geom_tufteboxplot(median.type = "line", whisker.type = 'line') +
   #facet_grid(Outcome~Retailers+Wholesalers,scales="free_y",labeller = "label_context")+
   facet_grid(~Merger,scales="free",labeller = label_context)+
-  xlab("# Incumbent Vertically Integrated Firms")+
+  xlab("# Vertically Integrated Firms")+
   ylab("Consumer Surplus (%)")+
   #ylab("Avg. Downstream Price Change (%)")+
   #ylab("Share-Weighted Downstream Price Change")+

### 2022-07-29 doc/VerticalArrangements.tex
Gloria cut the abstract to 100 words, finished the intro, added information on integrated mergers to section 2, and made an appendix/moved stuff from section 3 to the appendix.

@@
 
-\begin{center}
-Preliminary and Incomplete
-\end{center}
-
 \begin{abstract}
-\noindent We use a standard vertical supply-chain model to study the welfare impacts of mergers that have both horizontal and vertical aspects.
+\noindent We use a vertical supply-chain model to study the welfare impacts of mergers with both horizontal and vertical aspects.
 The model features logit Bertrand competition downstream and Nash Bargaining upstream.
-We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
-The breadth of options we include better captures the variety of merger structures observed in practice, compared to the typical ``horizontal vs.
-vertical'' dichotomy.
-We further extend our analysis to accommodate preexisting vertical integration by third-party firms and linear marginal costs.
+We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+We find that in for mergers with both horizontal and vertical aspects, consumers are typically net harmed.
 \end{abstract}
@@
 \section{Introduction}
-Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  Indeed, as the 2020 U.S.
-Department of Justice (DOJ) and Federal Trade Commission (FTC) Vertical Merger Guidelines (henceforth, the ``VMG'') state, mergers often have both horizontal and vertical aspects.
-However, the theories and models used to analyze mergers often fail to address these complexities.
-Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), while the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
-Yet, a given merger may combine all of these effects simultaneously, which raises the necessity of balancing their impacts in a unified framework.
+Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.
+However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.
+Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
+As a result, the antitrust community has little consensus for what welfare impacts one should expect from a merger that combines all of these effects simultaneously.
 
-In this paper, we use a merger simulation model to assess the welfare implications of these complex mergers.
+The horizontal versus vertical distinction has been enshrined in U.S.
+Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.
+The DOJ did not.
+}  Given these policy documents, it is not surprising that the agencies typically try to frame most mergers as either horizontal or vertical.
+However, as part of the 2022 effort by the U.S.
+Department of Justice (DOJ) and the Federal Trade Commission (FTC) to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically''\footnote{The full question is 1.g.
+in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.
+The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy?
+What aspects of modern market realities may be lost by focusing on these relationships categorically?
+Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition?
+If so, how?''.
+}  In this paper, our goal is to provide guidance to practitioners on how these mergers can be simulated in a familiar framework and to also show the range of welfare outcomes that are likely given this modeling environment.
+
+We use a standard merger simulation model to assess the welfare implications of these complex mergers.
 Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
@@
 In each of these simulations, we calculate the impacts on consumer welfare and firm profits.
-Furthermore, we examine how the presence of additional integrated rivals in the market changes our results.
-Our baseline simulations assume that units are produced with constant marginal costs, but we later extend our analysis to linear upward-sloping marginal costs.
+Furthermore, we examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
 
-%We find.
+Our simulations show certain patterns across mergers.
+For mixed upstream and mixed downstream mergers, we find that consumers are net harmed in most instances.
+The exceptions occur when upstream firms have more bargaining power than downstream firms, which is when we expect pre-merger input prices to be high and the gains from EDM to be the largest.
+Furthermore, we find that integrated mergers are typically net harmful to consumers as well.
+The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.
+Thus, although all of these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
+Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
+The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes.
 
-We then employ this model to investigate the 2021 merger of two solid waste management firms, Republic and Santek, in one geographic market: Chattanooga, Tennessee.
-The solid waste management industry has seen a large number of mergers with both horizontal and vertical elements.
-Broadly speaking, the supply chain of this sector involves haulers at the downstream level that bring trash to disposal facilities at the upstream level.
-A variety of firms participate in these services, with some owning their own hauling and disposal assets, while others are active in only a segment of the market.
-We find that ignoring vertical integration substantially understates harm to consumers in Chattanooga.
+We then employ this model to investigate the 2021 merger of two solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
+The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.
+Both Republic and Santek operated upstream and downstream, as did one of their largest rivals, Waste Connections.
+In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.
+Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.
+We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm.
+Furthermore, we show that modeling the merger as a purely horizontal downstream transaction would understate consumer harm substantially.
 
-%Our paper is related to the literature on mergers and vertical integration.
+Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
+The closest merger simulation article is our previous work in \citeN{ST2021}, which uses the same model we feature here.
+That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash bargaining model for the upstream market.
+\citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
+\citeN{MS2013} and \citeN{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
+However, none of these papers have results for the wide range of merger types that we study.
 
-%The paper proceeds as follows.
+Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \citeN{DKVB2010}.
+These models have been applied to several industries that are the frequent target of complex mergers, such as healthcare (e.g.
+\citeN{HL2017}) and video programming distribution (e.g.
+\citeN{CLWY2017}).
+These structural empirical papers explore the welfare effects of various competitive interactions in certain industries.
+Unlike these papers, our framework has been simplified for use with limited data, with the goal of being more accessible to practitioners.
+These simplifications allow us to provide numerical results over a wide range of market conditions, which complements the insights that come from the structural empirical literature on specific markets.
+
+The paper proceeds as follows.
+In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers in this framework.
+Section \ref{sec:sims} explains how we construct our numerical simulations and provides their results, and our application to Republic/Santek appears in Section \ref{sec:application}.
+Additional simulation results and the details of the data we use on the solid waste management markets appear in the Appendix.
+Section \ref{sec:concl} concludes.
 
@@
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We loosen this restriction to allow for linear increasing marginal cost as a robustness check.
+We loosen this restriction to allow for linear increasing marginal cost as a robustness check that appears in the Appendix.
 
@@
 Whether these additional incentives to raise prices will dominate the direct EDM impact is an empirical question.
-When considering a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.
+Furthermore, if we instead considered a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.
 
@@
 
-%Integrated rivals.
+The discussion thus far has not dealt with the possible existence of integrated rivals to the merging firms.
+If such rival firms exist, they would also have pricing and bargaining incentives similar to those captured in the above post-merger first order conditions.
+For example, if the integrated firm $rw$ were bargaining over what input price to charge a rival integrated firm $sv$, the first order condition would be
+\begin{equation}
+\begin{split}
+&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
+& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
+&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w,v\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right.\\
+&\left.- \underbrace{[p_{sv} - c^W_{sv} - c^R_{sv}] \Delta s_{sv}(\mathbb{W}^s \setminus \{w\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^v \setminus \{s\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^s \setminus \{w\})}_{\text{wholesale recapture leverage effect}}\right).
+\end{split}
+\label{eq: intmerger wh upstream foc}
+\end{equation}
+The existing integration of firm $sv$ gives it countervailing bargaining leverage that may counteract any change in leverage caused by a merger of retailer $r$ and wholesaler $w$.
+The extent of the net impact depends on how these leverage terms balance out in equilibrium.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Numerical Simulations}
-In this section, we use a similar setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and allow for the presence of integrated rival firms.
+\section{Numerical Simulations \label{sec:sims}}
+In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and also allow for the presence of integrated rival firms.
 
@@
 
+Throughout these simulations, we maintain the assumption that marginal costs are constant.
+However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.
+We explore the robustness of our results to the constant cost assumption in the Appendix, where we provide results that allow marginal costs to increase linearly.
+Our simulations show that having linearly increasing marginal costs tends to result in mergers that are more harmful.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
-Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total} depict box and whisker plots summarizing consumer harm and total harm as the number of incumbent integrated firms increases.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figures \ref{fig:CVvertincumbBW_consumer},  \ref{fig:CVvertincumbBW_total} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases.
+For the results on total welfare, see the Appendix.
+The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
 By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties.
-For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.
+For integrated mergers, the plot when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depicts a merger between two integrated firms, again with no integrated rivals.
 
 We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
-Absent incumbent integration by rivals (denoted by 0 integrated firms in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.
-The addition of a single rival integrated firm (1 integrated firm in the figure) decreases median consumer harm from about 0\% of pre-merger revenues to to 2\% while widening the distribution of outcomes slightly.
-By contrast, additional integrating rivals has little effect on median total harm.
-Adding more integrated rivals tends to increase the proportion of mergers that benefit consumers, while also increasing the proportion of mergers that reduce total surplus.
-Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers.
-Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.
-It appears that these rivals can therefore somewhat limit the extent of negative impacts that would otherwise occur through RRC.
+Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW_consumer}), about half of vertical mergers benefit consumers.
+The addition of a single rival integrated firm (denoted by 1 integrated firm in the figure) moves median consumer welfare from about 0\% of pre-merger revenues to a positive 2\% and widens the distribution of outcomes slightly.
+Adding more integrated rivals does not have much additional effect.
 
-Moving to the second panel from the left in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues.
-This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.
-Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, and median harm decreases modestly.
-Consumer welfare increases in almost 30\% of mergers and total welfare increases in about 14\% of mergers, but median consumer harm and total harm are roughly steady at approximately 4\% of pre-merger revenues.
+Moving to the second panel in Figure \ref{fig:CVvertincumbBW_consumer}, upstream mergers absent any incumbent vertical integration (0 integrated firms) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues.
+This result is to be expected, as upstream mergers without additional integration do not have the countervailing effects for welfare possible in the context of downstream mergers (where combined retails may bargain for lower input prices) or vertical mergers (where EDM can create benefits).
+Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear.
+Consumer welfare increases in almost 30\% of mergers, though median consumer harm is roughly steady at approximately 4\% of pre-merger revenues.
 The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.
-Just as we saw with downstream mergers, here we find that harm dominates in most simulations.
-Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- while only slightly changing median harm.
-Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.
+Here we find that harm dominates in most simulations.
+Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and somewhat trims the whiskers while having little impact on median harm.
 
-Next we examine the welfare impacts of downstream mergers.
-As with upstream mergers,  downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 6\% of pre-merger revenues.
+Next we examine the welfare impacts of downstream mergers in the third panel of Figure \ref{fig:CVvertincumbBW_consumer}.
+As with upstream mergers, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues.
 \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
-Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, and total welfare increases in about 2\%, but median consumer harm grows to about 12\% of pre-merger revenues, and median total harm grows to about 8\%.
+Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, but median consumer harm grows to about 12\% of pre-merger revenues.
 Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.
-On net, we find that the harms dominate in most instances.
-Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat decreases the range of outcomes and moves the distribution towards less harm.
-With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 34\% of markets and shrinks median consumer harm to about 5\% of pre-merger revenues.
-The inter-quartile range of total harm narrows, and median total harm falls to about 4\% of pre-merger revenues.
-As we saw with vertical mergers, it appears that the presence of third party integrated rivals can limit harm, though here the effect is a bit stronger.
+Similar to what we saw for upstream mergers, we find that the harms dominate in most instances.
+Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat moves the distribution towards less harm.
+With 3 integrated third parties (denoted by 4 integrated firms in the figure), the merger benefits consumers in about 34\% of markets and shrinks median consumer harm to about 5\% of pre-merger revenues.
 
-We finish by examining integrated mergers in the right-most panel of Figures \ref{fig:CVvertincumbBW_consumer} and\ref{fig:CVvertincumbBW_total}.
-Absent the presence of rival incumbent integrated firms (0 integrated firms in the figures), mergers between two integrated firms benefits consumers in about 6\% of simulated markets and are net beneficial in about 17\% of simulated markets.
-Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.
-Adding one additional rival incumbent integrated firm (1 integrated firms in the figures) lengthens the upper whisker, increasing the percentage of beneficial mergers, but has little effect on either median consumer or total welfare.
-Likewise, adding a second integrated rival (2 integrated firms in the figures) lengthens the upper whisker while also eliminating the most harmful integrated mergers and slightly increasing median consumer and total harms.
-Adding further integrated incumbents has modest incremental effects on the distribution of harm.
+We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW_consumer}.
+Absent the presence of rival incumbent integrated firms (0 integrated firms in the figure), mergers between two integrated firms benefit consumers in about 6\% of simulated markets.
+Thus, we find that mergers between two firms that are already integrated are harmful to consumers in the vast majority of cases.
+Adding one additional rival incumbent integrated firm (1 integrated firms in the figure) does not have a significant impact on the median or range of outcomes.
+Adding more integrated rivals (2-4 integrated firms in the figures) moves median consumer welfare to be slightly more negative.
 
-To summarize, our analysis of Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total} yield some overall conclusions.
-First, we find that mergers where one or both of the merging firms are already integrated (and no rivals are integrated) are only beneficial in a small proportion of instances.
-Second, the presence of integrated rivals can decrease the likelihood of observing the most harmful mergers, though the strength of that effect can vary widely across types of mergers.
-Third, with the exception of downstream mergers,additional integrated firms often tend to increase the incidence of with the exception of integrated mergers, additional integrated firms do not substantively affects median consumer or total surplus, but can greatly effect outcome distribution skewness.
+To summarize, our analysis of Figure \ref{fig:CVvertincumbBW_consumer} yields some overall conclusions.
+First, we find that mergers where one or both of the merging firms are already integrated (and all rivals are unintegrated) are harmful to consumers in most instances.
+Although these mergers have the possibility of creating benefits through EDM, these gains appear to be outweighed in the majority of the mergers we study.
+Second, the presence of integrated rivals has some impact on the distribution of consumer harm, but the observed patterns are noisy and the resulting box plots are largely similar to those where such rivals are absent.
+Therefore, having such rival firms in and of itself does not generate significantly better outcomes for consumers.
 
@@
 Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.
-We also show that the result persists for all types of mergers when there are rival non-merging integrated firms.
+We also show that the result persists when there are rival non-merging integrated firms.
 
-Figures \ref{fig:CVbargincumbent_consumer}  and \ref{fig:CVbargincumbent_total}  depict box and whisker plots summarizing the consumer and total welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
+Figure \ref{fig:CVbargincumbent_consumer} depicts box and whisker plots summarizing the consumer welfare effects for vertical, upstream, downstream, and integrated mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
 Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
-As in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}.
+For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
 For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
-By contrast, for upstream and downstream mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated firm when all third parties are unintegrated.
+For upstream and downstream mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated firm when all third parties are unintegrated.
 In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.
@@
 
-We begin by examining vertical mergers.
-We find that the presence of integrated incumbents has little impact on the relationship between relative bargaining power and welfare.
-Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit welfare when wholesalers have relatively more power, and harm welfare when retailers have relatively more power.
+We begin by examining vertical mergers in the leftmost panel of Figure \ref{fig:CVbargincumbent_consumer}.
+We focus first on the case when there are no incumbent integrated firms (the light blue plots).
+Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
 Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.
 This channel becomes less relevant as retailers gain more power.
-The impact on consumers is roughly zero when wholesalers and retailers have equal power.
-Furthermore, we find that consumer harm when there are 0 incumbent integrated firms first-order stochastically dominates harm when there are 4  incumbent integrated firms, with the difference in magnitudes decreasing as retailer bargaining power increases.
+The impact on consumers is zero in the neighborhood of retailer to wholesaler bargaining power ratios of 1 to 2/3.
+As we add 1 and then 4 integrated rival firms (shown by the darker and darkest blue box plots, respectively), the range of welfare outcomes narrows and shifts upwards, though we still see the negative relationship between consumer welfare and relative retailer bargaining power.
 Across all three integrated incumbent scenarios, median consumer harm moves from about -24\% of pre-merger expenditures when wholesalers have the advantage to 7\% when retailers have the advantage.
-Similarly, median total harm goes from from about -14\% of pre-merger expenditures when wholesales have the advantage to 2.9\% when retailers have the advantage.
 
-We turn next to upstream mergers in Figures \ref{fig:CVbargincumbent_consumer} and   \ref{fig:CVbargincumbent_total}.
-Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is unlike that of downstream and vertical mergers: median harm first increases and then decreases as wholesalers have relatively more bargaining power, with the largest depicted median harm-- 12\% for consumers, 18\% on net-- occurring when relative bargaining power 7/3.
-Increasing bargaining power beyond 7/3 evidently diminishes harm because pre-merger, wholesalers have already expropriated much of the retailers' surplus.
-By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figures), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
-This is intuitive, as now all three types of mergers create some form of EDM that is to varying degrees passed through to consumers.
+We turn next to upstream mergers in Figure \ref{fig:CVbargincumbent_consumer}.
+Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that of downstream and vertical mergers, as it does not have a strong monotonic relationship.
+By contrast, for mergers between an integrated firm and an unintegrated upstream supplier (the darker blue plots), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
+This is intuitive, as now all three types of mergers involve bargaining interactions that can lower input prices, the savings from which are then passed through to consumers to varying degrees.
 Median consumer harm goes from roughly -21\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Median total harm increases from about -24\% to more than 3\% of pre-merger expenditures.
-Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.
+Adding 3 incumbent integrated rivals (the darkest blue plots) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows and the medians shift up slightly.
 
-For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.
-Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figures \ref{fig:CVbargincumbent_consumer} and \ref{fig:CVbargincumbent_total} feature a decrease in welfare, which mirrors our findings in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVvertincumbBW_total}.
+For downstream mergers, like with vertical mergers, we find that higher retailer relative bargaining power leads to more negative impacts on welfare.
+Absent any incumbent integration (the light blue plots), all the outcomes shown in Figure \ref{fig:CVbargincumbent_consumer} harm consumers, which mirrors our findings in Figure \ref{fig:CVvertincumbBW_consumer}.
 Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 19\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Likewise, median total harm moves from about 3\% of pre-merger expenditures to about 8\%.
 In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.
@@
 
-Once we allow one of the merging firms to be integrated (denoted by 1 integrated firm in the figure), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.
-The box and whisker plots in Figure \ref{fig:CVbargincumbent_consumer} fold towards one another around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of consumer harm.
+Once we allow one of the merging firms to be integrated (the darker blue plots), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.
+The box and whisker plots in Figure \ref{fig:CVbargincumbent_consumer} now show a stronger relationship between bargaining power and extent of consumer harm.
 Median consumer harm goes from approximately -28\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.
-Median total harm goes from almost 5\% to about 11\% of pre-merger revenues.
 The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
-Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here neutrality happens between  3/2 and 7/3, when wholesalers have a bit more than 1.5 times the power of retailers.
+Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here the neutrality crossing shifts left, when wholesalers have a bit more than 1.5 times the power of retailers.
 It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.
 This is intuitive, as here the downstream merger also causes an additional lessening in horizontal competition.
-Once we add rival integrated firms (see 4 integrated firms in the figure) the box and whisker plots rotate counter-clockwise, weakening but not erasing the relationship between bargaining power and harm.
+Once we add 3 rival integrated firms (the darkest blue plots) the box and whisker plots shift up.
 The range of outcomes also shrinks.
-Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -35\% of pre-merger revenues to 13\% of pre-merger revenues, while median total harm goes from almost 3\% to 5\% of pre-merger revenues.
+This is consistent with what we see for vertical and upstream mergers.
+Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -35\% of pre-merger revenues to 13\% of pre-merger revenues.
 
-The last column in Figures \ref{fig:CVbargincumbent_consumer} and   \ref{fig:CVbargincumbent_total} display results for integrated mergers.
+The last panel in Figure \ref{fig:CVbargincumbent_consumer} displays results for integrated mergers.
 In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers.
@@
 Median consumer harm moves from about 0\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Likewise, median total harm moves from about -19\% when wholesalers have relatively more bargaining power to 15\% of pre-merger expenditures when retailers have relatively more bargaining power.
-A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 1 or 4 integrated firms in the figure) has modest impact on the relationship between bargaining power and harm.
-However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 or more.
+A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see the darker and darkest blue plots) has tends to narrow the range of outcomes and somewhat shift the distribution upwards.
+However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 times or more.
 The crossing at zero harm is shifted to the left.
 
-%Finally, Figure \ref{fig:CVbargincumbent_consumer} depicts box and whisker plots summarizing consumer, retailer, wholesaler and total surplus  for integrated mergers with 2,4, or 6 integrated firms.
-When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargincumbent_consumer} depicts a merger between two integrated firms: all rivals are unintegrated.
-Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.
-
-%\textbf{Describe the counterintuitive result that increasing retailer bargaining power harms retailers but help wholesalers.}
+In summary, our review of Figure \ref{fig:CVbargincumbent_consumer} has revealed some systematic patterns.
+First, we see that vertical mergers, upstream and downstream mergers where one of the merging firms is integrated, and integrated mergers all exhibit a negative relationship between consumer harm and increasing relative retailer bargaining power.
+Consumers seem most likely to benefit from these mergers when wholesalers have significant bargaining power, which are situations when we would expect the gains from EDM to be largest.
+Second, the point at which this monotonic relationship between consumer welfare and bargaining power crosses zero varies across merger types, and appears to be farthest to the left in integrated mergers.
+That is, wholesalers need to have a substantial edge in bargaining power over retailers in order for consumers to benefit from an integrated merger.
+Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift this distribution upwards.
+This pattern is much less noisy than in Figure \ref{fig:CVvertincumbBW_consumer}, where we did not condition on relative bargaining power.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Robustness to Alternative Cost Specifications}
-
-Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allow costs to increase linearly.
-Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).
-
-In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs.
-Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause the integrated wholesaler's prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
-Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.
-
-Our simulations confirm this hypothesis.
-See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.
-Overall, harm is greater for mergers with linear costs.
-The difference is largest for vertical mergers.
-When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 55\% of simulations and total surplus in about 29\% of simulations, but only increase consumer surplus in about 36\% of simulations and total surplus in about 2\% of simulations when all firms have linear costs.
-
-The bottom panel of Figure \ref{fig:surplussumcost} compares a situation where only the merging firms have constant marginal costs to one where only the merging firms have linear marginal costs.
-We see that the distributions in these bottom panels are largely the same as those in the top panels.
-Thus, it appears that the key driver of our results is whether or not the merging parties have constant marginal costs.
-When the merging firms' marginal costs are constant, the distribution of outcomes moves towards less harm.
-
-%Mergers with constant-cost firms increases consumer surplus in about 15\% of all mergers and total surplus in 10.5\% of all mergers, while mergers with linear-cost firms increase consumer surplus in only 2.3\% of mergers and total surplus in 0.8\% of all mergers.
-However, while the consumer and total harm distributions for each merger type under the all-firm linear marginal costs scenario first-order stochastically dominates their counterparts under the all-firm constant marginal costs scenario, the consumer and total harm distributions under the party linear marginal costs scenario only first-order stochastically dominates the their counterparts under the party constant marginal costs scenario for vertical and integrated mergers.
-The median consumer harm from markets with linear marginal costs is about 9.3\% of pre-merger total expenditures, about 1.3 times the magnitude of markets with constant marginal costs.
-
-%Changing the cost structure consistently yields the greatest differences in outcomes for vertical mergers.
-When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of all vertical merger simulations and total surplus in about 18\% of all vertical merger simulations, but only increase consumer surplus in about 1\% of vertical merger simulations and total surplus in about 0.9\% of vertical merger simulations when all firms have linear costs.
-Moreover, qualitatively similar results hold for vertical mergers when only the merging parties' products have constant marginal costs, compared to when only the merging parties' products have linear costs.
-
-%The differences between the constant and linear cost structures is more muted for both upstream and downstream mergers.
-Upstream mergers with all constant marginal cost firms benefit consumers in about 12\% of simulations, but only 0.2\% of simulations when firms have linear costs.
-Downstream mergers with all constant marginal cost firms benefit consumers in about 11\% of simulations, but only about 6\% of simulations when all firms have linear costs.
-Again, qualitatively similar results hold for vertical mergers when only the merging parties' products have constant marginal costs, compared to when only the merging parties' products have linear costs.
-
-%Finally, while there is a marked difference in outcomes between the constant and linear cost structures for integrated  mergers under the all-firms scenario, the differences in outcomes are more muted under the party scenario, especially for consumer surplus.
-In particular, median consumer harm when all firms face constant marginal costs is 13.6\%, compared to the median consumer harm of 17.5\% when all firms face linear marginal costs.
-By contrast, median consumer harm when only the merging parties' have constant costs is 15.1\%, compared to the median consumer harm of 16.5\% when the parties' have linear costs.
-%Integrated mergers benefit consumers in about 10\% of simulations when all firms have constant costs, but less than 1\% of simulations when all firms have linear costs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Empirical Application}
+\section{Empirical Application: Republic/Santek \label{sec:application}}
 
@@
 As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevant local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.
-
-The vertical supply chain in the solid waste industry is primarily comprised of waste collection operations or ``haulers'' and waste disposal facilities.
-Haulers collect municipal solid waste (MSW) from businesses and residences and must dispose of it at a lawful disposal site, predominantly landfills.
-Waste disposal (upstream) is a required input into waste collection services (downstream).
-Some haulers are vertically integrated and operate their own disposal facilities.
-Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
-Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens' waste themselves.
-Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
-Thus, mergers that combine hauling and disposal assets may incentivize the merged entity to raise its hauling rivals' cost of disposal in order to benefit its own collection operations.
-Whether or not the merged entity is both able and incentivized to undertake such action depends upon the extent of its market power in the local disposal market, the merging parties’ profit margins in each line of business, and their intensity of hauling competition with prospective disposal customers.
 
@@
 
-%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
-Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
-and Plaintiff States v.
-Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
-For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
-In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S.
+The DOJ filed a complaint and consent decree in the Republic/Santek case, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW)disposal markets.
+In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga region.\footnote{See U.S.
 and State of Alabama v.
 Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
-%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
-The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
-The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.
 
-\subsection{Combination of vertically integrated assets: Chattanooga}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Market Background}
 
-The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commercial waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers.
-The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
+The vertical supply chain in the solid waste industry is primarily comprised of waste collection operations or ``haulers'' and waste disposal facilities.
+Haulers collect MSW from businesses and residences and must dispose of it at a lawful disposal site, predominantly landfills.
+Waste disposal (upstream) is a required input into waste collection services (downstream).
+Some haulers are vertically integrated and operate their own disposal facilities.
+Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
+Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens' waste themselves.
+Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
+
+The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic/Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and SCCW collection as well as due to RRC in the SCCW collection market by raising the MSW disposal costs of independent haulers.
+The CIS notes that pre-merger, Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
 In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere.
@@
 In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, and was the parties' sole competitor in the MSW disposal market.
-Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
-Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consummation since the merger was completed before the filing of the CIS for the Republic-Santek merger.
+Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity (henceforth ``WM-ADS'') regardless of whether the data source predates the merger consummation since the merger was completed before the filing of the CIS for the Republic/Santek merger.
 The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.
+(We label this firm ``Regional.'')
 
-\subsubsection{MSW Disposal Volume Data}
-\label{chatt_disposal_vol}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Data}
 
-Data on MSW disposal comes from data collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+Data on MSW disposal comes from information collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
 This data identifies the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
@@
 
-Using this data in conjunction with information on collection market volumes, discussed in Section \ref{chatt_collection_share}, the supply relationships between the upstream disposal market participants (\textit{Republic, Santek, and Waste Connections}) and downstream collection market participants (\textit{Republic, Santek, Waste Connections, WM-ADS, and ``Regional''}) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Using these data in conjunction with information on collection market volumes, the supply relationships between the upstream disposal market participants (\textit{Republic, Santek, and Waste Connections}) and downstream collection market participants (\textit{Republic, Santek, Waste Connections, WM-ADS, and ``Regional''}) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
 Santek and Waste Connections exhibit excess disposal capacity.
 However, Republic's collection volumes are estimated to exceed their ability to self-supply disposal in the Chattanooga market.
-As a result, Republic will need to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
+As a result, Republic needs to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
 We assume that this disposal is purchased from both Santek and Waste connections according to their share of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
-We further assume that WM-ADS and Regional purchased disposal capacity from both Santek and Republic according to these shares.
+We further assume that WM-ADS and Regional purchase disposal capacity from both Santek and Republic according to these shares.
 The resulting relationships between the upstream and downstream market participants are captured in Figure \ref{fig:chatt_tree}.
 
-\subsubsection{MSW Disposal Price Data}
 MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
@@
 
-\subsubsection{Collection Market Share Data}
-\label{chatt_collection_share}
-
-The CIS states that in the Chattanooga Area the post-merger HHI for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
+The CIS states that in the Chattanooga Area the post-merger Herfindahl-Hirschman Index for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
 Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
@@
 
-However, the market shares required for implementation of the Sheu-Taragin (\textit{\textbf{add ref}}) framework are expressed as upstream-downstream market participant pairs.
-Following the discussion of supply relationships in Section \ref{chatt_disposal_vol}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal-Santek collection, Waste Connection disposal-Waste Connections collection) are assigned their full collection share.
-Capacity constrained, integrated firms (Republic disposal-Republic collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal-Republic collection, Waste Connections disposal-Republic collection).
+However, the market shares required for implementation of our model are expressed as upstream-downstream market participant pairs.
+Following the discussion of supply relationships above, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
+Capacity constrained, integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
 Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.
 
-\subsubsection{Cost, Margin and  Elasticity Data}
-
 Collection and disposal margins are calculated for Republic from data on revenue by line of service and components of cost of operations in the company's 2019 annual report.
-The data is reported at the company level and is not specific to the Chattanooga Area, but revenues are reported by collection segment.
-
+The data are reported at the company level and are not specific to the Chattanooga Area, but revenues are reported by collection segment.
 Collection costs in \$/Ton are estimated from Republic and Waste Management 2019 10k financial statements.
@@
 
-A survey of demand elasticity estimates for the collection market are present in Bel and Gradus (2016) with an average of -0.34.
-These estimates are not specific to the SCCW market, but available evidence suggests that the commercial segment should be more inelastic than other segments such that analysis using these estimates would tend to be conservative for estimating the size of merger induced price effects.\footnote{Bel, G.
-and R.
-Gradus, 2016, Effects of unit-based pricing on household waste collection demand: A meta-regression analysis, Resource and Energy Economics, 44, 169-182.}
-
-\subsubsection{Chattanooga Area Merger Simulation Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Merger Simulation Results}
 
@@
 Thus, on net the model predicts the merger would be harmful despite the presence of significant efficiencies from EDM that reduce prices upstream.
-These aggregate effects obscure the complicated competitive effects between these interdependent firms though.
 These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
@@
 The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections prices to its downstream rivals are estimated to increase 22\%.
-For the merged firm the price increase is driven by both the effects of increased bargaining leverage and upward pricing pressure, whereas, Waste Connections' price increase is driven only by the change in bargaining leverage.
-These large price increases are offset in the aggregate, volume-weighted estimate by the large decrease in the disposal price for the integrating pair post-merger due to the EDM effect, and the cost to Republic for volume disposed with Santek decreases by 60\%.
+These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, and the cost to Republic for volume disposed with Santek decreases by 60\%.
 
@@
 The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall.
-The pre-merger integrated Republic pair's price increases 21\% from the effect of upward pricing pressure alone and Santek's increases by 23\% from a combination of UPP and increased bargaining leverage.
+The pre-merger integrated Republic pair's price increases 21\% and Santek's increases by 23\% from a combination of UPP and increased bargaining leverage.
 The newly integrated Santek-Republic pair's price is estimated to increase by 9\%, demonstrating that the upstream efficiencies are not fully passed-through to consumers.
-Pre-merger integrated rival, Waste Connections, collection price increases by approximately 5\% and share increases by 6 percentage points.
-Unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and share increases by 2 percentage points each.
-
-\subsubsection{Merger Simulation of Collection Market Only}
+The integrated rival, Waste Connections, increases its collection price by approximately 5\% and its share increases by 6 percentage points.
+The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and share increases by 2 percentage points each.
 
@@
 
-The Republic-Santek merger was ultimately settled with the DOJ through a divestiture.
+The Republic/Santek merger was ultimately settled with the DOJ through a divestiture.
 According to the final judgement, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station in the Chattanooga area.\footnote{Final Judgment: U.S and Plaintiff States v.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+\section{Conclusion \label{sec:concl}}
 
-This paper relaxes an assumption made in \citeN{ST2021}: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated.
-Relaxing this constraint is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are already vertically integrated.
-Here, we show using numerical simulations that relaxing these assumptions can result in merger outcomes that are markedly different from those when these assumptions are maintained.
-This is further illustrated by the results from our simulations in the Republic/Santek merger, where we find that ignoring existing vertical relationships drastically understates harm to Chattanooga consumers.
-Our hope is that future research will focus on some of the other assumptions made in \citeN{ST2021}, such as how the presence of siloed firms, two-part tariffs, or large fixed costs affect the Nash bargaining game and therefore merger outcomes.
+This paper relaxes an assumption made in much of the merger simulation literature: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated.
+Removing this limitation is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are already vertically integrated.
+Using a series of numerical simulations and the model of \citeN{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
+The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be largest.
+We also find that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
+We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships drastically understates harm to Chattanooga consumers.
+
+Our results suggest that mergers that have horizontal aspects remain harmful for consumers in many instances, even when these mergers also have vertical aspects.  
+This appears to be especially true when retailers have relatively more bargaining power, which in our model connotes with situations where retailer margins are relatively large compared to wholesale margins.
+Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.
+However, we of course have the caveat that our results are specific to our modeling framework.
+An important question for future research would be to explore merger outcomes in in other contexts.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{table}
@@
 \centering
+\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
+\label{fig:CVbargincumbent_consumer}
+\end{sidewaysfigure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=1.0]{output/chattanooga_tree.png}
+\caption{Chattanooga Area MSW Disposal and Collection Market}
+\label{fig:chatt_tree}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=1.0]{output/TrashSimsFirm.png}
+\caption{Republic/Santek Simulation Results}
+\label{fig:TrashSimsFirms}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{output/TrashSimsCompare.png}
+\caption{Merger Simulation Results Comparison}
+\label{fig:TrashSimsCompare}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+\newpage
+\clearpage
+
+\numberwithin{equation}{section} 
+\numberwithin{figure}{section} 
+\numberwithin{table}{section}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Robustness to Increasing Marginal Costs}
+
+Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allow costs to increase linearly.
+Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).
+
+In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs.
+Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause the integrated wholesaler's prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
+Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.
+
+Our simulations confirm this hypothesis.
+See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.
+Overall, harm is greater for mergers with linear costs.
+The difference is largest for vertical mergers.
+When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 55\% of simulations and total surplus in about 29\% of simulations, but only increase consumer surplus in about 36\% of simulations and total surplus in about 2\% of simulations when all firms have linear costs.
+
+The bottom panel of Figure \ref{fig:surplussumcost} compares a situation where only the merging firms have constant marginal costs to one where only the merging firms have linear marginal costs.
+We see that the distributions in these bottom panels are largely the same as those in the top panels.
+Thus, it appears that the key driver of our results is whether or not the merging parties have constant marginal costs.
+When the merging firms' marginal costs are constant, the distribution of outcomes moves towards less harm.
+
+\begin{sidewaysfigure}
+\centering
+\includegraphics[scale=0.9]{output/surplussum_cost.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
+The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue,left) or linear marginal costs (orange, right).
+The boxes in the bottom row (``Party") either assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
+}
+\label{fig:surplussumcost}
+\end{sidewaysfigure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data for the Waste Management Markets}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Additional Figures}
+
+Figures \ref{fig:CVvertincumbBW_total} and \ref{fig:CVbargincumbent_total} are the analogous results as in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVbargincumbent_consumer}, but for total welfare.
+The patterns for total welfare are largely similar to those for consumer welfare.
+We see in Figure \ref{fig:CVvertincumbBW_total} that most upstream, downstream, and integrated mergers are harmful.
+Figure \ref{fig:CVbargincumbent_total} shows that net beneficial
+mergers typically only occur when wholesalers have relatively more bargaining power.  The presence of integrated rivals can narrow the range of outcomes.
+
+\begin{sidewaysfigure}
+\centering
 \includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
@@
 \end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\centering
+\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
+\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
+\label{fig:CVbargincumbent_total}
+\end{sidewaysfigure}
+
+\end{document}
+
+%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions.
+Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S.
+and Plaintiff States v.
+Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
+For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
+In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
+%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements.
+The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
+The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
 
@@
 
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargincumbent_consumer}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargincumbent_total}
-\end{sidewaysfigure}
-
 % \begin{sidewaysfigure}
@@
 % \end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/surplussum_cost.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
-The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue,left) or linear marginal costs (orange, right).
-The boxes in the bottom row (``Party") either assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-\label{fig:surplussumcost}
-\end{sidewaysfigure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=1.0]{output/chattanooga_tree.png}
-\caption{Chattanooga Area MSW Disposal and Collection Market}
-\label{fig:chatt_tree}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=1.0]{output/TrashSimsFirm.png}
-\caption{Republic/Santek Simulation Results}
-\label{fig:TrashSimsFirms}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{output/TrashSimsCompare.png}
-\caption{Merger Simulation Results Comparison}
-\label{fig:TrashSimsCompare}
-\end{figure}
-
-\end{document}

### 2022-08-03 doc/VerticalArrangements.tex
This version incorporates Gloria's 6/3 edits to the document

@@
 \begin{abstract}
-\noindent We use a vertical supply-chain model to study the welfare impacts of mergers with both horizontal and vertical aspects.
-The model features logit Bertrand competition downstream and Nash Bargaining upstream.
-We numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
-We find that in for mergers with both horizontal and vertical aspects, consumers are typically net harmed.
+\noindent We study the welfare impacts of mergers with both horizontal and vertical aspects.
+Our model features logit Bertrand competition downstream and Nash Bargaining upstream.
+We numerically simulate four merger types: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two integrated retailer/wholesaler pairs.
+We find that mergers with both horizontal and vertical characteristics typically harm consumers.
+We also apply the model to the Republic/Santek merger as a real-world example.
 \end{abstract}
@@
 }  Given these policy documents, it is not surprising that the agencies typically try to frame most mergers as either horizontal or vertical.
-However, as part of the 2022 effort by the U.S.
-Department of Justice (DOJ) and the Federal Trade Commission (FTC) to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically''\footnote{The full question is 1.g.
+However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically''\footnote{The full question is 1.g.
 in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.
@@
 
-We use a standard merger simulation model to assess the welfare implications of these complex mergers.
 Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
-We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
-In each of these simulations, we calculate the impacts on consumer welfare and firm profits.
+We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+In each of these simulations, we calculate the impacts on consumer welfare.
 Furthermore, we examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
@@
 Our simulations show certain patterns across mergers.
-For mixed upstream and mixed downstream mergers, we find that consumers are net harmed in most instances.
-The exceptions occur when upstream firms have more bargaining power than downstream firms, which is when we expect pre-merger input prices to be high and the gains from EDM to be the largest.
+For horizontal upstream and downstream mergers, we find that consumers are net harmed in most instances.
+The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be the largest.
 Furthermore, we find that integrated mergers are typically net harmful to consumers as well.
@@
 The paper proceeds as follows.
-In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers in this framework.
-Section \ref{sec:sims} explains how we construct our numerical simulations and provides their results, and our application to Republic/Santek appears in Section \ref{sec:application}.
-Additional simulation results and the details of the data we use on the solid waste management markets appear in the Appendix.
+In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers.
+Section \ref{sec:sims} explains how we construct our numerical simulations and provides our results, and our application to Republic/Santek appears in Section \ref{sec:application}.
+Additional simulation results appear in the Appendix.
 Section \ref{sec:concl} concludes.
@@
 Assume there is a set of consumers who each choose to buy a product sold by a single retailer.
-Retailers are indexed by $r$, while the wholesalers that supply these retailers are indexed by $w$.
+Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.
 Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.
@@
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We loosen this restriction to allow for linear increasing marginal cost as a robustness check that appears in the Appendix.
+We allow for linear increasing marginal cost as a robustness check that appears in the Appendix.
 
@@
 
-When the merged firm is bargaining with the unaffiliated wholesaler $v$ over what input price to pay as a retailer, the bargaining first order condition becomes
+When the merged firm is bargaining with the unaffiliated wholesaler $v$ over what input price to pay, the bargaining first order condition becomes
 \begin{equation}
@@
 
-Turning to wholesale prices, again consider a merger between a preexisting integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
+As for wholesale prices, again consider a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
 When wholesaler $w$ bargains with an unaffiliated retailer $u$ we have the first order condition given by,
@@
 - \underbrace{\sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^r \setminus \{v\})}_{\text{retail recapture leverage effect}}\\
-&-\underbrace{\left.\sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})\right)}_{\text{wholesale recapture leverage effect}}.
+&-\left.\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
 \end{split}
@@
 \section{Numerical Simulations \label{sec:sims}}
-In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and also allow for the presence of integrated rival firms.
+In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations.
 
@@
 We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.
-Finally, we assume that an integrated merger is between two firms that are both already vertically integrated pre-merger.
+Finally, an integrated merger is between two firms that are both already vertically integrated pre-merger.
 
 For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.
-}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7.
+}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers to 7.
 We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
@@
 The bargaining parameter ranges from from 0.1 (where wholesalers have the advantage) to 0.9 (where retailers have the advantage).
-We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$, which ranges from $9$ (wholesaler power is nine times greater than retailer power) to $1/9$ (retailer power is nine times greater than wholesaler power).
+We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$.
 The bargaining parameter is identical for all of the retailers in each simulation.
@@
 
-Table \ref{tab:simsum} provides summary statistics for our simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}  For reference, the 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with HHIs over 2,500 points.
+Table \ref{tab:simsum} provides summary statistics for our simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}  The 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with Herfindahl Hirschman Index (HHI) values over 2,500 points.
 The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.
-}
-
-The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.
+}  The vast majority of our mergers meet these conditions.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
+}  The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.
 Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
-Pre-merger HHIs range from 2,911 at the $25^{th}$ percentile to 4,981 at the $75^{th}$, with a median of 3,747.
-HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,048.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
-}  HHIs for downstream mergers increase by 2,128 points at the median, resulting in a median post-merger HHI equal to 5,153.
-HHIs for upstream mergers increase by 2,070 points at the median, resulting in a median post-merger HHI equal to 4,968.
-HHIs for integrated mergers increase by 2,803 points at the median, resulting in a median post-merger HHI equal to 6,998.
 
 Figure \ref{fig:surplussum} summarizes our results for welfare.
-Each of the four panels show how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).
+Each of the four panels shows how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).
 Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.
@@
 However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.
-We explore the robustness of our results to the constant cost assumption in the Appendix, where we provide results that allow marginal costs to increase linearly.
-Our simulations show that having linearly increasing marginal costs tends to result in mergers that are more harmful.
+In the Appendix, we provide results that allow marginal costs to increase linearly.
+We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.
 
@@
 
-Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
+We examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
 Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases.
@@
 By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties.
-For integrated mergers, the plot when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depicts a merger between two integrated firms, again with no integrated rivals.
+For integrated mergers, the plot when the number of incumbent integrated firms equals 0 depicts a merger between two integrated firms, again with no integrated rivals.
 
@@
 Moving to the second panel in Figure \ref{fig:CVvertincumbBW_consumer}, upstream mergers absent any incumbent vertical integration (0 integrated firms) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues.
-This result is to be expected, as upstream mergers without additional integration do not have the countervailing effects for welfare possible in the context of downstream mergers (where combined retails may bargain for lower input prices) or vertical mergers (where EDM can create benefits).
+This result is to be expected, as upstream mergers without additional integration do not have the countervailing effects for welfare possible in the context of downstream mergers (where combined retailers may bargain for lower input prices) or vertical mergers (where EDM can create benefits).
 Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear.
@@
 Thus, we find that mergers between two firms that are already integrated are harmful to consumers in the vast majority of cases.
-Adding one additional rival incumbent integrated firm (1 integrated firms in the figure) does not have a significant impact on the median or range of outcomes.
-Adding more integrated rivals (2-4 integrated firms in the figures) moves median consumer welfare to be slightly more negative.
+Adding one additional rival incumbent integrated firm (1 integrated firm in the figure) does not have a significant impact on the median or on the range of outcomes.
+Adding more integrated rivals (2-4 integrated firms in the figure) moves median consumer welfare to be slightly more negative.
 
@@
 
-We turn next to upstream mergers in Figure \ref{fig:CVbargincumbent_consumer}.
-Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that of downstream and vertical mergers, as it does not have a strong monotonic relationship.
+We turn next to upstream mergers in the second panel of Figure \ref{fig:CVbargincumbent_consumer}.
+Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that of downstream and vertical mergers, as it is not monotonic.
 By contrast, for mergers between an integrated firm and an unintegrated upstream supplier (the darker blue plots), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
@@
 The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
-Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here the neutrality crossing shifts left, when wholesalers have a bit more than 1.5 times the power of retailers.
+Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here the neutrality crossing shifts left, when wholesalers have a bit more than 3/2 times the power of retailers.
 It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.
@@
 Median consumer harm moves from about 0\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
-A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see the darker and darkest blue plots) has tends to narrow the range of outcomes and somewhat shift the distribution upwards.
+A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see the darker and darkest blue plots) tends to narrow the range of outcomes and somewhat shift the distribution upwards.
 However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 times or more.
@@
 The U.S.
-waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}}} In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors.
-However, solid waste companies tend not to be homogeneous in their degree of vertical integration across geographies.
-As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevant local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.
+waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{See Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/.}}} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, national competitors.
+Furthermore, there are a variety of vertical supply chain configurations across local markets, making the industry an excellent application to study the welfare impacts of mergers that have both horizontal and vertical aspects.
 
-In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{Competitive Impact Statement: U.S.
-and Plaintiff States v.
-Waste Management, Inc.
-and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impact Statement: U.S.
+Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{See Competitive Impact Statement: U.S.
 and State of Alabama v.
-Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{Waste Business Journal,\tiny{ \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
-
-The DOJ filed a complaint and consent decree in the Republic/Santek case, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW)disposal markets.
-In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga region.\footnote{See U.S.
+Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}.}  The DOJ filed a complaint and negotiated a settlement in the case, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW) disposal markets.
+In addition, vertical anti-competitive effects were alleged to arise from the combination of Republic and Santek integrated assets in the Chattanooga region.\footnote{See U.S.
 and State of Alabama v.
 Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.
 
@@
 The vertical supply chain in the solid waste industry is primarily comprised of waste collection operations or ``haulers'' and waste disposal facilities.
-Haulers collect MSW from businesses and residences and must dispose of it at a lawful disposal site, predominantly landfills.
+Haulers collect MSW from businesses and residences and must dispose of it at a lawful disposal site, usually a landfill.
 Waste disposal (upstream) is a required input into waste collection services (downstream).
@@
 
-The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic/Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and SCCW collection as well as due to RRC in the SCCW collection market by raising the MSW disposal costs of independent haulers.
+The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic/Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area,'' subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and SCCW collection as well as due to RRC in the SCCW collection market by raising the MSW disposal costs of independent haulers.
 The CIS notes that pre-merger, Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
-In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere.
+In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations before ultimately being disposed of in the parties’ landfills elsewhere.
 Thus, pre-merger both parties were large, vertically integrated competitors in the Chattanooga Area.
 
-In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, and was the parties' sole competitor in the MSW disposal market.
-Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
-Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity (henceforth ``WM-ADS'') regardless of whether the data source predates the merger consummation since the merger was completed before the filing of the CIS for the Republic/Santek merger.
-The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.
-(We label this firm ``Regional.'')
+In addition, another large, vertically integrated waste company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in the MSW disposal market.
+National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed below.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WM-ADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
+The final significant participant in the SCCW collection market was a major regional firm that is not identified by name in the CIS.
+(We label this firm ``Regional'' in what follows.)
 
@@
 
-Data on MSW disposal comes from information collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
-This data identifies the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
+Data on MSW disposal come from information collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+These data identify the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
 The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
@@
 Disposal volumes are measured in tons and have been combined across landfills owned by the same firm.
-The TDEQ data identifies landfills owned or operated by Republic, Santek, Waste Connections, and three other market participants (each with less than 0.5\% share) receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% combined.
-Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out of state landfill in GA.} These three fringe participants have been excluded from the analysis.
+The TDEQ data identify landfills owned or operated by Republic, Santek, Waste Connections, and three other market participants (each with less than 0.5\% share) receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% share combined.
+Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out-of-state landfill in Georgia.} These three fringe participants have been excluded from the analysis.
 After re-scaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.
 
-Using these data in conjunction with information on collection market volumes, the supply relationships between the upstream disposal market participants (\textit{Republic, Santek, and Waste Connections}) and downstream collection market participants (\textit{Republic, Santek, Waste Connections, WM-ADS, and ``Regional''}) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Using these data in conjunction with information on collection market volumes, the supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and downstream collection market participants (Republic, Santek, Waste Connections, WM-ADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
 Santek and Waste Connections exhibit excess disposal capacity.
@@
 MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
-The measure of price reflected is the ``gate rate'', which is the posted price at the landfill, measured in \$/Ton.\footnote{Disposal prices for large customer's may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
-The price used in the analysis for each is the volume weighted average for their landfills.
+The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in \$/Ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
+The price used in the analysis for each is the volume-weighted average for their landfills.
 
-The CIS states that in the Chattanooga Area the post-merger Herfindahl-Hirschman Index for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
+The CIS states that in the Chattanooga Area the post-merger HHI for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
 Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
@@
 Following the discussion of supply relationships above, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
-Capacity constrained, integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
+Capacity constrained integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
 Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.
@@
 Santek, Waste Connections, and Regional do not publicly produce comparable financial statements.
-Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure in this market due to their lack of internal disposal capacity.
+Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure due to their lack of internal disposal capacity.
+
+Given these data sources, we construct our final input data set, as seen in Table 2.
+As discussed above, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
+We expect that the DOJ would have had access to more accurate confidential information during their investigation.
+Therefore, we view our results as only illustrative of the types of analysis and results that could have been produced for this case.
+Note that, other than imposing that Republic only offers disposal to itself, we do not structurally incorporate capacity constraints into the simulation model.
+We assume that the changes in volume that would be caused by the merger would be large enough to shift which constraints would be binding.
 
@@
 
-Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of harm to consumers and total producer benefits of approximately \$13.4 million.
-Thus, on net the model predicts the merger would be harmful despite the presence of significant efficiencies from EDM that reduce prices upstream.
-These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
+Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of annual harm to consumers and total annual producer benefits of approximately \$13.4 million.
+Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek (1) and Republic (0.7) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.
+High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.
+(See Table 2.)}  These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
 
@@
 However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs).
-The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections prices to its downstream rivals are estimated to increase 22\%.
-These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, and the cost to Republic for volume disposed with Santek decreases by 60\%.
+The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 22\%.
+These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by 60\%.
 
@@
 The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall.
-The pre-merger integrated Republic pair's price increases 21\% and Santek's increases by 23\% from a combination of UPP and increased bargaining leverage.
-The newly integrated Santek-Republic pair's price is estimated to increase by 9\%, demonstrating that the upstream efficiencies are not fully passed-through to consumers.
+The Republic-Republic product's price increases 21\% and the Santek-Santek product's price increases by 23\% from a combination of UPP and increased bargaining leverage.
+The newly integrated Santek-Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
 The integrated rival, Waste Connections, increases its collection price by approximately 5\% and its share increases by 6 percentage points.
-The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and share increases by 2 percentage points each.
+The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
 
-We compare these results to those from a merger simulation considering only a merger in the collection market under logit demand, treating each upstream-downstream pair as if it were an independent firm, as shown in Figure \ref{fig:TrashSimsCompare}.
-Ignoring these vertical relationships results in an estimated 9.4\% increase in collection prices with a 15\% increase in the merging parties' price, consumer harm of \$9.6 million and a producer benefit of \$7.1 million.
+In Figure \ref{fig:TrashSimsCompare}, we compare these results to those from a merger simulation considering only a merger in the collection market, treating each upstream-downstream pair as if it were an independent firm.
+Ignoring these vertical relationships results in an estimated 9.4\% increase in collection prices with a 15\% increase in the merging parties' price, consumer harm of \$9.6 million annually and a producer benefit of \$7.1 million annually.
 While net harm is nearly the same under both models at about \$2.5 million, estimated consumer harm is 67\% higher in the model including vertical relationships and is offset by markedly larger producer benefits.
-Further, the collection-only model overestimates diversion to the unintegrated rivals and as a result underestimates the market price effect due to UPP from the merger.
-(See also Table \ref{tab:trash_merged}.)
+The collection-only model misses the substantial RRC effect experienced by the Santek-WM-ADS and Santek-Regional products.
+In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and rival prices increase alongside them.
+This causes more harm to consumers.
 
 The Republic/Santek merger was ultimately settled with the DOJ through a divestiture.
-According to the final judgement, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station in the Chattanooga area.\footnote{Final Judgment: U.S and Plaintiff States v.
+According to the final judgement, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station in the Chattanooga Area.\footnote{See Final Judgment: U.S and Plaintiff States v.
 Republic Services, Inc.
-and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
+and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}.} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
 
@@
 This paper relaxes an assumption made in much of the merger simulation literature: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated.
-Removing this limitation is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are already vertically integrated.
+Removing this limitation is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are vertically integrated.
 Using a series of numerical simulations and the model of \citeN{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
-The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be largest.
-We also find that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
+The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.
+We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
 We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships drastically understates harm to Chattanooga consumers.
@@
 Our results suggest that mergers that have horizontal aspects remain harmful for consumers in many instances, even when these mergers also have vertical aspects.  
-This appears to be especially true when retailers have relatively more bargaining power, which in our model connotes with situations where retailer margins are relatively large compared to wholesale margins.
+This appears to be especially true when retailers have relatively more bargaining power, which in our model connotes with situations where retailer margins are relatively large compared to wholesaler margins.
 Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.
 However, we of course have the caveat that our results are specific to our modeling framework.
-An important question for future research would be to explore merger outcomes in in other contexts.
+An important question for future research would be to explore merger outcomes in other contexts.
 
@@
 \small{\include{sumtable}}
-\caption{Summary Statistics}
+\caption{Summary Statistics for Numerical Simulations}
 \label{tab:simsum}
@@
 \small{\include{TrashData}}
-\small{\include{TrashSims}}
-
-\begin{table}
-\centering
-\small{\include{trash_merged_output}}
-\caption{Predicted Market Shares for Integrated Model vs. Collection Only Model}
-\label{tab:trash_merged}
-\end{table}
 
@@
 \includegraphics[scale=0.9]{output/CVvertincumbBW_consumer.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market change.
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market changes.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms changes.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
@@
 \includegraphics[scale=1.0]{output/chattanooga_tree.png}
-\caption{Chattanooga Area MSW Disposal and Collection Market}
+\caption{Structure of the Chattanooga Area Disposal and Collection Market}
 \label{fig:chatt_tree}
@@
 \includegraphics{output/TrashSimsCompare.png}
-\caption{Merger Simulation Results Comparison}
+\caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
 \label{fig:TrashSimsCompare}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Robustness to Increasing Marginal Costs}
+\section{Appendix: Increasing Marginal Costs}
 
-Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allow costs to increase linearly.
-Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties products (bottom row, orange).
+Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allows costs to increase linearly.
+Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties' products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties' products (bottom row, orange).
 
 In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs.
-Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause the integrated wholesaler's prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
+Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause the input prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
 Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data for the Waste Management Markets}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Additional Figures}
+\section{Appendix: Additional Figures}
 
@@
 \includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market change.
+\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market changes.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
+The different colored boxes display how outcomes change as the number of vertically  integrated firms changes.
 Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}

### 2022-08-03 doc/TrashData.tex
Fixed issues from Gloria's 8/2 email. Added '$' to Prices in Table 2, added table reference for table 2. Changed 'WM_ADS' to 'WM-ADS' in figures and tables. Creaetd a new version of figure 6 that includes upstream merger only results.

@@
 
-\caption{\label{tab:}Republic/Santek Merger Simulation Inputs}
+\caption{\label{tab:trashdata}Republic/Santek Merger Simulation Inputs}
 \centering
 \resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llrrrrr}
+\begin{tabular}[t]{rrccccc}
 \toprule
-Disposal Firm & Collection Firm & Volume (000s) & Disposal Price & Disposal Margin (\$) & Collection Margin (\$) & Collection Price\\
+Disposal Firm & Collection Firm & Volume (000s) & Disposal Price (\$) & Disposal Margin (\$) & Collection Margin (\$) & Collection Price (\$)\\
 \midrule
 Republic & Republic & 165 & 42 & 0 & 44 & 179\\
-Santek & Republic & 34 & 36 & 20 &  & 173\\
-Santek & Santek & 218 & 16 & 0 &  & 134\\
-Santek & WM\_ADS & 30 & 36 & 20 &  & 221\\
-Santek & Regional & 30 & 36 & 20 &  & 221\\
-\addlinespace
-WasteConn & Republic & 19 & 25 & 14 &  & 162\\
-WasteConn & WasteConn & 48 & 11 & 0 &  & 110\\
-WasteConn & WM\_ADS & 17 & 25 & 14 &  & 210\\
-WasteConn & Regional & 17 & 25 & 14 &  & 210\\
+\cmidrule{1-7}
+ & Republic & 34 & 36 & 20 &  & 173\\
+
+ & Santek & 218 & 16 & 0 &  & 134\\
+
+ & WM-ADS & 30 & 36 & 20 &  & 221\\
+
+\multirow{-4}{*}{\raggedleft\arraybackslash Santek} & Regional & 30 & 36 & 20 &  & 221\\
+\cmidrule{1-7}
+ & Republic & 19 & 25 & 14 &  & 162\\
+
+ & WasteConn & 48 & 11 & 0 &  & 110\\
+
+ & WM-ADS & 17 & 25 & 14 &  & 210\\
+
+\multirow{-4}{*}{\raggedleft\arraybackslash WasteConn} & Regional & 17 & 25 & 14 &  & 210\\
 \bottomrule

### 2022-08-03 doc/VerticalArrangements.tex
Fixed issues from Gloria's 8/2 email. Added '$' to Prices in Table 2, added table reference for table 2. Changed 'WM_ADS' to 'WM-ADS' in figures and tables. Creaetd a new version of figure 6 that includes upstream merger only results.

@@
 
-Given these data sources, we construct our final input data set, as seen in Table 2.
+Given these data sources, we construct our final input data set, as seen in Table \ref{tab:trashdata}.
 As discussed above, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
@@
 High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.
-(See Table 2.)}  These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
+(See Table \ref{tab:trashdata}.)}  These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
 
@@
 \begin{table}
-\small{\include{sumtable}}
 \caption{Summary Statistics for Numerical Simulations}
 \label{tab:simsum}
+\small{\include{sumtable}}
 \end{table}

### 2022-08-03 code/TrashSims.R

@@ -21,6 +21,7 @@ simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
   #mutate(disposal_volume=ifelse(disposal_firm_name != collection_firm_name,mean(disposal_volume[disposal_firm_name != collection_firm_name]),disposal_volume)) %>%
   ungroup() %>%
   mutate(#collection_paired_share=collection_paired_share/sum(collection_paired_share),
+         collection_firm_name=gsub("_","-",collection_firm_name),
          collection_paired_share=disposal_volume/sum(disposal_volume),
          disposal_dollar_margin=ifelse(disposal_firm_name=="Republic",
                                        disposal_percent_margin*disposal_price_republic,
@@ -367,17 +368,18 @@ print(kable(select(simdata,disposal_firm_name,
               rename("Disposal Firm"=disposal_firm_name,
                      "Collection Firm"=collection_firm_name,
                      "Volume (000s)"=disposal_volume,
-                     "Disposal Price"=disposal_price,
+                     "Disposal Price ($)"=disposal_price,
                      "Disposal Margin ($)"=disposal_dollar_margin,
                      "Collection Margin ($)"=collection_dollar_margin,
-                     "Collection Price"=priceDown_repmargin_2nd
+                     "Collection Price ($)"=priceDown_repmargin_2nd
                      ) %>%
   mutate(`Volume (000s)`=`Volume (000s)`/1e3) %>%
   mutate(across(where(is.numeric),round)),
-format = "latex",
+            format = "latex",
+            align = c("r","r",rep("c",5)),
             booktabs = TRUE,
-            caption = "Republic/Santek Merger Simulation Inputs"#,label = "siminputs"
-            ) %>% kable_styling(latex_options = "scale_down")
+            caption = "Republic/Santek Merger Simulation Inputs",label = "trashdata"
+            ) %>% kable_styling(latex_options = "scale_down") %>% collapse_rows(1, latex_hline = "major")
       ,digits=2)
 
 sink()
@@ -409,12 +411,13 @@ sink()
 # )
 
 vert_sum <- summary(simres_vert)
-
 base_sum <- summary(simres_noeff)
[…]

### 2022-08-22 doc/TrashData.tex
Gloria's 8/12 edits.

--- 2022-07-21 doc/TrashData.tex
@@
 
-\caption{\label{tab:}Republic/Santek Merger Simulation Inputs}
+\caption{\label{tab:trashdata}Republic/Santek Merger Simulation Inputs. Prices and margins are reported in dollars, while volume is reported in thousands of pounds.}
 \centering
 \resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llrrrrr}
+\begin{tabular}[t]{rrcccccr}
 \toprule
-Disposal Firm & Collection Firm & Volume (000s) & Disposal Price & Disposal Margin (\$) & Collection Margin (\$) & Collection Price\\
+Disposal Firm & Collection Firm & Volume & Disposal Price & Disposal Margin & Collection Margin & Collection Price & Volume (000s)\\
 \midrule
-Republic & Republic & 165 & 42 & 0 & 44 & 179\\
-Santek & Republic & 34 & 36 & 20 &  & 173\\
-Santek & Santek & 218 & 16 & 0 &  & 134\\
-Santek & WM\_ADS & 30 & 36 & 20 &  & 221\\
-Santek & Regional & 30 & 36 & 20 &  & 221\\
-\addlinespace
-WasteConn & Republic & 19 & 25 & 14 &  & 162\\
-WasteConn & WasteConn & 48 & 11 & 0 &  & 110\\
-WasteConn & WM\_ADS & 17 & 25 & 14 &  & 210\\
-WasteConn & Regional & 17 & 25 & 14 &  & 210\\
+Republic & Republic & 164728 & 42 & 0 & 44 & 179 & 165\\
+\cmidrule{1-8}
+ & Republic & 34008 & 36 & 20 &  & 173 & 34\\
+
+ & Santek & 217727 & 16 & 0 &  & 134 & 218\\
+
+ & WM-ADS & 30492 & 36 & 20 &  & 221 & 30\\
+
+\multirow{-4}{*}{\raggedleft\arraybackslash Santek} & Regional & 30492 & 36 & 20 &  & 221 & 30\\
+\cmidrule{1-8}
+ & Republic & 19289 & 25 & 14 &  & 162 & 19\\
+
+ & WasteConn & 47786 & 11 & 0 &  & 110 & 48\\
+
+ & WM-ADS & 17294 & 25 & 14 &  & 210 & 17\\
+
+\multirow{-4}{*}{\raggedleft\arraybackslash WasteConn} & Regional & 17294 & 25 & 14 &  & 210 & 17\\
 \bottomrule

### 2022-08-22 doc/VerticalArrangements.tex
Gloria's 8/12 edits.

@@
 For horizontal upstream and downstream mergers, we find that consumers are net harmed in most instances.
-The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be the largest.
+The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.
 Furthermore, we find that integrated mergers are typically net harmful to consumers as well.
 The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.
-Thus, although all of these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
+Thus, although these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
 Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
@@
 The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.
-Both Republic and Santek operated upstream and downstream, as did one of their largest rivals, Waste Connections.
+Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.
 In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.
@@
 Section \ref{sec:sims} explains how we construct our numerical simulations and provides our results, and our application to Republic/Santek appears in Section \ref{sec:application}.
+Section \ref{sec:concl} concludes.
 Additional simulation results appear in the Appendix.
-Section \ref{sec:concl} concludes.
 
@@
 \section{Theory\label{sec:theory}}
-We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2021}.
+We begin by describing our basic framework, taken from \citeN{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
 This framework allows us to study a variety of merger configurations.
@@
 \subsection{Downstream and Upstream Competition}
-Assume there is a set of consumers who each choose to buy a product sold by a single retailer.
+Assume there is a set of consumers who each choose to buy a single product sold by a retailer.
 Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.
@@
 Suppose that retailer $r$ and wholesaler $w$ were to merge.
-Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $w$ is given by
+Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $rw$ is given by
 \begin{equation}
@@
 Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $r$ raises its prices.
-EDM between $w$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $w$ sold by retailer $s$, those units earn a larger margin.
+EDM between $w$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $w$ sold by retailer $s$, those units may earn a larger margin.
 This impact is what we label the ``indirect EDM effect,'' which comes through the interaction of EDM with UPP.
@@
 }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers to 7.
-We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers).
-Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and vertically integrated retailers purchase inputs from wholesalers other than their integrated partner.
+We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (2 to 6 for integrated mergers).
+Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and integrated retailers purchase inputs from wholesalers other than their integrated partner.
 
@@
 The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
-By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties.
+By contrast, for upstream and downstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated wholesaler or retailer, respectively, with no integrated third parties.
 For integrated mergers, the plot when the number of incumbent integrated firms equals 0 depicts a merger between two integrated firms, again with no integrated rivals.
@@
 Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
+Results for total welfare are in the Appendix.
 For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
@@
 We turn next to upstream mergers in the second panel of Figure \ref{fig:CVbargincumbent_consumer}.
-Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that of downstream and vertical mergers, as it is not monotonic.
+Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that for downstream and vertical mergers, as it is not monotonic.
 By contrast, for mergers between an integrated firm and an unintegrated upstream supplier (the darker blue plots), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
@@
 Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift this distribution upwards.
-This pattern is much less noisy than in Figure \ref{fig:CVvertincumbBW_consumer}, where we did not condition on relative bargaining power.
+This pattern for integrated rivals is much less noisy than in Figure \ref{fig:CVvertincumbBW_consumer}, where we did not condition on relative bargaining power.
 
@@
 The U.S.
-waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{See Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/.}}} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, national competitors.
+waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{See Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}.}} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, national competitors.
 Furthermore, there are a variety of vertical supply chain configurations across local markets, making the industry an excellent application to study the welfare impacts of mergers that have both horizontal and vertical aspects.
@@
 
-In addition, another large, vertically integrated waste company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in the MSW disposal market.
+In addition, another large, vertically integrated company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in MSW disposal.
 National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed below.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
@@
 
-Data on MSW disposal come from information collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+Data on MSW disposal come from information collected by Tennessee Department of Environmental Quality (TDEQ) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
 These data identify the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
@@
 Note that, other than imposing that Republic only offers disposal to itself, we do not structurally incorporate capacity constraints into the simulation model.
-We assume that the changes in volume that would be caused by the merger would be large enough to shift which constraints would be binding.
+We assume that the changes in volume that would be caused by the merger would not be large enough to shift which constraints would be binding.\footnote{We also assume that the Waste Connections-Republic product remains on the market, which may only be true in the short run.
+}
 
@@
 Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of annual harm to consumers and total annual producer benefits of approximately \$13.4 million.
-Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek (1) and Republic (0.7) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.
+Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda=1$) and Republic ($\lambda=0.7$) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.
 High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.
@@
 The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall.
-The Republic-Republic product's price increases 21\% and the Santek-Santek product's price increases by 23\% from a combination of UPP and increased bargaining leverage.
+The Republic-Republic product's price increases 21\% and the Santek-Santek product's price increases by 23\%.
 The newly integrated Santek-Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
@@
 The collection-only model misses the substantial RRC effect experienced by the Santek-WM-ADS and Santek-Regional products.
-In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and rival prices increase alongside them.
+In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and other prices increase alongside them.
 This causes more harm to consumers.
@@
 \caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
-The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue,left) or linear marginal costs (orange, right).
+The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue, left) or linear marginal costs (orange, right).
 The boxes in the bottom row (``Party") either assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).

### 2022-08-22 code/TrashSims.R

@@ -367,18 +367,18 @@ print(kable(select(simdata,disposal_firm_name,
 
               rename("Disposal Firm"=disposal_firm_name,
                      "Collection Firm"=collection_firm_name,
-                     "Volume (000s)"=disposal_volume,
-                     "Disposal Price ($)"=disposal_price,
-                     "Disposal Margin ($)"=disposal_dollar_margin,
-                     "Collection Margin ($)"=collection_dollar_margin,
-                     "Collection Price ($)"=priceDown_repmargin_2nd
+                     "Volume"=disposal_volume,
+                     "Disposal Price"=disposal_price,
+                     "Disposal Margin"=disposal_dollar_margin,
+                     "Collection Margin"=collection_dollar_margin,
+                     "Collection Price"=priceDown_repmargin_2nd
                      ) %>%
-  mutate(`Volume (000s)`=`Volume (000s)`/1e3) %>%
+  mutate(`Volume (000s)`=`Volume`/1e3) %>%
   mutate(across(where(is.numeric),round)),
             format = "latex",
             align = c("r","r",rep("c",5)),
             booktabs = TRUE,
-            caption = "Republic/Santek Merger Simulation Inputs",label = "trashdata"
+            caption = "Republic/Santek Merger Simulation Inputs. Prices and margins are reported in dollars, while volume is reported in thousands of pounds.",label = "trashdata"
             ) %>% kable_styling(latex_options = "scale_down") %>% collapse_rows(1, latex_hline = "major")
       ,digits=2)
 

### 2022-08-22 doc/TrashData.tex
minor edits to trashsims table

--- 2022-08-03 doc/TrashData.tex
@@
 
-\caption{\label{tab:trashdata}Republic/Santek Merger Simulation Inputs}
+\caption{\label{tab:trashdata}Republic/Santek Merger Simulation Inputs. Prices and margins are reported in dollars, while volume is reported in thousands of pounds.}
 \centering
@@
 \toprule
-Disposal Firm & Collection Firm & Volume (000s) & Disposal Price (\$) & Disposal Margin (\$) & Collection Margin (\$) & Collection Price (\$)\\
+Disposal Firm & Collection Firm & Volume & Disposal Price & Disposal Margin & Collection Margin & Collection Price\\
 \midrule

### 2022-08-22 code/TrashSims.R

@@ -373,12 +373,12 @@ print(kable(select(simdata,disposal_firm_name,
                      "Collection Margin"=collection_dollar_margin,
                      "Collection Price"=priceDown_repmargin_2nd
                      ) %>%
-  mutate(`Volume (000s)`=`Volume`/1e3) %>%
+  mutate(`Volume`=`Volume`/1e3) %>%
   mutate(across(where(is.numeric),round)),
             format = "latex",
             align = c("r","r",rep("c",5)),
             booktabs = TRUE,
-            caption = "Republic/Santek Merger Simulation Inputs. Prices and margins are reported in dollars, while volume is reported in thousands of pounds.",label = "trashdata"
+            caption = "Republic/Santek Merger Simulation Inputs. Volume is reported in thousands of pounds, while prices and margins are reported in dollars.",label = "trashdata"
             ) %>% kable_styling(latex_options = "scale_down") %>% collapse_rows(1, latex_hline = "major")
       ,digits=2)
 

### 2022-10-04 code/TrashSims.R
minor edits to get the code to run on Board machines. These figures use a different strategy for fixing either upstream or downstream pre-merger prices in the antitrust package than before. However, the results look the same as the first method.

@@ -299,7 +299,8 @@ simres_vert <- with(simdata,
                               ownerPostUp=ownerPostUp,
                               labels = as.character(Name),
                               insideSize=sum(disposal_volume)
-                              ,constrain="wholesaler"
+                              ,constrain="wholesaler",
+                              chain_level="full"
 ))
 
 
@@ -327,6 +328,11 @@ simres_vert@down@mcPre <- mcPre$down
 simres_vert@down@mcPost <- mcPost$down
 simres_vert@up@mcPre <- mcPre$up
 simres_vert@up@mcPost <- mcPost$up
+
+simres_vert_nodown <- simres_vert_noup <- simres_vert
+simres_vert_nodown@chain_level="wholesaler"
+simres_vert_noup@chain_level="retailer"
+
 simres_vertsPre <- calcPrices(simres_vert, preMerger = TRUE)
 simres_vert@down@pricePre <- simres_vertsPre$down
 simres_vert@up@pricePre <- simres_vertsPre$up
@@ -334,6 +340,26 @@ simres_vertsPost <- calcPrices(simres_vert, preMerger = FALSE)
 simres_vert@down@pricePost <- simres_vertsPost$down
 simres_vert@up@pricePost <- simres_vertsPost$up
 
+simres_vertsPre <- calcPrices(simres_vert_noup, preMerger = TRUE)
+simres_vert_noup@down@pricePre <- simres_vertsPre$down
+simres_vert_noup@up@pricePre <- simres_vertsPre$up
+simres_vertsPost <- calcPrices(simres_vert_noup, preMerger = FALSE)
+simres_vert_noup@down@pricePost <- simres_vertsPost$down
+simres_vert_noup@up@pricePost <- simres_vertsPost$up
+
+simres_vertsPre <- calcPrices(simres_vert_nodown, preMerger = TRUE)
+simres_vert_nodown@down@pricePre <- simres_vertsPre$down
+simres_vert_nodown@up@pricePre <- simres_vertsPre$up
+simres_vertsPost <- calcPrices(simres_vert_nodown, preMerger = FALSE)
+simres_vert_nodown@down@pricePost <- simres_vertsPost$down
+simres_vert_nodown@up@pricePost <- simres_vertsPost$up
+
+cat("Market results Full model:\n")
+summary(simres_vert,market=TRUE)
+cat("Market results Downstream Only model:\n")
+summary(simres_vert_noup,market=TRUE)
[…]

### 2022-10-24 doc/VerticalArrangements.tex
minor edits to trashsims table
minor edits to get the code to run on Board machines. These figures use a different strategy for fixing either upstream or downstream pre-merger prices in the antitrust package than before. However, the results look the same as the first method.
compiled version with Peggy's edits

@@
 \usepackage{lscape}     % helps with rotating pages
-\usepackage{setspace}   % controls line spacing
 \usepackage{caption}    % controls  captions
@@
  \usepackage{xcolor}
+
+\graphicspath{{../output/}}
 
@@
 Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.
-We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm.
-Furthermore, we show that modeling the merger as a purely horizontal downstream transaction would understate consumer harm substantially.
+We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.
+Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would understate consumer harm substantially.
 
@@
 Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
-Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens' waste themselves.
+Disposal customers include private waste haulers without their own disposal assets and as well as local governments that collect their citizens' waste themselves.
 Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
@@
 In addition, another large, vertically integrated company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in MSW disposal.
-National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed below.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in the Appendix.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
 Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WM-ADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Data}
-
-Data on MSW disposal come from information collected by Tennessee Department of Environmental Quality (TDEQ) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
-These data identify the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
-The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
-Thus, the total market quantity is defined as all MSW volumes originating in Hamilton County, Tennessee where Chattanooga is located.
-
-Disposal volumes are measured in tons and have been combined across landfills owned by the same firm.
-The TDEQ data identify landfills owned or operated by Republic, Santek, Waste Connections, and three other market participants (each with less than 0.5\% share) receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% share combined.
-Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out-of-state landfill in Georgia.} These three fringe participants have been excluded from the analysis.
-After re-scaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.
-
-Using these data in conjunction with information on collection market volumes, the supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and downstream collection market participants (Republic, Santek, Waste Connections, WM-ADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Using data described in the Appendix, the supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and downstream collection market participants (Republic, Santek, Waste Connections, WM-ADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
 Santek and Waste Connections exhibit excess disposal capacity.
@@
 
-MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
-The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in \$/Ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
-The price used in the analysis for each is the volume-weighted average for their landfills.
-
-The CIS states that in the Chattanooga Area the post-merger HHI for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
-Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
-After re-scaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WM-ADS, 8.3\%, and Regional, 8.3\%.
-
-However, the market shares required for implementation of our model are expressed as upstream-downstream market participant pairs.
-Following the discussion of supply relationships above, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
-Capacity constrained integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
-Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.
-
-Collection and disposal margins are calculated for Republic from data on revenue by line of service and components of cost of operations in the company's 2019 annual report.
-The data are reported at the company level and are not specific to the Chattanooga Area, but revenues are reported by collection segment.
-Collection costs in \$/Ton are estimated from Republic and Waste Management 2019 10k financial statements.
-These costs are reported at the company level across all segments.
-The estimated share of these costs attributable to the Chattanooga market, based on the number of markets in which the companies operate, is divided by the tons disposed for each company.
-Santek, Waste Connections, and Regional do not publicly produce comparable financial statements.
-Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure due to their lack of internal disposal capacity.
-
-Given these data sources, we construct our final input data set, as seen in Table \ref{tab:trashdata}.
-As discussed above, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
+Our final input data set is presented in Table \ref{tab:trashdata}.
+As discussed above and in the Appendix, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
 We expect that the DOJ would have had access to more accurate confidential information during their investigation.
@@
 Note that, other than imposing that Republic only offers disposal to itself, we do not structurally incorporate capacity constraints into the simulation model.
-We assume that the changes in volume that would be caused by the merger would not be large enough to shift which constraints would be binding.\footnote{We also assume that the Waste Connections-Republic product remains on the market, which may only be true in the short run.
-}
+We assume that the changes in volume that would be caused by the merger would not be large enough to shift which constraints would be binding.\footnote{We also assume that the Waste Connections-Republic product remains on the market, which may only be true in the short run.}
 
@@
 The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall.
-The Republic-Republic product's price increases 21\% and the Santek-Santek product's price increases by 23\%.
+The Republic previously integrated product's price increases 21\% and the Santek previously integrated product's price increases by 23\%.
 The newly integrated Santek-Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
@@
 
-In Figure \ref{fig:TrashSimsCompare}, we compare these results to those from a merger simulation considering only a merger in the collection market, treating each upstream-downstream pair as if it were an independent firm.
-Ignoring these vertical relationships results in an estimated 9.4\% increase in collection prices with a 15\% increase in the merging parties' price, consumer harm of \$9.6 million annually and a producer benefit of \$7.1 million annually.
-While net harm is nearly the same under both models at about \$2.5 million, estimated consumer harm is 67\% higher in the model including vertical relationships and is offset by markedly larger producer benefits.
+In Figure \ref{fig:TrashSimsCompareAll} we compare the full vertical model results, which computes both post-merger downstream and upstream price effects, to two other vertical supply chain models that hold the prices of one level fixed, treating each upstream-downstream pair as if it were an independent firm.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``Downstream Only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``Upstream Only'' model computes dosposal price effects holding collection prices fixed at pre-merger levels.
+
+Ignoring these vertical relationships in the downstream-only model results in an estimated 11\% increase in collection prices with a greater than 16\% increase in the merging parties' price, consumer harm of \$10 million annually and net harm of approximately \$2 million.
+Estimated consumer harm is 60\% higher in the model including vertical relationships and is offset by markedly larger producer benefits.
 The collection-only model misses the substantial RRC effect experienced by the Santek-WM-ADS and Santek-Regional products.
@@
 This causes more harm to consumers.
+
+In the upstream-only model, however, downstream prices and therefore shares are fixed at their pre-merger levels.
+While disposal prices increase by 8.2\% post-merger the merger effect is a pure transfer between disposal and collection firms.
+Thus, both consumer and net harm are estimated to be zero.
 
@@
 \centering
-\includegraphics[scale=0.9]{output/surplussum.png}
+\includegraphics[scale=0.9]{surplussum.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus.
@@
 \centering
-\includegraphics[scale=0.9]{output/CVvertincumbBW_consumer.png}
+\includegraphics[scale=0.9]{CVvertincumbBW_consumer.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market changes.
@@
 \centering
-\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
+\includegraphics[scale=0.9]{CVbargincumbent_consumer.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \centering
-\includegraphics[scale=1.0]{output/chattanooga_tree.png}
+\includegraphics[scale=1.0]{chattanooga_tree.png}
 \caption{Structure of the Chattanooga Area Disposal and Collection Market}
@@
 \centering
-\includegraphics[scale=1.0]{output/TrashSimsFirm.png}
+\includegraphics[scale=1.0]{TrashSimsFirm.png}
 \caption{Republic/Santek Simulation Results}
@@
 \centering
-\includegraphics{output/TrashSimsCompare.png}
+\includegraphics{TrashSimsCompare.png}
 \caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
 \label{fig:TrashSimsCompare}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{TrashSimsCompareAll.png}
+\caption{Comparison between Collection-Only, Disposal-only and Full Vertical Simulations for Republic/Santek}
+\label{fig:TrashSimsCompareAll}
 \end{figure}
@@
 \centering
-\includegraphics[scale=0.9]{output/surplussum_cost.png}
+\includegraphics[scale=0.9]{surplussum_cost.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
@@
 \end{sidewaysfigure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\section{Appendix: Empirical Application Data}
+
+Data on MSW disposal come from information collected by Tennessee Department of Environmental Quality (TDEQ) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+These data identify the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
+The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
+Thus, the total market quantity is defined as all MSW volumes originating in Hamilton County, Tennessee where Chattanooga is located.
+
+Disposal volumes are measured in tons and have been combined across landfills owned by the same firm.
+The TDEQ data identify landfills owned or operated by Republic, Santek, Waste Connections, and three other market participants (each with less than 0.5\% share) receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% share combined.
+Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out-of-state landfill in Georgia.} These three fringe participants have been excluded from the analysis.
+After re-scaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.
+
+However, the market shares required for implementation of our model are expressed as upstream-downstream market participant pairs.
+Following the discussion of supply relationships above, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
+Capacity constrained integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
+Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.
+
+MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
+The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in \$/Ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
+The price used in the analysis for each is the volume-weighted average for their landfills.
+
+The CIS states that in the Chattanooga Area the post-merger HHI for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
+Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
+After re-scaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WM-ADS, 8.3\%, and Regional, 8.3\%.
+
+Collection and disposal margins are calculated for Republic from data on revenue by line of service and components of cost of operations in the company's 2019 annual report.
+The data are reported at the company level and are not specific to the Chattanooga Area, but revenues are reported by collection segment.
+Collection costs in \$/Ton are estimated from Republic and Waste Management 2019 10k financial statements.
+These costs are reported at the company level across all segments.
+The estimated share of these costs attributable to the Chattanooga market, based on the number of markets in which the companies operate, is divided by the tons disposed for each company.
+Santek, Waste Connections, and Regional do not publicly produce comparable financial statements.
+Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure due to their lack of internal disposal capacity.
 
@@
 \centering
-\includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
+\includegraphics[scale=0.9]{CVvertincumbBW_total.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market changes.
@@
 \centering
-\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
+\includegraphics[scale=0.9]{CVbargincumbent_total.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
-For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
+%For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets.
 In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S.
@@
 The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
-The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
+%The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
 
@@
 % \centering
-% \includegraphics[scale=0.9]{../output/CVvertincumb_updownBW.png}
+% \includegraphics[scale=0.9]{../CVvertincumb_updownBW.png}
 % \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
@@
 % \centering
-% \includegraphics[scale=0.9]{../output/CVvertincumb_vertincBW.png}
+% \includegraphics[scale=0.9]{../CVvertincumb_vertincBW.png}
 % \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
@@
 % \centering
-% \includegraphics[scale=0.9]{../output/CVbargupBW.png}
+% \includegraphics[scale=0.9]{../CVbargupBW.png}
 % \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 % \centering
-% \includegraphics[scale=0.9]{../output/CVbargdownBW.png}
+% \includegraphics[scale=0.9]{../CVbargdownBW.png}
 % \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 % \centering
-% \includegraphics[scale=0.9]{../output/CVbargvertBW.png}
+% \includegraphics[scale=0.9]{../CVbargvertBW.png}
 % \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 % \centering
-% \includegraphics[scale=0.85]{output/CVbargbothBW.png}
+% \includegraphics[scale=0.85]{CVbargbothBW.png}
 % \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.

### 2022-10-27 doc/VerticalArrangements.tex
mostly small edits to paper.

@@
 \title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
-Furthermore, the views expressed here should not be purported to reflect those of the Federal Trade Commission or the U.S.
+Furthermore, the views expressed here should not be purported to reflect those of the U.S.
 Department of Justice.
@@
 Department of Justice, margaret.loudermilk@usdoj.gov} \\  U.S Department of Justice \and Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov.
-} \\ Federal Reserve Board \and  Charles Taragin\footnote{Federal Trade Commission, ctaragin@ftc.gov}  \\  Federal Trade Commission }
+} \\ Federal Reserve Board \and  Charles Taragin\footnote{Board of Governors of the Federal Reserve System, charles.s.taragin@frb.gov}  \\  Federal Reserve Board }
 
@@
 \begin{abstract}
-\noindent We study the welfare impacts of mergers with both horizontal and vertical aspects.
+\noindent We study the welfare impacts of mergers in markets where some firms are already vertically integrated.
 Our model features logit Bertrand competition downstream and Nash Bargaining upstream.
 We numerically simulate four merger types: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two integrated retailer/wholesaler pairs.
-We find that mergers with both horizontal and vertical characteristics typically harm consumers.
+We find that mergers with both horizontal and vertical characteristics typically harm consumers, and that the presence of third-party integrated firms typically does not mitigate harm.
 We also apply the model to the Republic/Santek merger as a real-world example.
@@
 Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
-The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes.
+The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes, but does not substantially mitigate consumer harm.
 
@@
 Section \ref{sec:concl} concludes.
-Additional simulation results appear in the Appendix.
+Additional simulation results appear in Appendices \ref{app:mc} and \ref{app:total}.
 
@@
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We allow for linear increasing marginal cost as a robustness check that appears in the Appendix.
+We allow for linear increasing marginal cost as a robustness check that appears in Appendix \ref{app:mc}.
 
@@
 The bargaining parameter is identical for all of the retailers in each simulation.
-We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option.
+We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option which sets its Nash-Bertrand Equilibrium price to \$5, has 15\% market share and no marginal costs.
 All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.
+We assume that pre-merger, upstream marginal costs are 25\% of upstream firm margins and 10\% of downstream firm margins.
 We set the market size to 1.
@@
 For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+After eliminating situations where the merger is unprofitable to the merging firms, as well as downstream markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
 } Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.
@@
 
-Table \ref{tab:simsum} provides summary statistics for our simulations.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}  The 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with Herfindahl Hirschman Index (HHI) values over 2,500 points.
+Table \ref{tab:simsum} provides summary statistics for our simulations.
+The 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with Herfindahl Hirschman Index (HHI) values over 2,500 points.
 The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.
@@
 However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.
-In the Appendix, we provide results that allow marginal costs to increase linearly.
+In Appendix \ref{app:mc} , we provide results that allow marginal costs to increase linearly.
 We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.
@@
 Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases.
-For the results on total welfare, see the Appendix.
+For the results on total welfare, see Appendix \ref{app:total}.
 The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
@@
 Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
-Results for total welfare are in the Appendix.
+Results for total welfare are in Appendix \ref{app:total}.
 For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
@@
 Second, the point at which this monotonic relationship between consumer welfare and bargaining power crosses zero varies across merger types, and appears to be farthest to the left in integrated mergers.
-That is, wholesalers need to have a substantial edge in bargaining power over retailers in order for consumers to benefit from an integrated merger.
+That is, wholesalers need to have a substantial bargaining power advantage over retailers in order for consumers to benefit from an integrated merger.
 Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift this distribution upwards.
@@
 Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}
 
@@
 In addition, another large, vertically integrated company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in MSW disposal.
-National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in the Appendix.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
 Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WM-ADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
@@
 
-Using data described in the Appendix, the supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and downstream collection market participants (Republic, Santek, Waste Connections, WM-ADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Using data described in Appendix \ref{app:trashdata}, the supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and downstream collection market participants (Republic, Santek, Waste Connections, WM-ADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
 Santek and Waste Connections exhibit excess disposal capacity.
@@
 Our final input data set is presented in Table \ref{tab:trashdata}.
-As discussed above and in the Appendix, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
+As discussed above and in Appendix \ref{app:trashdata}, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
 We expect that the DOJ would have had access to more accurate confidential information during their investigation.
@@
 
-In Figure \ref{fig:TrashSimsCompareAll} we compare the full vertical model results, which computes both post-merger downstream and upstream price effects, to two other vertical supply chain models that hold the prices of one level fixed, treating each upstream-downstream pair as if it were an independent firm.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``Downstream Only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``Upstream Only'' model computes dosposal price effects holding collection prices fixed at pre-merger levels.
+In Figure \ref{fig:TrashSimsCompareAll} we compare the full vertical model results, which computes both post-merger downstream and upstream price effects, to two other vertical supply chain models that hold the prices of one level fixed, treating each upstream-downstream pair as if it were an independent firm.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``Downstream Only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``Upstream Only'' model computes disposal price effects holding collection prices fixed at pre-merger levels.
 
@@
 Our results suggest that mergers that have horizontal aspects remain harmful for consumers in many instances, even when these mergers also have vertical aspects.
-This appears to be especially true when retailers have relatively more bargaining power, which in our model connotes with situations where retailer margins are relatively large compared to wholesaler margins.
+This appears to be especially true when retailers have relatively more bargaining power, which in our model occurs when retailer margins are relatively large compared to wholesaler margins.
 Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Increasing Marginal Costs}
+\section{Appendix: Increasing Marginal Costs\label{app:mc}}
 
@@
 \newpage
-\section{Appendix: Empirical Application Data}
+\section{Appendix: Empirical Application Data\label{app:trashdata}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Additional Figures}
+\section{Appendix: Additional Figures\label{app:total}}
 

### 2022-11-02 code/TrashSims.R
added pre-merger prices to compareAll figure. Changed color scheme for better black and white printing.

@@ -497,6 +497,7 @@ up_sum <- separate(up_sum,name,sep=":",into=c("Disposal","Collector")) %>%
   arrange(Level,Effect,Disposal,desc(`Change (%)`)) %>%
   mutate(across(where(is.numeric),round)) %>% relocate(Level,Effect)
 
+
 compare <- bind_rows(
   mutate(vert_sum, Model="Full"),
   mutate(vertnoup_sum, Model="Downstream Only"),
@@ -538,23 +539,27 @@ compare <- compare %>% mutate(Name=interaction(Disposal,Collector,drop=TRUE,sep=
   pivot_longer(c(`Pre-merger` ,`Post-merger`),names_to = "Type",values_to = "value") %>%
   mutate(Type=factor(Type,levels=c("Pre-merger","Post-merger")),
          Model=factor(Model,levels=c("Full","Downstream Only","Upstream Only"))) %>%
-  filter(!grepl("Change|Pre",Type))
+  filter(!grepl("Change",Type))
 
 compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
+                                      !(Type=="Pre-merger" & Model!="Full") &
                                       !(Level=="Disposal" & Model=="Downstream Only") &
-                                      !(Level=="Collection" & Model =="Upstream Only")),
+                                      !(Level=="Collection" & Model =="Upstream Only")) %>%
+                        mutate(Model=ifelse(Type=="Pre-merger","Pre-merger",as.character(Model)),
+                               Model=factor(Model,levels=c("Pre-merger","Full","Upstream Only","Downstream Only"))),
                    aes(x=Name,y=value,fill=Model,label=value)) +
   facet_grid(~Level,scales = "free_x") + geom_bar(stat="identity",
                                                          position=position_dodge()
                                                          #position="stack"
   )  +
-  xlab("Disposal/Collector") + ylab("Equilibrium Post-merger Prices") + #geom_hline(yintercept = 0,linetype="dashed") +
-  scale_fill_brewer(type="qual",palette = "Paired") +#scale_fill_grey(start = .9, end = .1) +
+  xlab("Disposal/Collector") + ylab("Equilibrium Prices") + #geom_hline(yintercept = 0,linetype="dashed") +
+  scale_fill_brewer(type="qual",palette = "Dark2",direction=-1) +#scale_fill_grey(start = .9, end = .1) +
[…]

### 2022-11-03 doc/VerticalArrangements.tex
Incorporated Gloria's 11/2 edits to document and TrashData.tex. First Full run of simulations on FRB machine. inter-quartile ranges line up almost perfectly, min and maxes are very close. Examined pictures and they look the same as well. Fixed minor issues with extra row in sumtable.

@@
 Our model features logit Bertrand competition downstream and Nash Bargaining upstream.
-We numerically simulate four merger types: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and (4) integrated mergers between two integrated retailer/wholesaler pairs.
-We find that mergers with both horizontal and vertical characteristics typically harm consumers, and that the presence of third-party integrated firms typically does not mitigate harm.
-We also apply the model to the Republic/Santek merger as a real-world example.
+We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs.
+We find that mergers that have both horizontal and vertical characteristics typically harm consumers.
+We apply the model to the Republic/Santek merger as a real-world example.
 \end{abstract}
@@
 The DOJ did not.
-}  Given these policy documents, it is not surprising that the agencies typically try to frame most mergers as either horizontal or vertical.
-However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically''\footnote{The full question is 1.g.
+}  Given these policy documents, it is not surprising that the agencies typically frame most mergers as either horizontal or vertical.
+However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
 in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.
@@
 We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.
-Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would understate consumer harm substantially.
+Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts.
+
+As a caveat to our analysis, it is based on one specific model of competition and mergers.
+Certain alternative structures are likely to give different welfare results.
+Indeed, we found in \citeN{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
+This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
+Exploring other commonly used frameworks would be a natural extension in future research.
 
@@
 The closest merger simulation article is our previous work in \citeN{ST2021}, which uses the same model we feature here.
-That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash bargaining model for the upstream market.
+That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
 \citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
@@
 Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \citeN{DKVB2010}.
-These models have been applied to several industries that are the frequent target of complex mergers, such as healthcare (e.g.
+These models have been applied to several sectors that are the frequent target of complex mergers, such as healthcare (e.g.
 \citeN{HL2017}) and video programming distribution (e.g.
@@
 Section \ref{sec:concl} concludes.
-Additional simulation results appear in Appendices \ref{app:mc} and \ref{app:total}.
+Additional simulation results and information on the data used in the application appear in the Appendix.
 
@@
 Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.
-We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$, and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.
+We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$ and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.
 We divide the set $\mathbb{W}$ into overlapping subsets, each labeled $\mathbb{W}^r$, to indicate which wholesalers' products are carried by which retailers.
@@
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We allow for linear increasing marginal cost as a robustness check that appears in Appendix \ref{app:mc}.
+We allow for linear increasing marginal costs as a robustness check that appears in Appendix \ref{app:mc}.
 
@@
 
-For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.
+For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parametrize the Dirichlet distribution so it is equivalent to a uniform distribution.
 }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers to 7.
 We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (2 to 6 for integrated mergers).
-Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and integrated retailers purchase inputs from wholesalers other than their integrated partner.
+Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner, and integrated retailers purchase inputs from wholesalers other than their integrated partner.
 
-Our simulations focus on mergers that are more likely to have competitive effects and to therefore come under agency scrutiny.
+Our simulations focus on mergers that are more likely to have anti-competitive effects and to therefore come under agency scrutiny.
 For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger.
@@
 The bargaining parameter is identical for all of the retailers in each simulation.
-We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option which sets its Nash-Bertrand Equilibrium price to \$5, has 15\% market share and no marginal costs.
+We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option that sets its Bertrand equilibrium price to \$5, has 15\% market share, and has zero marginal costs.
 All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.
@@
 However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.
-In Appendix \ref{app:mc} , we provide results that allow marginal costs to increase linearly.
+In Appendix \ref{app:mc}, we provide results that allow marginal costs to increase linearly.
 We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.
@@
 The U.S.
-waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{See Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}.}} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, national competitors.
+waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{See Waste Dive, \tiny{\url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}.}} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, nationally active competitors.
 Furthermore, there are a variety of vertical supply chain configurations across local markets, making the industry an excellent application to study the welfare impacts of mergers that have both horizontal and vertical aspects.
@@
 Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
-Disposal customers include private waste haulers without their own disposal assets and as well as local governments that collect their citizens' waste themselves.
+Disposal customers include private waste haulers without their own disposal assets as well as local governments that collect their citizens' waste themselves.
 Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
@@
 The CIS notes that pre-merger, Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
-In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations before ultimately being disposed of in the parties’ landfills elsewhere.
+In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties' landfills within the area or passing through their transfer stations before ultimately being disposed of in the parties' landfills elsewhere.
 Thus, pre-merger both parties were large, vertically integrated competitors in the Chattanooga Area.
@@
 In addition, another large, vertically integrated company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in MSW disposal.
-National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
 Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WM-ADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
@@
 As a result, Republic needs to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
-We assume that this disposal is purchased from both Santek and Waste connections according to their share of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
+We assume that this disposal is purchased from both Santek and Waste connections according to their shares of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
 We further assume that WM-ADS and Regional purchase disposal capacity from both Santek and Republic according to these shares.
@@
 Turning to the downstream market, all collection prices increase post-merger.
-The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall.
-The Republic previously integrated product's price increases 21\% and the Santek previously integrated product's price increases by 23\%.
-The newly integrated Santek-Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
-The integrated rival, Waste Connections, increases its collection price by approximately 5\% and its share increases by 6 percentage points.
+The merging parties' post-merger collection prices increase by approximately 20\%, and their share decreases by about 8 percentage points overall.
+The Republic/Republic product's price increases 21\%, and the Santek/Santek product's price increases by 23\%.
+The newly integrated Santek/Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
+The integrated rival, Waste Connections, increases its collection price by approximately 5\%, and its share increases by 6 percentage points.
 The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
 
-In Figure \ref{fig:TrashSimsCompareAll} we compare the full vertical model results, which computes both post-merger downstream and upstream price effects, to two other vertical supply chain models that hold the prices of one level fixed, treating each upstream-downstream pair as if it were an independent firm.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``Downstream Only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``Upstream Only'' model computes disposal price effects holding collection prices fixed at pre-merger levels.
+In Figure \ref{fig:TrashSimsCompareAll} we compare the full vertical model results, which compute both post-merger downstream and upstream price effects, to two other models that hold the prices of one level fixed.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``downstream-only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``upstream-only'' model computes disposal price effects holding collection prices fixed at pre-merger levels.
 
-Ignoring these vertical relationships in the downstream-only model results in an estimated 11\% increase in collection prices with a greater than 16\% increase in the merging parties' price, consumer harm of \$10 million annually and net harm of approximately \$2 million.
-Estimated consumer harm is 60\% higher in the model including vertical relationships and is offset by markedly larger producer benefits.
-The collection-only model misses the substantial RRC effect experienced by the Santek-WM-ADS and Santek-Regional products.
+Ignoring these vertical relationships in the downstream-only model results in an estimated 11\% increase in collection prices with a greater than 16\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million.
+Estimated consumer harm is 60\% higher in the full vertical model and is offset by markedly larger producer benefits.
+The downstream-only model misses the substantial RRC effect experienced by the Santek/WM-ADS and Santek/Regional products.
 In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and other prices increase alongside them.
@@
 
-In the upstream-only model, however, downstream prices and therefore shares are fixed at their pre-merger levels.
-While disposal prices increase by 8.2\% post-merger the merger effect is a pure transfer between disposal and collection firms.
+In the upstream-only model, downstream prices and therefore shares are fixed at their pre-merger levels.
+We still allow for EDM for integrated products.
+Overall disposal prices increase by 8.2\% post-merger, and the merger effect is a pure transfer between firms.
 Thus, both consumer and net harm are estimated to be zero.
+The main impact of the merger is in changing the bargaining leverage for Santek, now that it is combined with Republic.
+This effect causes Santek to raise its upstream disposal prices to both WM-ADS and Regional.
+The price increase is even larger in the upstream-only model compared to the full model, as the lack of downstream price adjustments means that WM-ADS and Regional cannot shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.
+In this sense, the upstream-only model overstates the increase in leverage that Santek receives from the merger.
 
 The Republic/Santek merger was ultimately settled with the DOJ through a divestiture.
-According to the final judgement, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station in the Chattanooga Area.\footnote{See Final Judgment: U.S and Plaintiff States v.
+According to the Final Judgment, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station.\footnote{See Final Judgment: U.S and Plaintiff States v.
 Republic Services, Inc.
@@
 We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
-We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships drastically understates harm to Chattanooga consumers.
+We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships gives an incomplete picture of the merger's welfare impacts.
 
@@
 
+\begin{table}
+\caption{Republic/Santek Merger Simulation Inputs}
 \small{\include{TrashData}}
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.  Volume is reported in thousands of tons, whereas prices and margins are reported in dollars per ton.}
+\label{tab:trashdata}
+\end{table}
 
@@
 \centering
-\includegraphics[scale=0.9]{CVvertincumbBW_consumer.png}
+\includegraphics[scale=0.9]{output/CVvertincumbBW_consumer.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market changes.
@@
 \centering
-\includegraphics[scale=0.9]{CVbargincumbent_consumer.png}
+\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \centering
-\includegraphics[scale=1.0]{chattanooga_tree.png}
+\includegraphics[scale=1.0]{output/chattanooga_tree.png}
 \caption{Structure of the Chattanooga Area Disposal and Collection Market}
@@
 \centering
-\includegraphics[scale=1.0]{TrashSimsFirm.png}
 \caption{Republic/Santek Simulation Results}
+\includegraphics[scale=1.0]{output/TrashSimsFirm.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
 \label{fig:TrashSimsFirms}
@@
 \centering
-\includegraphics{TrashSimsCompare.png}
-\caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
-\label{fig:TrashSimsCompare}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{TrashSimsCompareAll.png}
-\caption{Comparison between Collection-Only, Disposal-only and Full Vertical Simulations for Republic/Santek}
+\caption{Comparison between Downstream-Only, Upstream-Only, and Full Vertical Simulations for Republic/Santek}
+\includegraphics{output/TrashSimsCompareAll.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
 \label{fig:TrashSimsCompareAll}
@@
 In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs.
-Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause the input prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
+Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause input prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
 Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.
@@
 \newpage
-\section{Appendix: Empirical Application Data\label{app:trashdata}}
+\section{Appendix: Data for Republic/Santek\label{app:trashdata}}
 
-Data on MSW disposal come from information collected by Tennessee Department of Environmental Quality (TDEQ) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
+Data on MSW disposal come from information collected by the Tennessee Department of Environmental Quality (TDEQ) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019.
 These data identify the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee.
-The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only).
+The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations that are an intermediate disposal site only).
 Thus, the total market quantity is defined as all MSW volumes originating in Hamilton County, Tennessee where Chattanooga is located.
@@
 However, the market shares required for implementation of our model are expressed as upstream-downstream market participant pairs.
-Following the discussion of supply relationships above, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
+Following the discussion of supply relationships in Section \ref{sec:application}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
 Capacity constrained integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
@@
 MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
-The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in \$/Ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
+The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in dollars per ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
 The price used in the analysis for each is the volume-weighted average for their landfills.
@@
 
-Collection and disposal margins are calculated for Republic from data on revenue by line of service and components of cost of operations in the company's 2019 annual report.
+Collection and disposal margins are calculated for Republic from data on revenue by line of service and from components of the cost of operations in the company's 2019 annual report.
 The data are reported at the company level and are not specific to the Chattanooga Area, but revenues are reported by collection segment.
-Collection costs in \$/Ton are estimated from Republic and Waste Management 2019 10k financial statements.
+Collection costs in dollars per ton are estimated from Republic and Waste Management 2019 10k financial statements.
 These costs are reported at the company level across all segments.
@@
 \centering
-\includegraphics[scale=0.9]{CVvertincumbBW_total.png}
+\includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market changes.
@@
 \centering
-\includegraphics[scale=0.9]{CVbargincumbent_total.png}
+\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
 \caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
@@
 \end{document}
+
+\begin{figure}
+\centering
+\includegraphics{output/TrashSimsCompare.png}
+\caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
+\label{fig:TrashSimsCompare}
+\end{figure}
 

### 2022-11-03 code/RunSims.R

@@ -22,7 +22,7 @@ simpath <- "./data"
 #profiler <- microbenchmark::microbenchmark(times=1,{
 #Rprof(filename = "Rprof.out", append = FALSE, interval = 2, line.profiling=TRUE)
 
-cl <- makeCluster(mc <- getOption("cl.cores", 75))
+cl <- makeCluster(mc <- getOption("cl.cores", 30))
 ## to make this reproducible
 clusterSetRNGStream(cl, 3141519)
 
@@ -344,6 +344,7 @@ sumtable <- spread(sumtable, quant, val) %>% select(-Max,Max) %>%
   mutate(across(where(is.numeric),.fns=prettyNum, digits=2,big.mark=",")) %>%
   rename(Variable=variable,Merger=merger,`50th`=p50,`25th`=p25,`75th`=p75) %>%arrange(Variable,Merger)
 
+sumtable <- filter(sumtable,!is.na(Variable))
 sink("./doc/sumtable.tex")
 print(kable(sumtable,format="latex",digits=0) %>% collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle"))
 sink()

### 2022-11-03 code/TrashSims.R

@@ -500,8 +500,8 @@ up_sum <- separate(up_sum,name,sep=":",into=c("Disposal","Collector")) %>%
 
 compare <- bind_rows(
   mutate(vert_sum, Model="Full"),
-  mutate(vertnoup_sum, Model="Downstream Only"),
-  mutate(vertnodown_sum, Model="Upstream Only")
+  mutate(vertnoup_sum, Model="Downstream-only"),
+  mutate(vertnodown_sum, Model="Upstream-only")
 )
 
 sink("./doc/TrashSims.tex")
@@ -538,15 +538,15 @@ compare <- compare %>% mutate(Name=interaction(Disposal,Collector,drop=TRUE,sep=
   mutate(Change=`Post-merger` - `Pre-merger`) %>%
   pivot_longer(c(`Pre-merger` ,`Post-merger`),names_to = "Type",values_to = "value") %>%
   mutate(Type=factor(Type,levels=c("Pre-merger","Post-merger")),
-         Model=factor(Model,levels=c("Full","Downstream Only","Upstream Only"))) %>%
+         Model=factor(Model,levels=c("Full","Downstream-only","Upstream-only"))) %>%
   filter(!grepl("Change",Type))
 
 compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
                                       !(Type=="Pre-merger" & Model!="Full") &
-                                      !(Level=="Disposal" & Model=="Downstream Only") &
-                                      !(Level=="Collection" & Model =="Upstream Only")) %>%
+                                      !(Level=="Disposal" & Model=="Downstream-only") &
+                                      !(Level=="Collection" & Model =="Upstream-only")) %>%
                         mutate(Model=ifelse(Type=="Pre-merger","Pre-merger",as.character(Model)),
-                               Model=factor(Model,levels=c("Pre-merger","Full","Upstream Only","Downstream Only"))),
+                               Model=factor(Model,levels=c("Pre-merger","Full","Upstream-only","Downstream-only"))),
                    aes(x=Name,y=value,fill=Model,label=value)) +
   facet_grid(~Level,scales = "free_x") + geom_bar(stat="identity",
                                                          position=position_dodge()
@@ -560,7 +560,7 @@ compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
   labs(fill="")
 
 compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="Disposal" & Effect=="Prices") &
-                                      Model!="Upstream Only"),
[…]

### 2022-11-03 code/genfigures.R

@@ -126,7 +126,7 @@ pvertincumb.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer","
   #geom_text(data=ann_text,label="Wholesale advantage")
   labs(colour="Outcome:")+
   labs(title =  "The Distributions  of Merger Outcomes as the Number of Integrated Firms Increases",
-       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures.\nHorizontal mergers occur between a vertically integrated and unintegrated firm."
+       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures."
        #subtitle = "1st and 2nd score auctions yields radically different predictions for downstream mergers,\n but similar predictions for upstream mergers",
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
   )
@@ -158,7 +158,7 @@ pvertincumb_c.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Consumer"
   #geom_text(data=ann_text,label="Wholesale advantage")
   labs(colour="Outcome:")+
   labs(title =  "The Distribution of Consumer Surplus Changes as the Number of Integrated Firms Increases",
-       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures.\nHorizontal mergers occur between a vertically integrated and unintegrated firm."
+       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures."
        #subtitle = "1st and 2nd score auctions yields radically different predictions for downstream mergers,\n but similar predictions for upstream mergers",
        #caption ="outMargin = 25\nshareOutDown = .15\nmcshare.up =.25\nmcshare.down = .1\nnfirms.up = 3"
   )
@@ -190,7 +190,7 @@ pvertincumb_t.bw <-  ggplot(res.nests.logit %>% filter(Outcome %in% c("Total") #
   #geom_text(data=ann_text,label="Wholesale advantage")
   labs(colour="Outcome:")+
   labs(title =  "The Distribution of Total Surplus Changes as the Number of Integrated Firms Increases",
-       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures.\nHorizontal mergers occur between a vertically integrated and unintegrated firm."
+       subtitle="Outcomes are reported as a percentage of pre-merger total expenditures."
[…]

### 2022-11-08 doc/VerticalArrangements.tex
another round of edits from gloria, prior to submitting to EAG for next week's seminar.

@@
 The U.S.
-waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{See Waste Dive, \tiny{\url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}.}} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, nationally active competitors.
+waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{See Waste Dive, \tiny{\url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}}.} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, nationally active competitors.
 Furthermore, there are a variety of vertical supply chain configurations across local markets, making the industry an excellent application to study the welfare impacts of mergers that have both horizontal and vertical aspects.
@@
 In the upstream-only model, downstream prices and therefore shares are fixed at their pre-merger levels.
-We still allow for EDM for integrated products.
+We still have upstream prices set to marginal cost for integrated products.
 Overall disposal prices increase by 8.2\% post-merger, and the merger effect is a pure transfer between firms.
@@
 This effect causes Santek to raise its upstream disposal prices to both WM-ADS and Regional.
-The price increase is even larger in the upstream-only model compared to the full model, as the lack of downstream price adjustments means that WM-ADS and Regional cannot shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.
+The price increase is even larger in the upstream-only model compared to the full model.
+The lack of downstream price adjustments means that WM-ADS and Regional cannot shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.
 In this sense, the upstream-only model overstates the increase in leverage that Santek receives from the merger.

### 2022-11-08 code/TrashSims.R
changed color scheme in figure 6 to make it prettier and more printer friendly

@@ -553,7 +553,7 @@ compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
                                                          #position="stack"
   )  +
   xlab("Disposal/Collector") + ylab("Equilibrium Prices") + #geom_hline(yintercept = 0,linetype="dashed") +
-  scale_fill_brewer(type="qual",palette = "Dark2",direction=-1) +#scale_fill_grey(start = .9, end = .1) +
+  scale_fill_brewer(type="seq",palette = "YlGnBu",direction=-1) +#scale_fill_grey(start = .9, end = .1) +
   geom_text( #color=ifelse(Type=="Post-merger","white","black"),
     hjust=1.5, position=position_dodge(width=.9),size =3) +  coord_flip()+ theme_bw() +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +

### 2022-12-09 doc/VerticalArrangements.tex
Gloria corrected an issue with the Chatanaoog network graph.
Gloria's 12/7 edits to the doc.

@@
 
-\bigbreak Keywords: bargaining models; merger simulation; vertical markets; vertical mergers
+\bigbreak Keywords: bargaining models; merger simulation; vertical mergers
 
@@
 \section{Introduction}
-Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.
+Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for mergers involving large firms that usually attract the most scrutiny.
 However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.
@@
 
-The horizontal versus vertical distinction has been enshrined in U.S.
+The horizontal versus vertical distinction has been enshrined in the U.S.
 Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.
@@
 Thus, although these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
-Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
-The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes, but does not substantially mitigate consumer harm.
+Furthermore, we find that the presence of integrated rival firms has a slight tendency to mitigate consumer harm.
 
@@
 Certain alternative structures are likely to give different welfare results.
-Indeed, we found in \citeN{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
-This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
+Indeed, we found in \citeN{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, tended to result in less consumer harm from mergers of two downstream firms.
+However, results were more similar between these models for vertical mergers and for mergers of upstream firms.
+We also showed in our previous paper that using a nested logit demand model instead of a standard logit tended to increase consumer harm from mergers between downstream firms, between upstream firms, and between a downstream and an upstream firm.\footnote{For these results, we assumed that these merging firms were located in the same nest.
+}  In this paper, given the large variety of merger types we study, we have opted to focus on a more limited set of modeling assumptions.
 Exploring other commonly used frameworks would be a natural extension in future research.
@@
 For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-After eliminating situations where the merger is unprofitable to the merging firms, as well as downstream markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
+After eliminating situations where the merger is unprofitable to the merging firms, as well as downstream markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a firm that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
 } Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.
@@
 Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.
-In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integrated mergers first-order stochastically dominates consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers.
+In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integrated mergers first-order stochastically dominate consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers.
 However, while consumer harm is typically greater under integrated mergers than downstream mergers, it is not first-order stochastically dominant.
@@
 For the results on total welfare, see Appendix \ref{app:total}.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+The plots in the first three panels when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}
+
+For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
 By contrast, for upstream and downstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated wholesaler or retailer, respectively, with no integrated third parties.
 For integrated mergers, the plot when the number of incumbent integrated firms equals 0 depicts a merger between two integrated firms, again with no integrated rivals.
+Further increases along the horizontal axis add integrated rivals.
+Note that as the value on the horizontal axis rises towards 4, the underlying sample of included simulations mechanically shifts towards less concentrated markets.
+This occurs because, in order to include as many as 4 integrated producers, a market must have at least that many firms prior to the merger, which rules out situations like mergers to monopoly.
 
-We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
+We begin with vertical mergers in Figure \ref{fig:CVvertincumbBW_consumer}, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
 Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW_consumer}), about half of vertical mergers benefit consumers.
@@
 Although these mergers have the possibility of creating benefits through EDM, these gains appear to be outweighed in the majority of the mergers we study.
-Second, the presence of integrated rivals has some impact on the distribution of consumer harm, but the observed patterns are noisy and the resulting box plots are largely similar to those where such rivals are absent.
+Second, the presence of integrated rivals has some impact on the distribution of consumer harm, but the observed patterns are noisy and the resulting box plots are largely similar to those where these rivals are absent.
 Therefore, having such rival firms in and of itself does not generate significantly better outcomes for consumers.
+This latter result is interesting, as one might have expected to see a marked decrease in the extent of consumer harm, given the tendency in our sample construction towards less concentrated markets as the number of integrated rivals rises.
 
@@
 Results for total welfare are in Appendix \ref{app:total}.
+
 For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
@@
 In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.
-Finally, in integrated mergers, when the number of incumbent integrated firms equals 0, this depicts a merger between two integrated firms with all rivals are unintegrated.
+Finally, in integrated mergers, when the number of incumbent integrated firms equals 0, this depicts a merger between two integrated firms with all rivals being unintegrated.
 Plots with either 1 or 4 integrated firms depict mergers between integrated incumbents when there are either 1 or 4 rival integrated firms in the market, respectively.
@@
 The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
-Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here the neutrality crossing shifts left, when wholesalers have a bit more than 3/2 times the power of retailers.
+Compared to vertical mergers, where impacts to consumers were roughly neutral at a 1 to 2/3 ratio of bargaining power, here the neutrality crossing shifts left, when wholesalers have between 7/3 and 3/2 times the power of retailers.
 It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.
@@
 That is, wholesalers need to have a substantial bargaining power advantage over retailers in order for consumers to benefit from an integrated merger.
-Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift this distribution upwards.
+Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift the distribution upwards.
 This pattern for integrated rivals is much less noisy than in Figure \ref{fig:CVvertincumbBW_consumer}, where we did not condition on relative bargaining power.
+The tightening and upward shift of the welfare distribution is expected, given the tendency in the sample towards less concentrated markets as the number of integrated rivals increases.
+In order to correct for this sampling issue, we also generated versions of Figure \ref{fig:CVbargincumbent_consumer} that hold the total number of upstream and downstream firms constant.\footnote{These figures are available upon request from the authors.
+}  These additional figures still show that the median of the welfare distribution tends to shift upwards as integrated rivals are added, though the range of the distribution no longer has a strong pattern.
 
@@
 The CIS notes that pre-merger, Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
-In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties' landfills within the area or passing through their transfer stations before ultimately being disposed of in the parties' landfills elsewhere.
+In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market.
 Thus, pre-merger both parties were large, vertically integrated competitors in the Chattanooga Area.
@@
 In addition, another large, vertically integrated company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in MSW disposal.
-National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by the MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms.
 Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WM-ADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
@@
 Santek and Waste Connections exhibit excess disposal capacity.
-However, Republic's collection volumes are estimated to exceed their ability to self-supply disposal in the Chattanooga market.
+In contrast, Republic's collection volumes are estimated to exceed their ability to self-supply disposal in the Chattanooga market.
 As a result, Republic needs to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
-We assume that this disposal is purchased from both Santek and Waste connections according to their shares of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
-We further assume that WM-ADS and Regional purchase disposal capacity from both Santek and Republic according to these shares.
+We assume that this disposal is purchased from both Santek and Waste Connections according to their shares of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
+We further assume that WM-ADS and Regional purchase disposal capacity from both Santek and Waste Connections according to these shares.
 The resulting relationships between the upstream and downstream market participants are captured in Figure \ref{fig:chatt_tree}.
@@
 Disposal prices for pre-merger integrated pairs are unchanged post-merger.
-However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs).
+In contrast, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs).
 The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 22\%.
-These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by 60\%.
+These large price increases are offset by the large decrease in the disposal price for the integrating pair due to EDM, as the cost to Republic for volume disposed with Santek decreases by 60\%.
 
@@
 Republic Services, Inc.
-and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}.} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
+and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}.} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the harm from the merger.
 
@@
 The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.
-We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
+We also show that having integrated non-merging rivals tends to make mergers slightly more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
 We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships gives an incomplete picture of the merger's welfare impacts.
@@
 \centering
-\includegraphics[scale=1.0]{output/chattanooga_tree.png}
-\caption{Structure of the Chattanooga Area Disposal and Collection Market}
+\includegraphics[scale=.7]{output/chattanooga_tree.png}
+\caption{Structure of the Chattanooga Area Disposal and Collection Markets}
 \label{fig:chatt_tree}
@@
 \includegraphics[scale=0.9]{output/surplussum_cost.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
+\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to different cost scenarios.
 The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue, left) or linear marginal costs (orange, right).
@@
 These costs are reported at the company level across all segments.
-The estimated share of these costs attributable to the Chattanooga market, based on the number of markets in which the companies operate, is divided by the tons disposed for each company.
+The estimated share of these costs attributable to the Chattanooga Area, based on the number of markets in which the companies operate, is divided by the tons disposed for each company.
 Santek, Waste Connections, and Regional do not publicly produce comparable financial statements.
-Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure due to their lack of internal disposal capacity.
+Instead, we assume that the cost structure is the same for the integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure due to their lack of internal disposal capacity.
 
@@
 We see in Figure \ref{fig:CVvertincumbBW_total} that most upstream, downstream, and integrated mergers are harmful.
-Figure \ref{fig:CVbargincumbent_total} shows that net beneficial
-mergers typically only occur when wholesalers have relatively more bargaining power.  The presence of integrated rivals can narrow the range of outcomes.
+Figure \ref{fig:CVbargincumbent_total} shows that net beneficial mergers typically only occur when wholesalers have relatively more bargaining power.
 

### 2022-12-14 doc/VerticalArrangements.tex
Formatted paper for submission to IJIO. Paper submitted to IJIO 12/14

--- 2022-11-08 doc/VerticalArrangements.tex
@@
-\documentclass[12pt]{article}
-\usepackage{chicago}    % bibliography package
+\documentclass[12pt,review,authoryear]{elsarticle}
+%\usepackage{ecrc}       % needed for elsarticle
+%\usepackage{chicago}    % bibliography package
 \usepackage{graphicx}   % insert PostScript figures
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
-Furthermore, the views expressed here should not be purported to reflect those of the U.S.
-Department of Justice.
-This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}
 
-\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
-\author{Margaret Loudermilk\footnote{U.S.
+%\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
+%\author{Margaret Loudermilk\footnote{U.S.
 Department of Justice, margaret.loudermilk@usdoj.gov} \\  U.S Department of Justice \and Gloria Sheu\footnote{Board of Governors of the Federal Reserve System, gloria.sheu@frb.gov.
@@
 
-\pagenumbering{roman}       % Roman numerals from abstract to text
-\maketitle                  % print title information
-\thispagestyle{empty}       % no page number on THIS page
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for the associated footnote;
+%% use the fnref command within \author or \address for footnotes;
+%% use the fntext command for the associated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for the associated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+%%
+%% \title{Title\tnoteref{label1}}
+%% \tnotetext[label1]{}
+%% \author{Name\corref{cor1}\fnref{label2}}
+%% \ead{email address}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \address{Address\fnref{label3}}
+%% \fntext[label3]{}
+
+%%\dochead{}
+%% Use \dochead if there is an article header, e.g. \dochead{Short communication}
+
+\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1} }
+\tnotetext[t1]{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
+Furthermore, the views expressed here should not be purported to reflect those of the U.S.
+Department of Justice.
+This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}
+
+%% use optional labels to link authors explicitly to addresses:
+%% \author[label1,label2]{<author name>}
+%% \address[label1]{<address>}
+%% \address[label2]{<address>}
+
+\author[1]{Margaret Loudermilk\fnref{fn1}}
+\ead{margaret.loudermilk@usdoj.gov}
+\affiliation[1]{organization={US Department of Justice}, addressline= {450 5th St. NW}, city={Washington, DC},postcode={20530}}
+\author[2]{Gloria Sheu\fnref{fn1}}
+\ead{gloria.sheu@frb.gov}
+\affiliation[2]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
+\author[3]{Charles Taragin\corref{cor1}\fnref{fn1}}
+ \cortext[cor1]{Corresponding author.}
+\ead{charles.s.taragin@frb.gov}
+\affiliation[3]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
+\fntext[fn1]{Declarations of interest: none.}
 
@@
 We apply the model to the Republic/Santek merger as a real-world example.
+
 \end{abstract}
 
-\bigbreak Keywords: bargaining models; merger simulation; vertical markets; vertical mergers
+\begin{keyword}
+bargaining models\sep merger simulation\sep vertical markets\sep vertical mergers
+\JEL  L13\sep L40\sep L41\sep L42
+\end{keyword}
 
-JEL classification: L13; L40; L41; L42
+\end{frontmatter}
 
-\newpage                    % start a new page
-\pagenumbering{arabic}      % Arabic page numbers from now on
-\doublespacing
+%\pagenumbering{roman}       % Roman numerals from abstract to text
+%\maketitle                  % print title information
+%\thispagestyle{empty}       % no page number on THIS page
+
+%\newpage                    % start a new page
+%\pagenumbering{arabic}      % Arabic page numbers from now on
+%\doublespacing
 
@@
 
-Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
+Our model is drawn directly from \cite{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
 We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
@@
 Certain alternative structures are likely to give different welfare results.
-Indeed, we found in \citeN{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
+Indeed, we found in \cite{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
 This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
@@
 Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
-The closest merger simulation article is our previous work in \citeN{ST2021}, which uses the same model we feature here.
-That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
-\citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
-\citeN{MS2013} and \citeN{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
+The closest merger simulation article is our previous work in \cite{ST2021}, which uses the same model we feature here.
+That paper in turn takes the setup from \cite{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
+\cite{DVDS2018} and \cite{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
+\cite{MS2013} and \cite{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
 However, none of these papers have results for the wide range of merger types that we study.
 
-Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \citeN{DKVB2010}.
+Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \cite{DKVB2010}.
 These models have been applied to several sectors that are the frequent target of complex mergers, such as healthcare (e.g.
-\citeN{HL2017}) and video programming distribution (e.g.
-\citeN{CLWY2017}).
+\cite{HL2017}) and video programming distribution (e.g.
+\cite{CLWY2017}).
 These structural empirical papers explore the welfare effects of various competitive interactions in certain industries.
@@
 \section{Theory\label{sec:theory}}
-We begin by describing our basic framework, taken from \citeN{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
+We begin by describing our basic framework, taken from \cite{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
 This framework allows us to study a variety of merger configurations.
@@
 Wholesale prices are set via Nash Bargaining between retailers and wholesalers.
-We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
+We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \cite{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
 }  Profits for wholesaler $w$ are given by $\pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M$, where $c^W_{rw}$ is the wholesale marginal cost when dealing with retailer $r$.
@@
 Together the series of downstream and upstream first order conditions determine market equilibrium.
-This model can be solved and calibrated as described in \citeN{ST2021}.
+This model can be solved and calibrated as described in \cite{ST2021}.
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
@@
 \section{Numerical Simulations \label{sec:sims}}
-In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations.
+In this section, we use a similar simulation setup to that in \cite{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations.
 
@@
 For the results on total welfare, see Appendix \ref{app:total}.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \cite{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \cite{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
 By contrast, for upstream and downstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated wholesaler or retailer, respectively, with no integrated third parties.
@@
 As with upstream mergers, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues.
-\citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
+\cite{ST2021} found the same result for mergers in downstream logit Bertrand markets.
 Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, but median consumer harm grows to about 12\% of pre-merger revenues.
@@
 \subsection{Mergers and Bargaining Power}
-\citeN{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power.
+\cite{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power.
 Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.
@@
 Results for total welfare are in Appendix \ref{app:total}.
-For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
+For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \cite{ST2021}.
 For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
@@
 We focus first on the case when there are no incumbent integrated firms (the light blue plots).
-Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
+Consistent with the patterns seen in \cite{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
 Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.
@@
 Removing this limitation is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are vertically integrated.
-Using a series of numerical simulations and the model of \citeN{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
+Using a series of numerical simulations and the model of \cite{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
 The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.
@@
 \newpage
-\bibliographystyle{chicago}
+%\bibliographystyle{chicago}
+\bibliographystyle{elsarticle-harv}
 \bibliography{master-bib}
@@
 \centering
-\includegraphics[scale=1.0]{output/chattanooga_tree.png}
+\includegraphics[scale=.7]{output/chattanooga_tree.png}
 \caption{Structure of the Chattanooga Area Disposal and Collection Market}

### 2023-06-02 doc/VerticalArrangements.tex
revised draft that addresses key comments from IJIO reviewers. Converted format of doc back to basic latex from elsevier

@@
-\documentclass[12pt,review,authoryear]{elsarticle}
+\documentclass[12pt]{article}
+%\documentclass[12pt,review,authoryear]{elsarticle}
 %\usepackage{ecrc}       % needed for elsarticle
@@
 
+% \begin{frontmatter}
+%
+% %% Title, authors and addresses
+%
+% %% use the tnoteref command within \title for footnotes;
+% %% use the tnotetext command for the associated footnote;
+% %% use the fnref command within \author or \address for footnotes;
+% %% use the fntext command for the associated footnote;
+% %% use the corref command within \author for corresponding author footnotes;
+% %% use the cortext command for the associated footnote;
+% %% use the ead command for the email address,
+% %% and the form \ead[url] for the home page:
+% %%
+% %% \title{Title\tnoteref{label1}}
+% %% \tnotetext[label1]{}
+% %% \author{Name\corref{cor1}\fnref{label2}}
+% %% \ead{email address}
+% %% \ead[url]{home page}
+% %% \fntext[label2]{}
+% %% \cortext[cor1]{}
+% %% \address{Address\fnref{label3}}
+% %% \fntext[label3]{}
+%
+% %%\dochead{}
+% %% Use \dochead if there is an article header, e.g. \dochead{Short communication}
+%
+% \title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1} }
+% \tnotetext[t1]{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
+Furthermore, the views expressed here should not be purported to reflect those of the U.S.
+Department of Justice.
+This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}
+%
+% %% use optional labels to link authors explicitly to addresses:
+% %% \author[label1,label2]{<author name>}
+% %% \address[label1]{<address>}
+% %% \address[label2]{<address>}
+%
+% \author[1]{Margaret Loudermilk\fnref{fn1}}
+% \ead{margaret.loudermilk@usdoj.gov}
+% \affiliation[1]{organization={US Department of Justice}, addressline= {450 5th St. NW}, city={Washington, DC},postcode={20530}}
+% \author[2]{Gloria Sheu\fnref{fn1}}
+% \ead{gloria.sheu@frb.gov}
+% \affiliation[2]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
+% \author[3]{Charles Taragin\corref{cor1}\fnref{fn1}}
+%  \cortext[cor1]{Corresponding author.}
+% \ead{charles.s.taragin@frb.gov}
+% \affiliation[3]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
+% \fntext[fn1]{Declarations of interest: none.}
+
+\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
+Furthermore, the views expressed here should not be purported to reflect those of the U.S.
+Department of Justice.
+This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}
+
+\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
+\author{Margaret Loudermilk\footnote{margaret.loudermilk@usdoj.gov} \\  U.S Department of Justice \and Gloria Sheu\footnote{gloria.sheu@frb.gov.
+} \\ Federal Reserve Board \and  Charles Taragin\footnote{charles.s.taragin@frb.gov (corresponding author).}  \\  Federal Reserve Board }
+
 \date{\today}
@@
 
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-%%\dochead{}
-%% Use \dochead if there is an article header, e.g. \dochead{Short communication}
-
-\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1} }
-\tnotetext[t1]{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
-Furthermore, the views expressed here should not be purported to reflect those of the U.S.
-Department of Justice.
-This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label2]{<address>}
-
-\author[1]{Margaret Loudermilk\fnref{fn1}}
-\ead{margaret.loudermilk@usdoj.gov}
-\affiliation[1]{organization={US Department of Justice}, addressline= {450 5th St. NW}, city={Washington, DC},postcode={20530}}
-\author[2]{Gloria Sheu\fnref{fn1}}
-\ead{gloria.sheu@frb.gov}
-\affiliation[2]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
-\author[3]{Charles Taragin\corref{cor1}\fnref{fn1}}
- \cortext[cor1]{Corresponding author.}
-\ead{charles.s.taragin@frb.gov}
-\affiliation[3]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
-\fntext[fn1]{Declarations of interest: none.}
+\pagenumbering{roman}       % Roman numerals from abstract to text
+\maketitle                  % print title information
+\thispagestyle{empty}       % no page number on THIS page
 
@@
 
-\begin{keyword}
-bargaining models\sep merger simulation\sep vertical markets\sep vertical mergers
-\JEL  L13\sep L40\sep L41\sep L42
-\end{keyword}
+% \begin{keyword}
+% bargaining models\sep merger simulation\sep vertical markets\sep vertical mergers
+% \JEL  L13\sep L40\sep L41\sep L42
+% \end{keyword}
+%
+% \end{frontmatter}
 
-\end{frontmatter}
+\bigbreak Keywords: bargaining models; merger simulation; vertical markets; vertical mergers
 
-%\pagenumbering{roman}       % Roman numerals from abstract to text
-%\maketitle                  % print title information
-%\thispagestyle{empty}       % no page number on THIS page
+JEL classification: L13; L40; L41; L42
 
-%\newpage                    % start a new page
-%\pagenumbering{arabic}      % Arabic page numbers from now on
-%\doublespacing
+\newpage                    % start a new page
+\pagenumbering{arabic}      % Arabic page numbers from now on
+\doublespacing
 
@@
 Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
-As a result, the antitrust community has little consensus for what welfare impacts one should expect from a merger that combines all of these effects simultaneously.
+As a result, the antitrust community has little consensus regarding how complex mergers that possess both horizontal and vertical aspects will affect welfare.
 
@@
 If so, how?''.
-}  In this paper, our goal is to provide guidance to practitioners on how these mergers can be simulated in a familiar framework and to also show the range of welfare outcomes that are likely given this modeling environment.
+}  In this paper, we address this question using a vertical supply chain model to explore the welfare implications of complex mergers.
 
 Our model is drawn directly from \cite{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
-We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
-In each of these simulations, we calculate the impacts on consumer welfare.
-Furthermore, we examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
+We use this model to explore the welfare implications of four types of complex mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
 
-Our simulations show certain patterns across mergers.
-For horizontal upstream and downstream mergers, we find that consumers are net harmed in most instances.
+Our model-based simulations show certain patterns across mergers.
+For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.
 The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.
-Furthermore, we find that integrated mergers are typically net harmful to consumers as well.
+Likewise, we find that integrated mergers are typically net harmful to consumers as well.
 The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.
-Thus, although these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
-Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
+Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
+
+We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
+We find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
 The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes, but does not substantially mitigate consumer harm.
 
-We then employ this model to investigate the 2021 merger of two solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
+We then employ this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
 The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.
@@
 
-As a caveat to our analysis, it is based on one specific model of competition and mergers.
-Certain alternative structures are likely to give different welfare results.
-Indeed, we found in \cite{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
-This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
-Exploring other commonly used frameworks would be a natural extension in future research.
-
 Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
-The closest merger simulation article is our previous work in \cite{ST2021}, which uses the same model we feature here.
+The closest merger simulation article is our previous work in \cite{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger.
 That paper in turn takes the setup from \cite{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
@@
 \cite{MS2013} and \cite{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
-However, none of these papers have results for the wide range of merger types that we study.
+However, none of these papers study complex mergers.
 
@@
 We begin by describing our basic framework, taken from \cite{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
-This framework allows us to study a variety of merger configurations.
+This framework allows us to study a variety of complex merger configurations.
 
@@
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We allow for linear increasing marginal costs as a robustness check that appears in Appendix \ref{app:mc}.
+We allow for linear increasing marginal costs as a robustness check that appears in \ref{app:mc}.
 
@@
 \subsection{Mergers}
-Here we describe the manner in which mergers are modeled in this framework.
+Here we describe the manner in which complex mergers are modeled in this framework.
 We begin by describing vertical mergers, as all cases we examine have some vertical aspects.
@@
 \section{Numerical Simulations \label{sec:sims}}
-In this section, we use a similar simulation setup to that in \cite{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations.
+In this section, we use a similar simulation setup to that in \cite{ST2021}, but allow for either the merging parties or their rivals to be vertically integrated before they merge.
+This modification allows us to explore the welfare implications of additional types of mergers beyond purely horizontal and purely vertical combinations.
 
@@
 \subsection{Data Generating Process}
-Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.
+Broadly speaking, we consider four categories of complex mergers: downstream, upstream, vertical, and integrated.
 We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.
@@
 This paper relaxes an assumption made in much of the merger simulation literature: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated.
-Removing this limitation is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are vertically integrated.
+Removing this limitation is important as competition authorities are routinely called upon to investigate complex mergers where either the merging parties or a third party are vertically integrated.
 Using a series of numerical simulations and the model of \cite{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
@@
 Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.
+
 However, we of course have the caveat that our results are specific to our modeling framework.
-An important question for future research would be to explore merger outcomes in other contexts.
+Certain alternative structures are likely to give different welfare results.
+Indeed, we found in \cite{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
+This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
+Exploring other commonly used frameworks would be a natural extension in future research.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Increasing Marginal Costs\label{app:mc}}
+\section{Increasing Marginal Costs\label{app:mc}}
 

### 2023-06-07 doc/VerticalArrangements.tex
fixed some more issues with removing the elseveir formatting
this version incorporates gloria's edits

@@
 %\usepackage{ecrc}       % needed for elsarticle
-%\usepackage{chicago}    % bibliography package
+\usepackage{chicago}    % bibliography package
 \usepackage{graphicx}   % insert PostScript figures
@@
 \begin{abstract}
-\noindent We study the welfare impacts of mergers in markets where some firms are already vertically integrated.
+\noindent We study the welfare impacts of mergers when some firms are already vertically integrated.
 Our model features logit Bertrand competition downstream and Nash Bargaining upstream.
 We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs.
-We find that mergers that have both horizontal and vertical characteristics typically harm consumers.
+In our model, we find that mergers with both horizontal and vertical characteristics typically harm consumers.
 We apply the model to the Republic/Santek merger as a real-world example.
@@
 Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
-As a result, the antitrust community has little consensus regarding how complex mergers that possess both horizontal and vertical aspects will affect welfare.
+As a result, the antitrust community has little consensus regarding how ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects, will affect welfare.
 
@@
 }  Given these policy documents, it is not surprising that the agencies typically frame most mergers as either horizontal or vertical.
-However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
+Similarly, most antitrust research maintains this separate categorization, leaving us with little systematic evidence for how complex mergers impact consumers.
+However, as part of the current effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
 in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.
@@
 Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition?
-If so, how?''.
-}  In this paper, we address this question using a vertical supply chain model to explore the welfare implications of complex mergers.
+If so, how?''  }  In this paper, we address this question using a vertical supply chain model to explore the welfare implications of complex mergers.
 
-Our model is drawn directly from \cite{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
-We use this model to explore the welfare implications of four types of complex mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+Our model is drawn from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
+We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+Relative to \citeN{ST2021}, the first category, vertical mergers, appears in the previous paper and here serves as a baseline for comparison.
+The other three cases, which feature at least one merging firm that is already integrated, are new.
 
@@
 Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
-
 We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
-We find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.
-The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes, but does not substantially mitigate consumer harm.
+We find that the existence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less harm.
 
@@
 
+There are a few caveats to our analysis.
+First, products in our merger simulations do not experience any changes in their qualities or costs, other than movements in their wholesale prices due to shifts in bargaining leverage.
+Therefore, we rule out merger efficiencies as they are conceptualized in, for example, \citeN{werden96}, and we instead focus on EDM as the main channel through which mergers may benefit consumers.
+Second, we model upstream bargaining as being over linear fees, rather than over two-part tariffs or other compensation schedules.
+Although this assumption means bargaining is inefficient, linear input prices are widely observed in some industries, perhaps because of difficulties with implementing or monitoring more complicated contracts.\footnote{One example of an industry with linear wholesale pricing that has been featured in several vertical merger proceedings is cable programming distribution, as detailed in \citeN{Rogerson2014}.}  Third, our model assumes that upstream and downstream prices are set treating all other prices as fixed.
+Although we make this assumption in part for simplicity, it also may be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.
+In contrast, \citeN{BBMM2023} allow downstream firms to treat their own prices as adjustable during upstream negotiations.
+The authors argue that this setup is useful for settings like grocery stores, where menu costs are low.
+}  Fourth, we rely on the logit for our demand function, which limits the ways in which consumers substitute between products.
+In \citeN{ST2012}, we found that assuming the merging firms operate in the same group within a nested logit model tends to increase the magnitude of consumer harm relative to the flat logit.
+
 Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
-The closest merger simulation article is our previous work in \cite{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger.
-That paper in turn takes the setup from \cite{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
-\cite{DVDS2018} and \cite{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
-\cite{MS2013} and \cite{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
+The closest merger simulation article is our previous work in \citeN{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger.
+That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
+\citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
+\citeN{MS2013} and \citeN{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
 However, none of these papers study complex mergers.
 
-Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \cite{DKVB2010}.
+Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \citeN{DKVB2010}.
 These models have been applied to several sectors that are the frequent target of complex mergers, such as healthcare (e.g.
-\cite{HL2017}) and video programming distribution (e.g.
-\cite{CLWY2017}).
+\citeN{HL2017}) and video programming distribution (e.g.
+\citeN{CLWY2017}).
 These structural empirical papers explore the welfare effects of various competitive interactions in certain industries.
 Unlike these papers, our framework has been simplified for use with limited data, with the goal of being more accessible to practitioners.
-These simplifications allow us to provide numerical results over a wide range of market conditions, which complements the insights that come from the structural empirical literature on specific markets.
+These simplifications allow us to provide numerical results over a wide range of market conditions, which complements the insights that come from the structural empirical literature on specific case studies.
 
@@
 \section{Theory\label{sec:theory}}
-We begin by describing our basic framework, taken from \cite{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
+We begin by describing our basic framework, taken from \citeN{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
 This framework allows us to study a variety of complex merger configurations.
@@
 Wholesale prices are set via Nash Bargaining between retailers and wholesalers.
-We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \cite{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
+We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
 }  Profits for wholesaler $w$ are given by $\pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M$, where $c^W_{rw}$ is the wholesale marginal cost when dealing with retailer $r$.
@@
 Together the series of downstream and upstream first order conditions determine market equilibrium.
-This model can be solved and calibrated as described in \cite{ST2021}.
+This model can be solved and calibrated as described in \citeN{ST2021}.
 In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We allow for linear increasing marginal costs as a robustness check that appears in \ref{app:mc}.
+We allow for linear increasing marginal costs as a robustness check that appears in Appendix \ref{app:mc}.
 
@@
 Here we describe the manner in which complex mergers are modeled in this framework.
+Throughout we assume that each product's quality, $\delta_{rw}$, and marginal costs, $c^R_{rw}$ and $c^W_{rw}$, remain unchanged after the merger.
 We begin by describing vertical mergers, as all cases we examine have some vertical aspects.
@@
 \section{Numerical Simulations \label{sec:sims}}
-In this section, we use a similar simulation setup to that in \cite{ST2021}, but allow for either the merging parties or their rivals to be vertically integrated before they merge.
-This modification allows us to explore the welfare implications of additional types of mergers beyond purely horizontal and purely vertical combinations.
+In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for either the merging parties or their rivals to be vertically integrated before they merge.
+This modification allows us to explore the welfare implications of complex mergers, rather than only purely horizontal or purely vertical combinations.
 
@@
 \subsection{Data Generating Process}
-Broadly speaking, we consider four categories of complex mergers: downstream, upstream, vertical, and integrated.
+Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.
 We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.
@@
 However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.
-In Appendix \ref{app:mc}, we provide results that allow marginal costs to increase linearly.
-We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.
+In Appendix \ref{app:mc}, we provide results that allow marginal costs to increase linearly.\footnote{We still maintain that these cost functions are not affected by the merger, which rules out merger cost efficiencies.
+}  We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.
 
@@
 For the results on total welfare, see Appendix \ref{app:total}.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \cite{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \cite{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
+The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{One difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
 By contrast, for upstream and downstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated wholesaler or retailer, respectively, with no integrated third parties.
@@
 As with upstream mergers, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues.
-\cite{ST2021} found the same result for mergers in downstream logit Bertrand markets.
+\citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
 Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, but median consumer harm grows to about 12\% of pre-merger revenues.
@@
 \subsection{Mergers and Bargaining Power}
-\cite{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power.
+\citeN{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power.
 Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.
@@
 Results for total welfare are in Appendix \ref{app:total}.
-For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \cite{ST2021}.
+For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
 For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
@@
 We focus first on the case when there are no incumbent integrated firms (the light blue plots).
-Consistent with the patterns seen in \cite{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
+Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
 Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.
@@
 This paper relaxes an assumption made in much of the merger simulation literature: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated.
-Removing this limitation is important as competition authorities are routinely called upon to investigate complex mergers where either the merging parties or a third party are vertically integrated.
-Using a series of numerical simulations and the model of \cite{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
+Removing this limitation is important as competition authorities are routinely called upon to investigate complex mergers where one or both of the merging parties are vertically integrated.
+Using a series of numerical simulations and the model of \citeN{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
 The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.
@@
 Certain alternative structures are likely to give different welfare results.
-Indeed, we found in \cite{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
+Indeed, we found in \citeN{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
 This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
@@
 \newpage
-%\bibliographystyle{chicago}
-\bibliographystyle{elsarticle-harv}
+\bibliographystyle{chicago}
+%\bibliographystyle{elsarticle-harv}
 \bibliography{master-bib}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Increasing Marginal Costs\label{app:mc}}
+\section{Appendix: Increasing Marginal Costs\label{app:mc}}
 

### 2024-07-26 doc/JIE/JINDECMS2023138-1Response.tex
Added JIE referee reports and first draft of response template.

\documentclass[12pt]{article}
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs
\usepackage{parskip}

\newcounter{para}
\newcommand\mypara{\par\textbf{\stepcounter{para}\thepara)\space}}

\begin{document}

\section{Response to Editor}

\subsection{Comments}

\mypara Incidentally, this section imposes a logit assumption in (1), but
is this really needed before turning to the simulation exercises in Sections 3 and 4?

\mypara In the end, I think R2 set a good bar for the simulations: the
point of doing the simulation [is] to try and provide some qualitative results about the
direction/nature of the net effect of the various countervailing forces and what this depends on?
If you can convincingly deliver on this, then the paper can inform practitioners on the key
elements to be sure to precisely capture in a merger simulation exercise.

\mypara When presenting your results, please consider whether the simulations should be presented as the
result of a statistical sampling (as is done now) or more in the vein of counterfactual analysis
where you vary one parameter at a time. The argument for the statistical sampling approach is
that parameters may interact---although you haven?t clearly established that in most of the
current results. The argument for the one parameter at a time approach is that it can be more
closely tied to the institution presented in the theory section. If you are tempted to do the
sampling approach because there are many parameters to vary, consider whether you should
instead focus the reader?s attention on the most important/illustrative results.

\subsection{TO DOs}
\setcounter{para}{0}
\mypara Be very clear on how the central point is parameterized/calibrated. Be clear about what
data is used to pin down what parameters. Also note what parameters you will vary
either because they are not well identified with available data or because you want to
consider sensitivity.
\mypara Be sure to compare simulations that allow for the complex nature of Republic/Santek
versus the more vanilla simulations considered in ST (2021). This is important to capture
this papers contribution.
\mypara Be sure to consider how the nature of bargaining and downstream diversion will
impact the results. For the later, you would need to go beyond a multinational logit
model, but a nested logit should be able to capture the intuition. I'm not asking you to
estimate substitution in this context since it doesn?t seem like you have the data to do so
(although I'd love to be shown wrong on this point!). I do want to know whether varying
substitution patterns which fit the data you do have are likely to alter the qualitative
outcomes of the merger simulation.

\section{Response to Referee 1}
\setcounter{para}{0}
\mypara I hope that the authors provide the code so that researchers can (a) run their own simulations
and (b) use the code for their own research. The authors should follow the AER/ReStud
guidelines that are becoming standard in the profession
( \url{https://aeadataeditor.github.io/aea-de-guidance/preparing-for-data-deposit.html} ). Basically,
make the code easily readable and post it to a code/data archive website like Zenodo (and link
to it from the author's webpage and the paper).
Note: Footnote 16 states that the antitrust R package contains the code to do the
simulations. I could not easily find the relevant parts of the package. It would be great if this
machinery was also added to antitrust , but I would like to see a stand-alone replication
package.
\mypara I think the authors could ditch the Republic/Santek example. The authors admit that the
example is for ``illustrative" purposes. Instead of a single example, the authors could choose a
few representative mergers and look at specific fake data that mimics those cases. Eg. many
upstream and few downstream or vise versa.
\mypara There is one thing I really like about Section 4: the comparison between a naive merger
analysis (eg. downstream/upstream horizontal effects only) and the more comprehensive
analysis that includes all of the vertical effects. I would like to see this type of analysis in Section

\mypara I think that the authors overclaim a bit. In the abstract (and this sentiment is repeated in the
paper, for example on page 2) the authors: ``In our model, we find that mergers with both
horizontal and vertical characteristics typically harm consumers."
First, I think the more important contribution is to point out that a naive analysis that ignores the
vertical aspect of a merger can mis-quantify the effects of the merger (see point 3 above) or that
the EDM benefits are typically small (a point already made).
Second, mergers arise in nature (and are prosecuted) endogenously. The median pre-merger
HHI in the simulations is between 2800 and 4300 and the median Delta HHI is over 1000. Most
of the mergers analyzed would be presumed harmful based on the classic Cournot intuition or
the more recent Nocke/Whinston intuition. I think the authors should avoid statements like in
their abstract that make it seem they are drawing generic conclusions about merger
enforcement.
Smaller Comments
\mypara  Table 1: It seems a bit suspicious that the price quantiles are all integers. Maybe I am missing
something, but the domain should be some subset of the real number line, making integers very
unlikely.
\mypara  Simulation details. I don't understand why shares are drawn from a Dirichlet distribution (or
any distribution), Shares are endogenous functions of model primitives. Do you mean that you
sampled $\delta$ from a distribution?
\mypara Simulation details. It might make more sense to calibrate to a market elasticity (for example
-1) instead of the way it is done in the paper.
\mypara  It might make sense if a retailer and manufacturer had meaning in the model. One suggestion
is to have $\delta$ be comprised of two components, and manufacturer effect, and a retailer effect,
where the retailer effect is perfectly correlated across manufacturer products within a retailer,
and the manufacturer effect is perfectly correlated across retailer for the same product. So
something like  $\delta_{rw} = \delta_{r} + \delta_{w}$
\mypara  In Section 2.2 (first paragraph), referencing a particular retailer with index r is confusing
because r is the generic index for a retailer.
\mypara In Equation 4, I think there is an additional effect in the first term that Luco and Marshall
(2020) would call the Edgeworth-Salinger effect.
\mypara  The authors mention on multiple occasions that a retailer could cease to trade with a
wholesaler (eg. page 8). I don't see how this is a possibility in the model.
\mypara  I don't think ``Vertical" and ``Horizontal" need to be quoted in the title.

\section{Response to Referee 2}
I recommend that the paper be shortened in two ways. First I think that it should report
the results for the simulations of complex mergers but should omit the results for the simulations
of mergers where the number of pre-existing rivals (that do not participate in the merger)
changes. That is, I think the authors should only include simulation results for the case when
there are no pre-existing vertically integrated rivals that do not participate in the merger and
report that the nature of the results does not change substantially when pre-existing vertically
integrated rivals that do not participate in the merger are allowed to exist. In addition to
shortening the paper, this will allow the authors to provide a simpler and much less confusing
description of the set of examples they consider in their simulations. Second, I think that the
material on generalizing the results to the case of increasing marginal costs should be omitted. I
think this a somewhat separate issue that deserves separate treatment if substantial enough results
can be derived.
\section{Response to Referee 3}
\setcounter{para}{0}
\mypara I could not find a contribution that would merit publication at the Journal of Industrial Economics. The paper is very similar to Sheu and Taragin (2021). The model and simulation are essentially the same. At times, the paper reads as an appendix of the previous article.

\mypara  The paper claims to simulate the welfare effects in a new set of mergers called ``complex mergers." Complex mergers are standard vertical and horizontal mergers where a firm is vertically integrated. They have been previously studied by Ho and Lee (2017) and Crawford et al. (2018), among others. (The related literature also needs to be improved.) So, the paper uses a stylized model to simulate standard mergers.

\mypara  Some of the modeling assumptions are very restrictive. For instance, the logit demand system, simultaneous bargaining-Bertrand game, and the linear fees in the bargaining model. While these assumptions seem reasonable for a policy paper for practitioners, they are too strong for the JINDEC, as they hinder the credibility of the main exercises in the paper. The logit demand, in particular, is unnecessarily restrictive in terms of substitution patterns and could be driving the welfare results. For example, vertically and non-vertically integrated firms are clearly different from the consumers' perspective. One would expect consumers self-selecting into the firms' types, which would affect the compensating variation computation. To the very least, random coefficients on price and on whether the firm is vertically integrated are necessary to obtain more credible results.

\mypara  The RRC discussion was perplexing. What the paper calls ``RRC" differs from the classic RRC in the literature in Salop and Scheffman (1983). In the model, it arises due to the greater bargaining power of the integrated firms. But the paper's assumption that ``negotiation for a given input price treats other input prices and all downstream prices as given" (p. 6) rules out the possibility of a standard RRC effect. So the paper contradicts Crawford et al. (2018) footnote 62. Is this an error? It is paramount that this issue is clarified in the future, even at a different journal. Please include the profit functions and derivations in the Appendix (e.g. equation 5 and subsequent).

\mypara  The simulation exercise in Section 3 does not implement the simulation of the model in Section 2. What is the mapping between the two? How do you ensure market shares are consistent with equation 1 and firm pricing decisions? The paper states that: ``we calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option that sets its Bertrand equilibrium price to \$5, has 15\% market share, and has zero marginal costs. All other goods are differenced." But how are prices consistent with the demand and profit maximization computed? Specifically, I could not understand how such a procedure fits the bargaining game upstream. They seem to be inconsistent. For instance, how do you ensure downstream (upstream) prices are consistent with equation 2 (3)? For these reasons, I could not follow the details of the simulation exercise.

\mypara Finally, I found the calibration exercise in Section 4 hard to follow. At times it reads as if the paper is collecting data. But it is closer to a calibration/simulation exercise motivated by the merger. But there is no estimation and the data is minimal. The theoretical model does not fit the setting in the case. First, the bargaining model is not present nor discussed in the DOJ Lawsuit. Second, the RRC discussed in the DOJ lawsuit refers to foreclosure/exclusionary conduct concerns. This feature could typically has different welfare implications. See, e.g., Ordover et al. (1990) and the references therein. I found the analysis in this section not credible.

\end{document}

### 2024-08-01 doc/VerticalArrangements.tex
reran trashsims to accomodate changes for JIE submission.

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Numerical Simulations \label{sec:sims}}
-In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for either the merging parties or their rivals to be vertically integrated before they merge.
-This modification allows us to explore the welfare implications of complex mergers, rather than only purely horizontal or purely vertical combinations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Data Generating Process}
-Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.
-We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.
-Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.
-We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.
-Finally, an integrated merger is between two firms that are both already vertically integrated pre-merger.
-
-For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parametrize the Dirichlet distribution so it is equivalent to a uniform distribution.
-}  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers to 7.
-We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (2 to 6 for integrated mergers).
-Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner, and integrated retailers purchase inputs from wholesalers other than their integrated partner.
-
-Our simulations focus on mergers that are more likely to have anti-competitive effects and to therefore come under agency scrutiny.
-For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger.
-Similarly, when simulating a vertical merger, we assign the products sold by the largest wholesaler and the largest retailer to a single firm post-merger.
-
-The bargaining parameter ranges from from 0.1 (where wholesalers have the advantage) to 0.9 (where retailers have the advantage).
-We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$.
-The bargaining parameter is identical for all of the retailers in each simulation.
-We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option that sets its Bertrand equilibrium price to \$5, has 15\% market share, and has zero marginal costs.
-All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.
-We assume that pre-merger, upstream marginal costs are 25\% of upstream firm margins and 10\% of downstream firm margins.
-We set the market size to 1.
-
-For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.
-After eliminating situations where the merger is unprofitable to the merging firms, as well as downstream markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%.
-} Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Overview of Simulated Output}
-
-Table \ref{tab:simsum} provides summary statistics for our simulations.
-The 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with Herfindahl Hirschman Index (HHI) values over 2,500 points.
-The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.
-}  The vast majority of our mergers meet these conditions.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.
-}  The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.
-Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.
-
-Figure \ref{fig:surplussum} summarizes our results for welfare.
-Each of the four panels shows how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).
-Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.
-Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger.
-The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median.
-Negative outcome values imply harm, and positive values imply benefits.
-
-We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.
-Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.
-In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integrated mergers first-order stochastically dominates consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers.
-However, while consumer harm is typically greater under integrated mergers than downstream mergers, it is not first-order stochastically dominant.
-Median consumer harm from integrated mergers is about 14\% of pre-merger expenditures, 1.3 times the magnitude of downstream mergers, and 3 times the magnitude from upstream mergers.
-
-Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas vertical mergers, downstream mergers, and integrated mergers almost always benefit retailers, upstream mergers harm retailers in about 60\% of all simulations.
-Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, while the retailer surplus distribution from vertical mergers stochastically dominates the distribution from upstream mergers.
-The median gain to retailers from integrated mergers is about 13\% of pre-merger expenditures,  1.5 times the magnitude of downstream mergers, and 1.1 times the magnitude of that from vertical mergers.
-
-Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 64\% of all upstream mergers, 11\% of vertical mergers, and about 16\% of downstream  and integrated mergers.
-Here, wholesaler surplus under integrated, downstream and upstream mergers each stochastically dominate the surplus from vertical mergers, but not one another.
-
-As for total welfare, approximately 29\% of vertical mergers and 17\% of integrated mergers are beneficial, whereas only 9\% of upstream mergers and 2\% of downstream mergers are beneficial.
-Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream  mergers, which dominates total harm from vertical mergers.
-Median consumer harm from integrated mergers is about 9\% of pre-merger total expenditures, 1.4 times the magnitude of that from downstream mergers, 2.6 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.
-
-Throughout these simulations, we maintain the assumption that marginal costs are constant.
-However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.
-In Appendix \ref{app:mc}, we provide results that allow marginal costs to increase linearly.\footnote{We still maintain that these cost functions are not affected by the merger, which rules out merger cost efficiencies.
-}  We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Mergers and Vertically Integrated Incumbent Firms}
-
-We examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.
-Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases.
-For the results on total welfare, see Appendix \ref{app:total}.
-The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{One difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated.
-By contrast, for upstream and downstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated wholesaler or retailer, respectively, with no integrated third parties.
-For integrated mergers, the plot when the number of incumbent integrated firms equals 0 depicts a merger between two integrated firms, again with no integrated rivals.
-
-We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.
-Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW_consumer}), about half of vertical mergers benefit consumers.
-The addition of a single rival integrated firm (denoted by 1 integrated firm in the figure) moves median consumer welfare from about 0\% of pre-merger revenues to a positive 2\% and widens the distribution of outcomes slightly.
-Adding more integrated rivals does not have much additional effect.
-
-Moving to the second panel in Figure \ref{fig:CVvertincumbBW_consumer}, upstream mergers absent any incumbent vertical integration (0 integrated firms) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues.
-This result is to be expected, as upstream mergers without additional integration do not have the countervailing effects for welfare possible in the context of downstream mergers (where combined retailers may bargain for lower input prices) or vertical mergers (where EDM can create benefits).
-Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear.
-Consumer welfare increases in almost 30\% of mergers, though median consumer harm is roughly steady at approximately 4\% of pre-merger revenues.
-The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.
-Here we find that harm dominates in most simulations.
-Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and somewhat trims the whiskers while having little impact on median harm.
-
-Next we examine the welfare impacts of downstream mergers in the third panel of Figure \ref{fig:CVvertincumbBW_consumer}.
-As with upstream mergers, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues.
-\citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.
-Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, but median consumer harm grows to about 12\% of pre-merger revenues.
-Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.
-Similar to what we saw for upstream mergers, we find that the harms dominate in most instances.
-Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat moves the distribution towards less harm.
-With 3 integrated third parties (denoted by 4 integrated firms in the figure), the merger benefits consumers in about 34\% of markets and shrinks median consumer harm to about 5\% of pre-merger revenues.
-
-We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW_consumer}.
-Absent the presence of rival incumbent integrated firms (0 integrated firms in the figure), mergers between two integrated firms benefit consumers in about 6\% of simulated markets.
-Thus, we find that mergers between two firms that are already integrated are harmful to consumers in the vast majority of cases.
-Adding one additional rival incumbent integrated firm (1 integrated firm in the figure) does not have a significant impact on the median or on the range of outcomes.
-Adding more integrated rivals (2-4 integrated firms in the figure) moves median consumer welfare to be slightly more negative.
-
-To summarize, our analysis of Figure \ref{fig:CVvertincumbBW_consumer} yields some overall conclusions.
-First, we find that mergers where one or both of the merging firms are already integrated (and all rivals are unintegrated) are harmful to consumers in most instances.
-Although these mergers have the possibility of creating benefits through EDM, these gains appear to be outweighed in the majority of the mergers we study.
-Second, the presence of integrated rivals has some impact on the distribution of consumer harm, but the observed patterns are noisy and the resulting box plots are largely similar to those where such rivals are absent.
-Therefore, having such rival firms in and of itself does not generate significantly better outcomes for consumers.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Mergers and Bargaining Power}
-\citeN{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power.
-Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.
-We also show that the result persists when there are rival non-merging integrated firms.
-
-Figure \ref{fig:CVbargincumbent_consumer} depicts box and whisker plots summarizing the consumer welfare effects for vertical, upstream, downstream, and integrated mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage).
-Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.
-Results for total welfare are in Appendix \ref{app:total}.
-For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}.
-For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated.
-For upstream and downstream mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated firm when all third parties are unintegrated.
-In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.
-Finally, in integrated mergers, when the number of incumbent integrated firms equals 0, this depicts a merger between two integrated firms with all rivals are unintegrated.
-Plots with either 1 or 4 integrated firms depict mergers between integrated incumbents when there are either 1 or 4 rival integrated firms in the market, respectively.
-
-We begin by examining vertical mergers in the leftmost panel of Figure \ref{fig:CVbargincumbent_consumer}.
-We focus first on the case when there are no incumbent integrated firms (the light blue plots).
-Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
-Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.
-This channel becomes less relevant as retailers gain more power.
-The impact on consumers is zero in the neighborhood of retailer to wholesaler bargaining power ratios of 1 to 2/3.
-As we add 1 and then 4 integrated rival firms (shown by the darker and darkest blue box plots, respectively), the range of welfare outcomes narrows and shifts upwards, though we still see the negative relationship between consumer welfare and relative retailer bargaining power.
-Across all three integrated incumbent scenarios, median consumer harm moves from about -24\% of pre-merger expenditures when wholesalers have the advantage to 7\% when retailers have the advantage.
-
-We turn next to upstream mergers in the second panel of Figure \ref{fig:CVbargincumbent_consumer}.
-Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that for downstream and vertical mergers, as it is not monotonic.
-By contrast, for mergers between an integrated firm and an unintegrated upstream supplier (the darker blue plots), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.
-This is intuitive, as now all three types of mergers involve bargaining interactions that can lower input prices, the savings from which are then passed through to consumers to varying degrees.
-Median consumer harm goes from roughly -21\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power.
-Adding 3 incumbent integrated rivals (the darkest blue plots) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows and the medians shift up slightly.
-
-For downstream mergers, like with vertical mergers, we find that higher retailer relative bargaining power leads to more negative impacts on welfare.
-Absent any incumbent integration (the light blue plots), all the outcomes shown in Figure \ref{fig:CVbargincumbent_consumer} harm consumers, which mirrors our findings in Figure \ref{fig:CVvertincumbBW_consumer}.
-Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 19\% of pre-merger expenditures when retailers have relatively more bargaining power.
-In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.
-Thus, there is little to counteract harms from decreased downstream competition.
-
-Once we allow one of the merging firms to be integrated (the darker blue plots), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.
-The box and whisker plots in Figure \ref{fig:CVbargincumbent_consumer} now show a stronger relationship between bargaining power and extent of consumer harm.
-Median consumer harm goes from approximately -28\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.
-The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.
-Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here the neutrality crossing shifts left, when wholesalers have a bit more than 3/2 times the power of retailers.
-It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.
-This is intuitive, as here the downstream merger also causes an additional lessening in horizontal competition.
-Once we add 3 rival integrated firms (the darkest blue plots) the box and whisker plots shift up.
-The range of outcomes also shrinks.
-This is consistent with what we see for vertical and upstream mergers.
-Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -35\% of pre-merger revenues to 13\% of pre-merger revenues.
-
-The last panel in Figure \ref{fig:CVbargincumbent_consumer} displays results for integrated mergers.
-In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers.
-Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm.
-Median consumer harm moves from about 0\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power.
-A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see the darker and darkest blue plots) tends to narrow the range of outcomes and somewhat shift the distribution upwards.
-However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 times or more.
-The crossing at zero harm is shifted to the left.
-
-In summary, our review of Figure \ref{fig:CVbargincumbent_consumer} has revealed some systematic patterns.
-First, we see that vertical mergers, upstream and downstream mergers where one of the merging firms is integrated, and integrated mergers all exhibit a negative relationship between consumer harm and increasing relative retailer bargaining power.
-Consumers seem most likely to benefit from these mergers when wholesalers have significant bargaining power, which are situations when we would expect the gains from EDM to be largest.
-Second, the point at which this monotonic relationship between consumer welfare and bargaining power crosses zero varies across merger types, and appears to be farthest to the left in integrated mergers.
-That is, wholesalers need to have a substantial bargaining power advantage over retailers in order for consumers to benefit from an integrated merger.
-Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift this distribution upwards.
-This pattern for integrated rivals is much less noisy than in Figure \ref{fig:CVvertincumbBW_consumer}, where we did not condition on relative bargaining power.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-\caption{Summary Statistics for Numerical Simulations}
-\label{tab:simsum}
-\small{\include{sumtable}}
-\end{table}
 
@@
 \end{table}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/surplussum.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-\label{fig:surplussum}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/CVvertincumbBW_consumer.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market changes.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-\label{fig:CVvertincumbBW_consumer}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms changes.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargincumbent_consumer}
-\end{sidewaysfigure}
 
@@
 \centering
-\caption{Comparison between Downstream-Only, Upstream-Only, and Full Vertical Simulations for Republic/Santek}
-\includegraphics{output/TrashSimsCompareAll.png}\\
+\caption{Simulated Welfare Effects from Mergers between  Republic/Santek, Santek/WasteConn, Santek/WM-ADS, and WM-ADS/Regional}
+\includegraphics{output/trashinterestingmergerbar.png}\\
 \footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
-\label{fig:TrashSimsCompareAll}
+\label{fig:TrashFakeOverall}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{Simulation Results from mergers between Santek/WasteConn, Santek/WM-ADS, and WM-ADS/Regional }
+\includegraphics{output/trashinterestingfirmbar.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\label{fig:TrashFakeFirms}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{The sensitivity of the Republic/Santek simulation results to changes in the bargaining power parameter.
+`All' changes the bargaining power paramtere for all non-vertically integrated firms.
+`Santek:Republic' changes the bargaining power parameter for just the Santek/Republic product.
+`WasteConn:Republic' changes the bargaining power parameter for just the WasteConn:Republic product.
+The Dashed line displays baseline results.}
+\includegraphics{output/trashbarg.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\label{fig:TrashBarg}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{The sensitivity of the Republic/Santek simulation results to changes in the outside share.
+The absolute value of the market elasticity corresponding to the change in ouside share is in parentheses.
+Baseline results assume 0 outside share.}
+\includegraphics{output/trashelast.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\label{fig:TrashElast}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{The sensitivity of the Republic/Santek simulation results to changes in the nesting parameter.`Integrated' assumes that all vertically integrated products are in a single nest, while `Landfill` assumes that all collectors that use the same landfill are in a single nest.
+A nesting parameter of 0 displays baseline results.
+}
+\includegraphics{output/trashnest.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\label{fig:TrashNest}
 \end{figure}
@@
 \numberwithin{table}{section}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Increasing Marginal Costs\label{app:mc}}
-
-Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allows costs to increase linearly.
-Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties' products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties' products (bottom row, orange).
-
-In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs.
-Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause input prices to increase more rapidly or downstream rival retailers to reduce output more quickly.
-Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.
-
-Our simulations confirm this hypothesis.
-See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.
-Overall, harm is greater for mergers with linear costs.
-The difference is largest for vertical mergers.
-When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 55\% of simulations and total surplus in about 29\% of simulations, but only increase consumer surplus in about 36\% of simulations and total surplus in about 2\% of simulations when all firms have linear costs.
-
-The bottom panel of Figure \ref{fig:surplussumcost} compares a situation where only the merging firms have constant marginal costs to one where only the merging firms have linear marginal costs.
-We see that the distributions in these bottom panels are largely the same as those in the top panels.
-Thus, it appears that the key driver of our results is whether or not the merging parties have constant marginal costs.
-When the merging firms' marginal costs are constant, the distribution of outcomes moves towards less harm.
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/surplussum_cost.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios.
-The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue, left) or linear marginal costs (orange, right).
-The boxes in the bottom row (``Party") either assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right).
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-\label{fig:surplussumcost}
-\end{sidewaysfigure}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Appendix: Additional Figures\label{app:total}}
-
-Figures \ref{fig:CVvertincumbBW_total} and \ref{fig:CVbargincumbent_total} are the analogous results as in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVbargincumbent_consumer}, but for total welfare.
-The patterns for total welfare are largely similar to those for consumer welfare.
-We see in Figure \ref{fig:CVvertincumbBW_total} that most upstream, downstream, and integrated mergers are harmful.
-Figure \ref{fig:CVbargincumbent_total} shows that net beneficial
-mergers typically only occur when wholesalers have relatively more bargaining power.  The presence of integrated rivals can narrow the range of outcomes.
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market changes.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-\label{fig:CVvertincumbBW_total}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\centering
-\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
-\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms changes.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-\label{fig:CVbargincumbent_total}
-\end{sidewaysfigure}
-
 \end{document}
 
-\begin{figure}
-\centering
-\includegraphics{output/TrashSimsCompare.png}
-\caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
-\label{fig:TrashSimsCompare}
-\end{figure}
+% \begin{figure}
+% \centering
+% \includegraphics{output/TrashSimsCompare.png}
+% \caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
+% \label{fig:TrashSimsCompare}
+% \end{figure}
 
@@
 %The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
-
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.9]{../CVvertincumb_updownBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-% \label{fig:CVvertincumbupdownBW}
-% \end{sidewaysfigure}
-%
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.9]{../CVvertincumb_vertincBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.
-}
-% \label{fig:CVvertincumbvertincBW}
-% \end{sidewaysfigure}
-
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.9]{../CVbargupBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-% \label{fig:CVbargupBW}
-% \end{sidewaysfigure}
-%
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.9]{../CVbargdownBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-% \label{fig:CVbargdownBW}
-% \end{sidewaysfigure}
-%
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.9]{../CVbargvertBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-% \label{fig:CVbargvertBW}
-% \end{sidewaysfigure}
-
-% \begin{sidewaysfigure}
-% \centering
-% \includegraphics[scale=0.85]{CVbargbothBW.png}
-% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes.
-The different colored boxes display how outcomes change as the number of vertically  integrated firms increases.
-Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
-% \label{fig:CVbargbothBW}
-% \end{sidewaysfigure}

### 2024-08-01 code/TrashSims.R

@@ -79,19 +79,20 @@ for (v in vertFirms){
 
 }
 
-sharesNoOut <- with(simdata,collection_paired_share)
-div <- tcrossprod(1/(1-sharesNoOut),sharesNoOut)*sharesNoOut
-diag(div) <- -sharesNoOut
 
 
 
 
-minD <- function(theta,is2nd=FALSE,margins=FALSE){
+
+minD <- function(theta,is2nd=FALSE,margins=FALSE,nests,nestParm=0,relevant){
 
   alpha <- theta[1]
-  barg <- theta[-1]
+  theta <- theta[-1]
+  barg <- theta
   nprods <- nrow(simdata)
 
+
+  pricesDown <- simdata$disposal_price
   marginsUp <- simdata$disposal_dollar_margin
   marginsDown <- simdata$collection_dollar_margin
   sharesDown   <- simdata$collection_paired_share
@@ -107,9 +108,11 @@ minD <- function(theta,is2nd=FALSE,margins=FALSE){
 
 
   #b[simdata$ownerPreUp != simdata$ownerPreDown] <- barg
+  if(missing(relevant)){
   b[simdata$ownerPreUp=="Santek" & simdata$ownerPreDown!="Santek"] <- barg[1]
   b[simdata$ownerPreUp=="WasteConn" & simdata$ownerPreDown!="WasteConn"] <- barg[2]
-
+  }
+  else{ b[relevant] <- barg}
   #b[simdata$ownerPreUp!="Republic" & simdata$ownerPreDown=="Republic"] <- barg[1]
   #b[simdata$ownerPreUp!="WM_ADS" & simdata$ownerPreDown=="WM_ADS"] <- barg[2]
   #b[simdata$ownerPreUp!="Regional" & simdata$ownerPreDown=="Regional"] <- barg[3]
@@ -161,11 +164,54 @@ minD <- function(theta,is2nd=FALSE,margins=FALSE){
   else{marginsCandDown <- simdata$marginDown_bert * simdata$alpha_bert[1]/alpha}
 
 
+  if(!missing(nests)){
+    nestMat <- tcrossprod(model.matrix(~-1+nests))
+  sharesBetween <- as.vector(tapply(sharesDown, nests, sum))
+  sharesBetween <- sharesBetween[nests]
+  sharesWithin <- sharesDown/sharesBetween
+
+
+  elastDiff.down <- -alpha*tcrossprod(sharesDown)
[…]

### 2024-08-01 code/genfigures.R

@@ -537,7 +537,7 @@ pfirmsdown_all.bw <- ggplot(filter(ungroup(res.nests.logit),merger =="down") , a
 
 
 pfirmsvert_all.bw <- ggplot(filter(ungroup(res.nests.logit),merger =="vertical" & Wholesalers != "1")%>%
-                              mutate(Retailers=factor(Retailers, labels=paste0("Retailers: ", levels(Retailers)))), aes(y=Outcome_value/mktrev.pre*100,
+                              mutate(Retailers=factor(Retailers, labels=paste0("Retailers: ", levels(droplevels(Retailers))))), aes(y=Outcome_value/mktrev.pre*100,
                                                                      #avgpricedelta/mktrev.pre*100,
                                                                      x=Wholesalers,color=Cost)) +
   #geom_boxplot(outlier.alpha = 0.1) +
@@ -1361,3 +1361,4 @@ dev.off()
 png("output/pprofits_barg.png",width = 10, height = 7, units = "in", res=300)
 print(pprofits_barg)
 dev.off()
+

### 2024-08-01 code/TrashSims.R
changed output path of files to output directory.

@@ -12,7 +12,7 @@ library(ggplot2)
 ## Load Data:
 
 library(readr)
-paired_data_4_22 <- read_csv("~/Projects/bargaining_convex/trash/paired_data_4_22.csv")
+paired_data_4_22 <- read_csv("~/Projects/bargaining_convex/output/paired_data_4_22.csv")
 
 disposal_price_republic <-  96 # source: 4/26 correspondence with Peggy
 
@@ -906,8 +906,8 @@ trashinterestingfirmbar <- ggplot(data= filter(mkt_mergersweep_firm,Unintegrated
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1))
 
 ggsave(filename="./output/trashinterestingfirmbar.png",trashinterestingfirmbar,width=7,height=7)
-write_csv(mkt_mergersweep,"./trash/mkt_mergersweep.csv",na="")
-write_csv(mkt_mergersweep_firm,"./trash/mkt_mergersweep_firm.csv",na="")
+write_csv(mkt_mergersweep,"./output/mkt_mergersweep.csv",na="")
+write_csv(mkt_mergersweep_firm,"./output/mkt_mergersweep_firm.csv",na="")
 
 # mkt_interesting <- filter(mkt_mergersweep,
 #                           (Acquirer=="Republic" & Target =="Santek" & Unintegrated %in% c("Base","Up","Down","Vertical")) |
@@ -929,8 +929,8 @@ write_csv(mkt_mergersweep_firm,"./trash/mkt_mergersweep_firm.csv",na="")
 #   theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.title=element_blank(),legend.position="bottom") +geom_hline(yintercept=0,linetype="dashed")
 #
 # ggsave("./output/trashinterestingmerger.png",trashinterestingplot,height = 7,width = 7)
-# write_csv(mkt_interesting,"./trash/mkt_interesting.csv",na="")
-# write_csv(mkt_interesting_firm,"./trash/mkt_interesting_firm.csv",na="")
+# write_csv(mkt_interesting,"./output/mkt_interesting.csv",na="")
+# write_csv(mkt_interesting_firm,"./output/mkt_interesting_firm.csv",na="")
 
 
  bargsweep <- function(thissim,barg=0.5,who=c("Santek:Republic","WasteConn:Republic","All","Base")){
[…]

### 2024-08-01 doc/IncreasingMarginalCostAppendix.tex
moved numerical simulations and marginal cost appendices to separate files.

\documentclass[12pt]{article}
%\documentclass[12pt,review,authoryear]{elsarticle}
%\usepackage{ecrc}       % needed for elsarticle
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs
%packages for kable tables
 \usepackage{booktabs}
 \usepackage{longtable}
 \usepackage{array}
 \usepackage{multirow}
 \usepackage{wrapfig}
 \usepackage{float}
 \usepackage{colortbl}
 \usepackage{pdflscape}
 \usepackage{tabu}
 \usepackage{threeparttable}
 \usepackage{threeparttablex}
 \usepackage[normalem]{ulem}
 \usepackage{makecell}
 \usepackage{xcolor}

\graphicspath{{../output/}}

\interfootnotelinepenalty=10000

\parskip     2.0mm       % space between paragraphs

\alph{footnote}         % make title footnotes alpha-numeric

\captionsetup[figure]{labelsep=space,labelfont=bf} % remove colon from figure name

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix: Increasing Marginal Costs\label{app:mc}}

Here, we explore the role that the constant marginal cost assumption plays in driving our results by comparing our simulations to an alternative set that allows costs to increase linearly. Figure \ref{fig:surplussumcost} displays box and whisker plots for the net effect on consumer and total welfare under four different scenarios: constant marginal costs for all firms (top row, blue), linear marginal costs for all firms (top row, orange), constant marginal costs for the merging parties' products but linear costs for the non-merging parties' products (bottom row, blue), and linear marginal costs for the merging parties' products but constant costs for the non-merging parties' products (bottom row, orange).

In theory, having linear marginal costs rather than constant costs could result in fewer gains from EDM, as efforts to increase sales due to a wholesale price reduction would be counteracted by rising costs. Likewise, in theory, RRC could be more profitable for the merging parties under linear compared to constant marginal costs, as linear marginal costs could cause input prices to increase more rapidly or downstream rival retailers to reduce output more quickly. Because impacts from EDM are plausibly less under linear marginal costs while those from RRC are plausibly more, one might expect that on net mergers are more harmful under linear rather than constant marginal costs.

Our simulations confirm this hypothesis.  See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.  Overall, harm is greater for mergers with linear costs. The difference is largest for vertical mergers.  When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 55\% of simulations and total surplus in about 29\% of simulations, but only increase consumer surplus in about 36\% of simulations and total surplus in about 2\% of simulations when all firms have linear costs.

The bottom panel of Figure \ref{fig:surplussumcost} compares a situation where only the merging firms have constant marginal costs to one where only the merging firms have linear marginal costs.  We see that the distributions in these bottom panels are largely the same as those in the top panels.  Thus, it appears that the key driver of our results is whether or not the merging parties have constant marginal costs.  When the merging firms' marginal costs are constant, the distribution of outcomes moves towards less harm.

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{output/surplussum_cost.png}
\caption{The figure displays box and whisker plots summarizing the extent to which merger outcomes change according to 4 different cost scenarios. The boxes in the top row (``All") either assume all firms face either constant marginal costs (blue, left) or linear marginal costs (orange, right). The boxes in the bottom row (``Party") either assume that the merging parties face constant marginal costs while other firms face linear marginal costs (blue, left), or the merging parties face linear marginal costs while other firms face constant marginal costs (orange, right). Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:surplussumcost}
\end{sidewaysfigure}

\end{document}

### 2024-08-01 doc/NumericalSimulationsAppendix.tex
moved numerical simulations and marginal cost appendices to separate files.

\documentclass[12pt]{article}
%\documentclass[12pt,review,authoryear]{elsarticle}
%\usepackage{ecrc}       % needed for elsarticle
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs
%packages for kable tables
 \usepackage{booktabs}
 \usepackage{longtable}
 \usepackage{array}
 \usepackage{multirow}
 \usepackage{wrapfig}
 \usepackage{float}
 \usepackage{colortbl}
 \usepackage{pdflscape}
 \usepackage{tabu}
 \usepackage{threeparttable}
 \usepackage{threeparttablex}
 \usepackage[normalem]{ulem}
 \usepackage{makecell}
 \usepackage{xcolor}

\graphicspath{{../output/}}

\interfootnotelinepenalty=10000

\parskip     2.0mm       % space between paragraphs

\alph{footnote}         % make title footnotes alpha-numeric

\captionsetup[figure]{labelsep=space,labelfont=bf} % remove colon from figure name

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations \label{sec:sims}}
In this section, we use a similar simulation setup to that in \citeN{ST2021}, but allow for either the merging parties or their rivals to be vertically integrated before they merge. This modification allows us to explore the welfare implications of complex mergers, rather than only purely horizontal or purely vertical combinations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, an integrated merger is between two firms that are both already vertically integrated pre-merger.

For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parametrize the Dirichlet distribution so it is equivalent to a uniform distribution.  }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers to 7. We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (2 to 6 for integrated mergers). Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner, and integrated retailers purchase inputs from wholesalers other than their integrated partner.

Our simulations focus on mergers that are more likely to have anti-competitive effects and to therefore come under agency scrutiny. For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger. Similarly, when simulating a vertical merger, we assign the products sold by the largest wholesaler and the largest retailer to a single firm post-merger.

The bargaining parameter ranges from from 0.1 (where wholesalers have the advantage) to 0.9 (where retailers have the advantage). We report our results in terms of relative bargaining power, $(1-\lambda)/\lambda$.  The bargaining parameter is identical for all of the retailers in each simulation.  We calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option that sets its Bertrand equilibrium price to \$5, has 15\% market share, and has zero marginal costs.  All other goods are differenced relative to this option, resulting in the mean zero outside good normalization.  We assume that pre-merger, upstream marginal costs are 25\% of upstream firm margins and 10\% of downstream firm margins. We set the market size to 1.

For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.  After eliminating situations where the merger is unprofitable to the merging firms, as well as downstream markets that do not pass the Hypothetical Monopolist Test, we have 1.6 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%. } Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Overview of Simulated Output}

Table \ref{tab:simsum} provides summary statistics for our simulations. The 2010 DOJ/FTC Horizontal Merger Guidelines categorize ``Highly Concentrated Markets'' as those with Herfindahl Hirschman Index (HHI) values over 2,500 points.  The Guidelines state that mergers with HHI changes greater than 200 points that result in Highly Concentrated markets are ``presumed likely to enhance market power.''\footnote{See \S 5.3 of the Guidelines.  }  The vast majority of our mergers meet these conditions.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  The median average wholesale pre-merger price is \$5.80, and the median average retail pre-merger price is \$15.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.

Figure \ref{fig:surplussum} summarizes our results for welfare.  Each of the four panels shows how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).  Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.  Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger. The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Negative outcome values imply harm, and positive values imply benefits.

We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.  Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.  In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from downstream and integrated mergers first-order stochastically dominates consumer harm from upstream mergers which in turn dominates consumer harm from vertical mergers. However, while consumer harm is typically greater under integrated mergers than downstream mergers, it is not first-order stochastically dominant. Median consumer harm from integrated mergers is about 14\% of pre-merger expenditures, 1.3 times the magnitude of downstream mergers, and 3 times the magnitude from upstream mergers.

Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas vertical mergers, downstream mergers, and integrated mergers almost always benefit retailers, upstream mergers harm retailers in about 60\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, while the retailer surplus distribution from vertical mergers stochastically dominates the distribution from upstream mergers. The median gain to retailers from integrated mergers is about 13\% of pre-merger expenditures,  1.5 times the magnitude of downstream mergers, and 1.1 times the magnitude of that from vertical mergers.

Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 64\% of all upstream mergers, 11\% of vertical mergers, and about 16\% of downstream  and integrated mergers. Here, wholesaler surplus under integrated, downstream and upstream mergers each stochastically dominate the surplus from vertical mergers, but not one another.

As for total welfare, approximately 29\% of vertical mergers and 17\% of integrated mergers are beneficial, whereas only 9\% of upstream mergers and 2\% of downstream mergers are beneficial. Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream  mergers, which dominates total harm from vertical mergers. Median consumer harm from integrated mergers is about 9\% of pre-merger total expenditures, 1.4 times the magnitude of that from downstream mergers, 2.6 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.

Throughout these simulations, we maintain the assumption that marginal costs are constant.  However, given that many of the mergers we study have vertical aspects where input cost considerations may drive the realization of EDM and RRC, it is reasonable to wonder how our results would change in an alternative marginal cost environment.  In Appendix \ref{app:mc}, we provide results that allow marginal costs to increase linearly.\footnote{We still maintain that these cost functions are not affected by the merger, which rules out merger cost efficiencies.  }  We find that this alternative assumption tends to result in mergers that are more harmful for consumers and total welfare.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers and Vertically Integrated Incumbent Firms}

We examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW_consumer} depicts box and whisker plots summarizing consumer welfare as the number of incumbent integrated firms increases. For the results on total welfare, see Appendix \ref{app:total}.  The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{One difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW_consumer} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plot when the number of incumbent integrated firms equals 1 summarizes market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for upstream and downstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated wholesaler or retailer, respectively, with no integrated third parties. For integrated mergers, the plot when the number of incumbent integrated firms equals 0 depicts a merger between two integrated firms, again with no integrated rivals.

We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.  Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW_consumer}), about half of vertical mergers benefit consumers.  The addition of a single rival integrated firm (denoted by 1 integrated firm in the figure) moves median consumer welfare from about 0\% of pre-merger revenues to a positive 2\% and widens the distribution of outcomes slightly. Adding more integrated rivals does not have much additional effect.

Moving to the second panel in Figure \ref{fig:CVvertincumbBW_consumer}, upstream mergers absent any incumbent vertical integration (0 integrated firms) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues. This result is to be expected, as upstream mergers without additional integration do not have the countervailing effects for welfare possible in the context of downstream mergers (where combined retailers may bargain for lower input prices) or vertical mergers (where EDM can create benefits).  Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear.  Consumer welfare increases in almost 30\% of mergers, though median consumer harm is roughly steady at approximately 4\% of pre-merger revenues.  The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.  Here we find that harm dominates in most simulations.  Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and somewhat trims the whiskers while having little impact on median harm.

Next we examine the welfare impacts of downstream mergers in the third panel of Figure \ref{fig:CVvertincumbBW_consumer}. As with upstream mergers, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues. \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.  Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 20\% of mergers, but median consumer harm grows to about 12\% of pre-merger revenues. Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.  Similar to what we saw for upstream mergers, we find that the harms dominate in most instances.  Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat moves the distribution towards less harm.  With 3 integrated third parties (denoted by 4 integrated firms in the figure), the merger benefits consumers in about 34\% of markets and shrinks median consumer harm to about 5\% of pre-merger revenues.

We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW_consumer}.  Absent the presence of rival incumbent integrated firms (0 integrated firms in the figure), mergers between two integrated firms benefit consumers in about 6\% of simulated markets. Thus, we find that mergers between two firms that are already integrated are harmful to consumers in the vast majority of cases.  Adding one additional rival incumbent integrated firm (1 integrated firm in the figure) does not have a significant impact on the median or on the range of outcomes. Adding more integrated rivals (2-4 integrated firms in the figure) moves median consumer welfare to be slightly more negative.

To summarize, our analysis of Figure \ref{fig:CVvertincumbBW_consumer} yields some overall conclusions.  First, we find that mergers where one or both of the merging firms are already integrated (and all rivals are unintegrated) are harmful to consumers in most instances.  Although these mergers have the possibility of creating benefits through EDM, these gains appear to be outweighed in the majority of the mergers we study.  Second, the presence of integrated rivals has some impact on the distribution of consumer harm, but the observed patterns are noisy and the resulting box plots are largely similar to those where such rivals are absent.  Therefore, having such rival firms in and of itself does not generate significantly better outcomes for consumers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers and Bargaining Power}
\citeN{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power. Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.  We also show that the result persists when there are rival non-merging integrated firms.

Figure \ref{fig:CVbargincumbent_consumer} depicts box and whisker plots summarizing the consumer welfare effects for vertical, upstream, downstream, and integrated mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage). Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.  Results for total welfare are in Appendix \ref{app:total}.  For vertical, upstream, and downstream mergers, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to results in \citeN{ST2021}. For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated. For upstream and downstream mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated firm when all third parties are unintegrated. In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.  Finally, in integrated mergers, when the number of incumbent integrated firms equals 0, this depicts a merger between two integrated firms with all rivals are unintegrated. Plots with either 1 or 4 integrated firms depict mergers between integrated incumbents when there are either 1 or 4 rival integrated firms in the market, respectively.

We begin by examining vertical mergers in the leftmost panel of Figure \ref{fig:CVbargincumbent_consumer}.  We focus first on the case when there are no incumbent integrated firms (the light blue plots).  Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.  Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.  This channel becomes less relevant as retailers gain more power.  The impact on consumers is zero in the neighborhood of retailer to wholesaler bargaining power ratios of 1 to 2/3.  As we add 1 and then 4 integrated rival firms (shown by the darker and darkest blue box plots, respectively), the range of welfare outcomes narrows and shifts upwards, though we still see the negative relationship between consumer welfare and relative retailer bargaining power.  Across all three integrated incumbent scenarios, median consumer harm moves from about -24\% of pre-merger expenditures when wholesalers have the advantage to 7\% when retailers have the advantage.

We turn next to upstream mergers in the second panel of Figure \ref{fig:CVbargincumbent_consumer}.  Without incumbent integrated firms in the market (the light blue plots), the relationship between harm and bargaining power is unlike that for downstream and vertical mergers, as it is not monotonic.  By contrast, for mergers between an integrated firm and an unintegrated upstream supplier (the darker blue plots), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.  This is intuitive, as now all three types of mergers involve bargaining interactions that can lower input prices, the savings from which are then passed through to consumers to varying degrees. Median consumer harm goes from roughly -21\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power. Adding 3 incumbent integrated rivals (the darkest blue plots) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows and the medians shift up slightly.

For downstream mergers, like with vertical mergers, we find that higher retailer relative bargaining power leads to more negative impacts on welfare.  Absent any incumbent integration (the light blue plots), all the outcomes shown in Figure \ref{fig:CVbargincumbent_consumer} harm consumers, which mirrors our findings in Figure \ref{fig:CVvertincumbBW_consumer}.  Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 19\% of pre-merger expenditures when retailers have relatively more bargaining power. In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.  Thus, there is little to counteract harms from decreased downstream competition.

Once we allow one of the merging firms to be integrated (the darker blue plots), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.  The box and whisker plots in Figure \ref{fig:CVbargincumbent_consumer} now show a stronger relationship between bargaining power and extent of consumer harm.  Median consumer harm goes from approximately -28\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.  The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.  Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here the neutrality crossing shifts left, when wholesalers have a bit more than 3/2 times the power of retailers.  It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.  This is intuitive, as here the downstream merger also causes an additional lessening in horizontal competition.  Once we add 3 rival integrated firms (the darkest blue plots) the box and whisker plots shift up. The range of outcomes also shrinks.  This is consistent with what we see for vertical and upstream mergers.  Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -35\% of pre-merger revenues to 13\% of pre-merger revenues.

The last panel in Figure \ref{fig:CVbargincumbent_consumer} displays results for integrated mergers.  In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers. Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm. Median consumer harm moves from about 0\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power. A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see the darker and darkest blue plots) tends to narrow the range of outcomes and somewhat shift the distribution upwards. However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 times or more. The crossing at zero harm is shifted to the left.

In summary, our review of Figure \ref{fig:CVbargincumbent_consumer} has revealed some systematic patterns.  First, we see that vertical mergers, upstream and downstream mergers where one of the merging firms is integrated, and integrated mergers all exhibit a negative relationship between consumer harm and increasing relative retailer bargaining power.  Consumers seem most likely to benefit from these mergers when wholesalers have significant bargaining power, which are situations when we would expect the gains from EDM to be largest.  Second, the point at which this monotonic relationship between consumer welfare and bargaining power crosses zero varies across merger types, and appears to be farthest to the left in integrated mergers.  That is, wholesalers need to have a substantial bargaining power advantage over retailers in order for consumers to benefit from an integrated merger.  Third, the presence of integrated rivals tends to shrink the range of welfare outcomes and to shift this distribution upwards.  This pattern for integrated rivals is much less noisy than in Figure \ref{fig:CVvertincumbBW_consumer}, where we did not condition on relative bargaining power.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliographystyle{chicago}
%\bibliographystyle{elsarticle-harv}
\bibliography{master-bib}
\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Tables and Figures
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{Summary Statistics for Numerical Simulations}
\label{tab:simsum}
\small{\include{sumtable}}
\end{table}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{output/surplussum.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer, retailer, wholesaler, and total surplus. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:surplussum}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{output/CVvertincumbBW_consumer.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer surplus as the number of vertically integrated firms present in a market changes.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:CVvertincumbBW_consumer}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{output/CVbargincumbent_consumer.png}
\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect consumer surplus as the bargaining power of wholesalers relative to retailers changes. The different colored boxes display how outcomes change as the number of vertically  integrated firms changes. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
\label{fig:CVbargincumbent_consumer}
\end{sidewaysfigure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\newpage
\clearpage

\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Appendix: Additional Figures\label{app:total}}

Figures \ref{fig:CVvertincumbBW_total} and \ref{fig:CVbargincumbent_total} are the analogous results as in Figures \ref{fig:CVvertincumbBW_consumer} and \ref{fig:CVbargincumbent_consumer}, but for total welfare. The patterns for total welfare are largely similar to those for consumer welfare.  We see in Figure \ref{fig:CVvertincumbBW_total} that most upstream, downstream, and integrated mergers are harmful.  Figure \ref{fig:CVbargincumbent_total} shows that net beneficial
mergers typically only occur when wholesalers have relatively more bargaining power.  The presence of integrated rivals can narrow the range of outcomes.

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{output/CVvertincumbBW_total.png}
\caption{The figure displays box and whisker plots summarizing the extent to which mergers affect total surplus as the number of vertically integrated firms present in a market changes.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
\label{fig:CVvertincumbBW_total}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\includegraphics[scale=0.9]{output/CVbargincumbent_total.png}
\caption{The figure displays box and whisker plots summarizing the extent to which vertical, upstream, downstream and integrated mergers affect total surplus as the bargaining power of wholesalers relative to retailers changes. The different colored boxes display how outcomes change as the number of vertically  integrated firms changes. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
\label{fig:CVbargincumbent_total}
\end{sidewaysfigure}

\end{document}
%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
%For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets. In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements. The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
%The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.

% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../CVvertincumb_updownBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
% \label{fig:CVvertincumbupdownBW}
% \end{sidewaysfigure}
%
% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../CVvertincumb_vertincBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers affect consumer (blue,left) and total (orange,right) surplus as the number of vertically integrated firms present in a market change.  Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median. }
% \label{fig:CVvertincumbvertincBW}
% \end{sidewaysfigure}

% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../CVbargupBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated wholesaler affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. The different colored boxes display how outcomes change as the number of vertically  integrated firms increases. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
% \label{fig:CVbargupBW}
% \end{sidewaysfigure}
%
% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../CVbargdownBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an integrated and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. The different colored boxes display how outcomes change as the number of vertically  integrated firms increases. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
% \label{fig:CVbargdownBW}
% \end{sidewaysfigure}
%
% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.9]{../CVbargvertBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among an unintegrated wholesaler and unintegrated retailer affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. The different colored boxes display how outcomes change as the number of vertically  integrated firms increases. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
% \label{fig:CVbargvertBW}
% \end{sidewaysfigure}

% \begin{sidewaysfigure}
% \centering
% \includegraphics[scale=0.85]{CVbargbothBW.png}
% \caption{The figure displays box and whisker plots summarizing the extent to which mergers among two integrated wholesalers and retailers affect consumer, retailer, wholesaler, and total surplus as the bargaining power of wholesalers relative to retailers changes. The different colored boxes display how outcomes change as the number of vertically  integrated firms increases. Whiskers depict the $5^{th}$ and $95^{th}$ percentiles of a particular outcome, boxes depict the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line depicts the median.}
% \label{fig:CVbargbothBW}
% \end{sidewaysfigure}

### 2024-08-01 trash/landfill/data_clean.Rmd
added code written by Romeo Ignacio to geocode landfills.

---
title: "Landfill Geographic Analysis"
output: html_document
date: "2024-01-11"
author: Charles Taragin and Romeo Ignacio
---

```{r setup, include=FALSE}
library(tidycensus)
library(tidyverse)
library(sf)
library(dplyr)
library(readxl)
library(knitr)
library(kableExtra)
library(vtable)
library(stringr)

knitr::opts_chunk$set(echo = F, message = F)
```

### Summary Statistics

```{r map_data, results = 'hide'}
### Step 1: Read in and graph data
csa_shp <- get_acs(
  geography = "combined statistical area",
  variables = "B19013_001",
  geometry = TRUE,
  survey = "acs1",
  year = 2019
) %>%
  select(GEOID,NAME,geometry)

cbsa_shp <- get_acs(
  geography = "cbsa",
  variables = "B19013_001",
  geometry = TRUE,
  survey = "acs1",
  year = 2019
) %>%
  select(GEOID,NAME,geometry)

# Deleted option: cbsa

county_shp <- get_acs(
  geography = "county",
  variables = "B19013_001",
  geometry = TRUE,
  survey = "acs1",
  year = 2019
) %>%
  select(GEOID,NAME,geometry)

raw <- read_xlsx("/fsr/project4/Charles/bargaining_convex/trash/Copy of WBJ_Disposal_Sites_2019.xlsx",skip=3)

## Clean up raw file names

raw_uniq_own_nms <- raw %>% 
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Owner Code`,`Owner Entity`,Type,Family) %>%
  unique() %>%
  mutate(`Owner Full Name`= ifelse(is.na(`Owner Code`),"",`Owner Entity`),
    `Owner Code`=ifelse(is.na(`Owner Code`),`Owner Entity`,`Owner Code`)) %>%
  group_by(`Owner Code`)  %>%
  select(`Owner Code`) %>%
  arrange(`Owner Code`) %>%
  unique()

write.csv(raw_uniq_own_nms,"/fsr/project4/Charles/bargaining_convex/trash/owner_nms_raw.csv")

raw_uniq_op_nms <- raw %>% 
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Operator Code`,`Operator Entity`,Type,Family) %>%
  unique() %>%
  mutate(
    `Operator Code`=ifelse(is.na(`Operator Code`),`Operator Entity`,`Operator Code`)) %>%
  group_by(`Operator Code`)  %>%
  select(`Operator Code`) %>%
  arrange(`Operator Code`) %>%
  unique()

write.csv(raw_uniq_op_nms,"/fsr/project4/Charles/bargaining_convex/trash/operator_nms_raw.csv")

# Clean for owner
raw_uniq_own_nms <- raw %>% 
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Owner Code`,`Owner Entity`,Type,Family) %>%
  unique() %>%
  mutate(`Owner Code`=ifelse(is.na(`Owner Code`),`Owner Entity`,`Owner Code`)) %>%
  group_by(`Owner Code`) %>%
  select(`Owner Code`) %>%
  mutate(`Owner Code`=str_replace_all(`Owner Code`,"Tidewater Fiber Corp","Tidewater Fibre Corp"),
         `Owner Code`=str_replace_all(`Owner Code`,"Town of Bethleham","Town of Bethlehem"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("ABC Disposal Systems"),"ABC"),
         `Owner Code`=toupper(`Owner Code`),
         `Owner Code`=str_replace_all(`Owner Code`,"\\.",""),
         `Owner Code`=str_replace_all(`Owner Code`,"\\,",""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" inc\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corp\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" co\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" company\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corporation\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" LLC\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" & "),"&"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("LP\\b"),""),
         `Owner Code`=str_squish(`Owner Code`)) %>%
  unique()

  #Clean up certain strings

raw <- raw %>%
  mutate(`Operator Code`=ifelse(is.na(`Operator Code`),`Operator Entity`,`Operator Code`)) %>%
  mutate(`Operator Code`=str_replace_all(`Operator Code`,"Tidewater Fiber Corp","Tidewater Fibre Corp"),
         `Operator Code`=str_replace_all(`Operator Code`,"Angelo's Recycled Materials - Tampa Recycling Facility","Angelo's Recycled Materials"),
         `Operator Code`=str_replace_all(`Operator Code`,regex("ABC Disposal Systems"),"ABC"),
         `Operator Code`=str_replace_all(`Operator Code`,regex("Athens Services"),"ATHEN"),
         `Operator Code`=str_replace(`Operator Code`,"Monadnock Disposal Service Inc.",	
"Monadnock Disposal Services"),
         `Operator Code`=str_replace(`Operator Code`,"Southern Disposal Services","Southern Disposal Service"),
         `Operator Code`=str_replace(`Operator Code`,"Vogel Disposal Service","Vogel Disposal"),
         `Operator Code`=str_replace(`Operator Code`,"Strategic Materials","Strategic Material"),
         `Operator Code`=toupper(`Operator Code`),
         `Operator Code`=str_replace_all(`Operator Code`,"\\.",""),
         `Operator Code`=str_replace_all(`Operator Code`,"\\,",""),
         `Operator Code`=str_replace_all(`Operator Code`,"-"," "),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" inc\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" corp\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" co\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" company\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" corporation\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" LLC\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" & "),"&"),
         `Operator Code`=str_replace_all(`Operator Code`,regex("LP\\b"),""),
         `Operator Code`=str_squish(`Operator Code`))

raw <- raw %>%
  mutate(`Owner Code`=ifelse(is.na(`Owner Code`),`Owner Entity`,`Owner Code`)) %>%
  mutate(`Owner Code`=str_replace_all(`Owner Code`,"Tidewater Fiber Corp","Tidewater Fibre Corp"),
         `Owner Code`=str_replace_all(`Owner Code`,"Town of Bethleham","Town of Bethlehem"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("ABC Disposal Systems"),"ABC"),
         `Owner Code`=toupper(`Owner Code`),
         `Owner Code`=str_replace_all(`Owner Code`,"\\.",""),
         `Owner Code`=str_replace_all(`Owner Code`,"\\,",""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" inc\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corp\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" co\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" company\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corporation\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" LLC\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" & "),"&"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("LP\\b"),""),
         `Owner Code`=str_squish(`Owner Code`))

#Bestway Disposal (Different owner???)
#Monadnock Disposal Services (Diferent owner???)
#Texas Disposal Systems vs Texas Disposal Systems Inc.
operator_names <- raw %>%
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Operator Entity`) %>%
  unique()
#Willimantic Waste Paper Co For operator (Different operator???)
#Bestway Disposal (Different operator???)

disposal_df <- raw %>%
  select(Code,Location,City,County,State,Zip,Longitude,Latitude,Code,Type,Family,`Waste Shed`,`Owner Code`,`Owner Entity`,`Operator Code`,`Operator Entity`,`Total of All Wastes (tons/day)`) %>%
  rename(Waste_Shed = `Waste Shed`,
         Owner_Code = `Owner Code`,
         Owner_Entity = `Owner Entity`,
         Operator_Code = `Operator Code`,
         Operator_Entity=`Operator Entity`) #%>%
  # mutate(Owner_Code=Owner_Entity,
  #        Operator_Code=Operator_Entity)

## Determine level of aggregation

```

```{r}
### Step 2: Merge in shape files
## Merge all disposal data
disposal_df_shp <- st_as_sf(disposal_df,coords=c("Longitude","Latitude"), crs=4326) %>% #crs=4269)
  st_transform(crs=4269)
############### CSA ###############
intersect_df <- st_join(disposal_df_shp,csa_shp,left=T,join=st_intersects) %>%
    rename(cbsa=NAME) %>%
    mutate(st_cnty=paste(County,State,sep = ":"),
    market=ifelse(is.na(cbsa),st_cnty,cbsa),
    iscsa=!is.na(cbsa)) %>%
    st_drop_geometry()
############### CBSA ###############
cbsa_intersect_df <- st_join(disposal_df_shp,cbsa_shp,left=T,join=st_intersects) %>%
    rename(cbsa=NAME) %>%
    mutate(st_cnty=paste(County,State,sep = ":"),
    market=ifelse(is.na(cbsa),st_cnty,cbsa),
    iscbsa=!is.na(cbsa)) %>%
    st_drop_geometry()

############### County ###############
county_intersect_df <- st_join(disposal_df_shp,county_shp) %>%
  st_drop_geometry() %>%
  rename(county_shp=NAME) %>%
  mutate(county_same = str_detect(county_shp,County)) %>%
  relocate(county_shp,County,county_same)

# prop.table(table(county_intersect_df$county_same==1))

## Create crosswalk
crosswalk_disposal_df <- intersect_df %>%
  select(Code,market)

write.csv(crosswalk_disposal_df,"/fsr/project4/Charles/bargaining_convex/trash/cbsa_disposal_cross.csv",row.names = F)
```

```{r, results ="hide"}
## Clean up data

intersect_df %>% filter(Type == "LF" & market == "Chattanooga, TN-GA Metro Area")

# Inquire about wasteshed and what to do if NA???
landfill_csa_df <- intersect_df %>%
  st_drop_geometry() %>%
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>% # Step 1: Keep only LF basde on either Type or Family
  filter(Waste_Shed!="Exclusive or Captive"|is.na(Waste_Shed)) %>% # Step 2: Drop all observations labelled "Exclusive or Captive" (1619 Observations)
  mutate(firm_name="") %>%
  select(Owner_Code,Owner_Entity,Operator_Code,Operator_Entity,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa)

landfill_cbsa_df <- cbsa_intersect_df %>%
  st_drop_geometry() %>%
  filter(Type=="LF"|Family=="LF") %>% # Step 1: Keep only LF basde on either Type or Family
  filter(Waste_Shed!="Exclusive or Captive"|is.na(Waste_Shed)) %>% # Step 2: Drop all observations labelled "Exclusive or Captive" (1619 Observations)
  select(Owner_Code,Owner_Entity,Operator_Code,Operator_Entity,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscbsa)

```

This is on the Owner Code-Operator Code level or Owner Entity-Operator Entity level if code does not exist.  Let's first look at number of markets that are in a CSA and number of markets that are county based as they do not fall in a CSA.  We see that only 11% of markets are CSA market based. 
```{r}
### Get number of landsfill sights that don't match
sum_landfill_all <- landfill_csa_df %>%
  mutate(coor_exis=ifelse(market!=st_cnty,"CSA","County")) %>%
  st_drop_geometry() #%>%
  # select(market,coor_exis) %>% # New line, can delete to get old calculation
  # unique()  # New line, can delete to get old calculation

sum_landfill <- sum_landfill_all  %>%
  select(coor_exis) %>%
  group_by(coor_exis) %>%
  summarise(count=n()) %>%
  ungroup() %>%
  mutate(per_count=100*prop.table(count))

#market!=st_cnty
sum_landfill %>%
  kbl(col.names=c("Type","Count","Percent"),
      digits=0,
      caption = "# Landfills Market Types") %>%
  kable_classic_2(full_width = F)

```
Now Let's decompose markets that have 1 landfill and > 1 landfill.  We see about 59% of markets have only one landfill.

```{r}
## Create summary statistics (On MSA-Owner-Operator Lvl)
landfill_sum_csa <- landfill_csa_df %>%
  ungroup() %>%
  group_by(market,Owner_Code,Operator_Code) %>%
  summarise(count=n(),
            total_waste=sum(`Total of All Wastes (tons/day)`), .groups="keep")

landfill_sum_cbsa <- landfill_cbsa_df %>%
  ungroup() %>%
  group_by(market,Owner_Code,Operator_Code) %>%
  summarise(count=n(),
            total_waste=sum(`Total of All Wastes (tons/day)`), .groups="keep")

write.csv(landfill_sum_csa,"/fsr/project4/Charles/bargaining_convex/trash/landfill_final.csv",row.names = F)

# On CSA Level
landfill_mkt_cnt_csa <- landfill_sum_csa %>%
  ungroup() %>%
  group_by(market) %>%
  summarise(count=n())
# On CBSA Level
landfill_mkt_cnt_cbsa <- landfill_sum_cbsa %>%
  ungroup() %>%
  group_by(market) %>%
  summarise(count=n())

landfill_prop_csa <- landfill_mkt_cnt_csa %>%
  ungroup %>%
  mutate(several_markets=ifelse(count>1,"Several Landfills","One Landfill"))

landfill_prop_cbsa <- landfill_mkt_cnt_cbsa %>%
  ungroup %>%
  mutate(several_markets=ifelse(count>1,"Several Landfills","One Landfill"))

# Get summary statistics of landfills
sum_landfill <- landfill_csa_df %>%
  mutate(coor_exis=ifelse(market!=st_cnty,"CSA","County")) %>%
  st_drop_geometry() %>%
  select(coor_exis) %>%
  group_by(coor_exis) %>%
  summarise(count=n()) %>%
  ungroup() %>%
  mutate(per_count=100*prop.table(count))

landfill_prop_csa %>%
  group_by(several_markets) %>%
  summarise(count=n()) %>%
  mutate(per_count=100*prop.table(count)) %>% 
  kbl(col.names=c("Type","Count","Percent"),
      digits=0,
      format.args = list(big.mark = ","),
      caption="Markets with as least one Landfill") %>%
  kable_classic_2(full_width = F)
# summary(landfill_sum$count)
# quantile(landfill_sum$count,seq(0,1,0.05))

```

```{r}
## Create HHI Level data
### On CSA Level ###
landfill_shrs_csa <- landfill_csa_df %>%
  ungroup() %>%
  group_by(Owner_Code,Operator_Code,market,iscsa) %>% # Aggregate on Market level, Owner Entity, Operator level
  summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
  ungroup() %>%
  group_by(market) %>% # Aggregate on Market
  mutate(total_waste = sum(waste_own),
         waste_shr = 100*(waste_own/total_waste),
         waste_hhi = sum(waste_shr^2))
### On CBSA Level ###
landfill_shrs_cbsa <- landfill_cbsa_df %>%
  ungroup() %>%
  group_by(Owner_Code,Operator_Code,market,iscbsa) %>% # Aggregate on Market level, Owner Entity, Operator level
  summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
  ungroup() %>%
  group_by(market) %>% # Aggregate on Market
  mutate(total_waste = sum(waste_own),
         waste_shr = 100*(waste_own/total_waste),
         waste_hhi = sum(waste_shr^2)) ## Have NA remove option

chat <- filter(intersect_df,grepl("Chattanooga",market))

write.csv(chat,"/fsr/project4/Charles/bargaining_convex/trash/chat.csv",row.names = F)

### CSA ### (921 markets)
landfill_hhi_csa_df  <- landfill_shrs_csa %>% 
  select(market,waste_hhi,iscsa) %>% # Keep market level information and filter out duplicates
  unique()

write.csv(landfill_hhi_csa_df,"/fsr/project4/Charles/bargaining_convex/trash/landfill_hhi_final.csv",row.names = F)

### Export Chatanooga CSA ###
chat_shrs <- landfill_shrs_csa %>%
  filter(market=="Chattanooga-Cleveland-Dalton, TN-GA CSA")

write.csv(chat_shrs,"/fsr/project4/Charles/bargaining_convex/trash/chat_share.csv",row.names=F)

### CBSA ### (1088 markets)
landfill_hhi_cbsa_df <- landfill_shrs_cbsa %>% 
  select(market,waste_hhi,iscbsa) %>% # Keep market level information and filter out duplicates
  unique()

```

Now we will look at summary stats of HHI landfill data on the CSA level.  We have `r nrow(landfill_hhi_csa_df)` total markets on the CSA level in the dataset but only `r sum(!is.na(landfill_hhi_csa_df$waste_hhi))` analyzed.  This can be explained as `r sum(is.na(landfill_hhi_csa_df$waste_hhi))` markets have no waste produced, which produces NAs when calculating for HHI.

```{r message=FALSE, warning=F}
graph_data <- landfill_hhi_csa_df %>% select(market,waste_hhi)
st(graph_data,out="kable",labels=c("Waste HHI"),
   summ=c('notNA(x)','min(x)','pctile(x)[25]' ,'median(x)','pctile(x)[75]', 'max(x)','mean(x)', 'sd(x)'),
   summ.names=c('N','Min','Pctile[25]' ,'Median','Pctile[75]', 'Max','Mean', 'Sd'),
   title = "CSA HHI Summary Statistics") %>%
  kable_classic_2(full_width = F)

```
Now we will look at summary stats of HHI conditional on if the market is in a CSA or if it's not.

```{r}
### Custom summarise function command

csa_grp_landfill <- landfill_hhi_csa_df %>% 
  mutate(`Type`=ifelse(iscsa==1,"CSA","County")) %>%
  group_by(Type) %>%
  drop_na %>%
  summarise(`N`=n(),
  `Min`=min(waste_hhi),
  `Pctile[25]`=pctile(waste_hhi)[25],
  `Median`=median(waste_hhi),
  `Pctile[75]`=pctile(waste_hhi)[75],
  `Max`=max(waste_hhi),
  `Mean`=mean(waste_hhi),
  `Sd`=sd(waste_hhi)) %>%
  kbl(digits=2,
      caption = "Market Types") %>%
  kable_classic_2(full_width = F)

csa_grp_landfill

# csa_grp_landfill <- landfill_hhi_csa_df
# st(csa_grp_landfill,group = "In CSA",group.long=T,out="kable",labels=c("Waste HHI"),title="CSA vs County HHIs Summary Statistics",summ=c('min(x)','pctile(x)[25]' ,'median(x)','pctile(x)[75]', 'max(x)','mean(x)', 'sd(x)'),factor.numeric=T) %>%
#   kable_classic_2(full_width = F)
```

```{r}

csa_grp_landfill <- landfill_hhi_csa_df %>% rename(`In CSA`=iscsa)

# tapply(landfill_hhi_csa_df$waste_hhi,landfill_hhi_csa_df$iscsa,summary)
```
Now let's look at the deciles of HHIs on the CSA level below:
```{r}

# quantile(landfill_hhi_csa_df$waste_hhi,seq(0,1,0.1),na.rm=T)

# quantile(landfill_hhi_cbsa_df$waste_hhi,seq(0,1,0.1),na.rm=T)
df_1 <- landfill_hhi_csa_df %>%
  mutate(`Type`="CSA") %>%
  select(-iscsa) %>%
  left_join(landfill_prop_csa)

df_2 <- landfill_hhi_cbsa_df %>%
  mutate(`Type`="CBSA") %>%
  select(-iscbsa) %>%
  left_join(landfill_prop_cbsa)

landfill_csa_cbsa_df <- bind_rows(df_1,df_2)

landfill_csa <- df_1

landfill_csa %>%
  group_by(`Type`) %>%
  summarise(q=list(quantile(waste_hhi,seq(0,1,0.1),na.rm=T))) %>%
  unnest_wider(q) %>% 
  arrange(desc(`Type`)) %>%
  kbl(full_wdith=F,
      digits=0,
      caption = "CSA HHIs") %>%
  kable_classic_2(full_width = F)

```

The count of markets on the CSA level is `r nrow(landfill_hhi_csa_df)`  (including county level markets if no CSA exists).

Now let's look at the HHIs of only markets with several landfills.  
```{r}
non_mono_landfill_csa_cbsa_df <- landfill_csa_cbsa_df %>%
  filter(several_markets=="Several Landfills")

non_mono_landfill_csa_df <- non_mono_landfill_csa_cbsa_df %>%
  filter(waste_hhi<10000 & Type == "CSA")

non_mono_landfill_cbsa_df <- non_mono_landfill_csa_cbsa_df %>%
  filter(several_markets=="Several Landfills" & Type == "CBSA")

non_mono_landfill_csa_df %>%
  group_by(`Type`) %>%
  summarise(q=list(quantile(waste_hhi,seq(0,1,0.1),na.rm=T))) %>%
  unnest_wider(q) %>% 
  arrange(desc(`Type`)) %>%
  kbl(full_wdith=F,
      digits=0,
      caption = "CSA HHIs") %>%
  kable_classic_2(full_width = F)
```

The count of markets with more than one landfill on the CSA level is `r nrow(non_mono_landfill_csa_df)`  (including county level markets if no CSA exists).

```{r}

## Data check if all percentages in a market equal 1
chk_lf_df <- landfill_shrs_csa %>% 
  group_by(market) %>%
  mutate(chk_per=sum(waste_shr)) %>%
  ungroup() %>%
  select(chk_per) %>%
  unique()
```

### Graphs
```{r}
## Count of firms
cnt_landfill_csa <- landfill_shrs_csa %>% 
  ungroup() %>%
  group_by(market) %>%
  summarise(landfills_cnt = n())

## CSA level bar chart data
bar_csa <- landfill_shrs_csa %>% 
  ungroup() %>%
  group_by(market) %>%
  summarise(landfills_cnt = n()) %>%
  ungroup() %>%
  group_by(landfills_cnt) %>%
  summarise(freq=n()) %>%
  mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
  ungroup() %>%
  group_by(grp_lbls) %>%
  summarise(freq=sum(freq)) %>%
  mutate(lvl="CSA")

## CBSA level bar chart data
bar_cbsa <- landfill_shrs_cbsa %>% 
  ungroup() %>%
  group_by(market) %>%
  summarise(landfills_cnt = n()) %>%
  ungroup() %>%
  group_by(landfills_cnt) %>%
  summarise(freq=n()) %>%
  mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
  ungroup() %>%
  group_by(grp_lbls) %>%
  summarise(freq=sum(freq)) %>%
  mutate(lvl="CBSA")

combine_csa_cbsa <- bind_rows(bar_csa,bar_cbsa)

csa_chart <- bar_csa
```

Below is a bar-chart of frequency of landfills across markets.  About `r max(bar_csa$freq)` markets across a CSA only have 1 landfill that serves the market.  The highest number of firms in a market is `r max(cnt_landfill_csa$landfills_cnt)` in the New York-Newark CSA.  `r  sum(cnt_landfill_csa$landfills_cnt>=20)` of these markets have a firm count greater than 20.
```{r}
positions = c(seq(1,9,1),"10+")
ggplot(csa_chart,aes(x=grp_lbls,y=freq)) +
  geom_bar(stat="identity") +
  theme_bw() + 
  geom_text(aes(label = freq), hjust = -0.05, colour = "black") +
  geom_hline(yintercept=0,color="gold",linetype='dashed') +
  coord_flip() +  
  scale_x_discrete(limits = positions) +
  xlab("Landfill Counts") +
  ylab("Frequency") + 
  ggtitle("Freq. of Landfills by CSA Based Market") +
  labs(caption = "Source: Waste Business Journal's: Directory of Waste Processing & Disposal Sites 2019")
## Bar chart CSA

# ggplot(data=bar_csa, aes(x=grp_lbls,y=freq)) +
#   geom_bar(stat="identity") + 
#   ggtitle("CSA Freq. of Landfills by Market") +
#   xlab("Landfill Counts") +
#   ylab("Frequency") + 
#   scale_x_discrete(limits = positions) + 
#   theme_bw()
```

### 2024-08-01 trash/landfill/data_clean_operator.Rmd
added code written by Romeo Ignacio to geocode landfills.

---
title: "Landfill Geographic Analysis: Operator Level"
output: html_document
date: "2024-01-11"
author: Charles Taragin and Romeo Ignacio
---

```{r setup, include=FALSE}
library(tidycensus)
library(tidyverse)
library(sf)
library(dplyr)
library(readxl)
library(knitr)
library(kableExtra)
library(vtable)
library(stringr)
library(gtsummary)

knitr::opts_chunk$set(echo = F, message = F)
```

### Summary Statistics

```{r map_data, results = 'hide'}
### Step 1: Read in and graph data
csa_shp <- get_acs(
  geography = "combined statistical area",
  variables = "B19013_001",
  geometry = TRUE,
  survey = "acs1",
  year = 2019
) %>%
  select(GEOID,NAME,geometry)

cbsa_shp <- get_acs(
  geography = "cbsa",
  variables = "B19013_001",
  geometry = TRUE,
  survey = "acs1",
  year = 2019
) %>%
  select(GEOID,NAME,geometry)

# Deleted option: cbsa

county_shp <- get_acs(
  geography = "county",
  variables = "B19013_001",
  geometry = TRUE,
  survey = "acs1",
  year = 2019
) %>%
  select(GEOID,NAME,geometry)

raw <- read_xlsx("/fsr/project4/Charles/bargaining_convex/trash/Copy of WBJ_Disposal_Sites_2019.xlsx",skip=3)

## Clean up raw file names

raw_uniq_own_nms <- raw %>% 
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Owner Code`,`Owner Entity`,Type,Family) %>%
  unique() %>%
  mutate(`Owner Full Name`= ifelse(is.na(`Owner Code`),"",`Owner Entity`),
    `Owner Code`=ifelse(is.na(`Owner Code`),`Owner Entity`,`Owner Code`),
    ) %>%
  group_by(`Owner Code`)  %>%
  select(`Owner Code`) %>%
  arrange(`Owner Code`) %>%
  group_by(`Owner Code`) %>%
  summarise(count=n())

write.csv(raw_uniq_own_nms,"/fsr/project4/Charles/bargaining_convex/trash/owner_nms_raw.csv")

### This is for names of entities that did not get assigned an owner code for checks

## New way to check data:  Have a flag regarding is_na, check if replaced operator name is similiar to owner name and vice versa

# 3492 obe
raw_uniq_op_nms <- raw %>% 
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Operator Code`,`Operator Entity`,Type,Family) %>%
  unique() %>%
  mutate(
    `Operator Code`=ifelse(is.na(`Operator Code`),`Operator Entity`,`Operator Code`)) %>%
  group_by(`Operator Code`)  %>%
  select(`Operator Code`) %>%
  arrange(`Operator Code`) %>%
  group_by(`Operator Code`) %>%
  summarise(count=n() )

tmp <- raw %>% 
  left_join(raw_uniq_op_nms) %>%
  filter(is.na(count))

write.csv(raw_uniq_op_nms,"/fsr/project4/Charles/bargaining_convex/trash/operator_nms_raw.csv")

# Clean for owner
# Original count 3787
raw_uniq_own_nms <- raw %>% 
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Owner Code`,`Owner Entity`,Type,Family,Ownership) %>%
  unique() %>%
  mutate(`Owner Code`=ifelse(is.na(`Owner Code`),`Owner Entity`,`Owner Code`)) %>%
  group_by(`Owner Code`) %>%
  select(`Owner Code`,Ownership) %>%
  mutate(`Owner Code`=str_replace_all(`Owner Code`,"Tidewater Fiber Corp","Tidewater Fibre Corp"),
         `Owner Code`=str_replace_all(`Owner Code`,"Town of Bethleham","Town of Bethlehem"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("ABC Disposal Systems"),"ABC"),
         `Owner Code`=toupper(`Owner Code`),
         `Owner Code`=str_replace_all(`Owner Code`,"\\.",""),
         `Owner Code`=str_replace_all(`Owner Code`,"\\,",""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" inc\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corp\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" co\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" company\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corporation\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" LLC\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" & "),"&"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("LP\\b"),""),
         `Owner Code`=str_squish(`Owner Code`)) %>%
  unique() %>%
  mutate(dup=n())

  #Clean up certain strings

raw <- raw %>%
  mutate(`Operator Code`=ifelse(is.na(`Operator Code`),`Operator Entity`,`Operator Code`)) %>%
  mutate(`Operator Code`=str_replace_all(`Operator Code`,"Tidewater Fiber Corp","Tidewater Fibre Corp"),
         `Operator Code`=str_replace_all(`Operator Code`,"Angelo's Recycled Materials - Tampa Recycling Facility","Angelo's Recycled Materials"),
         `Operator Code`=str_replace_all(`Operator Code`,regex("ABC Disposal Systems"),"ABC"),
         `Operator Code`=str_replace_all(`Operator Code`,regex("Athens Services"),"ATHEN"),
         `Operator Code`=str_replace(`Operator Code`,"Monadnock Disposal Service Inc.",	
"Monadnock Disposal Services"),
         `Operator Code`=str_replace(`Operator Code`,"Southern Disposal Services","Southern Disposal Service"),
         `Operator Code`=str_replace(`Operator Code`,"Vogel Disposal Service","Vogel Disposal"),
         `Operator Code`=str_replace(`Operator Code`,"Strategic Materials","Strategic Material"),
         `Operator Code`=toupper(`Operator Code`),
         `Operator Code`=str_replace_all(`Operator Code`,"\\.",""),
         `Operator Code`=str_replace_all(`Operator Code`,"\\,",""),
         `Operator Code`=str_replace_all(`Operator Code`,"-"," "),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" inc\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" corp\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" co\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" company\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" corporation\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" LLC\\b",ignore_case=T),""),
         `Operator Code`=str_replace_all(`Operator Code`,regex(" & "),"&"),
         `Operator Code`=str_replace_all(`Operator Code`,regex("LP\\b"),""),
         `Operator Code`=str_squish(`Operator Code`))

raw <- raw %>%
  mutate(`Owner Code`=ifelse(is.na(`Owner Code`),`Owner Entity`,`Owner Code`)) %>%
  mutate(`Owner Code`=str_replace_all(`Owner Code`,"Tidewater Fiber Corp","Tidewater Fibre Corp"),
         `Owner Code`=str_replace_all(`Owner Code`,"Town of Bethleham","Town of Bethlehem"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("ABC Disposal Systems"),"ABC"),
         `Owner Code`=toupper(`Owner Code`),
         `Owner Code`=str_replace_all(`Owner Code`,"\\.",""),
         `Owner Code`=str_replace_all(`Owner Code`,"\\,",""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" inc\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corp\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" co\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" company\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" corporation\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" LLC\\b",ignore_case=T),""),
         `Owner Code`=str_replace_all(`Owner Code`,regex(" & "),"&"),
         `Owner Code`=str_replace_all(`Owner Code`,regex("LP\\b"),""),
         `Owner Code`=str_squish(`Owner Code`)) 
## Look at names for data cleaning
# 1056 missing owner firms
own_count <- raw %>%
  group_by(`Owner Code`) %>%
  summarise(count=n())
# 622 missing operator firms
op_count <- raw %>%
  group_by(`Operator Code`) %>%
  summarise(count=n())
# 10 missing firms
own_r_op_count <- raw %>%
  mutate(is_na_own = is.na(`Owner Code`),
         `Owner Code` = ifelse(is.na(`Owner Code`),`Operator Code`,`Owner Code`)) %>%
  group_by(`Owner Code`) %>%
  summarise(count=n())
# 10 missing firms
op_r_own_count <- raw %>%
  mutate(is_na_op = is.na(`Operator Code`),
         `Operator Code` = ifelse(is.na(`Operator Code`),`Owner Code`,`Operator Code`)) %>%
  group_by(`Operator Code`,is_na_op) %>%
  summarise(count=n())

raw <- raw %>%
  mutate(`Operator Code` = ifelse(is.na(`Operator Code`),`Owner Code`,`Operator Code`),
         `Owner Code` = ifelse(is.na(`Owner Code`),`Operator Code`,`Owner Code`),
         `Operator Code` = ifelse(is.na(`Operator Code`),`Facility Name`,`Operator Code`),
         `Owner Code` = ifelse(is.na(`Owner Code`),`Facility Name`,`Owner Code`))

tmp <- raw %>%
  filter(is.na(`Operator Code`) & is.na(`Owner Code`))

#Bestway Disposal (Different owner???)
#Monadnock Disposal Services (Diferent owner???)
#Texas Disposal Systems vs Texas Disposal Systems Inc.
operator_names <- raw %>%
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  select(`Operator Entity`) %>%
  unique()
#Willimantic Waste Paper Co For operator (Different operator???)
#Bestway Disposal (Different operator???)
# 9499 observations before
disposal_df <- raw %>%
  select(Code,Location,City,County,State,Zip,Longitude,Latitude,Code,Type,Family,`Waste Shed`,`Operator Code`,`Total of All Wastes (tons/day)`,Operation) %>% # Make changes to entity type here
  mutate(st_cnty=paste(County,State,sep = ":")) %>%
  rename(Waste_Shed = `Waste Shed`,
         Operator_Code = `Operator Code`)

## Determine level of aggregation

```

```{r}
### Step 2: Merge in shape files
## Merge all disposal data
disposal_df_shp <- st_as_sf(disposal_df,coords=c("Longitude","Latitude"), crs=4326) %>% #crs=4269)
  st_transform(crs=4269)
############### CSA ###############
intersect_df <- st_join(disposal_df_shp,csa_shp,left=T,join=st_intersects) %>%
    rename(cbsa=NAME) %>%
    mutate(st_cnty=paste(County,State,sep = ":"),
    market=ifelse(is.na(cbsa),st_cnty,cbsa),
    iscsa=!is.na(cbsa)) %>%
    st_drop_geometry()
############### CBSA ###############
cbsa_intersect_df <- st_join(disposal_df_shp,cbsa_shp,left=T,join=st_intersects) %>%
    rename(cbsa=NAME) %>%
    mutate(st_cnty=paste(County,State,sep = ":"),
    market=ifelse(is.na(cbsa),st_cnty,cbsa),
    iscbsa=!is.na(cbsa)) %>%
    st_drop_geometry()

############### County ###############
county_intersect_df <- st_join(disposal_df_shp,county_shp) %>%
  st_drop_geometry() %>%
  rename(county_shp=NAME) %>%
  mutate(county_same = str_detect(county_shp,County)) %>%
  relocate(county_shp,County,county_same)

# prop.table(table(county_intersect_df$county_same==1))

## Create crosswalk
crosswalk_disposal_df <- intersect_df %>%
  select(Code,market)

write.csv(crosswalk_disposal_df,"/fsr/project4/Charles/bargaining_convex/trash/cbsa_disposal_cross.csv",row.names = F)
```

```{r, results ="hide"}
## Clean up data

intersect_df %>% filter(Type == "LF" & market == "Chattanooga, TN-GA Metro Area")

landfill_csa_df <- intersect_df %>%
  st_drop_geometry() %>%
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>% # Step 1: Keep only LF basde on either Type or Family
  filter(Waste_Shed!="Exclusive or Captive"|is.na(Waste_Shed)) %>% # Step 2: Drop all observations labelled "Exclusive or Captive" (1619 Observations)
  mutate(firm_name="") %>%
  select(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa) %>% # Change this line
  group_by(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa) %>% # Change this line
  summarise(`Total of All Wastes (tons/day)`=sum(`Total of All Wastes (tons/day)`))

# Inquire about wasteshed and what to do if NA???
# Old observation count 7193
# New observation number 6823
landfill_csa_df <- intersect_df %>%
  st_drop_geometry() %>%
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>% # Step 1: Keep only LF basde on either Type or Family
  filter(Waste_Shed!="Exclusive or Captive"|is.na(Waste_Shed)) %>% # Step 2: Drop all observations labelled "Exclusive or Captive" (1619 Observations)
  mutate(firm_name="") %>%
  select(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa,Operation) %>% # Change this line
  ungroup()

# Old observation count 1703
landfill_cbsa_df <- cbsa_intersect_df %>%
  st_drop_geometry() %>%
  filter(Type=="LF"|Family=="LF") %>% # Step 1: Keep only LF based on either Type or Family
  filter(Waste_Shed!="Exclusive or Captive"|is.na(Waste_Shed)) %>% # Step 2: Drop all observations labelled "Exclusive or Captive" (1619 Observations)
  select(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscbsa) %>% # Change this line
  ungroup()

```

```{r}
### Get number of landsfill sights that don't match
sum_landfill_all <- landfill_csa_df %>%
  mutate(coor_exis=ifelse(market!=st_cnty,"CSA","County")) %>%
  st_drop_geometry() #%>%
  # select(market,coor_exis) %>% # New line, can delete to get old calculation
  # unique()  # New line, can delete to get old calculation

sum_landfill <- sum_landfill_all  %>%
  select(coor_exis) %>%
  group_by(coor_exis) %>%
  summarise(count=n()) %>%
  ungroup() %>%
  mutate(per_count=100*prop.table(count))

```

This analysis is done on the Operator Code level or Operator Entity name if Operator Code does not exist.  Let's first look at number of markets that are in a CSA and number of markets that are county based as they do not fall in a CSA.  We see that only `r paste0(round(sum_landfill$per_count[1],0),"%")` of landfills are CSA market based
```{r}

#market!=st_cnty
sum_landfill %>%
  kbl(col.names=c("Type","Count","Percent"),
      digits=0,
      caption = "# Landfills by Markets") %>%
  kable_classic_2(full_width = F)

```
```{r}
## Create summary statistics (On MSA-Owner-Operator Lvl)
landfill_sum_csa <- landfill_csa_df %>%
  ungroup() %>%
  group_by(market,Operator_Code) %>%
  summarise(count=n(),
            total_waste=sum(`Total of All Wastes (tons/day)`), .groups="keep")

write.csv(landfill_sum_csa,"/fsr/project4/Charles/bargaining_convex/trash/landfill_final.csv",row.names = F)

# On CSA Level
landfill_mkt_cnt_csa <- landfill_sum_csa %>%
  ungroup() %>%
  group_by(market) %>%
  summarise(count=n())

landfill_prop_csa <- landfill_mkt_cnt_csa %>%
  ungroup %>%
  mutate(several_markets=ifelse(count>1,"Several Landfills","One Landfill"))

# Get summary statistics of landfills
sum_landfill <- landfill_csa_df %>%
  mutate(coor_exis=ifelse(market!=st_cnty,"CSA","County")) %>%
  st_drop_geometry() %>%
  select(coor_exis) %>%
  group_by(coor_exis) %>%
  summarise(count=n()) %>%
  ungroup() %>%
  mutate(per_count=100*prop.table(count))

sum_landfill_prop <- landfill_prop_csa %>%
  group_by(several_markets) %>%
  summarise(count=n()) %>%
  mutate(per_count=100*prop.table(count))

```

Now Let's decompose markets that have 1 landfill and > 1 landfill.  We see about `r paste0(round(sum_landfill_prop$per_count[1],0),"%")` of markets have only one landfill.

```{r}
sum_landfill_prop %>% 
  kbl(col.names=c("Type","Count","Percent"),
      digits=0,
      format.args = list(big.mark = ","),
      caption="Markets with as least one Landfill") %>%
  kable_classic_2(full_width = F)
```

```{r}
## Create HHI Level data
### On CSA Level ###
landfill_shrs_csa <- landfill_csa_df %>%
  ungroup() %>%
  group_by(Operator_Code,market,iscsa) %>% # Aggregate on Market level, Owner Entity, Operator level
  summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
  ungroup() %>%
  group_by(market) %>% # Aggregate on Market
  mutate(total_waste = sum(waste_own),
         waste_shr = 100*(waste_own/total_waste),
         waste_hhi = sum(waste_shr^2))
### On CBSA Level ###
landfill_shrs_cbsa <- landfill_cbsa_df %>%
  ungroup() %>%
  group_by(Operator_Code,market,iscbsa) %>% # Aggregate on Market level, Owner Entity, Operator level
  summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
  ungroup() %>%
  group_by(market) %>% # Aggregate on Market
  mutate(total_waste = sum(waste_own),
         waste_shr = 100*(waste_own/total_waste),
         waste_hhi = sum(waste_shr^2)) ## Have NA remove option

chat <- filter(intersect_df,grepl("Chattanooga",market))

write.csv(chat,"/fsr/project4/Charles/bargaining_convex/trash/chat.csv",row.names = F)

### CSA ### (921 markets)
landfill_hhi_csa_df  <- landfill_shrs_csa %>% 
  select(market,waste_hhi,iscsa) %>% # Keep market level information and filter out duplicates
  unique()

write.csv(landfill_hhi_csa_df,"/fsr/project4/Charles/bargaining_convex/trash/landfill_hhi_final.csv",row.names = F)

### Export Chatanooga CSA ###
chat_shrs <- landfill_shrs_csa %>%
  filter(market=="Chattanooga-Cleveland-Dalton, TN-GA CSA")

write.csv(chat_shrs,"/fsr/project4/Charles/bargaining_convex/trash/chat_share.csv",row.names=F)

### CBSA ### (1088 markets)
landfill_hhi_cbsa_df <- landfill_shrs_cbsa %>% 
  select(market,waste_hhi,iscbsa) %>% # Keep market level information and filter out duplicates
  unique()

```

Now we will look at summary stats of HHI landfill data on the CSA level.  We have `r nrow(landfill_hhi_csa_df)` total markets on the CSA level in the dataset but only `r sum(!is.na(landfill_hhi_csa_df$waste_hhi))` analyzed.  This can be explained as `r sum(is.na(landfill_hhi_csa_df$waste_hhi))` markets have no waste produced, which produces NAs when calculating for HHI.

```{r message=FALSE, warning=F}
graph_data <- landfill_hhi_csa_df %>% select(market,waste_hhi)
st(graph_data,out="kable",labels=c("Waste HHI"),
   summ=c('notNA(x)','min(x)','pctile(x)[25]' ,'median(x)','pctile(x)[75]', 'max(x)','mean(x)', 'sd(x)'),
   summ.names=c('N','Min','Pctile[25]' ,'Median','Pctile[75]', 'Max','Mean', 'Sd'),
   title = "CSA HHI Summary Statistics") %>%
  kable_classic_2(full_width = F)

```
Now we will look at summary stats of HHI conditional on if the market is in a CSA or if it's not.

```{r}
### Custom summarise function command

csa_grp_landfill <- landfill_hhi_csa_df %>% 
  mutate(`Type`=ifelse(iscsa==1,"CSA","County")) %>%
  group_by(Type) %>%
  drop_na %>%
  summarise(`N`=n(),
  `Min`=min(waste_hhi),
  `Pctile[25]`=pctile(waste_hhi)[25],
  `Median`=median(waste_hhi),
  `Pctile[75]`=pctile(waste_hhi)[75],
  `Max`=max(waste_hhi),
  `Mean`=mean(waste_hhi),
  `Sd`=sd(waste_hhi)) %>%
  kbl(digits=2,
      caption = "Market Types") %>%
  kable_classic_2(full_width = F)

csa_grp_landfill

# csa_grp_landfill <- landfill_hhi_csa_df
# st(csa_grp_landfill,group = "In CSA",group.long=T,out="kable",labels=c("Waste HHI"),title="CSA vs County HHIs Summary Statistics",summ=c('min(x)','pctile(x)[25]' ,'median(x)','pctile(x)[75]', 'max(x)','mean(x)', 'sd(x)'),factor.numeric=T) %>%
#   kable_classic_2(full_width = F)
```

```{r}

csa_grp_landfill <- landfill_hhi_csa_df %>% rename(`In CSA`=iscsa)

# tapply(landfill_hhi_csa_df$waste_hhi,landfill_hhi_csa_df$iscsa,summary)
```
Now let's look at the deciles of HHIs on the CSA level below:
```{r}

# quantile(landfill_hhi_csa_df$waste_hhi,seq(0,1,0.1),na.rm=T)

# quantile(landfill_hhi_cbsa_df$waste_hhi,seq(0,1,0.1),na.rm=T)
df_1 <- landfill_hhi_csa_df %>%
  mutate(`Type`="CSA") %>%
  select(-iscsa) %>%
  left_join(landfill_prop_csa)

landfill_csa <- df_1

# Table for HHIs on CSA level
landfill_csa %>%
  group_by(`Type`) %>%
  summarise(q=list(quantile(waste_hhi,seq(0,1,0.1),na.rm=T))) %>%
  unnest_wider(q) %>% 
  arrange(desc(`Type`)) %>%
  kbl(full_wdith=F,
      digits=0,
      caption = "CSA HHIs") %>%
  kable_classic_2(full_width = F)

```

The count of markets on the CSA level is `r nrow(landfill_hhi_csa_df)`  (including county level markets if no CSA exists).

Now let's look at the tabulation of number of firms per market (assuming equally sized firms) from the HHIs we calculated before.  We will use the ceiling function to round to the nearest integer.

```{r}
# summary(non_mono_and_mono_landfill_csa_df$waste_hhi)
non_mono_and_mono_landfill_csa_df <- landfill_csa %>% 
  mutate(eff_firms = 1e4/waste_hhi,
         old_eff_firms=eff_firms,
         old_eff_firms=ceiling(eff_firms),
         eff_firms = ifelse((ceiling(eff_firms)-eff_firms)>0.5,floor(eff_firms),ceiling(eff_firms)))
#1.8 median means there are 1.8 equally sized firms in the market (equivalent)
```

```{r}
data.frame(table(non_mono_and_mono_landfill_csa_df$old_eff_firms)) %>%
  pivot_wider(names_from=Var1, values_from=Freq) %>% 
  kbl(full_wdith=F,
      digits=0,
      caption = "Tabulations if Firms were equally sized (Based on HHIs)") %>%
  kable_classic_2(full_width = F)
```

Now let's do a similar exericse but correcting for rounding issues.

```{r}
data.frame(table(non_mono_and_mono_landfill_csa_df$eff_firms)) %>%
  pivot_wider(names_from=Var1, values_from=Freq) %>% 
  kbl(full_wdith=F,
      digits=0,
      caption = "Tabulations if Firms were equally sized (Based on HHIs. Corrected for rounding)") %>%
  kable_classic_2(full_width = F)
```

Now let's look at the HHIs of only markets with several landfills.  
```{r}
non_mono_landfill_csa_df <- landfill_csa %>%
  filter(several_markets=="Several Landfills") %>%
  filter(waste_hhi<10000 & Type == "CSA")

non_mono_landfill_csa_df %>%
  group_by(`Type`) %>%
  summarise(q=list(quantile(waste_hhi,seq(0,1,0.1),na.rm=T))) %>%
  unnest_wider(q) %>% 
  arrange(desc(`Type`)) %>%
  kbl(full_wdith=F,
      digits=0,
      caption = "CSA HHIs") %>%
  kable_classic_2(full_width = F)
```

The count of markets with more than one landfill that produce waste is `r nrow(non_mono_landfill_csa_df)`  (including CSA markets and county level markets if no CSA exists).

```{r}

## Data check if all percentages in a market equal 1
chk_lf_df <- landfill_shrs_csa %>% 
  group_by(market) %>%
  mutate(chk_per=sum(waste_shr)) %>%
  ungroup() %>%
  select(chk_per) %>%
  unique()
```

Now let's look at the tabulation of number of firms per market (assuming equally sized firms) from the HHIs we calculated before like we did in the previous section.  Notice that the tabulations are quite similar when we looked at all markets.  The first table calculates this based on a ceiling function while the second one calculates this with some correction for rounding (using either a floor or ceiling function dependent on the number)

```{r}
# summary(non_mono_landfill_csa_df$waste_hhi)
non_mono_landfill_csa_df <- non_mono_landfill_csa_df %>% 
  mutate(eff_firms = 1e4/waste_hhi,
         old_eff_firms=eff_firms,
         old_eff_firms=ceiling(eff_firms),
         eff_firms = ifelse((ceiling(eff_firms)-eff_firms)>0.5,floor(eff_firms),ceiling(eff_firms)))
#1.8 median means there are 1.8 equally sized firms in the market (equivalent)
```

```{r}
data.frame(table(non_mono_landfill_csa_df$old_eff_firms)) %>%
  pivot_wider(names_from=Var1, values_from=Freq) %>% 
  kbl(full_wdith=F,
      digits=0,
      caption = "Tabulations if Firms were equally sized (Based on HHIs)") %>%
  kable_classic_2(full_width = F)
```

```{r}
data.frame(table(non_mono_landfill_csa_df$eff_firms)) %>%
  pivot_wider(names_from=Var1, values_from=Freq) %>% 
  kbl(full_wdith=F,
      digits=0,
      caption = "Tabulations if Firms were equally sized (Based on HHIs. Corrected for rounding)") %>%
  kable_classic_2(full_width = F)
```

### Chatanooga County

```{r}
chat_sum <- chat_shrs %>%
  ungroup() %>%
  select(Operator_Code,waste_own,waste_shr,waste_hhi)

```

For an example case, let's look at the share of waste by owner in this market.  The HHI of this market is  `r chat_sum$waste_hhi[1] `

```{r}
chat_sum %>%
  select(Operator_Code,waste_own,waste_shr) %>%
  arrange(desc(waste_shr)) %>%
  kbl(col.names=c("Operator Code","Total Waste","Waste Share"),
      digits=0,
      caption = "# Landfills by Markets") %>%
  kable_classic_2(full_width = F)
```

### Graphs of landfills
```{r}
## Count of firms
cnt_landfill_csa <- landfill_shrs_csa %>% 
  ungroup() %>%
  group_by(market) %>%
  summarise(landfills_cnt = n())

## CSA level bar chart data
bar_csa_freq <- landfill_shrs_csa %>% 
  ungroup() %>%
  group_by(market) %>%
  summarise(landfills_cnt = n()) %>%
  ungroup() %>%
  group_by(landfills_cnt) %>%
  summarise(freq=n()) %>%
  mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
  ungroup() %>%
  group_by(grp_lbls) %>%
  summarise(freq=sum(freq)) %>%
  mutate(lvl="CSA",
         var="Freq.")
bar_csa_prop <- landfill_shrs_csa %>% 
  ungroup() %>%
  group_by(market) %>%
  summarise(landfills_cnt = n()) %>%
  group_by(landfills_cnt) %>%
  summarise(freq=n()) %>%
  mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
  ungroup() %>%
  mutate(prop=100*freq/sum(freq)) %>%
  group_by(grp_lbls) %>%
  summarise(freq=sum(prop)) %>%
  mutate(lvl="CSA",
         var="Prop.")

## CBSA level bar chart data
# bar_cbsa <- landfill_shrs_cbsa %>% 
#   ungroup() %>%
#   group_by(market) %>%
#   summarise(landfills_cnt = n()) %>%
#   ungroup() %>%
#   group_by(landfills_cnt) %>%
#   summarise(freq=n()) %>%
#   mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
#   ungroup() %>%
#   group_by(grp_lbls) %>%
#   summarise(freq=sum(freq)) %>%
#   mutate(lvl="CBSA")

# combine_csa_cbsa <- bind_rows(bar_csa,bar_cbsa)
combine_freq_prop <- bind_rows(bar_csa_prop,bar_csa_freq) %>% 
  mutate(var=factor(var, levels=c('Prop.','Freq.')))

```

Below is a bar-chart of frequency of landfills across markets.  About `r max(bar_csa_freq$freq)` markets across a CSA only have 1 landfill that serves the market.  The highest number of firms in a market is `r max(cnt_landfill_csa$landfills_cnt)` in the New York-Newark CSA.  `r  sum(cnt_landfill_csa$landfills_cnt>=20)` of these markets have a firm count greater than 20.

```{r,fig.width=14}
positions = c(seq(1,9,1),"10+")
ggplot(bar_csa_prop,aes(x=grp_lbls,y=freq)) +
  geom_bar(stat="identity") +
  theme_bw() + 
  geom_text(aes(label = paste0(round(100*freq/sum(freq),0),"%")), hjust = -0.05, colour = "black") +
  geom_hline(yintercept=0,color="gold",linetype='dashed') +
  coord_flip() +  
  scale_x_discrete(limits = positions) +
  xlab("Landfill Counts") +
  ylab("Frequency") + 
  ggtitle("Freq. of Landfills by CSA Based Market") +
  theme(plot.caption = element_text(hjust = 0),
        plot.margin = unit(c(0.15, 0.15, 0.15, 0.15), 
                                "inches")) +
  labs(caption = "Source: Waste Business Journal's: Directory of Waste Processing & Disposal Sites 2019")
## Bar chart CSA

# ggplot(data=bar_csa, aes(x=grp_lbls,y=freq)) +
#   geom_bar(stat="identity") + 
#   ggtitle("CSA Freq. of Landfills by Market") +
#   xlab("Landfill Counts") +
#   ylab("Frequency") + 
#   scale_x_discrete(limits = positions) + 
#   theme_bw()
```

### Cross tabulation of firms

```{r}
### Check to see if they are the same
chk_vertical <- disposal_df %>%
  filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
  mutate(is_municp=grepl("COUNTY|CITY|TOWN|VILLAGE",Operator_Code),
         is_municp_pub=ifelse(Operation=="Public",TRUE,FALSE),
         is_equal=is_municp==is_municp_pub)

landfill_shrs_csa_operation <- landfill_csa_df %>%
  ungroup() %>%
  group_by(Operator_Code,market,iscsa,Operation) %>% # Aggregate on Market level, Owner Entity, Operator level
  summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
  ungroup() %>%
  group_by(market) %>% # Aggregate on Market
  mutate(total_waste = sum(waste_own),
         waste_shr = 100*(waste_own/total_waste),
         waste_hhi = sum(waste_shr^2))
```

Now we will do a cross tabulation of number of vertically and non-vertically integrated firms per market.  Let's first look at the two different ways we determined which firm is vertically integrated and how correlated they are.  The first way we identified vertically integrated firms was through key words such as "Town", "City", "County" or "Village" in their operating code name.  The other way we identified vertically integrated firms is through the a variable "Operation" that classified firms as public or private.  We see the correlation between these two measures is `r round(cor(chk_vertical$is_municp,chk_vertical$is_municp_pub),2)`.  
```{r}

vertical_firms <- landfill_shrs_csa %>%
  mutate(is_municp=grepl("COUNTY|CITY|TOWN|VILLAGE",Operator_Code)) %>%
  group_by(is_municp,market) %>%
  summarise(freq=n()) %>%
  mutate(is_municp= factor(is_municp,levels=c(TRUE,FALSE),labels=c("Vertically Integrated","Non-Vertically Integrated"))) %>%
  pivot_wider(names_from=is_municp, values_from=freq) %>%
  mutate(`Non-Vertically Integrated`=ifelse(is.na(`Non-Vertically Integrated`),0,`Non-Vertically Integrated`),
         `Vertically Integrated`=ifelse(is.na(`Vertically Integrated`),0,`Vertically Integrated`),
         `Non-Vertically Integrated`=factor(ifelse(`Non-Vertically Integrated`>=10,"10+",`Non-Vertically Integrated`),levels=c(seq(0,9),"10+"),labels=paste(c(seq(0,9),"10+"))),
         `Vertically Integrated`=factor(ifelse(`Vertically Integrated`>=10,"10+",`Vertically Integrated`),levels=c(seq(0,9),"10+"),labels=paste(c(seq(0,9),"10+"))))
```

```{r}
# tab1 <- 100*prop.table(table(vertical_firms$`Non-Vertically Integrated`)) %>% round(4)
# data.frame(tab1) %>% 
#   pivot_wider(names_from=Var1,values_from=Freq) %>%
#   kbl(caption = "Proportions of Non-Vertically Integrated Firms") %>%
#   kable_classic_2(full_width = F) 
```

```{r}
# tab2 <- 100*prop.table(table(vertical_firms$`Vertically Integrated`)) %>% round(4)
# data.frame(tab2) %>%
#   pivot_wider(names_from=Var1,values_from=Freq) %>%
#   kbl(caption = "Proportions of Vertically Integrated Firms") %>%
#   kable_classic_2(full_width = F)
```

```{r}
tab3 <- 100*addmargins(prop.table(table(vertical_firms$`Vertically Integrated`,vertical_firms$`Non-Vertically Integrated`))) %>% round(2)

write.csv(tab3,"/fsr/project4/Charles/bargaining_convex/trash/vertical_nonvertical_firms_operator.csv")
tab3 %>%
  kbl(caption = "Proportions of Vertically Integrated and Non-Vertically Integrated Firms",
      ) %>%
  pack_rows("Vertically", start_row = 1, end_row = 12,bold=T) %>% 
  add_header_above(c(" " = 1,"Non-Vertically" = 12),bold=T) %>%
  kable_classic_2(full_width = F)

```
```{r}
# 100*addmargins(prop.table(table(vertical_firms$`Non-Vertically Integrated`,vertical_firms$`Vertically Integrated`)))[1,]
```

### 2024-08-01 trash/landfill/data_clean_owner.Rmd
added code written by Romeo Ignacio to geocode landfills.

--- 2024-08-01 trash/landfill/data_clean_operator.Rmd
@@
 ---
-title: "Landfill Geographic Analysis: Operator Level"
+title: "Landfill Geographic Analysis: Owner Level"
 output: html_document
@@
 disposal_df <- raw %>%
-  select(Code,Location,City,County,State,Zip,Longitude,Latitude,Code,Type,Family,`Waste Shed`,`Operator Code`,`Total of All Wastes (tons/day)`,Operation) %>% # Make changes to entity type here
+  select(Code,Location,City,County,State,Zip,Longitude,Latitude,Code,Type,Family,`Waste Shed`,`Owner Code`,`Total of All Wastes (tons/day)`,Operation) %>% # Make changes to entity type here
   mutate(st_cnty=paste(County,State,sep = ":")) %>%
   rename(Waste_Shed = `Waste Shed`,
-         Operator_Code = `Operator Code`)
+         Owner_Code = `Owner Code`)
 
@@
   mutate(firm_name="") %>%
-  select(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa) %>% # Change this line
-  group_by(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa) %>% # Change this line
+  select(Owner_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa) %>% # Change this line
+  group_by(Owner_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa) %>% # Change this line
   summarise(`Total of All Wastes (tons/day)`=sum(`Total of All Wastes (tons/day)`))
@@
   mutate(firm_name="") %>%
-  select(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa,Operation) %>% # Change this line
+  select(Owner_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscsa,Operation) %>% # Change this line
   ungroup()
@@
   filter(Waste_Shed!="Exclusive or Captive"|is.na(Waste_Shed)) %>% # Step 2: Drop all observations labelled "Exclusive or Captive" (1619 Observations)
-  select(Operator_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscbsa) %>% # Change this line
+  select(Owner_Code,market,st_cnty,County,State,`Total of All Wastes (tons/day)`,iscbsa) %>% # Change this line
   ungroup()
@@
 
-This analysis is done on the Operator Code level or Operator Entity name if Operator Code does not exist.
+This analysis is done on the Owner Code level or Owner Entity name if Owner Code does not exist.
 Let's first look at number of markets that are in a CSA and number of markets that are county based as they do not fall in a CSA.
@@
   ungroup() %>%
-  group_by(market,Operator_Code) %>%
+  group_by(market,Owner_Code) %>%
   summarise(count=n(),
@@
   ungroup() %>%
-  group_by(Operator_Code,market,iscsa) %>% # Aggregate on Market level, Owner Entity, Operator level
+  group_by(Owner_Code,market,iscsa) %>% # Aggregate on Market level, Owner Entity, Operator level
   summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
@@
   ungroup() %>%
-  group_by(Operator_Code,market,iscbsa) %>% # Aggregate on Market level, Owner Entity, Operator level
+  group_by(Owner_Code,market,iscbsa) %>% # Aggregate on Market level, Owner Entity, Operator level
   summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
@@
 ```{r}
-### Custom summarise function command
 
@@
 
-Now let's look at the tabulation of number of firms per market (assuming equally sized firms) from the HHIs we calculated before.  We will use the ceiling function to round to the nearest integer.
-
-```{r}
-# summary(non_mono_and_mono_landfill_csa_df$waste_hhi)
-non_mono_and_mono_landfill_csa_df <- landfill_csa %>% 
-  mutate(eff_firms = 1e4/waste_hhi,
-         old_eff_firms=eff_firms,
-         old_eff_firms=ceiling(eff_firms),
-         eff_firms = ifelse((ceiling(eff_firms)-eff_firms)>0.5,floor(eff_firms),ceiling(eff_firms)))
-#1.8 median means there are 1.8 equally sized firms in the market (equivalent)
-```
-
-```{r}
-data.frame(table(non_mono_and_mono_landfill_csa_df$old_eff_firms)) %>%
-  pivot_wider(names_from=Var1, values_from=Freq) %>% 
-  kbl(full_wdith=F,
-      digits=0,
-      caption = "Tabulations if Firms were equally sized (Based on HHIs)") %>%
-  kable_classic_2(full_width = F)
-```
-
-Now let's do a similar exericse but correcting for rounding issues.
-
-```{r}
-data.frame(table(non_mono_and_mono_landfill_csa_df$eff_firms)) %>%
-  pivot_wider(names_from=Var1, values_from=Freq) %>% 
-  kbl(full_wdith=F,
-      digits=0,
-      caption = "Tabulations if Firms were equally sized (Based on HHIs. Corrected for rounding)") %>%
-  kable_classic_2(full_width = F)
-```
-
 Now let's look at the HHIs of only markets with several landfills.  
@@
 ```
-
-Now let's look at the tabulation of number of firms per market (assuming equally sized firms) from the HHIs we calculated before like we did in the previous section.
-Notice that the tabulations are quite similar when we looked at all markets.
-The first table calculates this based on a ceiling function while the second one calculates this with some correction for rounding (using either a floor or ceiling function dependent on the number)
-
-```{r}
-# summary(non_mono_landfill_csa_df$waste_hhi)
-non_mono_landfill_csa_df <- non_mono_landfill_csa_df %>% 
-  mutate(eff_firms = 1e4/waste_hhi,
-         old_eff_firms=eff_firms,
-         old_eff_firms=ceiling(eff_firms),
-         eff_firms = ifelse((ceiling(eff_firms)-eff_firms)>0.5,floor(eff_firms),ceiling(eff_firms)))
-#1.8 median means there are 1.8 equally sized firms in the market (equivalent)
-```
-
-```{r}
-data.frame(table(non_mono_landfill_csa_df$old_eff_firms)) %>%
-  pivot_wider(names_from=Var1, values_from=Freq) %>% 
-  kbl(full_wdith=F,
-      digits=0,
-      caption = "Tabulations if Firms were equally sized (Based on HHIs)") %>%
-  kable_classic_2(full_width = F)
-```
-
-```{r}
-data.frame(table(non_mono_landfill_csa_df$eff_firms)) %>%
-  pivot_wider(names_from=Var1, values_from=Freq) %>% 
-  kbl(full_wdith=F,
-      digits=0,
-      caption = "Tabulations if Firms were equally sized (Based on HHIs. Corrected for rounding)") %>%
-  kable_classic_2(full_width = F)
-```
-
 ### Chatanooga County
@@
   ungroup() %>%
-  select(Operator_Code,waste_own,waste_shr,waste_hhi)
+  select(Owner_Code,waste_own,waste_shr,waste_hhi)
 
@@
 chat_sum %>%
-  select(Operator_Code,waste_own,waste_shr) %>%
+  select(Owner_Code,waste_own,waste_shr) %>%
   arrange(desc(waste_shr)) %>%
-  kbl(col.names=c("Operator Code","Total Waste","Waste Share"),
+  kbl(col.names=c("Owner Code","Total Waste","Waste Share"),
       digits=0,
@@
 ## CBSA level bar chart data
-# bar_cbsa <- landfill_shrs_cbsa %>% 
-#   ungroup() %>%
-#   group_by(market) %>%
-#   summarise(landfills_cnt = n()) %>%
-#   ungroup() %>%
-#   group_by(landfills_cnt) %>%
-#   summarise(freq=n()) %>%
-#   mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
-#   ungroup() %>%
-#   group_by(grp_lbls) %>%
-#   summarise(freq=sum(freq)) %>%
-#   mutate(lvl="CBSA")
+bar_cbsa <- landfill_shrs_cbsa %>% 
+  ungroup() %>%
+  group_by(market) %>%
+  summarise(landfills_cnt = n()) %>%
+  ungroup() %>%
+  group_by(landfills_cnt) %>%
+  summarise(freq=n()) %>%
+  mutate(grp_lbls = ifelse(landfills_cnt>=10,"10+",landfills_cnt)) %>%
+  ungroup() %>%
+  group_by(grp_lbls) %>%
+  summarise(freq=sum(freq)) %>%
+  mutate(lvl="CBSA")
 
@@
   filter(Type=="LF"|Family=="LF"|Type=="TS"|Family=="TS") %>%
-  mutate(is_municp=grepl("COUNTY|CITY|TOWN|VILLAGE",Operator_Code),
+  mutate(is_municp=grepl("COUNTY|CITY|TOWN|VILLAGE",Owner_Code),
          is_municp_pub=ifelse(Operation=="Public",TRUE,FALSE),
@@
   ungroup() %>%
-  group_by(Operator_Code,market,iscsa,Operation) %>% # Aggregate on Market level, Owner Entity, Operator level
+  group_by(Owner_Code,market,iscsa,Operation) %>% # Aggregate on Market level, Owner Entity, Operator level
   summarise(waste_own = sum(`Total of All Wastes (tons/day)`), .groups="keep") %>%
@@
 vertical_firms <- landfill_shrs_csa %>%
-  mutate(is_municp=grepl("COUNTY|CITY|TOWN|VILLAGE",Operator_Code)) %>%
+  mutate(is_municp=grepl("COUNTY|CITY|TOWN|VILLAGE",Owner_Code)) %>%
   group_by(is_municp,market) %>%

### 2024-08-02 code/TrashSims.R
added back in TrashSimsCompareAll.png/ Other minor nits.

@@ -12,7 +12,7 @@ library(ggplot2)
 ## Load Data:
 
 library(readr)
-paired_data_4_22 <- read_csv("~/Projects/bargaining_convex/output/paired_data_4_22.csv")
+paired_data_4_22 <- read_csv("~/Projects/bargaining_convex/trash/paired_data_4_22.csv")
 
 disposal_price_republic <-  96 # source: 4/26 correspondence with Peggy
 

### 2024-08-09 doc/VerticalArrangements.tex
added back in TrashSimsCompareAll.png/ Other minor nits.
modified picutres per Gloria's 8/6 comments

@@
 \centering
+\caption{Republic/Santek Simulations: Full vs. Partial Models}
+\includegraphics[scale=1.0]{output/TrashSimsCompareAll.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\label{fig:TrashSimsFirms}
+\end{figure}
+
+\begin{figure}
+\centering
 \caption{Simulated Welfare Effects from Mergers between  Republic/Santek, Santek/WasteConn, Santek/WM-ADS, and WM-ADS/Regional}
@@
 \centering
-\caption{The sensitivity of the Republic/Santek simulation results to changes in the bargaining power parameter.
-`All' changes the bargaining power paramtere for all non-vertically integrated firms.
+\caption{The sensitivity of consumer welfare in Santek/Republic merger simulations to changes in the bargaining power parameter.
+%`All' changes the bargaining power paramtere for all non-vertically integrated firms.
 `Santek:Republic' changes the bargaining power parameter for just the Santek/Republic product.
 `WasteConn:Republic' changes the bargaining power parameter for just the WasteConn:Republic product.
-The Dashed line displays baseline results.}
+The Dashed line displays baseline results.
+}
 \includegraphics{output/trashbarg.png}\\
@@
 \centering
-\caption{The sensitivity of the Republic/Santek simulation results to changes in the outside share.
-The absolute value of the market elasticity corresponding to the change in ouside share is in parentheses.
+\caption{The sensitivity of consumer welfare in the Republic/Santek simulations to changes in the outside share.
+The absolute value of the market elasticity corresponding to the change in outside share is in parentheses.
 Baseline results assume 0 outside share.}
@@
 \centering
-\caption{The sensitivity of the Republic/Santek simulation results to changes in the nesting parameter.`Integrated' assumes that all vertically integrated products are in a single nest, while `Landfill` assumes that all collectors that use the same landfill are in a single nest.
+\caption{The sensitivity of consumer welfare in the Republic/Santek simulations to changes in the nesting parameter.`Integrated' assumes that all vertically integrated products are in a single nest, while `Landfill` assumes that all collectors that use the same landfill are in a single nest.
 A nesting parameter of 0 displays baseline results.

### 2024-08-09 code/TrashSims.R

@@ -581,23 +581,22 @@ vert_sum <- mutate(vert_sum,Name=interaction(Disposal,Collector,drop=TRUE,sep="/
                     Name=reorder(Name,`Post-merger`* as.numeric(Effect=="Prices") * as.numeric(Level=="Collection")))
 
 firmplot <- ggplot(data=  vert_sum  %>%
-         mutate(Change=`Post-merger` - `Pre-merger`)
-
-                %>%
+         mutate(Change=`Post-merger` - `Pre-merger`)%>%
          pivot_longer(c(`Pre-merger` ,`Post-merger`),names_to = "Type",values_to = "value") %>%
-         mutate(Type=factor(Type,levels=c("Pre-merger","Post-merger"))) %>%
-         filter(!grepl("Change",Type)),
+         mutate(Type=factor(Type,levels=c("Pre-merger","Post-merger")),
+                Name=factor(as.character(Name))) %>%
+         filter(!grepl("Change",Type)) %>% arrange(Name),
 aes(x=Name,y=value,fill=Type,label=value)) +
   facet_grid(~Level+Effect,scales = "free_x") + geom_bar(stat="identity",
-                                                             position=position_dodge()
+                                                             position=position_dodge2(reverse=TRUE)
                                                              #position="stack"
                                                          )  +
   xlab("Disposal/Collector") + ylab("Equilibrium Levels") + #geom_hline(yintercept = 0,linetype="dashed") +
   scale_fill_brewer(type="qual",palette = "Paired") +#scale_fill_grey(start = .9, end = .1) +
   geom_text( #color=ifelse(Type=="Post-merger","white","black"),
-                hjust=1.5, position=position_dodge(width=.9),size =3) +  coord_flip()+ theme_bw() +
-  theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1))
-
+                hjust=1.5, position=position_dodge2(width=.9,reverse=TRUE),size =3) +  coord_flip()+ theme_bw() +
+  theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_x_discrete(limits=rev)
 
 
 compare <- compare %>% mutate(Name=interaction(Disposal,Collector,drop=TRUE,sep="/"),
@@ -890,20 +889,21 @@ trashinterestingfirmbar <- ggplot(data= filter(mkt_mergersweep_firm,Unintegrated
[…]

### 2024-08-09 doc/VerticalArrangements.tex
minor edits to figure labels. Fixed color-ordering of bars in compareplor so that pre-merger is first.

@@
 \centering
-\caption{The sensitivity of consumer welfare in the Republic/Santek simulations to changes in the nesting parameter.`Integrated' assumes that all vertically integrated products are in a single nest, while `Landfill` assumes that all collectors that use the same landfill are in a single nest.
+\caption{The sensitivity of consumer welfare in the Republic/Santek simulations to changes in the nesting parameter.`Integrated' assumes that all vertically integrated products are in a single nest, while `Landfill' assumes that all collectors that use the same landfill are in a single nest.
 A nesting parameter of 0 displays baseline results.

### 2024-08-09 code/TrashSims.R

@@ -615,13 +615,13 @@ compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
                                Model=factor(Model,levels=c("Pre-merger","Full","Upstream-only","Downstream-only"))),
                    aes(x=Name,y=value,fill=Model,label=value)) +
   facet_grid(~Level,scales = "free_x") + geom_bar(stat="identity",
-                                                         position=position_dodge()
+                                                         position=position_dodge2(reverse = TRUE)
                                                          #position="stack"
   )  +
   xlab("Disposal/Collector") + ylab("Equilibrium Prices") + #geom_hline(yintercept = 0,linetype="dashed") +
   scale_fill_brewer(type="seq",palette = "YlGnBu",direction=-1) +#scale_fill_grey(start = .9, end = .1) +
   geom_text( #color=ifelse(Type=="Post-merger","white","black"),
-    hjust=1.5, position=position_dodge(width=.9),size =3) +  coord_flip()+ theme_bw() +
+    hjust=1.5, position=position_dodge2(width=.9,reverse=TRUE),size =3) +  coord_flip()+ theme_bw() +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(fill="")
 

### 2024-08-09 code/TrashSims.R
changed products to be alphabetically ordered.

@@ -612,7 +612,8 @@ compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
                                       !(Level=="Disposal" & Model=="Downstream-only") &
                                       !(Level=="Collection" & Model =="Upstream-only")) %>%
                         mutate(Model=ifelse(Type=="Pre-merger","Pre-merger",as.character(Model)),
-                               Model=factor(Model,levels=c("Pre-merger","Full","Upstream-only","Downstream-only"))),
+                               Model=factor(Model,levels=c("Pre-merger","Full","Upstream-only","Downstream-only")),
+                               Name=factor(as.character(Name))),
                    aes(x=Name,y=value,fill=Model,label=value)) +
   facet_grid(~Level,scales = "free_x") + geom_bar(stat="identity",
                                                          position=position_dodge2(reverse = TRUE)
@@ -623,7 +624,8 @@ compareplot <- ggplot(data=  filter(compare,Effect!="Shares" &
   geom_text( #color=ifelse(Type=="Post-merger","white","black"),
     hjust=1.5, position=position_dodge2(width=.9,reverse=TRUE),size =3) +  coord_flip()+ theme_bw() +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(fill="")
+  labs(fill="")+
+  scale_x_discrete(limits=rev)
 
 compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="Disposal" & Effect=="Prices") &
                                       Model!="Upstream-only"),

### 2024-08-19 doc/VerticalArrangements.tex
this version incorporates Gloria's edits to the intro and theory section. Includes changes to lit review and bibliography

@@
 
-\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
+\title{Beyond Horizontal and Vertical:\\ Simulating the Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
 Furthermore, the views expressed here should not be purported to reflect those of the U.S.
@@
 \begin{abstract}
-\noindent We study the welfare impacts of mergers when some firms are already vertically integrated.
+\noindent We study the simulation of mergers that have both horizontal and vertical aspects.
 Our model features logit Bertrand competition downstream and Nash Bargaining upstream.
-We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs.
-In our model, we find that mergers with both horizontal and vertical characteristics typically harm consumers.
+We apply the model to the Republic/Santek merger as a real-world example.
+%We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs.
 We apply the model to the Republic/Santek merger as a real-world example.
@@
 \section{Introduction}
-Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.
+Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely horizontal or purely vertical.
+This is particularly true for the mergers involving large firms that usually attract the most scrutiny.
 However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.
 Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
-As a result, the antitrust community has little consensus regarding how ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects, will affect welfare.
+As a result, the antitrust community has done little research into ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects.
 
-The horizontal versus vertical distinction has been enshrined in U.S.
-Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.
-The DOJ did not.
-}  Given these policy documents, it is not surprising that the agencies typically frame most mergers as either horizontal or vertical.
-Similarly, most antitrust research maintains this separate categorization, leaving us with little systematic evidence for how complex mergers impact consumers.
-However, as part of the current effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
+The horizontal versus vertical distinction had been enshrined in U.S.
+Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The most recent Horizontal Merger Guidelines is from 2010, while the most recent Vertical Merger Guidelines is from 2020.
+They have both been recently superseded by a joint document.
+}  Given these policy documents, it is not surprising that the agencies have historically framed most mergers as either horizontal or vertical.
+Similarly, most antitrust research maintains this separate categorization, leaving us with little guidance for how our merger simulations and related tools apply to complex mergers.
+However, in recognition of these issues, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical aspects of competition.
+As part of the revision effort, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
 in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.
@@
 Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition?
-If so, how?''  }  In this paper, we address this question using a vertical supply chain model to explore the welfare implications of complex mergers.
+If so, how?''  }  In this paper, we address this question by exploring the implications that a typical merger simulation model has for the welfare effects of complex mergers.
 
-Our model is drawn from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
-We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+%We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
 Relative to \citeN{ST2021}, the first category, vertical mergers, appears in the previous paper and here serves as a baseline for comparison.
@@
 
-Our model-based simulations show certain patterns across mergers.
+%Our model-based simulations show certain patterns across mergers.
 For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.
@@
 
-We then employ this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
+Our model is drawn from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
+We use this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
 The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.
@@
 }  Fourth, we rely on the logit for our demand function, which limits the ways in which consumers substitute between products.
-In \citeN{ST2012}, we found that assuming the merging firms operate in the same group within a nested logit model tends to increase the magnitude of consumer harm relative to the flat logit.
+We also provide additional results using the nested logit.
 
-Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
-The closest merger simulation article is our previous work in \citeN{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger.
+Our paper is related to the large literature on competition and vertical integration that grew out of a number of theory articles in the 1980s and 1990s.
+\citeN{Riordan2008} has an excellent summary.
+This research explores how incentives to disadvantage rivals could arise alongside the efficiencies (such as EDM) associated with vertical integration.
+Some of these incentives result in RRC effects (e.g.
+\citeN{SalopScheffman1983} or \citeN{Riordan1998}) or in the foreclosure of competitors from input markets (e.g.
+\citeN{Salinger1988} or \citeN{OSS1990}).
+Unlike in our paper, most of these models are not framed explicitly as bargaining problems, instead assuming firms making take-it-or-leave-it offers.
+However, many of the incentives towards RRC, which usually involve trading off profits from input sales against those from downstream sales, are similar to those that appear in our model.
+These theory papers tend to focus on pure vertical mergers rather than complex mergers, in order to maintain mathematical tractability.
+
+Meanwhile, the structural empirical industrial organization literature has also studied the competitive effects of vertical integration as part of a greater exploration of the impacts of bargaining power in vertical supply chains.
+Many of these papers are based on the model developed in \citeN{HW1988} modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.
+These models have been applied to several sectors that are the frequent target of vertical integration, such as healthcare (e.g.
+\citeN{HL2017}) and video programming distribution (e.g.
+\citeN{CLWY2017}).
+These structural empirical papers explore the welfare effects of various competitive interactions, including complex mergers, in certain industries.\footnote{There is a companion empirical literature that relies more on regression analysis that identified changes in market outcomes due to vertical integration (e.g.
+\citeN{Chipty2001} on video programming distribution).
+}  Our framework is similar to the models appearing in these papers, but it has been simplified for use with limited data, with the goal of being more accessible to practitioners and policymakers.
+
+Our work is also related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
+The closest merger simulation article is our previous research in \citeN{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger, ruling out complex mergers.
 That paper in turn takes the setup from \citeN{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.
 \citeN{DVDS2018} and \citeN{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.
-\citeN{MS2013} and \citeN{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.
-However, none of these papers study complex mergers.
-
-Our model draws upon the empirical industrial organization literature on bargaining in supply chains, typified by \citeN{DKVB2010}.
-These models have been applied to several sectors that are the frequent target of complex mergers, such as healthcare (e.g.
-\citeN{HL2017}) and video programming distribution (e.g.
-\citeN{CLWY2017}).
-These structural empirical papers explore the welfare effects of various competitive interactions in certain industries.
-Unlike these papers, our framework has been simplified for use with limited data, with the goal of being more accessible to practitioners.
-These simplifications allow us to provide numerical results over a wide range of market conditions, which complements the insights that come from the structural empirical literature on specific case studies.
+\citeN{MS2013}, \citeN{Rogerson2020}, and \citeN{DSS2024} develop UPP-style tools for assessing the impact of vertical mergers.
+However, these papers do not focus on complex mergers.
 
@@
 In Section \ref{sec:theory}, we introduce the theoretical framework and explain how we model mergers.
-Section \ref{sec:sims} explains how we construct our numerical simulations and provides our results, and our application to Republic/Santek appears in Section \ref{sec:application}.
+Our application to Republic/Santek appears in Section \ref{sec:application}.
 Section \ref{sec:concl} concludes.
@@
 \section{Theory\label{sec:theory}}
-We begin by describing our basic framework, taken from \citeN{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.
-This framework allows us to study a variety of complex merger configurations.
+We begin by describing our basic framework, based on \citeN{ST2021}, which uses a downstream Bertrand model embedded in an upstream Nash Bargaining setup.
+We extend this framework allows to study a variety of complex merger configurations.
 
@@
 Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.
-Prior to any mergers taking place, each wholesaler offers only one product, although each retailer can purchase from multiple wholesalers.
-We denote the set of all retailers by $\mathbb{R}=\{1, \dots, \left\vert{\mathbb{R}}\right\vert\}$ and the set of all wholesalers by $\mathbb{W}=\{1, \dots, \left\vert{\mathbb{W}}\right\vert\}$.
-We divide the set $\mathbb{W}$ into overlapping subsets, each labeled $\mathbb{W}^r$, to indicate which wholesalers' products are carried by which retailers.
-Similarly, we divide the set of retailers $\mathbb{R}$ into overlapping subsets, each labeled $\mathbb{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.
+Prior to any mergers taking place, each wholesaler offers only one product, although each retailer may purchase from multiple wholesalers.
+Therefore, unique products (meaning a wholesale product offered at a given retail outlet) are denoted by pairs of the retailer and wholesaler indices, $rw$.
+We denote the set of all retailers by $\mathcal{R}=\{1, \dots, \left\vert{\mathcal{R}}\right\vert\}$ and the set of all wholesalers by $\mathcal{W}=\{1, \dots, \left\vert{\mathcal{W}}\right\vert\}$.
+We divide the set $\mathcal{W}$ into potentially overlapping subsets, each labeled $\mathcal{W}^r$, to indicate which wholesalers' products are carried by which retailers.
+Similarly, we divide the set of retailers $\mathcal{R}$ into potentially overlapping subsets, each labeled $\mathcal{R}^w$, to indicate the retailers that carry the product sold by each wholesaler.
+Note that throughout we take as given the set of retailer-wholesaler combinations being offered in the market.
+We do not model the process of firm entry and exit or the way in which retailers choose which wholesale products to offer.
 
-The share of consumers that choose product $w$ sold by retailer $r$ has the logit form,
+Let the share of consumers that choose product $w$ sold by retailer $r$ be denoted by $s_{rw}$.
+This share is a function of the vector of retail prices $P$ offered across all available products.\footnote{We omit the retail price argument for $s_{rw}(P)$ in what follows for brevity's sake.}  In our simulations we model the share using both logit and nested logit functions.
+The random coefficients logit is another commonly used option.
+The profit for retailer $r$ is given by $\pi^r = \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx} M$, where $p^W_{rx}$ is the unit fee charged by wholesaler $x$ to retailer $r$, $c^R_{rx}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.
+Downstream prices are set in Bertrand equilibrium according to
 \begin{equation}
-s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{t \in \mathbb{R}} \sum_{x \in \mathbb{W}^t} \exp(\delta_{tx} - \alpha p_{tx})},
-\label{eq: nb share}
-\end{equation}
-where $\delta_{rw}$ is a quality parameter and $\alpha$ captures sensitivity to price $p_{rw}$.
-There is an outside good whose quality parameter and price have been normalized to zero.
-The retailer's profit is given by $\pi^r = \sum_{w \in \mathbb{W}^r} [p_{rw} - p^W_{rw} - c^R_{rw}] s_{rw} M$, where $p^W_{rw}$ is the unit fee charged by wholesaler $w$ to retailer $r$, $c^R_{rw}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.
-Downstream prices are set in Bertrand equilibrium, according to
-\begin{equation}
-\sum_{x \in \mathbb{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}
+\sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}
 {\partial p_{rw}} + s_{rw} = 0,
@@
 
-Wholesale prices are set via Nash Bargaining between retailers and wholesalers.
+Wholesale prices are set via bilaterial Nash Bargaining between retailers and wholesalers.
 We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
-}  Profits for wholesaler $w$ are given by $\pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M$, where $c^W_{rw}$ is the wholesale marginal cost when dealing with retailer $r$.
+}  Profits for wholesaler $w$ are given by $\pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M$, where $c^W_{xw}$ is the wholesale marginal cost when dealing with retailer $x$.
 The first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
@@
 \begin{split}
-&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
-&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right),
+&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
+&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right),
 \end{split}
@@
 where $\lambda \in [0,1]$ captures the bargaining power of the retailer relative to the wholesaler.
-The $\Delta s_{tx}(\mathbb{W}^r \setminus \{w\}) \equiv s_{tx}(\mathbb{W}^r \setminus \{w\}) - s_{tx}$ is the difference in the share of good $x$ sold by retailer $t$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.
+The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) \equiv s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.
 Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.
@@
 Together the series of downstream and upstream first order conditions determine market equilibrium.
-This model can be solved and calibrated as described in \citeN{ST2021}.
-In our baseline configuration, we assume that wholesale and retail marginal costs are constant.
-We allow for linear increasing marginal costs as a robustness check that appears in Appendix \ref{app:mc}.
+After assuming a functional form for shares, the model can be solved and calibrated along the lines described in \citeN{ST2021}.
 
@@
 Here we describe the manner in which complex mergers are modeled in this framework.
-Throughout we assume that each product's quality, $\delta_{rw}$, and marginal costs, $c^R_{rw}$ and $c^W_{rw}$, remain unchanged after the merger.
+Throughout we assume that all marginal costs, $c^R_{rw}$ and $c^W_{rw}$, and any non-price product characteristics remain unchanged after the merger, which rules out merger efficiencies.
 We begin by describing vertical mergers, as all cases we examine have some vertical aspects.
-Suppose that retailer $r$ and wholesaler $w$ were to merge.
-Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $rw$ is given by
+Suppose that retailer $t$ and wholesaler $v$ were to merge.
+Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $tv$ is given by
 \begin{equation}
 \begin{split}
-&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{EDM effect}} \\
-&+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} = 0.
+&\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} + s_{tv} + \overbrace{[p_{tv} - c^W_{tv} - c^R_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}^{\text{EDM effect}} \\
+&+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_{\text{upstream UPP effect}} = 0.
 \end{split}
@@
 The pricing problem balances two effects.
-On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $r$ being able to access product $w$ at marginal cost.
-This force would tend to lower the resulting price.
-On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to wholesaler $w$.
+On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $t$ being able to access product $v$ at marginal cost.
+This force would tend to lower the resulting price.\footnote{However, this lower marginal cost raises the potential for a larger margin, which could also cause a multi-product retailer to raise prices on its other goods $x \in \mathcal{W}^t \setminus \{v\}$, in order to divert sales to product $tv$.
+This is the Edgeworth-Salinger effect as in \citeN{Salinger1991}.}  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $t$ in order to divert sales to wholesaler $v$.
 This force would tend to raise the resulting consumer price.
 
-Turning to input prices, when wholesaler $w$ bargains with unaffiliated retailer $s$, the first order condition becomes
+Turning to input prices, when wholesaler $v$ bargains with unaffiliated retailer $j$, the first order condition becomes
 \begin{equation}
 \begin{split}
-&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
-& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
-&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
+&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
+& - \overbrace{\overbrace{[p_{tv}-c^W_{tv}-c^R_{tv}] \Delta s_{tv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^j \setminus \{v\})}^{\text{RRC effect}}=\\
+&\frac{1-\lambda}{\lambda} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
 \end{split}
@@
 \end{equation}
-which reflects the change in the disagreement payoff coming from the merger with retailer $r$.
-Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $s$, these losses are softened due to a potential for diversion to retailer $r$, which we label the ``RRC effect.''  Furthermore, the margin on product $w$ sold by retailer $r$ is potentially higher due to EDM, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
-These impacts tend to raise the resulting input price.
+which reflects the change in the disagreement payoff coming from the merger with retailer $t$.
+Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $j$, these losses are softened due to a potential for diversion to retailer $t$, which we label the ``RRC effect.''\footnote{The assumption of simultaneous downstream price setting rules out RRC effects in the style of \citeN{SalopScheffman1983}, in that here the wholesaler does not have an incentive to raise prices in order to directly impact the pricing strategies of rival retailers.
+Rather, our model has the RRC effect seen in \citeN{CLWY2017} (Section 6.2.2, p.
+934), where the wholesaler has an incentive to raise prices to rival retailers due to the possibility of increased sales volume for their merged downstream retail partner.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
+These impacts tend to raise the resulting input price to retailer $j$.
 
-When the merged firm is bargaining with the unaffiliated wholesaler $v$ over what input price to pay, the bargaining first order condition becomes
+When the merged firm is bargaining with the unaffiliated wholesaler $k$ over what input price to pay, the bargaining first order condition becomes
 \begin{equation}
 \begin{split}
-&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
-&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
-&\left.- \underbrace{[p_{rw} - c^W_{rw} - c^R_{rw}] \Delta s_{rw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
+&[p^W_{tk} - c^W_{tk}]s_{tk} - \sum_{x \in \mathcal{R}^k \setminus \{t\}} [p^W_{xk} - c^W_{xk}] \Delta s_{xk}(\mathcal{W}^t \setminus \{k\}) = \\
+&\frac{1-\lambda}{\lambda} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
+&\left.- \underbrace{[p_{tv} - c^W_{tv} - c^R_{tv}] \Delta s_{tv}(\mathcal{W}^t \setminus \{k\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{wholesale recapture leverage effect}}\right).
 \end{split}
@@
 \end{equation}
-In this case, the merged firm has two channels for potential additional profits should it cease to trade with wholesaler $v$.
-First, if retail sales are diverted to product $w$ sold by retailer $r$, those sales will earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $v$ carried by retailer $r$ could increase sales by wholesaler $w$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.
+In this case, the merged firm has two channels for potential additional profits should it cease to trade with wholesaler $k$.
+First, if retail sales are diverted to product $v$ sold by retailer $t$, those sales will earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $k$ carried by retailer $t$ could increase sales by wholesaler $v$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.
 
-Next consider a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
+Next consider a merger between an integrated retailer/wholesaler $tv$ and a standalone retailer $s$.
 Such a combination has a vertical component and a horizontal component.
-The first order condition for setting the downstream price $p_{rw}$ becomes
+The first order condition for setting the downstream price $p_{tv}$ becomes
 \begin{equation}
 \begin{split}
-&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{direct EDM effect}}+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} \\ &+\underbrace{\underbrace{[p_{sw} - c^W_{sw} - c^R_{sw}]\frac{\partial s_{sw}}{\partial p_{rw}}}_{\text{indirect EDM Effect}} + \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \frac{\partial s_{sx}} {\partial p_{rw}}}_{\text{downstream UPP effect}}= 0.
+&\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} + s_{tv} + \overbrace{[p_{tv} - c^W_{tv} - c^R_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}^{\text{direct EDM effect}}+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t, s\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_{\text{upstream UPP effect}} \\ &+\underbrace{\underbrace{[p_{sv} - c^W_{sv} - c^R_{sv}]\frac{\partial s_{sv}}{\partial p_{tv}}}_{\text{indirect EDM Effect}} + \sum_{x \in \mathcal{W}^s \setminus \{v\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \frac{\partial s_{sx}} {\partial p_{tv}}}_{\text{downstream UPP effect}}= 0.
 \end{split}
@@
 \end{equation}
-Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $r$ raises its prices.
-EDM between $w$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $w$ sold by retailer $s$, those units may earn a larger margin.
-This impact is what we label the ``indirect EDM effect,'' which comes through the interaction of EDM with UPP.
+Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $t$ raises its prices.
+EDM between $v$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $v$ sold by retailer $s$, those units may earn a larger margin.
+This impact, which comes through the interaction of EDM with UPP, follows the logic of \citeN{Salinger1991}, where EDM on one product can cause a firm to raise prices on other products in order to drive sales to the lower cost good.
+We label this effect the ``indirect EDM effect.''
 
-If instead an integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the resulting first order condition for product $rw$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc}.
-However,  the downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of wholesaler $v$.
-The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain product $v$ at marginal cost, which in turn would raise the value of diverted sales to that product.
+If instead an integrated retailer/wholesaler $tv$ were to merge with a standalone wholesaler, then the resulting first order condition for product $tv$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc} with some changes.
+The downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of the new partner wholesaler.
+The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain an additional wholesale product at marginal cost, which in turn would raise the value of diverted sales to that product.
 Whether these additional incentives to raise prices will dominate the direct EDM impact is an empirical question.
@@
 
-As for wholesale prices, again consider a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$.
-When wholesaler $w$ bargains with an unaffiliated retailer $u$ we have the first order condition given by,
+As for wholesale prices, again consider a merger between an integrated retailer/wholesaler $tv$ and a standalone retailer $s$.
+When wholesaler $v$ bargains with an unaffiliated retailer $j$ we have the first order condition given by,
 \begin{equation}
 \begin{split}
-&[p^W_{uw} - c^W_{uw}]s_{uw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s,u\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})\\
-& - \overbrace{\sum_{t \in \{r,s\}}\left(\overbrace{[p_{tw}-c^W_{tw}-c^R_{tw}] \Delta s_{tw}(\mathbb{W}^u \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^t \setminus \{w\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathbb{W}^u \setminus \{w\})\right)}^{\text{RRC effect}}=\\
-&\frac{1-\lambda}{\lambda} \left([p_{uw} - p^W_{uw} - c^R_{uw}]s_{uw} - \sum_{x \in \mathbb{W}^u \setminus \{w\}} [p_{ux} - p^W_{ux} - c^R_{ux}] \Delta s_{ux}(\mathbb{W}^u \setminus \{w\})\right).
+&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,s,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
+& - \overbrace{\sum_{x \in \{t,s\}}\left(\overbrace{[p_{xv}-c^W_{xv}-c^R_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{y \in \mathcal{W}^x \setminus \{v\}} [p_{xy} - p^W_{xy} - c^R_{xy}] \Delta s_{xy}(\mathcal{W}^j \setminus \{v\})\right)}^{\text{RRC effect}}=\\
+&\frac{1-\lambda}{\lambda} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
 \end{split}
@@
 \end{equation}
-The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $r$, both of which may potentially recapture sales should retailer $u$ lose access to product $w$.
+The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $t$, both of which may potentially recapture sales should retailer $j$ lose access to product $v$.
 The merged firm has higher bargaining leverage as a result.
-If instead the integrated retailer/wholesaler $rw$ were to merge with a standalone wholesaler $v$, then the profits earned by firm $rw$ should the negotiation with retailer $u$ fail are augmented with the earnings of wholesaler $v$ rather than of retailer $s$.
-This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with wholesaler $w$, that retailer's contract with wholesaler $v$ remains in place.
-This assumption can be loosened.
+If instead the integrated retailer/wholesaler $tv$ were to merge with a standalone wholesaler, then the profits earned by firm $tv$ should the negotiation with retailer $j$ fail are augmented with the earnings of the additional wholesaler rather than with the earnings of retailer $s$.
+This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with one wholesaler, that retailer's contracts with all other wholesalers remain in place.
 }    In the case of a merger between two integrated retailer/wholesaler firms, all of these effects would appear.
 
-Returning to a merger between an integrated retailer/wholesaler $rw$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $v$ to supply retailer $r$, the first order condition becomes
+Returning to a merger between an integrated retailer/wholesaler $tv$ and a standalone retailer $s$, when the merged firm bargains with an unaffiliated wholesaler $k$ to supply retailer $t$, the first order condition becomes
 \begin{equation}
 \begin{split}
-&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
-&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
-&- \underbrace{\sum_{t \in \{r,s\}}[p_{tw} - c^W_{tw} - c^R_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}
-- \underbrace{\sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^r \setminus \{v\})}_{\text{retail recapture leverage effect}}\\
-&-\left.\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right).
+&[p^W_{tk} - c^W_{tk}]s_{tk} - \sum_{x \in \mathcal{R}^k \setminus \{t\}} [p^W_{xk} - c^W_{xk}] \Delta s_{xk}(\mathcal{W}^t \setminus \{k\}) = \\
+&\frac{1-\lambda}{\lambda} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
+&- \underbrace{\sum_{x \in \{t,s\}}[p_{xv} - c^W_{xv} - c^R_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{EDM recapture effect}}
+- \underbrace{\sum_{x \in \mathcal{W}^s \setminus \{v\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathcal{W}^t \setminus \{k\})}_{\text{retail recapture leverage effect}}\\
+&-\left.\underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t,s\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{wholesale recapture leverage effect}}\right).
 \end{split}
@@
 \end{equation}
-Compared to equation \eqref{eq: vmerger ret upstream foc}, now the EDM recapture effect applies to sales of product $w$ at both affiliated retailers $r$ and $s$.
-Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $r$ lose access to product $v$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture leverage effect.
+Compared to equation \eqref{eq: vmerger ret upstream foc}, now the EDM recapture effect applies to sales of product $v$ at both affiliated retailers $t$ and $s$.
+Furthermore, there is also the possibility that sales will be diverted to retailer $s$ should retailer $t$ lose access to product $k$, which creates an additional ``retail recapture leverage effect'' alongside the wholesale recapture leverage effect.
 All of these additional terms tend to increase the bargaining leverage of the merged firm.
-If we instead examined a merger between retailer/wholesaler $rw$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $r$ for two merged wholesalers, rather than sales through two merged retailers.
+If we instead examined a merger between retailer/wholesaler $tv$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $t$ for two merged wholesalers, rather than sales for wholesaler $v$ through two merged retailers.
 Similarly, the retail recapture leverage effect would be replaced with an additional wholesale recapture leverage effect for the newly merged wholesaler.
 For a merger between two integrated retailer/wholesaler firms, all of these effects would enter.
-
-The discussion thus far has not dealt with the possible existence of integrated rivals to the merging firms.
-If such rival firms exist, they would also have pricing and bargaining incentives similar to those captured in the above post-merger first order conditions.
-For example, if the integrated firm $rw$ were bargaining over what input price to charge a rival integrated firm $sv$, the first order condition would be
-\begin{equation}
-\begin{split}
-&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
-& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
-&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w,v\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right.\\
-&\left.- \underbrace{[p_{sv} - c^W_{sv} - c^R_{sv}] \Delta s_{sv}(\mathbb{W}^s \setminus \{w\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^v \setminus \{s\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^s \setminus \{w\})}_{\text{wholesale recapture leverage effect}}\right).
-\end{split}
-\label{eq: intmerger wh upstream foc}
-\end{equation}
-The existing integration of firm $sv$ gives it countervailing bargaining leverage that may counteract any change in leverage caused by a merger of retailer $r$ and wholesaler $w$.
-The extent of the net impact depends on how these leverage terms balance out in equilibrium.
 

### 2024-08-29 doc/VerticalArrangements.tex
peggy's rewrite of the republic/santek merger section. copy emailed to charles on 20240829

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Empirical Application: Republic/Santek \label{sec:application}}
+\section{Empirical Application: the Chattanooga Waste Industry \label{sec:application}}
 
@@
 Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we begin by applying our model to simulate the effects of the Republic/Santek merger in Chattanooga.
+We then compare these results to those of counterfactual merger simulations in which we examine how competition might have been affected by the combination of other market competitors that exhibit different vertical supply chain arrangements.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.} By focusing on one market at a single point in time, we are better able to interrogate the welfare implications of these structural differences.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Merger Simulation Results}
+\subsection{Republic/Santek Merger Simulation Results}
 
-Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of annual harm to consumers and total annual producer benefits of approximately \$13.4 million.
+Overall, the merger between Republic and Santek is estimated to lower upstream prices by \hl{2.8\%} and increase downstream prices by \hl{12\%}, resulting in \$16.6 million of annual harm to consumers and total annual producer benefits of approximately \$14 million.
 Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda=1$) and Republic ($\lambda=0.7$) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.
 High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.
-(See Table \ref{tab:trashdata}.)}  These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
+(See Table \ref{tab:trashdata}.)}  These effects are illustrated in more detail in
+Figure \ref{fig:TrashSimsFirms}.
 
@@
 However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs).
-The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 22\%.
-These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by 60\%.
+The merged firms' disposal prices to downstream rivals are estimated to increase over 91\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 24\%.
+These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by approximately 61\%.
 
 Turning to the downstream market, all collection prices increase post-merger.
-The merging parties' post-merger collection prices increase by approximately 20\%, and their share decreases by about 8 percentage points overall.
+The merging parties' post-merger collection prices increase by approximately \hl{20\%}, and their share decreases by about 8 percentage points overall.
 The Republic/Republic product's price increases 21\%, and the Santek/Santek product's price increases by 23\%.
 The newly integrated Santek/Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
-The integrated rival, Waste Connections, increases its collection price by approximately 5\%, and its share increases by 6 percentage points.
-The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
+The integrated rival, Waste Connections, increases its collection price by approximately 6\%, and its share increases by 6 percentage points.
+The unintegrated rivals' collection prices each increase by about \hl{9\%}, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
 
@@
 
-Ignoring these vertical relationships in the downstream-only model results in an estimated 11\% increase in collection prices with a greater than 16\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million.
-Estimated consumer harm is 60\% higher in the full vertical model and is offset by markedly larger producer benefits.
+Ignoring these vertical relationships in the downstream-only model results in an estimated
+\hl{11\%} increase in collection prices with a greater than \hl{16\%} increase in the merging parties' price, consumer harm of \hl{\$10 million} annually, and net harm of approximately \hl{\$2 million}.
+Estimated consumer harm is \hl{60\%} higher in the full vertical model and is offset by markedly larger producer benefits.
 The downstream-only model misses the substantial RRC effect experienced by the Santek/WM-ADS and Santek/Regional products.
@@
 In the upstream-only model, downstream prices and therefore shares are fixed at their pre-merger levels.
-We still have upstream prices set to marginal cost for integrated products.
-Overall disposal prices increase by 8.2\% post-merger, and the merger effect is a pure transfer between firms.
-Thus, both consumer and net harm are estimated to be zero.
+Upstream prices are still set to marginal cost for integrated products.
+Overall disposal prices increase by \hl{8.2\%} post-merger, and the merger effect is a pure transfer between firms.
+Thus, both \hl{consumer and net harm are estimated to be zero}.
 The main impact of the merger is in changing the bargaining leverage for Santek, now that it is combined with Republic.
@@
 and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}.} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
+
+\subsection{Counterfactual Merger Simulations}
+
+In addition to applying our model to estimate the effects of the Republic/Santek merger on the Chattanooga market, in this section we simulate the effects of the hypothetical mergers of three other pairs (Santek/Waste Connections, Santek/WM-ADS, and WM-ADS/Regional) that would have resulted in companies with different vertical supply configurations combining.
+The estimated aggregate welfare effects of these counterfactual mergers are illustrated alongside those of the Reublic/Santek merger in Figure \ref{fig:TrashFakeOverall}.
+
+The pre-merger Herfindahl-Hirschman Index (HHI) for the Chattanooga disposal market is \hl{xxxx} and collection market is \hl{yyyy}.
+The merger of WM-ADS with the Regional collection firm is a straightforward ``horizontal'' merger between two relatively small downstream competitors.
+This merger results in a post-merger change in the collection market HHI of \hl{zzz}.
+Unsurprisingly, the resulting estimated welfare effects are small and consumer harm and producer benefits approximately cancel each other out.
+The more detailed results reported in Figure \ref{fig:TrashFakeFirms} show only minimal changes in prices or collection shares post-merger.
+
+Turning to the Santek/WM-ADS merger simulation, we consider the effects of a small downstream rival combining with a vertically integrated firm.
+Here the merger results in a post-merger change in the collection market HHI of \hl{aaa}, but consumer harm from increased collection prices of \hl{x\%} is significantly dampened by price decreases in collection of \hl{y\%} from EDM.
+This results in net harm of approximately \hl{\$z million}.
+
+Finally, we examine the potential merger of Santek and Waste Connections.
+While this merger is similar to that of Santek and Republic in that it is a merger between veritcally integratted rivals, Santek/WasteConn differs in it's effects on the Chattanooga market because both parties exclusively self-supplied for their disposal input pre-merger.
+This diminishes the possibility for EDM effects.
+In addition, while Santek/WasteConn significantly raise disposal prices to their downstream rivals, the presence of the larger, vertically integrated (even if capacity constrained) Republic limits the merged firm's ability to pass this through to the collection market.
+Thus, net harm to the Chattanooga market remains relatively small at \hl{\$d million}.
+This is in spite of post-merger changes in the HHIs of the disposal market of \hl{xxxx} and collection market of \hl{yyyy}.
+
+\subsection{Robustness}
+
+The robustness of our simulation results depends critically on a few key parameter values.
+One of these is the assumption of the bargaining power parameter, $\lambda$.
+Sheu and Taragin (2021) shows that mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
+Larger wholesaler bargaining power offers more possibilities for EDM, as premerger input prices are likely to be high.
+This channel becomes less relevant as retailers gain more power.
+Figure \ref{fig:TrashBarg} illustrates his relationship between consumer welfare and the value of the bargaining power parameter for the Republic/Santek merger simulation.
+Harm increases rapidly as bargaining power increases to $0.4$ and then becomes nearly constant for values above $0.5$.
+Thus, assuming an equal split between collection and disposal would yield an estimate of a near upper bound on consumer harm for this merger.
+
+In addition to specification of the bargaining power parameter, our results are highly sensitive to assumptions about the market elasticity which is specified through the share of consumers choosing to purchase goods outside the market (the outside good share).
+Our baseline results assume that the outside share is zero because there are essentially no reasonable substitutes for SCCW collection or MSW disposal.\footnote{See Competitive Impact Statement: U.S and Plaintiff States v.
+Republic Services, Inc.
+and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/dl?inline}} However, this assumption leads to the largest possible consumer harm in our model as shown in Figure \ref{fig:TrashElast}.
+Often when the size of the outside share is unknown a market elasticity of 1.0 is assumed.
+For the Republic/Santek merger example, this would lead to consumer harm approximately 10 percentage points smaller.
 
@@
 \footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
-\label{fig:TrashSimsFirms}
+\label{fig:TrashSimsCompareAll}
 \end{figure}

### 2024-08-29 doc/VerticalArrangements.tex
added HHI caclulations to code and paper, per Peggy's request. Need to check code in antitrust package responsible for calculating overall prices.

@@
 
-Overall, the merger between Republic and Santek is estimated to lower upstream prices by \hl{2.8\%} and increase downstream prices by \hl{12\%}, resulting in \$16.6 million of annual harm to consumers and total annual producer benefits of approximately \$14 million.
+Overall, the merger between Republic and Santek is estimated to lower upstream prices by almost 3\% and increase downstream prices by 12.5\%, resulting in \$16.6 million of annual harm to consumers and total annual producer benefits of approximately \$14 million.
 Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda=1$) and Republic ($\lambda=0.7$) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.
@@
 Turning to the downstream market, all collection prices increase post-merger.
-The merging parties' post-merger collection prices increase by approximately \hl{20\%}, and their share decreases by about 8 percentage points overall.
+The merging parties' post-merger collection prices increase by 19.5\%, and their share decreases by about 16 percentage points overall.
 The Republic/Republic product's price increases 21\%, and the Santek/Santek product's price increases by 23\%.
-The newly integrated Santek/Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
-The integrated rival, Waste Connections, increases its collection price by approximately 6\%, and its share increases by 6 percentage points.
-The unintegrated rivals' collection prices each increase by about \hl{9\%}, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
+The newly integrated Santek/Republic product's price is estimated to increase by almost 9\%, demonstrating that EDM is not fully passed-through to consumers.
+The integrated rival, Waste Connections, increases its collection price by 5.3\%, and its share increases by almost 10 percentage points.
+The unintegrated rivals' collection prices each increase by about 8\%, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
 
@@
 Ignoring these vertical relationships in the downstream-only model results in an estimated
-\hl{11\%} increase in collection prices with a greater than \hl{16\%} increase in the merging parties' price, consumer harm of \hl{\$10 million} annually, and net harm of approximately \hl{\$2 million}.
-Estimated consumer harm is \hl{60\%} higher in the full vertical model and is offset by markedly larger producer benefits.
+9.2\% increase in collection prices with an 16.4\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million.
+Estimated consumer harm is 60\% higher in the full vertical model and is offset by markedly larger producer benefits.
 The downstream-only model misses the substantial RRC effect experienced by the Santek/WM-ADS and Santek/Regional products.
@@
 Upstream prices are still set to marginal cost for integrated products.
-Overall disposal prices increase by \hl{8.2\%} post-merger, and the merger effect is a pure transfer between firms.
-Thus, both \hl{consumer and net harm are estimated to be zero}.
+Overall disposal prices increase by 11.2\% post-merger, and the merger effect is a pure transfer between firms.
+Thus, both consumer and net harm are estimated to be zero.
 The main impact of the merger is in changing the bargaining leverage for Santek, now that it is combined with Republic.
@@
 In addition to applying our model to estimate the effects of the Republic/Santek merger on the Chattanooga market, in this section we simulate the effects of the hypothetical mergers of three other pairs (Santek/Waste Connections, Santek/WM-ADS, and WM-ADS/Regional) that would have resulted in companies with different vertical supply configurations combining.
-The estimated aggregate welfare effects of these counterfactual mergers are illustrated alongside those of the Reublic/Santek merger in Figure \ref{fig:TrashFakeOverall}.
+The estimated aggregate welfare effects of these counterfactual mergers are illustrated alongside those of the Republic/Santek merger in Figure \ref{fig:TrashFakeOverall}.
 
-The pre-merger Herfindahl-Hirschman Index (HHI) for the Chattanooga disposal market is \hl{xxxx} and collection market is \hl{yyyy}.
+The pre-merger Herfindahl-Hirschman Index (HHI) for the Chattanooga disposal market is 2,646 and collection market is 3,035.
 The merger of WM-ADS with the Regional collection firm is a straightforward ``horizontal'' merger between two relatively small downstream competitors.
-This merger results in a post-merger change in the collection market HHI of \hl{zzz}.
+This merger results in a post-merger change in the collection market HHI of 136s.
 Unsurprisingly, the resulting estimated welfare effects are small and consumer harm and producer benefits approximately cancel each other out.
@@
 Turning to the Santek/WM-ADS merger simulation, we consider the effects of a small downstream rival combining with a vertically integrated firm.
-Here the merger results in a post-merger change in the collection market HHI of \hl{aaa}, but consumer harm from increased collection prices of \hl{x\%} is significantly dampened by price decreases in collection of \hl{y\%} from EDM.
-This results in net harm of approximately \hl{\$z million}.
+Here the merger results in a post-merger change in the collection market HHI of 620, but consumer harm from increased collection prices of 1.3\% is significantly dampened by price decreases in collection of 2.4\% from EDM.
+This results in net consumer harm of approximately \$1.5 million.
 
 Finally, we examine the potential merger of Santek and Waste Connections.
-While this merger is similar to that of Santek and Republic in that it is a merger between veritcally integratted rivals, Santek/WasteConn differs in it's effects on the Chattanooga market because both parties exclusively self-supplied for their disposal input pre-merger.
+While this merger is similar to that of Santek and Republic in that it is a merger between vertically integrated rivals, Santek/WasteConn differs in it's effects on the Chattanooga market because both parties exclusively self-supplied for their disposal input pre-merger.
 This diminishes the possibility for EDM effects.
 In addition, while Santek/WasteConn significantly raise disposal prices to their downstream rivals, the presence of the larger, vertically integrated (even if capacity constrained) Republic limits the merged firm's ability to pass this through to the collection market.
-Thus, net harm to the Chattanooga market remains relatively small at \hl{\$d million}.
-This is in spite of post-merger changes in the HHIs of the disposal market of \hl{xxxx} and collection market of \hl{yyyy}.
+Thus, net consumer harm to the Chattanooga market remains relatively small at \$6.6 million.
+This is in spite of post-merger changes in the HHIs of the disposal market of 1,896 and collection market of 620.
 
@@
 One of these is the assumption of the bargaining power parameter, $\lambda$.
-Sheu and Taragin (2021) shows that mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
-Larger wholesaler bargaining power offers more possibilities for EDM, as premerger input prices are likely to be high.
+\citeN{ST2021} shows that mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power.
+Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.
 This channel becomes less relevant as retailers gain more power.
@@
 \caption{The sensitivity of consumer welfare in Santek/Republic merger simulations to changes in the bargaining power parameter.
-%`All' changes the bargaining power paramtere for all non-vertically integrated firms.
+%`All' changes the bargaining power parameter for all non-vertically integrated firms.
 `Santek:Republic' changes the bargaining power parameter for just the Santek/Republic product.

### 2024-08-29 code/TrashSims.R

@@ -67,6 +67,23 @@ ownerPreUpMat=tcrossprod(ownerPreUpMat)
 ownerPreDownMat=model.matrix(~-1+ownerPreDown,data=simdata)
 ownerPreDownMat=tcrossprod(ownerPreDownMat)
 
+#disposal HHI
+hhiPreUp <- HHI(owner=ownerPreUpMat,shares=simdata$collection_paired_share)
+#collection HHI
+hhiPreDown <- HHI(owner=ownerPreDownMat,shares=simdata$collection_paired_share)
+#WM-ADS/Regional HHI Delta
+hhiDeltawmads_regional <- tapply(simdata$collection_paired_share*100,simdata$collection_firm_name,sum,na.rm=TRUE)
+hhiDeltawmads_regional <- 2*prod(hhiDeltawmads_regional[c("Regional","WM-ADS")])
+#WM-ADS/Santek HHI Delta
+hhiDeltawmads_santek <- tapply(simdata$collection_paired_share*100,simdata$collection_firm_name,sum,na.rm=TRUE)
+hhiDeltawmads_santek <- 2*prod(hhiDeltawmads_santek[c("Santek","WM-ADS")])
+#WasteConn/Santek HHI Delta
+hhiDeltaUpwasteconn_santek <- tapply(simdata$collection_paired_share*100,simdata$disposal_firm_name,sum,na.rm=TRUE)
+hhiDeltaUpwasteconn_santek  <- 2*prod(hhiDeltaUpwasteconn_santek [c("Santek","WasteConn")])
+hhiDeltaDownwasteconn_santek <- tapply(simdata$collection_paired_share*100,simdata$collection_firm_name,sum,na.rm=TRUE)
+hhiDeltaDownwasteconn_santek  <- 2*prod(hhiDeltaDownwasteconn_santek [c("Santek","WasteConn")])
+
+
 vertFirms <- intersect(simdata$ownerPreUp,simdata$ownerPreDown)
 
 for (v in vertFirms){

### 2024-08-30 code/TrashSims.R
fixed a multitude of issues with non-baseline fake merger sims. Specifically, except for instances where firm is vertically integrated pre-merger, I have set the pre- and post-merger bargaining power parameter equal to 0.73. I have also fixed pre- and post-merger marginal costs equal to the marginal costs from the Baseline. I have then recomputed pre-merger equilibrium prices based on ownership and bargaining power parameters.

@@ -685,7 +685,7 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
 
 
  mergersweep <- function(thissim,acquirer=c("Republic","Santek","WasteConn","WM-ADS"),
-                         target=c("Santek","WasteConn","WM-ADS","Regional"),
+                         target=c("Republic","Santek","WasteConn","WM-ADS","Regional"),
                          unintegrated=c("Base","Vertical","Up","Down"),
                          partial=FALSE){
 
@@ -702,7 +702,8 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
    if(unintegrated !="Base"){
 
    smallestBarg <- min(thissim@up@bargpowerPre)
-   thissim@up@bargpowerPre[thissim@up@ownerPre==thissim@down@ownerPre] <- smallestBarg
+   #thissim@up@bargpowerPre[thissim@up@ownerPre==thissim@down@ownerPre] <- smallestBarg
+   thissim@up@bargpowerPre <- rep(smallestBarg,length(thissim@up@bargpowerPre))
    thissim@up@bargpowerPost<-   thissim@up@bargpowerPre
    }
 
@@ -716,11 +717,10 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
      thissim@down@ownerPre <- paste0(thissim@down@ownerPre,"Ind")
      thissim@down@ownerPost <- paste0(thissim@down@ownerPost,"Ind")
      if(partial){
-       isIntegratedPre <- thissim@down@ownerPre == paste0(acquirer,"Ind")
-       isIntegratedPost <- thissim@down@ownerPost == paste0(acquirer,"Ind")
+       isIntegratedPost <- isIntegratedPre <- thissim@down@ownerPre == paste0(acquirer,"Ind") & thissim@up@ownerPre == acquirer
        thissim@down@ownerPre[isIntegratedPre] <- acquirer
        thissim@down@ownerPost[isIntegratedPost] <- acquirer
-       thissim@up@bargpowerPre[isIntegratedPre] <- 1
+       thissim@up@bargpowerPre[isIntegratedPre ] <- 1
        thissim@up@bargpowerPost[isIntegratedPost] <- 1
      }
    }
@@ -730,10 +730,7 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
[…]

### 2024-09-09 code/TrashSims.R
added new figure summarizing interesting variations of Republic/Santek merger.

@@ -899,9 +899,9 @@ ggsave("./output/trashfakemergeralt.png",trashfakemergerplotalt,height = 7,width
 trashinterestingmergerbar <- ggplot(data=filter(mkt_mergersweep,!(Acquirer=="Santek" & Target=="Republic" & Unintegrated == "Base") &
                                                   !(Acquirer=="WM-ADS" & Target=="Regional" & Unintegrated == "Partial Down") & #Unintegrated=="Base" &
                                                   name %in% c("Consumer Harm","Disposal Benefit","Collection Benefit")) %>% rename(Type=name) %>%
-                                      mutate(Merger=paste(Acquirer,Target,sep="/"),Unintegrated=factor(Unintegrated, levels=rev(c("Base","Vertical","Up","Partial Up","Down","Partial Down")) )) %>%
-                                      filter(Merger %in% c("Republic/Santek","Santek/Republic","Santek/WasteConn","Santek/WM-ADS", "WM-ADS/Regional")),aes(y=Unintegrated,x=value,fill=Type)) +
-  facet_grid(~Merger,scales = "free") +
+                                      mutate(Merger=paste(Acquirer,Target,sep="/"),Unintegrated=factor(Unintegrated, levels=c("Base","Vertical","Up","Partial Up","Down","Partial Down")) ) %>%
+                                      filter(Merger %in% c("Republic/Santek","Santek/Republic","Santek/WasteConn","Santek/WM-ADS", "WM-ADS/Regional")),aes(y=Merger,x=value,fill=Type)) +
+  facet_grid(~Unintegrated,scales = "free") +
   geom_bar(stat="identity",#fill="lightgrey",
                                                          position=position_dodge()
                                                          #position="stack"
@@ -912,12 +912,43 @@ trashinterestingmergerbar <- ggplot(data=filter(mkt_mergersweep,!(Acquirer=="San
             #hjust=1.1,
             vjust=0.9,
             color="black",
-   position=position_dodge(width=.9),size =3) + coord_flip()+
+   position=position_dodge(width=.9),size =3) +# coord_flip()+
   theme_bw() +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_y_discrete(limits=rev)
 
 ggsave(filename="./output/trashinterestingmergerbar.png",trashinterestingmergerbar,width = 6,height=4)
+
+
+
+
[…]

### 2024-09-10 code/TrashSims.R
improved look of interesting series of figures

@@ -897,12 +897,15 @@ ggsave("./output/trashfakemergeralt.png",trashfakemergerplotalt,height = 7,width
 
 
 trashinterestingmergerbar <- ggplot(data=filter(mkt_mergersweep,!(Acquirer=="Santek" & Target=="Republic" & Unintegrated == "Base") &
-                                                  !(Acquirer=="WM-ADS" & Target=="Regional" & Unintegrated == "Partial Down") & #Unintegrated=="Base" &
+                                                  !(Acquirer=="WM-ADS" & Target=="Regional" & Unintegrated == "Partial Down") & Unintegrated=="Base" &
                                                   name %in% c("Consumer Harm","Disposal Benefit","Collection Benefit")) %>% rename(Type=name) %>%
                                       mutate(Merger=paste(Acquirer,Target,sep="/"),Unintegrated=factor(Unintegrated, levels=c("Base","Vertical","Up","Partial Up","Down","Partial Down")) ) %>%
-                                      filter(Merger %in% c("Republic/Santek","Santek/Republic","Santek/WasteConn","Santek/WM-ADS", "WM-ADS/Regional")),aes(y=Merger,x=value,fill=Type)) +
-  facet_grid(~Unintegrated,scales = "free") +
-  geom_bar(stat="identity",#fill="lightgrey",
+                                      filter(Merger %in% c("Republic/Santek",#"Santek/Republic",
+                                                           "Santek/WasteConn","Santek/WM-ADS", "WM-ADS/Regional")),aes(y=Merger,x=value#,fill=Type
+                                                                                                                       )) +
+  #facet_grid(~Unintegrated,scales = "free") +
+  facet_grid(~Type,scales = "fixed") +
+  geom_bar(stat="identity",fill="lightgrey",
                                                          position=position_dodge()
                                                          #position="stack"
   )  +
@@ -927,13 +930,13 @@ trashinterestingsantrep <- ggplot(data=filter(mkt_mergersweep,Acquirer %in% c("S
                                                 !(Acquirer %in% c("Republic") & Target %in% c("Santek") & Unintegrated %in% c("Partial Up","Partial Down","Base")) ) %>% rename(Type=name) %>%
                                       mutate(Merger=ifelse(Acquirer %in% c("Republic") & Target %in% c("Santek"),paste(Acquirer,Target,sep="/"),
                                                            paste(Target,Acquirer,sep="/")),
[…]

### 2024-09-10 code/TrashSims.R
fixed ownership bug in downstream partial mergers

@@ -730,7 +730,7 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
      thissim@up@ownerPre <- paste0(thissim@up@ownerPre,"Ind")
      thissim@up@ownerPost <- paste0(thissim@up@ownerPost,"Ind")
      if(partial){
-       isIntegratedPost <-isIntegratedPre <- thissim@up@ownerPre == paste0(acquirer,"Ind") & thissim@down@ownerPre == acquirer
+       isIntegratedPost <-isIntegratedPre <- thissim@up@ownerPre == paste0(acquirer,"Ind") & thissim@down@ownerPre %in% c(acquirer,target)
        isIntegratedPost[thissim@up@ownerPre == paste0(acquirer,"Ind") & thissim@down@ownerPre == target] <- TRUE
        thissim@up@ownerPre[isIntegratedPre] <- acquirer
        thissim@up@ownerPost <- thissim@up@ownerPre
@@ -951,7 +951,7 @@ trashinterestingsantrep <- ggplot(data=filter(mkt_mergersweep,Acquirer %in% c("S
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_y_discrete(limits=rev) +ylab("Pre-merger Arrangement")
 
-ggsave(filename="./output/trashinterestingsantrep.png",trashinterestingsantrep,width = 6,height=4)
+ggsave(filename="./output/trashinterestingsantrep.png",trashinterestingsantrep,width = 7,height=4)
 
 mkt_mergersweep_firm$Product <- rownames(mkt_mergersweep_firm )
 mkt_mergersweep_firm$Product <- gsub("\\..*$","",mkt_mergersweep_firm$Product,perl=TRUE)

### 2024-09-10 code/TrashSims.R
fixed ownership bug in downstream partial mergers

@@ -717,9 +717,10 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
      thissim@down@ownerPre <- paste0(thissim@down@ownerPre,"Ind")
      thissim@down@ownerPost <- paste0(thissim@down@ownerPost,"Ind")
      if(partial){
-       isIntegratedPost <- isIntegratedPre <- thissim@down@ownerPre == paste0(acquirer,"Ind") & thissim@up@ownerPre == acquirer
+       isIntegratedPost <- isIntegratedPre <- thissim@down@ownerPre == paste0(acquirer,"Ind") & thissim@up@ownerPre %in% c(acquirer,target)
+       isIntegratedPost[thissim@down@ownerPre == paste0(acquirer,"Ind") & thissim@up@ownerPre == target] <- TRUE
        thissim@down@ownerPre[isIntegratedPre] <- acquirer
-       thissim@down@ownerPost[isIntegratedPost] <- acquirer
+       thissim@down@ownerPost <- thissim@down@ownerPre
        thissim@up@bargpowerPre[isIntegratedPre ] <- 1
        thissim@up@bargpowerPost[isIntegratedPost] <- 1
      }
@@ -927,7 +928,9 @@ ggsave(filename="./output/trashinterestingmergerbar.png",trashinterestingmergerb
 
 trashinterestingsantrep <- ggplot(data=filter(mkt_mergersweep,Acquirer %in% c("Santek","Republic") & Target %in% c("Santek","Republic") &
                                                   name %in% c("Consumer Harm","Disposal Benefit","Collection Benefit") &
-                                                !(Acquirer %in% c("Republic") & Target %in% c("Santek") & Unintegrated %in% c("Partial Up","Partial Down","Base")) ) %>% rename(Type=name) %>%
+                                                !(Acquirer %in% c("Republic") & Target %in% c("Santek") & Unintegrated %in% c("Partial Up","Partial Down","Base"))
+                                                #Acquirer %in% c("Republic") & Target %in% c("Santek")
+                                                ) %>% rename(Type=name) %>%
                                       mutate(Merger=ifelse(Acquirer %in% c("Republic") & Target %in% c("Santek"),paste(Acquirer,Target,sep="/"),
                                                            paste(Target,Acquirer,sep="/")),
                                                            Unintegrated=reorder(Unintegrated,-1*value*(Type=="Consumer Harm"))#,

### 2024-09-10 code/TrashSims.R
fixed ANOTHER ownership bug in downstream partial mergers

@@ -731,9 +731,9 @@ compareplot_noup <- ggplot(data=  filter(compare,Type!="Pre-merger" & !(Level=="
      thissim@up@ownerPre <- paste0(thissim@up@ownerPre,"Ind")
      thissim@up@ownerPost <- paste0(thissim@up@ownerPost,"Ind")
      if(partial){
-       isIntegratedPost <-isIntegratedPre <- thissim@up@ownerPre == paste0(acquirer,"Ind") & thissim@down@ownerPre %in% c(acquirer,target)
+       isIntegratedPost <-isIntegratedPre <- thissim@up@ownerPre == paste0(acquirer,"Ind") & thissim@down@ownerPre %in% c(acquirer)
        isIntegratedPost[thissim@up@ownerPre == paste0(acquirer,"Ind") & thissim@down@ownerPre == target] <- TRUE
-       thissim@up@ownerPre[isIntegratedPre] <- acquirer
+       thissim@up@ownerPre[isIntegratedPost] <- acquirer
        thissim@up@ownerPost <- thissim@up@ownerPre
        thissim@up@bargpowerPre[isIntegratedPre] <- 1
        thissim@up@bargpowerPost[isIntegratedPost] <- 1

### 2024-09-11 code/TrashSims.R
changed WM-ADS to WMADS and now referto all product pairs with a hyphen rather than a slash.

@@ -21,7 +21,7 @@ simdata <- filter(paired_data_4_22,collection_paired_share>0) %>%
   #mutate(disposal_volume=ifelse(disposal_firm_name != collection_firm_name,mean(disposal_volume[disposal_firm_name != collection_firm_name]),disposal_volume)) %>%
   ungroup() %>%
   mutate(#collection_paired_share=collection_paired_share/sum(collection_paired_share),
-         collection_firm_name=gsub("_","-",collection_firm_name),
+         collection_firm_name=gsub("_","",collection_firm_name),
          collection_paired_share=disposal_volume/sum(disposal_volume),
          disposal_dollar_margin=ifelse(disposal_firm_name=="Republic",
                                        disposal_percent_margin*disposal_price_republic,
@@ -71,12 +71,12 @@ ownerPreDownMat=tcrossprod(ownerPreDownMat)
 hhiPreUp <- HHI(owner=ownerPreUpMat,shares=simdata$collection_paired_share)
 #collection HHI
 hhiPreDown <- HHI(owner=ownerPreDownMat,shares=simdata$collection_paired_share)
-#WM-ADS/Regional HHI Delta
+#WMADS/Regional HHI Delta
 hhiDeltawmads_regional <- tapply(simdata$collection_paired_share*100,simdata$collection_firm_name,sum,na.rm=TRUE)
-hhiDeltawmads_regional <- 2*prod(hhiDeltawmads_regional[c("Regional","WM-ADS")])
-#WM-ADS/Santek HHI Delta
+hhiDeltawmads_regional <- 2*prod(hhiDeltawmads_regional[c("Regional","WMADS")])
+#WMADS/Santek HHI Delta
 hhiDeltawmads_santek <- tapply(simdata$collection_paired_share*100,simdata$collection_firm_name,sum,na.rm=TRUE)
-hhiDeltawmads_santek <- 2*prod(hhiDeltawmads_santek[c("Santek","WM-ADS")])
+hhiDeltawmads_santek <- 2*prod(hhiDeltawmads_santek[c("Santek","WMADS")])
 #WasteConn/Santek HHI Delta
 hhiDeltaUpwasteconn_santek <- tapply(simdata$collection_paired_share*100,simdata$disposal_firm_name,sum,na.rm=TRUE)
 hhiDeltaUpwasteconn_santek  <- 2*prod(hhiDeltaUpwasteconn_santek [c("Santek","WasteConn")])
[…]

### 2024-09-30 doc/TrashData.tex
minor edits to get the code to run on Board machines. These figures use a different strategy for fixing either upstream or downstream pre-merger prices in the antitrust package than before. However, the results look the same as the first method.
Incorporated Gloria's 11/2 edits to document and TrashData.tex. First Full run of simulations on FRB machine. inter-quartile ranges line up almost perfectly, min and maxes are very close. Examined pictures and they look the same as well. Fixed minor issues with extra row in sumtable.
reran trashsims to accomodate changes for JIE submission.
modified pi

@@
-\begin{table}
-
-\caption{\label{tab:trashdata}Republic/Santek Merger Simulation Inputs. Prices and margins are reported in dollars, while volume is reported in thousands of pounds.}
+\caption{\label{tab:trashdata}Republic/Santek Merger Simulation Inputs. Volume is reported in thousands of pounds, while prices and margins are reported in dollars.}
 \centering
@@
 \toprule
-Disposal Firm & Collection Firm & Volume & Disposal Price & Disposal Margin & Collection Margin & Collection Price\\
+Disposal Firm & Collection Firm & Volume & Disposal Price & Disposal Margin & Collection Margin & Collection Cost\\
 \midrule
-Republic & Republic & 165 & 42 & 0 & 44 & 179\\
+Republic & Republic & 165 & 42 & 0 & 44 & 93\\
 \cmidrule{1-7}
- & Republic & 34 & 36 & 20 &  & 173\\
+ & Republic & 34 & 36 & 20 &  & 93\\
 
- & Santek & 218 & 16 & 0 &  & 134\\
+ & Santek & 218 & 16 & 0 &  & 74\\
 
- & WM-ADS & 30 & 36 & 20 &  & 221\\
+ & WMADS & 30 & 36 & 20 &  & 148\\
 
-\multirow{-4}{*}{\raggedleft\arraybackslash Santek} & Regional & 30 & 36 & 20 &  & 221\\
+\multirow{-4}{*}{\raggedleft\arraybackslash Santek} & Regional & 30 & 36 & 20 &  & 148\\
 \cmidrule{1-7}
- & Republic & 19 & 25 & 14 &  & 162\\
+ & Republic & 19 & 25 & 14 &  & 93\\
 
- & WasteConn & 48 & 11 & 0 &  & 110\\
+ & WasteConn & 48 & 11 & 0 &  & 63\\
 
- & WM-ADS & 17 & 25 & 14 &  & 210\\
+ & WMADS & 17 & 25 & 14 &  & 148\\
 
-\multirow{-4}{*}{\raggedleft\arraybackslash WasteConn} & Regional & 17 & 25 & 14 &  & 210\\
+\multirow{-4}{*}{\raggedleft\arraybackslash WasteConn} & Regional & 17 & 25 & 14 &  & 148\\
 \bottomrule
 \end{tabular}}
-\end{table}

### 2024-09-30 code/TrashSims.R

@@ -471,7 +471,9 @@ print(kable(select(simdata,disposal_firm_name,
                    disposal_price,
                    disposal_dollar_margin,
                    collection_dollar_margin,
-                   priceDown_repmargin_2nd
+                   #priceDown_repmargin_2nd
+                   `collection cost`
+
 ) %>%
 
               rename("Disposal Firm"=disposal_firm_name,
@@ -480,7 +482,8 @@ print(kable(select(simdata,disposal_firm_name,
                      "Disposal Price"=disposal_price,
                      "Disposal Margin"=disposal_dollar_margin,
                      "Collection Margin"=collection_dollar_margin,
-                     "Collection Price"=priceDown_repmargin_2nd
+                     #"Collection Price"=priceDown_repmargin_2nd
+                     "Collection Cost"=`collection cost`
                      ) %>%
   mutate(`Volume`=`Volume`/1e3) %>%
   mutate(across(where(is.numeric),round)),

### 2024-09-30 doc/JIE/JINDECMS2023138-1Response.tex
incorporates Gloria's edits and reqrites in response to JIE R&R.

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Preliminaries
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title{Detailed Responses to the Referees for \\ ``Beyond Horizontal and Vertical'' \\MS 2023138}
+\author{Margaret Loudermilk, Gloria Sheu, and Charles Taragin}
+\date{October 2024}
 \begin{document}
+\maketitle
 
-\section{Response to Editor}
+We appreciate the opportunity to revise our paper, and we have made substantial changes in response to the suggestions from the editor and referees, which we found quite helpful.
+The largest change is that we have removed the results for sampled mergers and now instead focus on the Republic/Santek merger example.
+In this document, the editor and referee comments are in italics, followed by our response.
 
-\subsection{Comments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Editor
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Response to the Editor}
+\emph{My biggest concern is that I don’t understand why the range (much less distribution) of parameters you consider is interesting.
+Therefore, I propose you combine Sections 3 and 4 using the Republic/Santek merger as a ``central'' point of calibration and exploring perturbations of that point to establish how changes to the underlying model assumptions impact outcomes of interest.}
 
-\mypara Incidentally, this section imposes a logit assumption in (1), but
-is this really needed before turning to the simulation exercises in Sections 3 and 4?
+We understand the concern about the sampling distribution of our merger simulations in the previous draft of the paper.
+In response, we took your suggestion of combining our simulation exercises into one section centered on the Republic/Santek example.
+This is now Section 3 of the paper.
 
-\mypara In the end, I think R2 set a good bar for the simulations: the
-point of doing the simulation [is] to try and provide some qualitative results about the
-direction/nature of the net effect of the various countervailing forces and what this depends on?
-If you can convincingly deliver on this, then the paper can inform practitioners on the key
-elements to be sure to precisely capture in a merger simulation exercise.
+\emph{In particular, I will be looking for you to deliver on the following:}
+\begin{enumerate}
+\item  \emph{Be very clear on how the central point is parameterized/calibrated.
+Be clear about what data is used to pin down what parameters.
+Also note what parameters you will vary either because they are not well identified with available data or because you want to consider sensitivity.}
 
-\mypara When presenting your results, please consider whether the simulations should be presented as the
-result of a statistical sampling (as is done now) or more in the vein of counterfactual analysis
-where you vary one parameter at a time. The argument for the statistical sampling approach is
-that parameters may interact---although you haven?t clearly established that in most of the
-current results. The argument for the one parameter at a time approach is that it can be more
-closely tied to the institution presented in the theory section. If you are tempted to do the
-sampling approach because there are many parameters to vary, consider whether you should
-instead focus the reader?s attention on the most important/illustrative results.
+We have created new Section 3.2 that explains our calibration.
+We tie our explanation directly to the data inputs, which appear in Table 1, and to the set of equations we use for identification, which are the firm first order conditions in Section 2.
+We have considered in Section 3.3 the sensitivity of our results to the share of the outside good/market elasticity, to adding a nesting parameter, and to the size of the bargaining parameter, which we discuss further below.
 
-\subsection{TO DOs}
-\setcounter{para}{0}
-\mypara Be very clear on how the central point is parameterized/calibrated. Be clear about what
-data is used to pin down what parameters. Also note what parameters you will vary
-either because they are not well identified with available data or because you want to
-consider sensitivity.
-\mypara Be sure to compare simulations that allow for the complex nature of Republic/Santek
-versus the more vanilla simulations considered in ST (2021). This is important to capture
-this papers contribution.
-\mypara Be sure to consider how the nature of bargaining and downstream diversion will
-impact the results. For the later, you would need to go beyond a multinational logit
-model, but a nested logit should be able to capture the intuition. I'm not asking you to
-estimate substitution in this context since it doesn?t seem like you have the data to do so
-(although I'd love to be shown wrong on this point!). I do want to know whether varying
-substitution patterns which fit the data you do have are likely to alter the qualitative
-outcomes of the merger simulation.
+\item \emph{Be sure to compare simulations that allow for the complex nature of Republic/Santek versus the more vanilla simulations considered in ST (2021).
+This is important to capture this papers contribution.}
 
+This was an extremely useful comment in helping us focus our simulations on the most interesting cases.
+We have added a new Section 3.5 that walks through examples of complex mergers alongside similar ``non-complex'' mergers like those in ST (2021).
+For example, we show how our results change when moving from a pure vertical merger to a merger where both firms are already integrated and from a pure horizontal merger to a merger where one of the firms is already integrated.
+This allows us to, for example, show how layering in additional vertical aspects to a horizontal merger can impact welfare.
+
+\item \emph{Be sure to consider how the nature of bargaining and downstream diversion will impact the results.
+For the later, you would need to go beyond a multinational logit model, but a nested logit should be able to capture the intuition.
+I'm not asking you to estimate substitution in this context since it doesn't seem like you have the data to do so (although I'd love to be shown wrong on this point!).
+I do want to know whether varying substitution patterns which fit the data you do have are likely to alter the qualitative outcomes of the merger simulation.}
+
+We have added three new sensitivity analyses, at the end of Section 3.3, in order to address this point.
+We show how our simulated consumer welfare results are impacted by (1) the extent of substitution outside the market, (2) the use of the logit versus the nested logit, and (3) the bargaining power of retailers vesus wholesalers.
+In each of these exercises, we vary the relevant parameter in smooth increments and show how welfare changes relative to the baseline.
+The other aspects of the simulations are held fixed, allowing for cleaner comparisons.
+We find that our results change in an intuitive manner that we believe will help practitioners understand the impact of the key parameters on the model's conclusions.
+
+As part of this change, we removed the logit demand function from the theory discussion in Section 2, to make clear that the model can accommodate multiple forms of demand.
+We introduce our baseline logit assumption and the alternative nested logit assumption later in Section 3 when we come to our specific empirical application.
+
+\end{enumerate}
+
+\emph{When presenting your results, please consider whether the simulations should be presented as the result of a statistical sampling (as is done now) or more in the vein of counterfactual analysis where you vary one parameter at a time.
+The argument for the statistical sampling approach is that parameters may interact---although you haven't clearly established that in most of the
+current results.
+The argument for the one parameter at a time approach is that it can be more closely tied to the institution presented in the theory section.
+If you are tempted to do the sampling approach because there are many parameters to vary, consider whether you should instead focus the reader's attention on the most important/illustrative results.}
+
+As mentioned in our responses above, we have shifted away from the statistical sampling approach.
+In the sensitivity analysis, we vary one parameter at a time to facilitate comparisons.
+In comparing across different mergers, we have chosen a set of illustrative cases that give the reader concrete examples to focus on.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% R1
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Response to Referee 1}
-\setcounter{para}{0}
-\mypara I hope that the authors provide the code so that researchers can (a) run their own simulations
-and (b) use the code for their own research. The authors should follow the AER/ReStud
-guidelines that are becoming standard in the profession
-( \url{https://aeadataeditor.github.io/aea-de-guidance/preparing-for-data-deposit.html} ). Basically,
-make the code easily readable and post it to a code/data archive website like Zenodo (and link
-to it from the author's webpage and the paper).
-Note: Footnote 16 states that the antitrust R package contains the code to do the
-simulations. I could not easily find the relevant parts of the package. It would be great if this
-machinery was also added to antitrust , but I would like to see a stand-alone replication
-package.
-\mypara I think the authors could ditch the Republic/Santek example. The authors admit that the
-example is for ``illustrative" purposes. Instead of a single example, the authors could choose a
-few representative mergers and look at specific fake data that mimics those cases. Eg. many
-upstream and few downstream or vise versa.
-\mypara There is one thing I really like about Section 4: the comparison between a naive merger
-analysis (eg. downstream/upstream horizontal effects only) and the more comprehensive
-analysis that includes all of the vertical effects. I would like to see this type of analysis in Section
 
-\mypara I think that the authors overclaim a bit. In the abstract (and this sentiment is repeated in the
-paper, for example on page 2) the authors: ``In our model, we find that mergers with both
-horizontal and vertical characteristics typically harm consumers."
-First, I think the more important contribution is to point out that a naive analysis that ignores the
-vertical aspect of a merger can mis-quantify the effects of the merger (see point 3 above) or that
-the EDM benefits are typically small (a point already made).
-Second, mergers arise in nature (and are prosecuted) endogenously. The median pre-merger
-HHI in the simulations is between 2800 and 4300 and the median Delta HHI is over 1000. Most
-of the mergers analyzed would be presumed harmful based on the classic Cournot intuition or
-the more recent Nocke/Whinston intuition. I think the authors should avoid statements like in
-their abstract that make it seem they are drawing generic conclusions about merger
-enforcement.
-Smaller Comments
-\mypara  Table 1: It seems a bit suspicious that the price quantiles are all integers. Maybe I am missing
-something, but the domain should be some subset of the real number line, making integers very
-unlikely.
-\mypara  Simulation details. I don't understand why shares are drawn from a Dirichlet distribution (or
-any distribution), Shares are endogenous functions of model primitives. Do you mean that you
-sampled $\delta$ from a distribution?
-\mypara Simulation details. It might make more sense to calibrate to a market elasticity (for example
--1) instead of the way it is done in the paper.
-\mypara  It might make sense if a retailer and manufacturer had meaning in the model. One suggestion
-is to have $\delta$ be comprised of two components, and manufacturer effect, and a retailer effect,
-where the retailer effect is perfectly correlated across manufacturer products within a retailer,
-and the manufacturer effect is perfectly correlated across retailer for the same product. So
-something like  $\delta_{rw} = \delta_{r} + \delta_{w}$
-\mypara  In Section 2.2 (first paragraph), referencing a particular retailer with index r is confusing
-because r is the generic index for a retailer.
-\mypara In Equation 4, I think there is an additional effect in the first term that Luco and Marshall
-(2020) would call the Edgeworth-Salinger effect.
-\mypara  The authors mention on multiple occasions that a retailer could cease to trade with a
-wholesaler (eg. page 8). I don't see how this is a possibility in the model.
-\mypara  I don't think ``Vertical" and ``Horizontal" need to be quoted in the title.
+\begin{enumerate}
 
+    \item \emph{I hope that the authors provide the code so that researchers can (a) run their own simulations and (b) use the code for their own research.
+The authors should follow the AER/ReStud guidelines that are becoming standard in the profession (\footnotesize{\url{https://aeadataeditor.github.io/aea-de-guidance/preparing-for-data-deposit.html}}).
+Basically, make the code easily readable and post it to a code/data archive website like Zenodo (and link to it from the author's webpage and the paper).}
+
+    \emph{Note: Footnote 16 states that the antitrust R package contains the code to do the simulations.
+I could not easily find the relevant parts of the package.
+It would be great if this machinery was also added to antitrust , but I would like to see a stand-alone replication package.}
+
+     Much of the coding machinery we used has already been incorporated into the next version of the \texttt{antitrust} R package.
+A new function, \texttt{vertical.barg}, has been added which uses observed shares, prices and some margins to calibrate demand parameters and simulate the effect of complex mergers.
+We understand that having a replication package would also be helpful for readers, and we are planning to post a package once the paper has been finalized.
+Aside from containing all the code and data needed to run all the simulations presented in the paper, the replication package will contain code that allows calibration when downstream costs but not prices are observed.
+
+    \item \emph{I think the authors could ditch the Republic/Santek example.
+The authors admit that the example is for ``illustrative" purposes.
+Instead of a single example, the authors could choose a few representative mergers and look at specific fake data that mimics those cases.
+Eg.
+many upstream and few downstream or vise versa.}
+
+    Based on feedback from the editor, we ended up removing the other simulations and focusing on the Republic/Santek example.
+However, in the spirit of this comment, we chose a few representative mergers within the Republic/Santek context to provide the reader with additional intuition.
+
+    \item \emph{There is one thing I really like about Section 4: the comparison between a naive merger analysis (eg.
+downstream/upstream horizontal effects only) and the more comprehensive analysis that includes all of the vertical effects.
+I would like to see this type of analysis in Section 3.}
+
+    Thank you for your feedback.
+We have kept that comparison with the downstream- and upstream-only models, and have moved it into Section 3 to go with the additional examples we have added for Republic/Santek.
+
+    \item \emph{I think that the authors overclaim a bit.
+In the abstract (and this sentiment is repeated in the paper, for example on page 2) the authors: ``In our model, we find that mergers with both horizontal and vertical characteristics typically harm consumers."}
+
+    \emph{First, I think the more important contribution is to point out that a naive analysis that ignores the vertical aspect of a merger can mis-quantify the effects of the merger (see point 3 above) or that the EDM benefits are typically small (a point already made).}
+
+    \emph{Second, mergers arise in nature (and are prosecuted) endogenously.
+The median pre-merger HHI in the simulations is between 2800 and 4300 and the median Delta HHI is over 1000.
+Most of the mergers analyzed would be presumed harmful based on the classic Cournot intuition or the more recent Nocke/Whinston intuition.
+I think the authors should avoid statements like in their abstract that make it seem they are drawing generic conclusions about merger enforcement.}
+
+    We understand this concern and have now removed those generic-sounding statements.
+Any statements we make now are tied to the specific Republic/Santek example, and we have tried to make that clear in the text.
+
+\textbf{Smaller Comments}
+
+    \item \emph{Table 1: It seems a bit suspicious that the price quantiles are all integers.
+Maybe I am missing something, but the domain should be some subset of the real number line, making integers very unlikely.}
+
+    This occurred because we rounded all those numbers to integers.
+Sometimes we make choices in the name of clarity that backfire.
+That table has now been removed, since those simulations no longer appear in the paper.
+
+    \item \emph{Simulation details.
+I don't understand why shares are drawn from a Dirichlet distribution (or any distribution), Shares are endogenous functions of model primitives.
+Do you mean that you sampled $\delta$ from a distribution?}
+
+    We no longer include the simulations based on sampling, so the Dirichlet distribution is not used in this draft.
+To clarify on the use of market shares (which we do still use to calibrate the Republic/Santek exercise), we agree that they are endogenous to the model but they are also the type of data that practitioners and researchers would observe about a market.
+Therefore, shares are commonly used for calibration.
+It is helpful to draw the analogy from the structural estimation context: in recovering the parameters for a structural demand system, an econometrician usually uses variation in market shares and prices (often along with other data for instrumental variables) in order to estimate the model, even though the shares and prices are also functions of the model.
+For antitrust practitioners calibrating a model, they also use shares and prices.
+In both cases, we are trying to find parameters of the model that are consistent with the market outcomes that we observe in the data.
+We have provided some of this background in our calibration discussion in Section 3.2.
+
+    \item \emph{Simulation details. It might make more sense to calibrate to a market elasticity (for example -1) instead of the way it is done in the paper.}
+
+    Since we have removed the simulations that were formerly in Section 3, this part of the calibration is no longer used.
+In the Republic/Santek example, we now start by assuming no substitution outside of the waste management market (due to the strict government regulations mandating that commercial customers dispose of waste in proper landfills), and then vary the market elasticity in a sensitivity check that appears in Section 3.3.
+
+    \item  \emph{It might make sense if a retailer and manufacturer had meaning in the model.
+One suggestion is to have $\delta$ be comprised of two components, and manufacturer effect, and a retailer effect, where the retailer effect is perfectly correlated across manufacturer products within a retailer, and the manufacturer effect is perfectly correlated across retailer for the same product.
+So something like  $\delta_{rw} = \delta_{r} + \delta_{w}$}
+
+    We tried something in this spirit in our nested logit sensitivity exercise for the Republic/Santek simulations.
+We did one version where we grouped together all the products that use the same firm for collection and disposal (``integrated'' in Figure 4) and another version where we grouped together all products that use the same landfill (``landfill'' in Figure 4).
+
+    \item  \emph{In Section 2.2 (first paragraph), referencing a particular retailer with index r is confusing because r is the generic index for a retailer.}
+
+    We have shifted to referring to the merging firms in Section 2.2 using index $t$ for the retailer and $v$ for the wholesaler.
+
+    \item \emph{In Equation 4, I think there is an additional effect in the first term that Luco and Marshall (2020) would call the Edgeworth-Salinger effect.}
+
+    This is a good point.  We have added a footnote to the discussion of that equation (now equation ([8 XXXX])) mentioning the Salinger (1991) result and the Luco and Marshall (2020) paper.
+
+    \item \emph{The authors mention on multiple occasions that a retailer could cease to trade with a wholesaler (eg. page 8). I don't see how this is a possibility in the model.}
+
+    We agree that this is confusing and have removed those statements.
+We now say near the beginning of Section 2.1 that we are taking the observed set of retailer-wholesaler combinations offered as given.
+
+    \item  \emph{I don't think ``Vertical" and ``Horizontal" need to be quoted in the title.}
+
+    We have removed the quotation marks from the title.
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% R2
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Response to Referee 2}
-I recommend that the paper be shortened in two ways. First I think that it should report
-the results for the simulations of complex mergers but should omit the results for the simulations
-of mergers where the number of pre-existing rivals (that do not participate in the merger)
-changes. That is, I think the authors should only include simulation results for the case when
-there are no pre-existing vertically integrated rivals that do not participate in the merger and
-report that the nature of the results does not change substantially when pre-existing vertically
-integrated rivals that do not participate in the merger are allowed to exist. In addition to
-shortening the paper, this will allow the authors to provide a simpler and much less confusing
-description of the set of examples they consider in their simulations. Second, I think that the
-material on generalizing the results to the case of increasing marginal costs should be omitted. I
-think this a somewhat separate issue that deserves separate treatment if substantial enough results
-can be derived.
+\emph{I recommend that the paper be shortened in two ways.
+First I think that it should report the results for the simulations of complex mergers but should omit the results for the simulations of mergers where the number of pre-existing rivals (that do not participate in the merger) changes.
+That is, I think the authors should only include simulation results for the case when
+there are no pre-existing vertically integrated rivals that do not participate in the merger and report that the nature of the results does not change substantially when pre-existing vertically integrated rivals that do not participate in the merger are allowed to exist.
+In addition to shortening the paper, this will allow the authors to provide a simpler and much less confusing description of the set of examples they consider in their simulations.
+Second, I think that the material on generalizing the results to the case of increasing marginal costs should be omitted.
+I think this a somewhat separate issue that deserves separate treatment if substantial enough results can be derived.}
+
+Thank you for this helpful feedback.  We have removed both sets of material, pivoted the draft to focusing on the Republic/Santek merger, which has greatly streamlined the paper.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% R3
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Response to Referee 3}
-\setcounter{para}{0}
-\mypara I could not find a contribution that would merit publication at the Journal of Industrial Economics.
+\begin{enumerate}
+    \item \emph{I could not find a contribution that would merit publication at the Journal of Industrial Economics.
 The paper is very similar to Sheu and Taragin (2021).
 The model and simulation are essentially the same.
-At times, the paper reads as an appendix of the previous article.
+At times, the paper reads as an appendix of the previous article.}
 
-\mypara  The paper claims to simulate the welfare effects in a new set of mergers called ``complex mergers." Complex mergers are standard vertical and horizontal mergers where a firm is vertically integrated.
+    We agree that we could have been clearer in our last draft about the distinction between this paper and our previous research.
+We believe that removing the old Section 3 and changing the focus to the Republic/Santek merger helps address this issue.
+In the new Section 3, we show how moving from pure vertical and pure horizontal mergers (as appear in Sheu and Taragin (2021)) to complex mergers where one or both of the firms are already integrated prior to the merger (new for this paper) impacts the results.
+
+    \item \emph{The paper claims to simulate the welfare effects in a new set of mergers called ``complex mergers." Complex mergers are standard vertical and horizontal mergers where a firm is vertically integrated.
 They have been previously studied by Ho and Lee (2017) and Crawford et al.
 (2018), among others.
-(The related literature also needs to be improved.) So, the paper uses a stylized model to simulate standard mergers.
+(The related literature also needs to be improved.) So, the paper uses a stylized model to simulate standard mergers.}
 
-\mypara  Some of the modeling assumptions are very restrictive.
+    We agree that there has been past research on complex mergers, but we also believe that these mergers have been understudied relative to pure vertical and pure horizontal transactions.
+Furthermore, we also believe that research showing how a more complicated structural model can be simplified for use by practitioners, keeping in mind the data and time constraints they face, is important.
+We have personal experience with being told Crawford et al.
+(2018) was too complicated to be used as part of the court testimony we were preparing in a merger litigation.
+
+    We have added to the literature review in Section 1 to make clear that there are past papers that have studied complex mergers, including the two you mention above.
+We also have added a discussion of earlier theory literature on vertical integration to provide more context for our paper.
+
+    \item \emph{Some of the modeling assumptions are very restrictive.
 For instance, the logit demand system, simultaneous bargaining-Bertrand game, and the linear fees in the bargaining model.
@@
 One would expect consumers self-selecting into the firms' types, which would affect the compensating variation computation.
-To the very least, random coefficients on price and on whether the firm is vertically integrated are necessary to obtain more credible results.
+To the very least, random coefficients on price and on whether the firm is vertically integrated are necessary to obtain more credible results.}
 
-\mypara  The RRC discussion was perplexing.
+    We have removed the discussion of logit demand from the theory explanation in Section 2, to make clear that it is not a necessary part of the model.
+In our merger simulations in Section 3.3, we present both logit and nested logit results, the latter with one option where all the integrated products are in the same nest separate from the non-integrated products (in a similar idea to what you suggest in your comment) and a second option where products that use the same landfill are in the same nest.
+We agree that the logit assumption is restrictive, but we also note that it is commonly used in antitrust contexts, as the logit fits with the antitrust emphasis on market shares.
+We have added a discussion of these issues in Section 3.2.
+
+    \item \emph{The RRC discussion was perplexing.
 What the paper calls ``RRC" differs from the classic RRC in the literature in Salop and Scheffman (1983).
@@
 Please include the profit functions and derivations in the Appendix (e.g.
-equation 5 and subsequent).
+equation 5 and subsequent).}
 
-\mypara  The simulation exercise in Section 3 does not implement the simulation of the model in Section 2.
+    We understand the confusion around this point.
+Our meaning of the term agrees with Crawford et al.
+(2018).
+We use the term ``RRC'' in the way that they use it, to mean a shift in bargaining leverage that allows a vertically integrated firm to raise input prices to a downstream rival.
+Crawford et al.
+(2018) refer to this as ``raising rivals' cost.'' See, for example, their discussion in Section 6.2.2, pages 934-936, of their paper.
+Footnote 62 in their paper explains that this RRC effect is not motivated by the likelihood that downstream firms will raise their retail prices in response to higher wholesale prices.
+The behavior is instead motivated by the possibility that consumers will switch who they purchase from, at fixed prices.
+Additional discussion is provided in their footnote 36.
+We and Crawford et al.
+(2018) do not use RRC to mean the incentive modeled by Salop and Scheffman (1983), where a dominant firm chooses an action in order to directly impact downstream retail prices.
+We have added footnote [10 XXXX] in the discussion around equation ([9 XXXX]) to clarify this distinction.
+
+    We have included the profit functions in Section 2.1 and added an explanation for how they appear in the firm first order conditions.
+
+    \item \emph{The simulation exercise in Section 3 does not implement the simulation of the model in Section 2.
 What is the mapping between the two?
@@
 For instance, how do you ensure downstream (upstream) prices are consistent with equation 2 (3)?
-For these reasons, I could not follow the details of the simulation exercise.
+For these reasons, I could not follow the details of the simulation exercise.}
 
-\mypara Finally, I found the calibration exercise in Section 4 hard to follow.
+    We have removed those simulations from the paper.
+All the simulations and calibration we have done rely on the first order conditions of the model in Section 2, to ensure that the resulting prices and market share are consistent with the theory.
+We have added a discussion clarifying this point in Section 3.2.
+
+    \item \emph{Finally, I found the calibration exercise in Section 4 hard to follow.
 At times it reads as if the paper is collecting data.
@@
 (1990) and the references therein.
-I found the analysis in this section not credible.
+I found the analysis in this section not credible.}
+
+    We have designed this exercise to closely mirror the types of data and calibration exercises that are frequently used by antitrust agencies and practitioners in merger review.
+We have added an explanation of this purpose at the beginning of Section 3.
+
+    The DOJ complaint in Republic/Santek at paragraph 47 explains that the DOJ was alleging harm due to increased upstream disposal prices to rival downstream collection firms, not due to total exclusion or foreclosure of these collection firms.
+Specifically, the complaint states that the merged firm would have an ``incentive to raise the MSW disposal costs of independent haulers because Republic---no longer confronting competition from Santek in SCCW collection---would capture more of the business lost by independent haulers in the SCCW collection market.''  We agree that the complaint does not put this story in terms of bargaining, but the incentive discussed in this passage is very similar to the mechanism that leads to increased bargaining leverage and hence higher wholesale prices in our model.
+The knowledge that downstream customers have fewer independent rivals from which to purchase collection is what allows the merged firm to negotiate for a higher upstream disposal rate.
+
+\end{enumerate}
 

### 2024-09-30 doc/VerticalArrangements.tex
incorporates Gloria's edits and reqrites in response to JIE R&R.

--- 2024-08-19 doc/VerticalArrangements.tex
@@
 Department of Justice.
-This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}
+This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson, along with seminar participants at the U.S.
+Department of Justice and the Federal Trade Commission.}}
 
@@
 \author{Margaret Loudermilk\footnote{margaret.loudermilk@usdoj.gov} \\  U.S Department of Justice \and Gloria Sheu\footnote{gloria.sheu@frb.gov.
-} \\ Federal Reserve Board \and  Charles Taragin\footnote{charles.s.taragin@frb.gov (corresponding author).}  \\  Federal Reserve Board }
+} \\ Federal Reserve Board \and  Charles Taragin\footnote{charles.s.taragin@frb.gov (corresponding author)}  \\  Federal Reserve Board }
 
-\date{\today}
+%\date{\today}
+\date{October 2024}
 
@@
 \begin{abstract}
-\noindent We study the simulation of mergers that have both horizontal and vertical aspects.
-Our model features logit Bertrand competition downstream and Nash Bargaining upstream.
-We apply the model to the Republic/Santek merger as a real-world example.
-%We numerically simulate four merger types: vertical mergers between an unintegrated retailer and an unintegrated wholesaler, downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler, upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler, and integrated mergers between two integrated retailer/wholesaler pairs.
-We apply the model to the Republic/Santek merger as a real-world example.
+\noindent We study simulated mergers that have both horizontal and vertical aspects using a model with Bertrand competition downstream and Nash Bargaining upstream.
+We apply this framework to the Republic/Santek merger in waste management services, comparing the actual merger to different counterfactual mergers involving various vertical and horizontal combinations.
+Using this example, we find almost no instances where the elimination of double marginalization outweighs merger harms to consumers.
+Furthermore, we show how analyzing the upstream and downstream markets separately can lead to inaccurate results.
 
@@
 Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely horizontal or purely vertical.
-This is particularly true for the mergers involving large firms that usually attract the most scrutiny.
-However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.
+This is especially true for the mergers involving large firms that attract the most scrutiny.
+However, for both historical and practical reasons, antitrust research and policy analysis has generally maintained this dichotomy.
 Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
 As a result, the antitrust community has done little research into ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects.
+We have little guidance for how our merger simulation models and related tools apply to these complex mergers.
 
 The horizontal versus vertical distinction had been enshrined in U.S.
-Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The most recent Horizontal Merger Guidelines is from 2010, while the most recent Vertical Merger Guidelines is from 2020.
-They have both been recently superseded by a joint document.
-}  Given these policy documents, it is not surprising that the agencies have historically framed most mergers as either horizontal or vertical.
-Similarly, most antitrust research maintains this separate categorization, leaving us with little guidance for how our merger simulations and related tools apply to complex mergers.
-However, in recognition of these issues, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical aspects of competition.
-As part of the revision effort, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
+Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The most recent Horizontal Merger Guidelines are from 2010, while the most recent Vertical Merger Guidelines are from 2020.
+They have both been superseded by a joint document issued in 2023.
+}  However, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical aspects of competition.
+As part of the revision effort, these agencies requested public comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
 in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.
@@
 Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition?
-If so, how?''  }  In this paper, we address this question by exploring the implications that a typical merger simulation model has for the welfare effects of complex mergers.
+If so, how?''  }  In this paper, we address this question by exploring the implications that a standard merger simulation model has for the welfare effects of complex mergers.
 
-%We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
-Relative to \citeN{ST2021}, the first category, vertical mergers, appears in the previous paper and here serves as a baseline for comparison.
-The other three cases, which feature at least one merging firm that is already integrated, are new.
+Our model is drawn from \citeN{ST2021} and features a Bertrand downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
+Our goal is to make our analysis accessible to practitioners, so we use a model based in the merger simulation literature that can be calibrated with limited data.
+We then simulate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
+Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.
+The DOJ negotiated a settlement package meant to address harms in both the downstream and upstream markets.
 
-%Our model-based simulations show certain patterns across mergers.
-For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.
-The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.
-Likewise, we find that integrated mergers are typically net harmful to consumers as well.
-The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.
-Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
-We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
-We find that the existence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less harm.
-
-Our model is drawn from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.
-We use this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee.
-The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.
-Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.
-In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.
 Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.
 We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.
-Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts.
+We highlight how the results are effected by several key parameters and assumptions, namely the market elasticity, the use of flat logit versus nested demand preferences, and the bargaining parameter.
+Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts and misrepresent the extent of consumer harm.
+
+This empirical example allows us to explore the implications of the model for complex mergers.
+Starting from the Republic/Santek merger as proposed, we compare our baseline simulation results to those for counterfactual mergers between other firms in the market and to those for counterfactual mergers between different combinations of the upstream and downstream components of Republic and Santek.
+This exercise facilitates comparison between how the model behaves for complex mergers versus the more straightforward vertical and horizontal mergers that appear in, for instance, \citeN{ST2021}.
+Based on these simulations, we find that the complex mergers we study all result in net harm to consumers, despite the presence of large EDM effects in certain cases.
+ Only the counterfactual pure vertical merger between the Santek upstream division and the Republic downstream division provides net benefits to consumers.
 
@@
 The authors argue that this setup is useful for settings like grocery stores, where menu costs are low.
-}  Fourth, we rely on the logit for our demand function, which limits the ways in which consumers substitute between products.
-We also provide additional results using the nested logit.
+}  Fourth, we rely on logit demand for our baseline results, which limits the ways in which consumers substitute between products.
+We also provide sensitivity analyses using the nested logit.
 
-Our paper is related to the large literature on competition and vertical integration that grew out of a number of theory articles in the 1980s and 1990s.
+Our paper is related to the literature on competition and vertical integration that grew out of a number of theory articles in the 1980s and 1990s.
 \citeN{Riordan2008} has an excellent summary.
-This research explores how incentives to disadvantage rivals could arise alongside the efficiencies (such as EDM) associated with vertical integration.
-Some of these incentives result in RRC effects (e.g.
-\citeN{SalopScheffman1983} or \citeN{Riordan1998}) or in the foreclosure of competitors from input markets (e.g.
-\citeN{Salinger1988} or \citeN{OSS1990}).
-Unlike in our paper, most of these models are not framed explicitly as bargaining problems, instead assuming firms making take-it-or-leave-it offers.
+Broadly speaking, this research explores how incentives to disadvantage rivals could arise with vertical integration, and in turn shows how vertical mergers could negatively impact consumer welfare.
+See, for example, \citeN{SalopScheffman1983}, \citeN{Salinger1988}, \citeN{OSS1990}, and \citeN{Riordan1998}.
+In a number of these models, vertically integrated firms raise input prices to competitors or exit the input market entirely.
+Unlike in our paper, most of these models are not framed explicitly as bargaining problems, instead assuming posted or non-negotiated prices.
 However, many of the incentives towards RRC, which usually involve trading off profits from input sales against those from downstream sales, are similar to those that appear in our model.
@@
 Meanwhile, the structural empirical industrial organization literature has also studied the competitive effects of vertical integration as part of a greater exploration of the impacts of bargaining power in vertical supply chains.
-Many of these papers are based on the model developed in \citeN{HW1988} modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.
-These models have been applied to several sectors that are the frequent target of vertical integration, such as healthcare (e.g.
+Many of these papers are based on the model developed in \citeN{HW1988}, modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.
+These models have been applied to several sectors that are frequent targets of vertical integration, such as healthcare (e.g.
 \citeN{HL2017}) and video programming distribution (e.g.
 \citeN{CLWY2017}).
-These structural empirical papers explore the welfare effects of various competitive interactions, including complex mergers, in certain industries.\footnote{There is a companion empirical literature that relies more on regression analysis that identified changes in market outcomes due to vertical integration (e.g.
+These structural empirical papers explore the welfare effects of various competitive interactions, including complex mergers, in specific industries.\footnote{There is a companion empirical literature that relies more on regression analysis to identify changes in market outcomes due to vertical integration (e.g.
 \citeN{Chipty2001} on video programming distribution).
@@
 
-Our work is also related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
+Our work is also related to the literature on practical methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.
 The closest merger simulation article is our previous research in \citeN{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger, ruling out complex mergers.
@@
 Section \ref{sec:concl} concludes.
-Additional simulation results and information on the data used in the application appear in the Appendix.
+Additional information on the data used in the empirical application appears in the Appendix.
 
@@
 We begin by describing our basic framework, based on \citeN{ST2021}, which uses a downstream Bertrand model embedded in an upstream Nash Bargaining setup.
-We extend this framework allows to study a variety of complex merger configurations.
+This framework allows us to study a variety of complex merger configurations.
 
@@
 Retailers are indexed by $r$, and the wholesalers that supply these retailers are indexed by $w$.
-Prior to any mergers taking place, each wholesaler offers only one product, although each retailer may purchase from multiple wholesalers.
+Each wholesaler offers only one product, although each retailer may purchase from multiple wholesalers.
 Therefore, unique products (meaning a wholesale product offered at a given retail outlet) are denoted by pairs of the retailer and wholesaler indices, $rw$.
@@
 
-Let the share of consumers that choose product $w$ sold by retailer $r$ be denoted by $s_{rw}$.
-This share is a function of the vector of retail prices $P$ offered across all available products.\footnote{We omit the retail price argument for $s_{rw}(P)$ in what follows for brevity's sake.}  In our simulations we model the share using both logit and nested logit functions.
+Let the share of consumers that choose to purchase product $w$ sold by retailer $r$ be denoted by $s_{rw}$.
+This share is a function of the vector of retail prices $P$ offered across all available products.\footnote{We omit the retail price argument for $s_{rw}(P)$ in what follows for brevity's sake.}  In our simulations we model the share using both the logit and nested logit functions.
 The random coefficients logit is another commonly used option.
-The profit for retailer $r$ is given by $\pi^r = \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx} M$, where $p^W_{rx}$ is the unit fee charged by wholesaler $x$ to retailer $r$, $c^R_{rx}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.
+The profit for retailer $r$ is given by
+\begin{equation}
+\pi^r = \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx} M,
+\label{eq: ret profit}
+\end{equation}
+where $p^W_{rx}$ is the unit fee charged by wholesaler $x$ to retailer $r$, $c^R_{rx}$ captures any additional marginal costs borne by the retailer, and $M$ is the market size.
 Downstream prices are set in Bertrand equilibrium according to
@@
 \end{equation}
-which is the first order condition for product $w$ sold by retailer $r$.
+which is the first order condition for product $w$ sold by retailer $r$, found by maximizing the profit function $\pi^r$ with respect to $p_{rw}$.
 
+Turning to the upstream market, profits for wholesaler $w$ are given by
+\begin{equation}
+\pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M,
+\label{eq: wh profit}
+\end{equation}
+where $c^W_{xw}$ is the wholesale marginal cost when dealing with retailer $x$.
 Wholesale prices are set via bilaterial Nash Bargaining between retailers and wholesalers.
+The bargaining problem for wholesale price $p^W_{rw}$  takes the form,
+\begin{equation}
+\max_{p^W_{rw}} (\pi^r - d^r (\mathcal{W}^r \setminus \{w\}))^{\lambda_w} (\pi^w - d^w (\mathcal{R}^w \setminus \{r\}))^{1 - \lambda_w},
+\label{eq: bargain}
+\end{equation}
+where $\lambda_w \in [0,1]$ captures the bargaining power of retailers relative to wholesaler $w$.
+The $d^r (\mathcal{W}^r \setminus \{w\})$ is the disagreement payoff for the retailer, and $d^w (\mathcal{R}^w\setminus \{r\})$ is the disagreement payoff for the wholesaler.
+The disagreement payoff for the retailer is the profits it would earn if it ceases to sell this wholesale product,
+\begin{equation}
+d^r (\mathcal{W}^r \setminus \{w\}) = \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}(\mathcal{W}^r \setminus \{w\}) M,
+\end{equation}
+where $s_{rx}(\mathcal{W}^r \setminus \{w\})$ is the realized downstream market share for product $rx$ when retailer $r$ does not sell wholesaler $w$'s product.
+Similarly, the disagreement payoff of the wholesaler is
+\begin{equation}
+d^w (\mathcal{R}^w \setminus \{r\}) = \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] s_{xw}(\mathcal{W}^r \setminus \{w\}) M,
+\end{equation}
+which is earned if wholesaler $w$ does not offer its product to retailer $r$.
+
 We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
-}  Profits for wholesaler $w$ are given by $\pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M$, where $c^W_{xw}$ is the wholesale marginal cost when dealing with retailer $x$.
-The first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
+}  Then the first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
 \begin{equation}
@@
 &\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
-&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right),
+&\frac{1-\lambda_w}{\lambda_w} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right).
 \end{split}
@@
 \end{equation}
-where $\lambda \in [0,1]$ captures the bargaining power of the retailer relative to the wholesaler.
 The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) \equiv s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.
@@
 Together the series of downstream and upstream first order conditions determine market equilibrium.
-After assuming a functional form for shares, the model can be solved and calibrated along the lines described in \citeN{ST2021}.
+After assuming a functional form for shares, the model can be solved and calibrated along the lines described in \citeN{ST2021}, using these first order conditions.
 
@@
 Here we describe the manner in which complex mergers are modeled in this framework.
-Throughout we assume that all marginal costs, $c^R_{rw}$ and $c^W_{rw}$, and any non-price product characteristics remain unchanged after the merger, which rules out merger efficiencies.
+Throughout we assume that all marginal costs, $c^R_{rw}$ and $c^W_{rw}$, and any non-price product characteristics remain unchanged after the merger.
 We begin by describing vertical mergers, as all cases we examine have some vertical aspects.
+In modeling a merger, we assume that the combined firms maximize the sum of their profits, meaning we add together the profit functions in equations \eqref{eq: ret profit} and \eqref{eq: wh profit} for any merging retailers and wholesalers.
 Suppose that retailer $t$ and wholesaler $v$ were to merge.
@@
 On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $t$ being able to access product $v$ at marginal cost.
-This force would tend to lower the resulting price.\footnote{However, this lower marginal cost raises the potential for a larger margin, which could also cause a multi-product retailer to raise prices on its other goods $x \in \mathcal{W}^t \setminus \{v\}$, in order to divert sales to product $tv$.
-This is the Edgeworth-Salinger effect as in \citeN{Salinger1991}.}  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $t$ in order to divert sales to wholesaler $v$.
-This force would tend to raise the resulting consumer price.
+This force would tend to lower the resulting price.\footnote{However, this lower marginal cost may also encourage a multi-product retailer to raise prices on its other goods $x \in \mathcal{W}^t \setminus \{v\}$, in order to divert sales to product $tv$.
+This incentive is modeled in \citeN{Salinger1991} and studied empirically in \citeN{LM2020}, who call it the ``Edgeworth-Salinger effect.''}  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $t$ in order to divert sales to wholesaler $v$.\footnote{\citeN{Chen2001} models such an effect.}  This force would tend to raise the resulting consumer price.
 
@@
 & - \overbrace{\overbrace{[p_{tv}-c^W_{tv}-c^R_{tv}] \Delta s_{tv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^j \setminus \{v\})}^{\text{RRC effect}}=\\
-&\frac{1-\lambda}{\lambda} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
+&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
 \end{split}
@@
 Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $j$, these losses are softened due to a potential for diversion to retailer $t$, which we label the ``RRC effect.''\footnote{The assumption of simultaneous downstream price setting rules out RRC effects in the style of \citeN{SalopScheffman1983}, in that here the wholesaler does not have an incentive to raise prices in order to directly impact the pricing strategies of rival retailers.
-Rather, our model has the RRC effect seen in \citeN{CLWY2017} (Section 6.2.2, p.
-934), where the wholesaler has an incentive to raise prices to rival retailers due to the possibility of increased sales volume for their merged downstream retail partner.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
+Rather, our model has the RRC effect seen in \citeN{CLWY2017}, where the wholesaler has better bargaining leverage over rival retailers due to the possibility of increased sales volume (conditional on prices) for their merged downstream retail partner should negotiations break down.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
 These impacts tend to raise the resulting input price to retailer $j$.
@@
 &[p^W_{tk} - c^W_{tk}]s_{tk} - \sum_{x \in \mathcal{R}^k \setminus \{t\}} [p^W_{xk} - c^W_{xk}] \Delta s_{xk}(\mathcal{W}^t \setminus \{k\}) = \\
-&\frac{1-\lambda}{\lambda} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
+&\frac{1-\lambda_k}{\lambda_k} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
 &\left.- \underbrace{[p_{tv} - c^W_{tv} - c^R_{tv}] \Delta s_{tv}(\mathcal{W}^t \setminus \{k\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{wholesale recapture leverage effect}}\right).
@@
 In this case, the merged firm has two channels for potential additional profits should it cease to trade with wholesaler $k$.
-First, if retail sales are diverted to product $v$ sold by retailer $t$, those sales will earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $k$ carried by retailer $t$ could increase sales by wholesaler $v$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.
+First, if retail sales are diverted to product $v$ sold by retailer $t$, those sales could earn a higher margin due to lower marginal costs stemming from what we call the ``EDM recapture effect.''  Second, the loss of product $k$ carried by retailer $t$ could increase sales by wholesaler $v$ through other retailers, which we call the ``wholesale recapture leverage effect.''  Both of these effects would tend to lower the resulting input price.
 
@@
 EDM between $v$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $v$ sold by retailer $s$, those units may earn a larger margin.
-This impact, which comes through the interaction of EDM with UPP, follows the logic of \citeN{Salinger1991}, where EDM on one product can cause a firm to raise prices on other products in order to drive sales to the lower cost good.
-We label this effect the ``indirect EDM effect.''
+This impact, which comes through the interaction of EDM with UPP, we label the ``indirect EDM effect.''
 
@@
 The downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of the new partner wholesaler.
-The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain an additional wholesale product at marginal cost, which in turn would raise the value of diverted sales to that product.
+The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain an additional wholesale product at marginal cost, which in turn would tend to raise the value of diverted sales to that product.
 Whether these additional incentives to raise prices will dominate the direct EDM impact is an empirical question.
@@
 & - \overbrace{\sum_{x \in \{t,s\}}\left(\overbrace{[p_{xv}-c^W_{xv}-c^R_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{y \in \mathcal{W}^x \setminus \{v\}} [p_{xy} - p^W_{xy} - c^R_{xy}] \Delta s_{xy}(\mathcal{W}^j \setminus \{v\})\right)}^{\text{RRC effect}}=\\
-&\frac{1-\lambda}{\lambda} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
+&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
 \end{split}
@@
 &[p^W_{tk} - c^W_{tk}]s_{tk} - \sum_{x \in \mathcal{R}^k \setminus \{t\}} [p^W_{xk} - c^W_{xk}] \Delta s_{xk}(\mathcal{W}^t \setminus \{k\}) = \\
-&\frac{1-\lambda}{\lambda} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
+&\frac{1-\lambda_k}{\lambda_k} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
 &- \underbrace{\sum_{x \in \{t,s\}}[p_{xv} - c^W_{xv} - c^R_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{EDM recapture effect}}
@@
 All of these additional terms tend to increase the bargaining leverage of the merged firm.
-If we instead examined a merger between retailer/wholesaler $tv$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $t$ for two merged wholesalers, rather than sales for wholesaler $v$ through two merged retailers.
+If we instead examine a merger between retailer/wholesaler $tv$ and an unaffiliated wholesaler, that would augment the EDM recapture effect with sales of retailer $t$ for two merged wholesalers, rather than sales for wholesaler $v$ through two merged retailers.
 Similarly, the retail recapture leverage effect would be replaced with an additional wholesale recapture leverage effect for the newly merged wholesaler.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Empirical Application: Republic/Santek \label{sec:application}}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Empirical Application: the Chattanooga Waste Industry \label{sec:application}}
 The U.S.
 waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{See Waste Dive, \tiny{\url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}}.} In recent years the industry has experienced significant merger activity, including several acquisitions between vertically integrated, nationally active competitors.
-Furthermore, there are a variety of vertical supply chain configurations across local markets, making the industry an excellent application to study the welfare impacts of mergers that have both horizontal and vertical aspects.
-
-Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{See Competitive Impact Statement: U.S.
-and State of Alabama v.
-Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}.}  The DOJ filed a complaint and negotiated a settlement in the case, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW) disposal markets.
-In addition, vertical anti-competitive effects were alleged to arise from the combination of Republic and Santek integrated assets in the Chattanooga region.\footnote{See U.S.
-and State of Alabama v.
-Republic Services, Inc.
-and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}
+Furthermore, there are a variety of vertical supply chain configurations across local markets, making the industry an excellent application to study mergers that have both horizontal and vertical aspects.
+In this section, we use one such merger, between the firms Republic and Santek, to explore how our model captures and weighs EDM, RRC, and UPP effects in a real world setting.
+Throughout this exercise, our focus is on mirroring the types of analysis usually used by antitrust agencies in reviewing mergers.
+Given this context and the types of constraints that agencies often face, we assume that a limited amount of data are available and tailor our calibration methods accordingly.
 
@@
 \subsection{Market Background}
+Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{See the complaint: U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}, at paragraph 7.}  The DOJ filed a complaint, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW) disposal markets spread across multiple states.\footnote{See the complaint: U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC at paragraphs 37-43.}   In addition, vertical anti-competitive effects were alleged to arise from the combination of Republic and Santek integrated assets in the Chattanooga region.\footnote{See the complaint: U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC at paragraphs 44-47.}  The merger was ultimately settled with the DOJ through divestitures.
+As described in the Final Judgment filed with the settlement package, the parties were required to divest Santek's SCCW collection assets as well as two landfills and a transfer station in the Chattanooga region.\footnote{See the Final Judgment: U.S and Plaintiff States v.
+Republic Services, Inc.
+and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}.} In this section, we apply our model to simulate the effects of this merger in Chattanooga.
 
@@
 Disposal customers include private waste haulers without their own disposal assets as well as local governments that collect their citizens' waste themselves.
-Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.
 
-The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic/Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area,'' subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and SCCW collection as well as due to RRC in the SCCW collection market by raising the MSW disposal costs of independent haulers.
-The CIS notes that pre-merger, Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
+The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic/Santek settlement describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area,'' subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and SCCW collection as well as due to RRC in the SCCW collection market by raising the MSW disposal costs of independent haulers.\footnote{See the CIS: U.S.
+and State of Alabama v.
+Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}.} The CIS notes that pre-merger, Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors.
 In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties' landfills within the area or passing through their transfer stations before ultimately being disposed of in the parties' landfills elsewhere.
 Thus, pre-merger both parties were large, vertically integrated competitors in the Chattanooga Area.
+The DOJ complaint says that the merger would have raised the Chattanooga Area collection Herfindahl-Hirschman Index (HHI) by 2,660 points to 5,551 and the disposal HHI by 3,018 points to 6,980.\footnote{See the complaint, U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC at paragraphs 38 and 42.}
 
-In addition, another large, vertically integrated company, Waste Connections, existed in the market at the time of the merger and was the parties' sole competitor in MSW disposal.
-National firms Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
-Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WM-ADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
+In addition, another large, vertically integrated company, Waste Connections, existed in Chattanooga at the time of the merger and was the parties' sole competitor in MSW disposal.
+National firms Waste Management and ADS also owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WMADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
 The final significant participant in the SCCW collection market was a major regional firm that is not identified by name in the CIS.
@@
 
-Using data described in Appendix \ref{app:trashdata}, the supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and downstream collection market participants (Republic, Santek, Waste Connections, WM-ADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
-Santek and Waste Connections exhibit excess disposal capacity.
-However, Republic's collection volumes are estimated to exceed their ability to self-supply disposal in the Chattanooga market.
-As a result, Republic needs to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
-We assume that this disposal is purchased from both Santek and Waste connections according to their shares of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
-We further assume that WM-ADS and Regional purchase disposal capacity from both Santek and Republic according to these shares.
+Based on the available data for disposal capacity and volumes, the pre-merger supply relationships between the upstream disposal market participants (Republic, Santek, and Waste Connections) and the downstream collection market participants (Republic, Santek, Waste Connections, WMADS, and Regional) can be inferred under the assumption that vertically-integrated haulers first dispose of waste at their own disposal facilities and then sell any residual disposal capacity to rival haulers.
+Santek and Waste Connections exhibited excess disposal capacity.
+However, Republic's collection volumes were estimated to exceed their ability to self-supply disposal in the Chattanooga market.
+As a result, Republic needed to purchase additional disposal from Santek and Waste Connections, presumably at a higher marginal cost.
+Other than imposing that Republic only offers disposal to itself, we do not structurally incorporate capacity constraints into the simulation model.
+We assume that the changes in volume that would be caused by the merger would not be large enough to shift which constraints would be binding.\footnote{We also assume that the Waste Connections-Republic product remains post-merger, which may only be true in the short run.}
+
 The resulting relationships between the upstream and downstream market participants are captured in Figure \ref{fig:chatt_tree}.
+There are a total of 9 wholesaler-retailer products in the market.
+The complex nature of the merger between Republic and Santek is immediately apparent.
+There are potential benefits from EDM for the purchases of disposal by Republic from Santek, but also potential harm from RRC by Santek when supplying disposal to rivals WMADS and Regional, alongside a loss of horizontal competition between Republic and Santek in downstream collection.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Simulation Inputs and Calibration}
+Before simulating the merger, we need to specify a form for consumer demand.
+We use the multinomial logit.
+Although the logit makes some restrictive assumptions, it is often used in merger analysis because of the heavy reliance on market shares in antitrust proceedings.\footnote{\citeN{MS2021} note that given the highly specific scope of the markets used for antitrust analysis, the logit can be a reasonable approximation, and has appeared in past merger reviews.
+}  The logit implies that substitution, and hence competition, between products is proportional to market shares, which dovetails intellectually with the antitrust emphasis on closely scrutinizing mergers between firms with large market shares.\footnote{For example, the 2023 DOJ/FTC Merger Guidelines at \S 2.1 state that mergers raise a presumption of illegality when the change in the HHI is more than 100 points and results in a market concentration of at least 1,800 or in a merged firm with at least 30\% market share.
+Given that the change in the HHI is equal to two times the product of the merging firms' market shares, these presumptions are likely to screen in mergers when both firms have high shares.}  Nevertheless, we also provide some sensitivity analysis using the nested logit.
+
+The logit demand implies the following form for the share of product $rw$,
+\begin{equation}
+s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{x \in \mathcal{R}} \sum_{y \in \mathcal{W}^x} \exp(\delta_{xy} - \alpha p_{xy})},
+\label{eq: logit share}
+\end{equation}
+where $\delta_{rw}$ is a quality parameter and $\alpha$ captures sensitivity to price.
+There is an outside good whose quality parameter and price have been normalized to zero.
+We use this share equation to substitute into the first order conditions described in Section \ref{sec:theory}, which then form the set of equations used for calibration.
+
+For our sensitivity analysis using nested logit demand, we divide the set of retailers into a series of nests, with each nest indexed by $g$.
+The set of retailers within each nest is denoted by $\mathcal{R}_g$, and the set of all nests is denoted by $\mathcal{G}$.
+Then the share for product $rw$ is given by
+\begin{equation}
+s_{rw} = \frac{\exp((\delta_{rw}-\alpha p_{rw})/(1-\sigma))}{D_g^{\sigma} \sum_{f \in \mathcal{G}} D_f^{1-\sigma}},
+\label{eq: nl share}
+\end{equation}
+where
+\begin{equation}
+D_g = \sum_{r \in \mathcal{R}_g} \sum_{w \in \mathcal{W}^r} \exp((\delta_{rw} - \alpha p_{rw})/(1-\sigma)) \quad \forall g \in \mathcal{G},
+\end{equation}
+and where $\sigma$ is a nesting parameter such that $0 \leq \sigma < 1$.  When $\sigma=0$, the model reverts to the logit.
+
+Returning to the logit version of the model, in order to calibrate it, we need enough data to identify the model parameters and primitives, which consist of the vector of product qualities $\mathbf{\delta}$, the price coefficient $\alpha$, the bargaining parameters $\lambda$, and any unobserved upstream or downstream marginal costs.
+In a scholarly research setting, one would typically proceed by gathering a dataset of prices, market shares, and instrumental variables with enough variation to estimate the consumer demand parameters.
+Then given that demand system and the firm pricing first order conditions of the model, one would recover estimates of marginal costs.
+However, antitrust practitioners usually do not have access to the larger cross section or panel datasets necessary for demand estimation, instead observing only shares and prices for one or two markets, usually at only a single point in time.
+In our context, this means having a snapshot of data for the nine products in the Chattanooga Area.
+Nevertheless, antitrust practitioners can often supplement these data with some information on profit margins.
+Margins are helpful in that, combined with the firm pricing first order conditions, they provide information on the implied price elasticities and demand parameters that caused firms to choose those margins.
+
+We follow calibration methods similar to those described in \citeN{ST2021}.
+The derivatives of logit demand are functions only of market shares and the price coefficient $\alpha$, as $\partial s_{rx}/\partial p_{rw} = \alpha s_{rx} s_{rw}$  for $x \neq w$ and $\partial s_{rw}/\partial p_{rw} = -\alpha s_{rw} (1-s_{rw})$ for $x=w$.
+Furthermore, the logit also implies that $\Delta s_{xy} (\mathbb{W}^r \setminus \{w\}) = s_{rw} \left(s_{xy}/(1-s_{rw})\right)$.
+Therefore, the downstream pricing first order conditions and the upstream bargaining first order conditions described in Section \ref{sec:theory} form a system of equations that are functions only of market shares, margins, the downstream price coefficient, and the upstream bargaining parameters.
+Given this system of equations, our model is identified if we observe the upstream margins for all products and a downstream margin for a single product, along with the full vector of market shares, upstream prices, and downstream marginal costs.
+Calibration proceeds by numerically solving for the unknown objects in the system of first order condition equations described in Section \ref{sec:theory}, given the input data.
+Once the price coefficient $\alpha$ is calibrated, downstream pre-merger equilibrium prices can be recovered from the appropriate downstream pricing first order condition, and  the quality parameters can be recovered from the market shares in equation \eqref{eq: logit share}.
+After recovering all parameters and pre-merger prices, we use code from the \texttt{antitrust} R package to run the merger simulations.\footnote{When downstream prices rather than costs are observed, the \texttt{vertical.barg} function may be used to calibrate bargaining and demand parameters and simulate merger effects for all complex mergers discussed here.
+See  \url{https://CRAN.R-project.org/package=antitrust}.}  We again use the first order conditions from Section 2 to perform these simulations, but now assume that any merged firms jointly maximize their profits.
+
+The intuition behind the identification of the model is as follows.
+The process for identifying the downstream parameters is similar to that seen in \citeN{WF1994}.
+Conditional on the other data, knowledge of a downstream margin determines the how sensitive retail customers are to price, as a higher margin implies less elastic demand.
+Given this implied relationship, optimal downstream prices can be recovered using the firm first order conditions with data on shares and costs.
+Then knowledge of downstream market shares identifies the quality of each product, as higher shares conditional on prices implies higher quality.
+Turning to the upstream model, the remaining unknown objects are the bargaining parameters.
+Conditional on the data and the downstream model, bargaining power is identified by the relative sizes of the upstream and downstream profit margins.
+When downstream firms have higher margins than downstream firms, that is an indicator that retailers have more bargaining power.
+
+Our data comes from several public sources.
+We obtain information on disposal volumes and landfill usage from the Tennessee Department of Environmental Quality.
+We assume that Republic purchases disposal from both Santek and Waste connections according to their shares of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
+We further assume that WMADS and Regional purchase disposal capacity from both Santek and Republic according to these shares.
+The DOJ complaint states that due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal.\footnote{See the complaint: U.S.
+and State of Alabama v.
+Republic Services, Inc.
+and Santek Waste Services, LLC at paragraph 25.}  We interpret this statement as implying that the share of the outside good in our simulations is zero.
+We gather information on disposal prices from the 2019 Waste Business Journal’s Directory of Waste Processing \& Disposal Sites.
+Margin and downstream cost information comes from annual reports and public financial statements, under the assumption that the national averages reported there are reasonable approximations for the Chattanooga market.
+Our data sources and construction are described in more detail in Appendix \ref{app:trashdata}.
 
@@
 As discussed above and in Appendix \ref{app:trashdata}, we need to make several simplifying assumptions in order to transform the available information into the data required for our model.
-We expect that the DOJ would have had access to more accurate confidential information during their investigation.
-Therefore, we view our results as only illustrative of the types of analysis and results that could have been produced for this case.
-Note that, other than imposing that Republic only offers disposal to itself, we do not structurally incorporate capacity constraints into the simulation model.
-We assume that the changes in volume that would be caused by the merger would not be large enough to shift which constraints would be binding.\footnote{We also assume that the Waste Connections-Republic product remains on the market, which may only be true in the short run.}
+We expect that the DOJ would have had access to confidential information during their investigation, which could have allowed them to generate more precise results.
+Nevertheless, we believe our results are qualitatively accurate, as they are consistent with the anti-competitive concerns raised in the Republic/Santek complaint and CIS.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Merger Simulation Results}
-
-Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of annual harm to consumers and total annual producer benefits of approximately \$13.4 million.
-Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda=1$) and Republic ($\lambda=0.7$) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.
+\subsection{Republic/Santek Merger Simulation Results}
+The results of our merger simulation for prices and market shares appear in Figure \ref{fig:TrashSimsFirms}.
+The merger between Republic and Santek is estimated to lower upstream prices by almost 3\% and increase downstream prices by 12.5\%, resulting in \$16.6 million of annual harm to consumers and total annual producer benefits of approximately \$14 million.
+Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda_{Santek}=1$) and Waste Connections ($\lambda_{WasteConn}=0.73$) give a substantial edge to downstream firms, which suggests consumers are likely to be harmed because EDM benefits are probably small.
 High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.
-(See Table \ref{tab:trashdata}.)}  These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.
+(See Table \ref{tab:trashdata}.)}
 
@@
 However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs).
-The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 22\%.
-These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by 60\%.
+Santek's disposal prices to downstream rivals are estimated to increase over 88\% post-merger, reflecting RRC.
+These higher Santek prices allow Waste Connections to also increase its prices to downstream rivals by 22\%.
+These large price increases are somewhat offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by approximately 63\%.
 
 Turning to the downstream market, all collection prices increase post-merger.
-The merging parties' post-merger collection prices increase by approximately 20\%, and their share decreases by about 8 percentage points overall.
-The Republic/Republic product's price increases 21\%, and the Santek/Santek product's price increases by 23\%.
-The newly integrated Santek/Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers.
-The integrated rival, Waste Connections, increases its collection price by approximately 5\%, and its share increases by 6 percentage points.
-The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.
+The merging parties' post-merger collection prices increase more than 19\%, and their share decreases by about 16 percentage points overall.
+The Republic-Republic product's price increases 21\%, and the Santek-Santek product's price increases by 23\%.
+The newly integrated Santek-Republic product's price is estimated to increase by almost 9\%, demonstrating that EDM is not fully passed-through to consumers.
+The integrated rival product, Waste Connections-Waste Connections, increases its collection price by 5\%, and its market share increases by almost 10 percentage points.
+The unintegrated rivals' collection prices each increase by about 8\%, which again demonstrates incomplete pass-through of the change in upstream costs.
+Given that all downstream prices increase, consumers are harmed by the merger.
+Therefore, although the merger results in a large EDM effect for Santek disposal sold to Republic, it is outweighed by RRC effects in the disposal market and the loss of horizontal competition in the collection market.
 
-In Figure \ref{fig:TrashSimsCompareAll} we compare the full vertical model results, which compute both post-merger downstream and upstream price effects, to two other models that hold the prices of one level fixed.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``downstream-only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``upstream-only'' model computes disposal price effects holding collection prices fixed at pre-merger levels.
+This simulation admittedly relies on a simplified model, particularly with regards to the use of logit demand.
+The logit makes two strong assumptions: (1) that overall consumer welfare relies heavily on the extent of substitution to the outside good and (2) that substitution within the market is proportional to market shares.
+We can examine the sensitivity of our results to the first issue by varying our chosen value for the share of the outside good.
+The form of the logit implies that this share is proportional to the market elasticity of demand.\footnote{The market elasticity with respect to a price change for all inside products is $ -\alpha \overline{p} s_{00}$, where $s_{00}$ is the share of the outside good and $\overline{p}$ is the share-weighted average of prices among the inside goods.
+}  Our baseline results assume that the outside share is zero, following the assertion made by the DOJ that there are essentially no reasonable substitutes for SCCW collection or MSW disposal.\footnote{See the CIS: U.S and Plaintiff States v.
+Republic Services, Inc.
+and Santek, LLC.}  There are strict laws that dictate where and how commercial customers can dispose of waste.
+Such an assumption naturally tends to increase the harm from the merger, because consumers do not exit the market in response to a price increase.
+Results for alternative assumptions appear in  Figure \ref{fig:TrashElast}.
+The baseline results are those for the perfectly inelastic market appearing as the leftmost point.
+As we increase the share of the outside good (which, equivalently, increases the absolute value of the market elasticity of demand), the estimated harm to consumers decreases monotonically.
+However, the harm is still about a third of the baseline results when the market is unit elastic.
+Consumer harm remains over 5\% of pre-merger expenditures when the market elasticity is as high as 1.3.
 
-Ignoring these vertical relationships in the downstream-only model results in an estimated 11\% increase in collection prices with a greater than 16\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million.
+In order to examine how sensitive our results are to the substitution proportional to share assumption, we can model demand instead as a nested logit.
+We present two alternatives.
+In the first option (which we call ``integrated'' nesting), we organize the nests into two groups, one for the integrated products, meaning Republic-Republic, Santek-Santek, and Waste Connections-Waste Connections, and the second for all other products where the disposal and collection are performed by different firms.
+In the second option (which we call ``landfill'' nesting), we have three nests, one each for Republic, Santek, and Waste Connections, where each nest collects together all products that use the same landfill.
+We run these simulations for different possible values of the nesting parameter $\sigma$ between 0 and 0.9, where 0 is equivalent to the baseline logit.
+As $\sigma$ increases towards 1, the nests become more differentiated.
+
+Our results appear in Figure \ref{fig:TrashNest}.
+The baseline simulation is represented by the point at a nesting parameter of 0 on the left of in figure.
+As $\sigma$ increases, consumer harm under integrated nesting increases and consumer harm under landfill nesting decreases.
+These findings are intuitive, as with integrated nesting, the two highest selling products, Republic-Republic and Santek-Santek, are in a nest with only one other competitor, the much less popular Waste Connections-Waste Connections.
+As the nesting parameter tends towards one, the merger of Republic and Santek approaches a merger to monopoly in this nest, which causes a lot of harm at high nesting parameters, as evidenced by the jump in harm at a nesting parameter of 0.9.
+On the other hand, with landfill nesting, these two best selling products are separated into different nests, resulting in less horizontal competition between the merging firms and hence less consumer harm, especially when the nesting parameter is large.
+However, consumer harm remains noticeable under landfill nesting, at around 10\% of pre-merger expenditures, when the nesting parameter is around 0.5.
+
+Another key part of our model is the bargaining power parameters $\lambda$.
+As discussed in \citeN{ST2021}, vertical mergers tend to benefit consumers when wholesalers have relatively more power versus when retailers have relatively more power.
+Larger wholesaler bargaining power offers more possibilities for EDM, because pre-merger input prices are likely to be high.
+This channel becomes less relevant as retailers gain more power.
+Our baseline merger simulation uses relatively large bargaining power parameters of 1 for retailers negotiating with Santek and 0.73 for negotiating with Waste Connections, which reflect the higher relative margin we observe for collection versus disposal.
+We test the sensitivity of our results to the bargaining power parameter by varying it in alternative simulations.\footnote{In these simulations, we set all parameters for bargaining power between all non-integrated wholesaler-retailer pairs to be identical.
+We found that introducing some variation across wholesaler-retailer combinations did not meaningfully change our results.
+}  Figure \ref{fig:TrashBarg} provides our results, with the value for consumer harm in the baseline simulation denoted by the horizontal dashed line towards the top of the graph.
+As expected, consumer harm is larger when the bargaining power parameter is higher, similar to what is seen in \citeN{ST2021} for vertical mergers.
+Interestingly, consumer harm is close to the baseline value for bargaining parameters as low as 0.4.
+Below that, EDM effects become more pronounced, and they even result in net consumer benefits when retailer bargaining power falls down to 0.1.
+This result suggests that only in a counterfactual world where retail margins were observed to be much lower than wholesale margins should the DOJ have expected to find EDM effects to outweigh other harms.
+The data we have in Table \ref{tab:trashdata} suggests the opposite was true.
+
+In summary, our simulations estimate that the complex merger between Republic and Santek would have caused significant harm to consumers if it had been allowed to proceed without divestitures.
+Although the data inputs into the model are admittedly crude (though, in our experience, using data of comparable quality for merger analysis is not uncommon given the constraints that antitrust agencies face), we find that nontrivial consumer harm remains even after significant departures from our baseline parameter values.
+Furthermore, even when varying these parameters causes the results to meaningfully change, they shift in ways that are predictable and intuitive, as demonstrated by the monotonic relationships seen in Figures \ref{fig:TrashElast}, \ref{fig:TrashNest}, and \ref{fig:TrashBarg}.
+These results suggest that the model is useful for analyzing complex mergers in a policy contex.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparison to Alternative Models}
+In the interests of expediency or due to data and time constraints, it is common for antitrust practitioners to use merger simulations that do not account for the vertical aspects of a given market.
+In order to examine what impact such an assumption has on the simulations, we can compare the full vertical model results, which compute both post-merger downstream and upstream price effects, to two other models that hold the prices of one level fixed.\footnote{This approach uses the same calibrated demand and costs parameters across all three models.} The ``downstream-only'' model computes collection price effects holding disposal prices fixed at pre-merger levels, and the ``upstream-only'' model computes disposal price effects holding collection prices fixed at pre-merger levels.
+
+The results of this comparison are summarized in  Figure \ref{fig:TrashSimsCompareAll}.
+Ignoring these vertical relationships in the downstream-only model gives in an estimated 9.2\% increase in collection prices with an 16.4\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million.
 Estimated consumer harm is 60\% higher in the full vertical model and is offset by markedly larger producer benefits.
-The downstream-only model misses the substantial RRC effect experienced by the Santek/WM-ADS and Santek/Regional products.
+The downstream-only model misses the substantial RRC effect experienced by the Santek-WMADS and Santek-Regional products.
 In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and other prices increase alongside them.
@@
 In the upstream-only model, downstream prices and therefore shares are fixed at their pre-merger levels.
-We still have upstream prices set to marginal cost for integrated products.
-Overall disposal prices increase by 8.2\% post-merger, and the merger effect is a pure transfer between firms.
+Upstream prices are still set to marginal cost for integrated products.
+Overall disposal prices increase by 11.2\% post-merger, and the merger effect is a pure transfer between firms.
 Thus, both consumer and net harm are estimated to be zero.
 The main impact of the merger is in changing the bargaining leverage for Santek, now that it is combined with Republic.
-This effect causes Santek to raise its upstream disposal prices to both WM-ADS and Regional.
+This effect causes Santek to raise its upstream disposal prices to both WMADS and Regional.
 The price increase is even larger in the upstream-only model compared to the full model.
-The lack of downstream price adjustments means that WM-ADS and Regional cannot shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.
+The lack of downstream price adjustments means that WMADS and Regional cannot change prices in order to shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.
 In this sense, the upstream-only model overstates the increase in leverage that Santek receives from the merger.
 
-The Republic/Santek merger was ultimately settled with the DOJ through a divestiture.
-According to the Final Judgment, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station.\footnote{See Final Judgment: U.S and Plaintiff States v.
-Republic Services, Inc.
-and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}.} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
+Our results suggest that a divestiture in the collection market or in the disposal market alone likely would not have sufficiently remedied the anti-competitive effects of the merger.
+The DOJ ultimately required divestitures at both levels of the vertical chain.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparison to Alternative Mergers}
+The analysis above shows some useful aspects of the model for analyzing the complex merger between Republic and Santek.
+However, there are many other forms that a complex merger can take.
+In order to explore some of these other scenarios, we simulate counterfactual mergers between firms in the market other than Republic with Santek, specifically Santek/Waste Connections, Santek/WMADS, and WMADS/Regional.
+Santek/Waste Connections is an example of two fully integrated firms merging, and, unlike Republic, both firms sell their excess capacity to other downstream haulers.
+Santek/WMADS is a merger involving an integrated firm buying a pure downstream competitor.
+Filling out the set, WMADS/Regional is a horizontal merger between two firms that only operate downstream.
+The estimated aggregate welfare effects of these counterfactual mergers are illustrated alongside those of the Republic/Santek merger in Figure \ref{fig:TrashFakeOverall}.
+The results for prices and market shares appear in Figure \ref{fig:TrashFakeFirms}.
+
+We begin with the merger of Santek and Waste Connections, displayed in the second row of Figure \ref{fig:TrashFakeOverall} and the top panel of Figure \ref{fig:TrashFakeFirms}.
+The post-merger change in the HHI for the disposal market is 1,896 and for the collection market is 620, respectively.
+We see in Figure \ref{fig:TrashFakeOverall} that compared to Republic/Santek, the Santek/Waste Connections merger would have generated lower consumer harm and lower producer benefits.
+The harm result is interesting, because both Santek and Waste Connections exclusively self-supply their disposal input pre-merger.
+This eliminates the possibility for positive EDM effects.
+Furthermore, we see from Figure \ref{fig:TrashFakeFirms} that Santek and Waste Connections significantly raise disposal prices to their downstream rivals, Republic, WMADS, and Regional.
+However, the Republic-Republic product has a large downstream market share and does not face an input price increase, which limits the extent of price increases in the collection market.
+Thus, net consumer harm to the Chattanooga market is \$6.6 million, lower than that estimated for Republic/Santek.
+
+Turning to the Santek/WMADS simulation, this would have been a merger of a small downstream competitor combining with a vertically integrated firm.
+The merger results in a post-merger change in the collection market HHI of 620.
+Since Santek supplies WMADS with disposal, there is the possibility for EDM.
+As can be seen in Figure \ref{fig:TrashFakeOverall}, the welfare impacts are more muted relative to both the Republic/Santek and Santek/Waste Connections mergers.
+This is driven by the small size of the WMADS products relative to more popular options like Republic-Republic and Waste Connections-Waste Connections.
+Turning to Figure \ref{fig:TrashFakeFirms}, we see that EDM greatly lowers the disposal price between Santek and WMADS, while RRC slightly raises the disposal prices that Santek charges Republic and Regional.
+This nets out into the small shift in profits from upstream to downstream firms seen in Figure \ref{fig:TrashFakeOverall}.
+Prices in downstream collection rise slightly, which gives a relatively small loss of consumer welfare.
+
+As for the WMADS/Regional simulation, this would have been the combination of two downstream horizontal competitors, neither of which was vertically integrated.
+This merger would have resulted in a post-merger change in the HHI for the collection market of 136, which is much lower than any of the other combinations we have considered.
+Unsurprisingly, the resulting estimated welfare effects in Figure \ref{fig:TrashFakeOverall} are small, with consumer harm and producer benefits approximately canceling each other out.
+The more detailed results reported in Figure \ref{fig:TrashFakeFirms} show only minimal changes in prices or market shares post-merger.
+These results are consistent with the findings for downstream horizontal mergers seen in \citeN{ST2021}.
+
+Another way to explore the implications of the model for complex mergers is to construct certain counterfactual firms as merger candidates.
+Specifically, we take the upstream and downstream divisions of Republic, Santek, and Waste Connections, and unless otherwise indicated, assume they are independent firms.\footnote{We set their bargaining power parameters equal to $0.73$, which is what we calibrated for Waste Connections in the baseline model.
+}  This setup allows us to study additional complex mergers involving the most popular products, those sold by Republic and Santek.
+Specifically, we examine the following mergers: (1) a pure vertical merger between Santek's upstream disposal and Republic's downstream collection, labeled ``Vertical'' in the figure; (2) a pure upstream merger between Santek's disposal and Republic's disposal, labeled ``Up''; (3) a pure downstream merger between Santek's collection and Republic's collection, labeled ``Down''; (4) a partial upstream merger between a vertically integrated Santek and Republic's disposal, labeled ``Partial Up''; and (5) a partial downstream merger between a vertically integrated Santek and Republic's collection, labeled ``Partial Down.''  The last two are complex mergers, while the first three are similar to the non-complex cases studied in \citeN{ST2021}.
+In any of the instances where only one part of Republic or Santek are merging, we assume that the other part is operating independently.
+
+ The welfare results for these simulations are in Figure \ref{fig:TrashFakeRepSan}.
+The baseline simulation for the Republic/Santek merger as actually proposed is provided at the top for comparison.
+The last three rows echo findings in \citeN{ST2021} for non-complex mergers, which showed that our model tends to find consumer harm for most horizontal mergers, whereas vertical mergers have more scope for net consumer benefits.
+Starting with the pure vertical case at the bottom of the figure, we see that consumers actually benefit from this merger.
+This result occurs because the EDM for Santek selling disposal to Republic is so large, with the disposal price falling by approximately 63\%.
+Recall from Figure \ref{fig:TrashSimsFirms}, this was also seen in the baseline simulation.
+However, here we find that the RRC effects are much smaller, than in the baseline with Santek raising its disposal prices to WMADS and Regional by about 16\% versus 88\% in the baseline.
+This difference occurs because in the pure vertical case, the highly popular Santek-Santek and Republic-Republic products are not fully owned by the merging firms, which means that some of the benefits from customers induced to divert from withholding disposal would accrue to the independent Santek collection and Republic disposal firms.
+As a result, the EDM effect dominates and most downstream prices fall, which results in lower collection profits and net consumer benefits.
+
+ Turning to the pure upstream merger in the fifth row of Figure \ref{fig:TrashFakeRepSan}, the combination of the disposal assets of Santek and Republic harms consumers due to a loss in upstream competition.
+Both firms raise all of their disposal prices by over 30\%, which harms collection firms who in turn partially pass on these higher costs to final customers.
+It is interesting to compare this case to a partial upstream merger between an integrated Santek and Republic's disposal division, which appears in the second row of Figure \ref{fig:TrashFakeRepSan}.
+Compared to the pure upstream scenario, the partial upstream merger adds an RRC incentive.\footnote{There is no new EDM incentive, because Republic does not provide disposal to Santek.
+}  Now when Republic disposal negotiates with Republic collection, some of the customers that could switch away from the Republic-Republic product could choose a Santek collection option, including the Santek-Santek integrated product that benefits from EDM.
+As a result, the price of Republic disposal increases by about 70\%.
+Furthermore, the equilibrium prices for Santek disposal sold to rivals are significantly higher pre-merger than in the pure upstream merger case, and they still rise by nearly 30\% after the merger.
+This results in high post-merger downstream prices, shifting some of the increased profits from upstream to downstream firms and increasing harm to consumers.
+
+The results for the pure downstream merger between the collection divisions of Republic and Santek appear in the fourth row of Figure \ref{fig:TrashFakeRepSan}.
+In this case, consumers are harmed due to the loss of downstream competition, which increases prices in the collection market.
+Interestingly, we find that upstream disposal prices charged to the merged Republic and Santek fall slightly, due to increased bargaining leverage against input suppliers, which manifests itself in the decrease in upstream profits seen in the third column of Figure \ref{fig:TrashFakeRepSan}.
+This countervailing effect is also documented in \citeN{ST2021}.
+As is also seen in that previous paper, here we find that the lower input prices do not translate into net benefits for consumers.
+
+When we compare the pure downstream case to the partial downstream merger between an integrated Santek and Republic's collection assets, we see that consumer harm increases.
+This occurs even though the move to the partial merger introduces a large EDM effect for Santek's sales of disposal to Republic.
+That input price falls by over 70\%.
+The merged firm continues to negotiate a lower input price from Republic's disposal division, due to increased bargaining leverage.
+However, Santek also raises prices on disposal to WMADS and Regional due to RRC, which, combined with the loss in downstream competition, increases the harm to consumers.
+
+The preceding analysis demonstrates how various horizontal and vertical effects of mergers interact in the model.
+We find that in our empirical application, EDM only results in net benefits for consumers in the pure vertical case, whereas all other complex mergers result in consumer harm.
+Indeed, we see with the downstream and partial downstream Republic/Santek examples that moving from a pure horizontal merger to a complex merger that adds EDM effects can actually result in more harm to consumers.
+In the complex mergers we study, we find that EDM benefits are often outweighted by RRC effecs and the loss of horizontal competition.
 
@@
 Removing this limitation is important as competition authorities are routinely called upon to investigate complex mergers where one or both of the merging parties are vertically integrated.
-Using a series of numerical simulations and the model of \citeN{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
-The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.
-We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
-We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships gives an incomplete picture of the merger's welfare impacts.
 
-Our results suggest that mergers that have horizontal aspects remain harmful for consumers in many instances, even when these mergers also have vertical aspects.
-This appears to be especially true when retailers have relatively more bargaining power, which in our model occurs when retailer margins are relatively large compared to wholesaler margins.
-Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.
+Our results suggest that mergers that have horizontal aspects can remain harmful for consumers in a number of instances, even when these mergers also have vertical aspects.
+In our examples, the loss in horizontal competition between Republic and Santek and RRC behavior by Santek outweighs even large decreases in input prices from EDM.
+This suggests that antitrust agencies have reason to be cautious when reviewing mergers between large firms that tout the possibility of countervailing EDM benefits.
 
-However, we of course have the caveat that our results are specific to our modeling framework.
+However, we of course have the caveat that our results are specific to our modeling framework and calibrated parameters.
 Certain alternative structures are likely to give different welfare results.
-Indeed, we found in \citeN{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
+Indeed, \citeN{ST2021} find that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.
 This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.
@@
 \small{\include{TrashData}}
-\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.  Volume is reported in thousands of tons, whereas prices and margins are reported in dollars per ton.}
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.  Volume is reported in thousands of tons, whereas prices, costs, and margins are reported in dollars per ton.}
 \label{tab:trashdata}
@@
 \centering
-\includegraphics[scale=.7]{output/chattanooga_tree.png}
-\caption{Structure of the Chattanooga Area Disposal and Collection Market}
+\includegraphics[scale=.9]{output/chattanooga_tree.png}
+\caption{Structure of the Chattanooga Area Disposal and Collection Markets}
 \label{fig:chatt_tree}
@@
 \centering
-\caption{Republic/Santek Simulation Results}
+\caption{Republic/Santek Merger Simulation Results}
 \includegraphics[scale=1.0]{output/TrashSimsFirm.png}\\
@@
 \centering
-\caption{Republic/Santek Simulations: Full vs. Partial Models}
-\includegraphics[scale=1.0]{output/TrashSimsCompareAll.png}\\
-\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
-\label{fig:TrashSimsFirms}
+\caption{Sensitivity of Consumer Harm from Republic/Santek to the Outside Share}
+\includegraphics[scale=.9]{output/trashelast.png}\\
+\footnotesize{Notes: The absolute value of the market elasticity corresponding to the change in outside good share is in parentheses. Baseline results assume 0 outside share.  }
+\label{fig:TrashElast}
 \end{figure}
@@
 \centering
-\caption{Simulated Welfare Effects from Mergers between  Republic/Santek, Santek/WasteConn, Santek/WM-ADS, and WM-ADS/Regional}
-\includegraphics{output/trashinterestingmergerbar.png}\\
+\caption{Sensitivity of Consumer Harm from Republic/Santek to the Nesting Parameter}
+\includegraphics[scale=.9]{output/trashnest.png}\\
+\footnotesize{Notes: Simulations use nested logit demand.
+``Integrated'' assumes that all vertically integrated products are in the same nest, while ``Landfill'' assumes that all collectors that use the same landfill are in the same nest.
+Baseline results assume a nesting parameter of 0.
+}
+\label{fig:TrashNest}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{Sensitivity of Consumer Harm from Republic/Santek to Bargaining Power}
+\includegraphics[scale=.9]{output/trashbarg.png}\\
+\footnotesize{Notes: The bargaining power parameter is set to be the same across all non-integrated retailer-wholesaler pairs.  The dashed horizontal line denotes the baseline value.}
+\label{fig:TrashBarg}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{Republic/Santek Merger Simulations: Full vs. Partial Models}
+\includegraphics[scale=1.0]{output/TrashSimsCompareAll.png}\\
 \footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\label{fig:TrashSimsCompareAll}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{Simulated Welfare Effects for Other Mergers}
+\includegraphics[scale=1]{output/trashinterestingmergerbar.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections. }
 \label{fig:TrashFakeOverall}
@@
 \centering
-\caption{Simulation Results from mergers between Santek/WasteConn, Santek/WM-ADS, and WM-ADS/Regional }
-\includegraphics{output/trashinterestingfirmbar.png}\\
-\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
+\caption{Price and Market Share Simulation Results for Other Mergers }
+\includegraphics[scale=.9]{output/trashinterestingfirmbar.png}\\
+\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.  The three mergers studied are Santek/Waste Connections (top panel), Santek/WMADS (middle panel), and WMADS/Regional (bottom panel).  }
 \label{fig:TrashFakeFirms}
@@
 \centering
-\caption{The sensitivity of consumer welfare in Santek/Republic merger simulations to changes in the bargaining power parameter.
-%`All' changes the bargaining power paramtere for all non-vertically integrated firms.
-`Santek:Republic' changes the bargaining power parameter for just the Santek/Republic product.
-`WasteConn:Republic' changes the bargaining power parameter for just the WasteConn:Republic product.
-The Dashed line displays baseline results.
-}
-\includegraphics{output/trashbarg.png}\\
-\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
-\label{fig:TrashBarg}
-\end{figure}
-
-\begin{figure}
-\centering
-\caption{The sensitivity of consumer welfare in the Republic/Santek simulations to changes in the outside share.
-The absolute value of the market elasticity corresponding to the change in outside share is in parentheses.
-Baseline results assume 0 outside share.}
-\includegraphics{output/trashelast.png}\\
-\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
-\label{fig:TrashElast}
-\end{figure}
-
-\begin{figure}
-\centering
-\caption{The sensitivity of consumer welfare in the Republic/Santek simulations to changes in the nesting parameter.`Integrated' assumes that all vertically integrated products are in a single nest, while `Landfill' assumes that all collectors that use the same landfill are in a single nest.
-A nesting parameter of 0 displays baseline results.
-}
-\includegraphics{output/trashnest.png}\\
-\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
-\label{fig:TrashNest}
+\caption{Simulation Results for Republic/Santek under Different Firm Configurations  }
+\includegraphics[scale=.9]{output/trashinterestingsantrep.png}\\
+\footnotesize{Notes: ``Base'' is the baseline merger of Republic/Santek as actually proposed.
+``Partial Up'' is the merger between an integrated Santek and the upstream Republic division.
+``Partial Down'' is the merger between an integrated Santek and the downstream Republic division.
+``Down'' is the merger between the downstream divisions of both Santek and Republic.
+``Up'' is the merger between the upstream divisions of both Santek and Republic.
+``Vertical'' is the merger between the upstream division of Santek and the downstream division of Republic.}
+\label{fig:TrashFakeRepSan}
 \end{figure}
@@
 Following the discussion of supply relationships in Section \ref{sec:application}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal and collection, Waste Connection disposal and collection) are assigned their full collection share.
-Capacity constrained integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
+Capacity constrained integrated firms (Republic disposal and collection) are assigned their collection share up to their available capacity, with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal and Republic collection, Waste Connections disposal and Republic collection).
 Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.
@@
 The CIS states that in the Chattanooga Area the post-merger HHI for SCCW collection would be approximately 5,551 post-merger with an increase of 2,660 points and that the combined market share of the merging parties is 73\%.
-Taking these figures as given we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size, and 2) the non-merging parties are of equal size.
-After re-scaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WM-ADS, 8.3\%, and Regional, 8.3\%.
+Taking these figures as given, we can recover the collection market shares under the assumptions that 1) the merging parties are of equal size and 2) the non-merging parties are of equal size.
+After re-scaling, this produces downstream market shares for Republic, 37.6\%, Santek, 37.6\%, Waste Connections, 8.3\%, WMADS, 8.3\%, and Regional, 8.3\%.
 
 Collection and disposal margins are calculated for Republic from data on revenue by line of service and from components of the cost of operations in the company's 2019 annual report.
-The data are reported at the company level and are not specific to the Chattanooga Area, but revenues are reported by collection segment.
+These data are reported at the company level and are not specific to the Chattanooga Area, but revenues are reported by collection segment.
 Collection costs in dollars per ton are estimated from Republic and Waste Management 2019 10k financial statements.
@@
 Santek, Waste Connections, and Regional do not publicly produce comparable financial statements.
-Instead, we assume that the cost structure is the same for the vertically-integrated companies, taking into account their individual tons disposed, and that WM-ADS and Regional share the same cost structure due to their lack of internal disposal capacity.
+Instead, we assume that the cost structure is the same for the integrated companies, taking into account their individual tons disposed, and that WMADS and Regional share the same cost structure due to their lack of internal disposal capacity.
 
@@
 
-% \begin{figure}
-% \centering
-% \includegraphics{output/TrashSimsCompare.png}
-% \caption{Comparison between Collection-Only and Full Vertical Simulations for Republic/Santek}
-% \label{fig:TrashSimsCompare}
-% \end{figure}
+%We use this model to explore the welfare implications of four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.
+Relative to \citeN{ST2021}, the first category, vertical mergers, appears in the previous paper and here serves as a baseline for comparison.
+The other three cases, which feature at least one merging firm that is already integrated, are new.
+
+%Our model-based simulations show certain patterns across mergers.
+For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.
+The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.
+Likewise, we find that integrated mergers are typically net harmful to consumers as well.
+The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.
+Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.
+We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.
+We find that the existence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less harm.
 
@@
 %The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
+
+%Using a series of numerical simulations and the model of \citeN{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.
+The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.
+We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm.
+We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships gives an incomplete picture of the merger's welfare impacts.

### 2024-10-07 doc/JIE/JINDECMS2023138-1Response.tex
Gloria's 10/3 edits

@@
 
-We have created new Section 3.2 that explains our calibration.
+We have created a new Section 3.2 that explains our calibration.
 We tie our explanation directly to the data inputs, which appear in Table 1, and to the set of equations we use for identification, which are the firm first order conditions in Section 2.
@@
 \item \emph{Be sure to compare simulations that allow for the complex nature of Republic/Santek versus the more vanilla simulations considered in ST (2021).
-This is important to capture this papers contribution.}
+This is important to capture this paper's contribution.}
 
@@
 We have added three new sensitivity analyses, at the end of Section 3.3, in order to address this point.
-We show how our simulated consumer welfare results are impacted by (1) the extent of substitution outside the market, (2) the use of the logit versus the nested logit, and (3) the bargaining power of retailers vesus wholesalers.
+We show how our simulated consumer welfare results are impacted by (1) the extent of substitution outside the market, (2) the use of the logit versus the nested logit, and (3) the bargaining power of retailers versus wholesalers.
 In each of these exercises, we vary the relevant parameter in smooth increments and show how welfare changes relative to the baseline.
@@
 I could not easily find the relevant parts of the package.
-It would be great if this machinery was also added to antitrust , but I would like to see a stand-alone replication package.}
+It would be great if this machinery was also added to antitrust, but I would like to see a stand-alone replication package.}
 
      Much of the coding machinery we used has already been incorporated into the next version of the \texttt{antitrust} R package.
-A new function, \texttt{vertical.barg}, has been added which uses observed shares, prices and some margins to calibrate demand parameters and simulate the effect of complex mergers.
+A new function, \texttt{vertical.barg}, has been added that uses observed shares, prices, and some margins to calibrate demand parameters and simulate the effect of complex mergers.
 We understand that having a replication package would also be helpful for readers, and we are planning to post a package once the paper has been finalized.
-Aside from containing all the code and data needed to run all the simulations presented in the paper, the replication package will contain code that allows calibration when downstream costs but not prices are observed.
+Aside from containing all the code and data needed to run all the simulations presented in the paper, the replication package will contain code that allows calibration when downstream costs rather than prices are observed.
 
@@
     \item  \emph{It might make sense if a retailer and manufacturer had meaning in the model.
-One suggestion is to have $\delta$ be comprised of two components, and manufacturer effect, and a retailer effect, where the retailer effect is perfectly correlated across manufacturer products within a retailer, and the manufacturer effect is perfectly correlated across retailer for the same product.
-So something like  $\delta_{rw} = \delta_{r} + \delta_{w}$}
+One suggestion is to have $\delta$ be comprised of two components, a manufacturer effect, and a retailer effect, where the retailer effect is perfectly correlated across manufacturer products within a retailer, and the manufacturer effect is perfectly correlated across retailers for the same product.
+So something like  $\delta_{rw} = \delta_{r} + \delta_{w}$.}
 
@@
 We did one version where we grouped together all the products that use the same firm for collection and disposal (``integrated'' in Figure 4) and another version where we grouped together all products that use the same landfill (``landfill'' in Figure 4).
+The discussion of these results is in Section 3.3.
 
@@
 
-    This is a good point.  We have added a footnote to the discussion of that equation (now equation ([8 XXXX])) mentioning the Salinger (1991) result and the Luco and Marshall (2020) paper.
+    This is a good point.  We have added a footnote to the discussion of that equation (now equation (8)) mentioning the Salinger (1991) result and the Luco and Marshall (2020) paper.
 
@@
 
-Thank you for this helpful feedback.  We have removed both sets of material, pivoted the draft to focusing on the Republic/Santek merger, which has greatly streamlined the paper.
+Thank you for this helpful feedback.  We have removed both sets of material and pivoted the draft to focusing on the Republic/Santek merger, which has greatly streamlined the paper.
 
@@
 We have personal experience with being told Crawford et al.
-(2018) was too complicated to be used as part of the court testimony we were preparing in a merger litigation.
+(2018) was too complicated to be used as part of the court testimony we were helping to prepare in a merger litigation.
 
-    We have added to the literature review in Section 1 to make clear that there are past papers that have studied complex mergers, including the two you mention above.
+    We have added to the literature review in the Introduction to make clear that there are past papers that have studied complex mergers, including the two you mention above.
 We also have added a discussion of earlier theory literature on vertical integration to provide more context for our paper.
@@
 We have added a discussion of these issues in Section 3.2.
+
+    We understand that simultaneous pricing and linear fees may not be appropriate for all applications.
+We believe they are useful for some industries, like cable programming, where prices are fixed for periods of time and linear fee contracts are common.
+We have added a discussion of these limitations in the Introduction.
 
@@
 (2018).
-We use the term ``RRC'' in the way that they use it, to mean a shift in bargaining leverage that allows a vertically integrated firm to raise input prices to a downstream rival.
+We use the term ``RRC'' in the way that they use it, to mean a shift in bargaining leverage that causes a vertically integrated firm to raise input prices to a downstream rival.
 Crawford et al.
@@
 (2018) do not use RRC to mean the incentive modeled by Salop and Scheffman (1983), where a dominant firm chooses an action in order to directly impact downstream retail prices.
-We have added footnote [10 XXXX] in the discussion around equation ([9 XXXX]) to clarify this distinction.
+We have added footnote 10 in the discussion around equation (9) to clarify this distinction.
 
@@
     We have removed those simulations from the paper.
-All the simulations and calibration we have done rely on the first order conditions of the model in Section 2, to ensure that the resulting prices and market share are consistent with the theory.
+All the simulations and calibration we have done rely on the first order conditions of the model in Section 2, to ensure that the resulting prices and market shares are consistent with the theory.
 We have added a discussion clarifying this point in Section 3.2.
@@
 The theoretical model does not fit the setting in the case.
-First, the bargaining model is not present nor discussed in the DOJ Lawsuit.
+First, the bargaining model is not present nor discussed in the DOJ lawsuit.
 Second, the RRC discussed in the DOJ lawsuit refers to foreclosure/exclusionary conduct concerns.
-This feature could typically has different welfare implications.
+This feature could typically have different welfare implications.
 See, e.g., Ordover et al.
@@
 
-    The DOJ complaint in Republic/Santek at paragraph 47 explains that the DOJ was alleging harm due to increased upstream disposal prices to rival downstream collection firms, not due to total exclusion or foreclosure of these collection firms.
+    The DOJ complaint in Republic/Santek at paragraph 46 explains that the DOJ was alleging harm due to increased upstream disposal prices to rival downstream collection firms, not just due to the possibility of total exclusion or foreclosure of these collection firms.
 Specifically, the complaint states that the merged firm would have an ``incentive to raise the MSW disposal costs of independent haulers because Republic---no longer confronting competition from Santek in SCCW collection---would capture more of the business lost by independent haulers in the SCCW collection market.''  We agree that the complaint does not put this story in terms of bargaining, but the incentive discussed in this passage is very similar to the mechanism that leads to increased bargaining leverage and hence higher wholesale prices in our model.

### 2024-10-07 doc/VerticalArrangements.tex
Gloria's 10/3 edits

@@
 
-\title{Beyond Horizontal and Vertical:\\ Simulating the Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
+\title{Beyond Horizontal and Vertical:\\ Measuring the Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.
 Furthermore, the views expressed here should not be purported to reflect those of the U.S.
@@
 \begin{abstract}
-\noindent We study simulated mergers that have both horizontal and vertical aspects using a model with Bertrand competition downstream and Nash Bargaining upstream.
-We apply this framework to the Republic/Santek merger in waste management services, comparing the actual merger to different counterfactual mergers involving various vertical and horizontal combinations.
-Using this example, we find almost no instances where the elimination of double marginalization outweighs merger harms to consumers.
+\noindent We study mergers that have both horizontal and vertical aspects using a simulation model with Bertrand competition downstream and Nash Bargaining upstream.
+We apply this framework to the 2021 Republic/Santek merger in waste management services, comparing the actual merger to different counterfactual mergers involving various vertical and horizontal combinations.
+Within this example, we find almost no instances where the elimination of double marginalization outweighs merger harms to consumers.
 Furthermore, we show how analyzing the upstream and downstream markets separately can lead to inaccurate results.
@@
 However, for both historical and practical reasons, antitrust research and policy analysis has generally maintained this dichotomy.
-Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
+Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).
 As a result, the antitrust community has done little research into ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects.
-We have little guidance for how our merger simulation models and related tools apply to these complex mergers.
+We have scarce guidance for how our merger simulation models and related tools apply to these complex mergers.
 
@@
 They have both been superseded by a joint document issued in 2023.
-}  However, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical aspects of competition.
+}  However, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical competition.
 As part of the revision effort, these agencies requested public comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g.
@@
 We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.
-We highlight how the results are effected by several key parameters and assumptions, namely the market elasticity, the use of flat logit versus nested demand preferences, and the bargaining parameter.
+We highlight how the results are affected by several key parameters and assumptions, namely the market elasticity, the use of flat logit versus nested demand preferences, and the bargaining parameter.
 Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts and misrepresent the extent of consumer harm.
 
-This empirical example allows us to explore the implications of the model for complex mergers.
+This empirical example allows us to study the implications of the model for complex mergers.
 Starting from the Republic/Santek merger as proposed, we compare our baseline simulation results to those for counterfactual mergers between other firms in the market and to those for counterfactual mergers between different combinations of the upstream and downstream components of Republic and Santek.
-This exercise facilitates comparison between how the model behaves for complex mergers versus the more straightforward vertical and horizontal mergers that appear in, for instance, \citeN{ST2021}.
+This exercise facilitates comparison between how the model behaves for complex mergers versus the more straightforward vertical and horizontal mergers that appear in \citeN{ST2021}.
 Based on these simulations, we find that the complex mergers we study all result in net harm to consumers, despite the presence of large EDM effects in certain cases.
@@
 Although this assumption means bargaining is inefficient, linear input prices are widely observed in some industries, perhaps because of difficulties with implementing or monitoring more complicated contracts.\footnote{One example of an industry with linear wholesale pricing that has been featured in several vertical merger proceedings is cable programming distribution, as detailed in \citeN{Rogerson2014}.}  Third, our model assumes that upstream and downstream prices are set treating all other prices as fixed.
-Although we make this assumption in part for simplicity, it also may be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.
+Although we make this assumption in part for simplicity, it also can be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.
+\citeN{CLWY2017} use a similar assumption.
 In contrast, \citeN{BBMM2023} allow downstream firms to treat their own prices as adjustable during upstream negotiations.
@@
 \citeN{Riordan2008} has an excellent summary.
-Broadly speaking, this research explores how incentives to disadvantage rivals could arise with vertical integration, and in turn shows how vertical mergers could negatively impact consumer welfare.
+Broadly speaking, this research explores how incentives to disadvantage rivals could arise with vertical integration, and as a corollary shows how vertical mergers could negatively impact consumer welfare.
 See, for example, \citeN{SalopScheffman1983}, \citeN{Salinger1988}, \citeN{OSS1990}, and \citeN{Riordan1998}.
@@
 
-Meanwhile, the structural empirical industrial organization literature has also studied the competitive effects of vertical integration as part of a greater exploration of the impacts of bargaining power in vertical supply chains.
+Meanwhile, the structural empirical industrial organization literature has studied the competitive effects of vertical integration as part of a greater exploration of bargaining power in vertical supply chains.
 Many of these papers are based on the model developed in \citeN{HW1988}, modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.
@@
 \citeN{CLWY2017}).
-These structural empirical papers explore the welfare effects of various competitive interactions, including complex mergers, in specific industries.\footnote{There is a companion empirical literature that relies more on regression analysis to identify changes in market outcomes due to vertical integration (e.g.
+These papers explore the welfare effects of various competitive interactions, including complex mergers, in specific industries.\footnote{There is a companion empirical literature that relies more on regression analysis to identify changes in market outcomes due to vertical integration (e.g.
 \citeN{Chipty2001} on video programming distribution).
@@
 where $\lambda_w \in [0,1]$ captures the bargaining power of retailers relative to wholesaler $w$.
-The $d^r (\mathcal{W}^r \setminus \{w\})$ is the disagreement payoff for the retailer, and $d^w (\mathcal{R}^w\setminus \{r\})$ is the disagreement payoff for the wholesaler.
+The $d^r (\mathcal{W}^r \setminus \{w\})$ is the disagreement payoff for retailer $r$, and $d^w (\mathcal{R}^w\setminus \{r\})$ is the disagreement payoff for wholesaler $w$.
 The disagreement payoff for the retailer is the profits it would earn if it ceases to sell this wholesale product,
@@
 \end{equation}
-where $s_{rx}(\mathcal{W}^r \setminus \{w\})$ is the realized downstream market share for product $rx$ when retailer $r$ does not sell wholesaler $w$'s product.
+where $s_{rx}(\mathcal{W}^r \setminus \{w\})$ is the realized downstream market share for product $rx$ when retailer $r$ does not carry wholesaler $w$'s product.
 Similarly, the disagreement payoff of the wholesaler is
@@
 \end{equation}
-which is earned if wholesaler $w$ does not offer its product to retailer $r$.
+which is earned if wholesaler $w$ does not offer its product through retailer $r$.
 
-We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
+We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{This assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.
 }  Then the first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
@@
 \end{equation}
-The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) \equiv s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.
+The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) = s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.
 Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.
@@
 Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $j$, these losses are softened due to a potential for diversion to retailer $t$, which we label the ``RRC effect.''\footnote{The assumption of simultaneous downstream price setting rules out RRC effects in the style of \citeN{SalopScheffman1983}, in that here the wholesaler does not have an incentive to raise prices in order to directly impact the pricing strategies of rival retailers.
-Rather, our model has the RRC effect seen in \citeN{CLWY2017}, where the wholesaler has better bargaining leverage over rival retailers due to the possibility of increased sales volume (conditional on prices) for their merged downstream retail partner should negotiations break down.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
+Rather, our model has the RRC effect seen in \citeN{CLWY2017}, where the wholesaler has better bargaining leverage over rival retailers due to the possibility of increased sales volume (conditional on prices) for the merged downstream retail partner should negotiations break down.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.
 These impacts tend to raise the resulting input price to retailer $j$.
@@
 \end{equation}
-Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $t$ raises its prices.
+Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $t$ raises its price.
 EDM between $v$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $v$ sold by retailer $s$, those units may earn a larger margin.
@@
 The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain an additional wholesale product at marginal cost, which in turn would tend to raise the value of diverted sales to that product.
-Whether these additional incentives to raise prices will dominate the direct EDM impact is an empirical question.
+Whether these additional incentives to raise prices dominate the direct EDM impact is an empirical question.
 Furthermore, if we instead considered a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.
@@
 The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $t$, both of which may potentially recapture sales should retailer $j$ lose access to product $v$.
-The merged firm has higher bargaining leverage as a result.
+The merged firm has additional bargaining leverage as a result.
 If instead the integrated retailer/wholesaler $tv$ were to merge with a standalone wholesaler, then the profits earned by firm $tv$ should the negotiation with retailer $j$ fail are augmented with the earnings of the additional wholesaler rather than with the earnings of retailer $s$.
@@
 Some haulers are vertically integrated and operate their own disposal facilities.
-Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
+Integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity.
 Disposal customers include private waste haulers without their own disposal assets as well as local governments that collect their citizens' waste themselves.
@@
 In addition, another large, vertically integrated company, Waste Connections, existed in Chattanooga at the time of the merger and was the parties' sole competitor in MSW disposal.
-National firms Waste Management and ADS also owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Appendix \ref{app:trashdata}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
+National firms Waste Management and ADS also owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed below.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter into contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated.
 Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single entity (henceforth ``WMADS'') because they merged with one another in 2020, and that transaction was completed before the filing of the CIS for Republic/Santek.
@@
 For our sensitivity analysis using nested logit demand, we divide the set of retailers into a series of nests, with each nest indexed by $g$.
-The set of retailers within each nest is denoted by $\mathcal{R}_g$, and the set of all nests is denoted by $\mathcal{G}$.
+The set of retailers within each nest is $\mathcal{R}_g$, and the set of all nests is  $\mathcal{G}$.
 Then the share for product $rw$ is given by
@@
 Then given that demand system and the firm pricing first order conditions of the model, one would recover estimates of marginal costs.
-However, antitrust practitioners usually do not have access to the larger cross section or panel datasets necessary for demand estimation, instead observing only shares and prices for one or two markets, usually at only a single point in time.
+However, antitrust practitioners frequently do not have access to the large cross section or panel datasets necessary for demand estimation, instead observing only shares and prices for one or two markets, usually at only a single point in time.
 In our context, this means having a snapshot of data for the nine products in the Chattanooga Area.
@@
 We follow calibration methods similar to those described in \citeN{ST2021}.
-The derivatives of logit demand are functions only of market shares and the price coefficient $\alpha$, as $\partial s_{rx}/\partial p_{rw} = \alpha s_{rx} s_{rw}$  for $x \neq w$ and $\partial s_{rw}/\partial p_{rw} = -\alpha s_{rw} (1-s_{rw})$ for $x=w$.
-Furthermore, the logit also implies that $\Delta s_{xy} (\mathbb{W}^r \setminus \{w\}) = s_{rw} \left(s_{xy}/(1-s_{rw})\right)$.
+The derivatives of logit demand are functions only of market shares and the price coefficient $\alpha$, as $\partial s_{rx}/\partial p_{rw} = \alpha s_{rx} s_{rw}$  for $x \neq w$, $\forall r$ and $\partial s_{rw}/\partial p_{rw} = -\alpha s_{rw} (1-s_{rw})$ for $x=w$.
+Furthermore, the logit also means that $\Delta s_{xy} (\mathbb{W}^r \setminus \{w\}) = s_{rw} \left(s_{xy}/(1-s_{rw})\right)$.
 Therefore, the downstream pricing first order conditions and the upstream bargaining first order conditions described in Section \ref{sec:theory} form a system of equations that are functions only of market shares, margins, the downstream price coefficient, and the upstream bargaining parameters.
@@
 The process for identifying the downstream parameters is similar to that seen in \citeN{WF1994}.
-Conditional on the other data, knowledge of a downstream margin determines the how sensitive retail customers are to price, as a higher margin implies less elastic demand.
+Conditional on the other data, knowledge of a downstream margin determines how sensitive retail customers are to price, as a higher margin implies less elastic demand.
 Given this implied relationship, optimal downstream prices can be recovered using the firm first order conditions with data on shares and costs.
-Then knowledge of downstream market shares identifies the quality of each product, as higher shares conditional on prices implies higher quality.
+Then knowledge of downstream market shares identifies the quality of each product, as higher shares conditional on prices imply higher quality.
 Turning to the upstream model, the remaining unknown objects are the bargaining parameters.
 Conditional on the data and the downstream model, bargaining power is identified by the relative sizes of the upstream and downstream profit margins.
-When downstream firms have higher margins than downstream firms, that is an indicator that retailers have more bargaining power.
+When downstream firms have higher margins than upstream firms, that is an indicator that retailers have more bargaining power.
 
 Our data comes from several public sources.
-We obtain information on disposal volumes and landfill usage from the Tennessee Department of Environmental Quality.
+We obtain 2019 information on disposal volumes and landfill usage from the Tennessee Department of Environmental Quality.
 We assume that Republic purchases disposal from both Santek and Waste connections according to their shares of available excess MSW disposal capacity of 63.8\% and 36.2\%, respectively.
@@
 We gather information on disposal prices from the 2019 Waste Business Journal’s Directory of Waste Processing \& Disposal Sites.
-Margin and downstream cost information comes from annual reports and public financial statements, under the assumption that the national averages reported there are reasonable approximations for the Chattanooga market.
+Margin and downstream cost information comes from 2019 annual reports and public financial statements, under the assumption that the national averages reported there are reasonable approximations for the Chattanooga market.
 Our data sources and construction are described in more detail in Appendix \ref{app:trashdata}.
@@
 We expect that the DOJ would have had access to confidential information during their investigation, which could have allowed them to generate more precise results.
-Nevertheless, we believe our results are qualitatively accurate, as they are consistent with the anti-competitive concerns raised in the Republic/Santek complaint and CIS.
+Nevertheless, we believe our results are qualitatively accurate, as they are consistent with several of the anti-competitive concerns raised in the Republic/Santek complaint and CIS.
 
@@
 We can examine the sensitivity of our results to the first issue by varying our chosen value for the share of the outside good.
-The form of the logit implies that this share is proportional to the market elasticity of demand.\footnote{The market elasticity with respect to a price change for all inside products is $ -\alpha \overline{p} s_{00}$, where $s_{00}$ is the share of the outside good and $\overline{p}$ is the share-weighted average of prices among the inside goods.
+The form of the logit implies that this share is proportional to the market elasticity of demand.\footnote{The market elasticity with respect to a price change for all inside products is $ -\alpha \overline{p} s_{00}$, where $s_{00}$ is the share of the outside good and $\overline{p}$ is the share-weighted average price among the inside goods.
 }  Our baseline results assume that the outside share is zero, following the assertion made by the DOJ that there are essentially no reasonable substitutes for SCCW collection or MSW disposal.\footnote{See the CIS: U.S and Plaintiff States v.
@@
 Our results appear in Figure \ref{fig:TrashNest}.
-The baseline simulation is represented by the point at a nesting parameter of 0 on the left of in figure.
+The baseline simulation is represented by the point at a nesting parameter of 0 on the left in the figure.
 As $\sigma$ increases, consumer harm under integrated nesting increases and consumer harm under landfill nesting decreases.
@@
 
-Another key part of our model is the bargaining power parameters $\lambda$.
+Another key part of our model is the bargaining power parameters.
 As discussed in \citeN{ST2021}, vertical mergers tend to benefit consumers when wholesalers have relatively more power versus when retailers have relatively more power.
@@
 Furthermore, even when varying these parameters causes the results to meaningfully change, they shift in ways that are predictable and intuitive, as demonstrated by the monotonic relationships seen in Figures \ref{fig:TrashElast}, \ref{fig:TrashNest}, and \ref{fig:TrashBarg}.
-These results suggest that the model is useful for analyzing complex mergers in a policy contex.
+These results suggest that the model is useful for analyzing complex mergers in a policy context.
 
@@
 The results of this comparison are summarized in  Figure \ref{fig:TrashSimsCompareAll}.
-Ignoring these vertical relationships in the downstream-only model gives in an estimated 9.2\% increase in collection prices with an 16.4\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million.
+Ignoring vertical relationships in the downstream-only model gives in an estimated 9.2\% increase in collection prices with an 16.4\% increase in the merging parties' price, consumer harm of \$10 million annually, and net overall harm of approximately \$2 million.
 Estimated consumer harm is 60\% higher in the full vertical model and is offset by markedly larger producer benefits.
@@
 This nets out into the small shift in profits from upstream to downstream firms seen in Figure \ref{fig:TrashFakeOverall}.
-Prices in downstream collection rise slightly, which gives a relatively small loss of consumer welfare.
+Prices in downstream collection rise slightly, which gives a relatively small loss in consumer welfare.
 
@@
 }  This setup allows us to study additional complex mergers involving the most popular products, those sold by Republic and Santek.
-Specifically, we examine the following mergers: (1) a pure vertical merger between Santek's upstream disposal and Republic's downstream collection, labeled ``Vertical'' in the figure; (2) a pure upstream merger between Santek's disposal and Republic's disposal, labeled ``Up''; (3) a pure downstream merger between Santek's collection and Republic's collection, labeled ``Down''; (4) a partial upstream merger between a vertically integrated Santek and Republic's disposal, labeled ``Partial Up''; and (5) a partial downstream merger between a vertically integrated Santek and Republic's collection, labeled ``Partial Down.''  The last two are complex mergers, while the first three are similar to the non-complex cases studied in \citeN{ST2021}.
+Specifically, we examine the following mergers: (1) a pure vertical merger between Santek's upstream disposal and Republic's downstream collection, labeled ``Vertical'' in Figure \ref{fig:TrashFakeRepSan}; (2) a pure upstream merger between Santek's disposal and Republic's disposal, labeled ``Up''; (3) a pure downstream merger between Santek's collection and Republic's collection, labeled ``Down''; (4) a partial upstream merger between a vertically integrated Santek and Republic's disposal, labeled ``Partial Up''; and (5) a partial downstream merger between a vertically integrated Santek and Republic's collection, labeled ``Partial Down.''  The last two are complex mergers, while the first three are similar to the non-complex cases studied in \citeN{ST2021}.
+Furthermore, treating Waste Connections as separate upstream and downstream firms makes the overall environment more comparable to \citeN{ST2021}.
 In any of the instances where only one part of Republic or Santek are merging, we assume that the other part is operating independently.
@@
 Recall from Figure \ref{fig:TrashSimsFirms}, this was also seen in the baseline simulation.
-However, here we find that the RRC effects are much smaller, than in the baseline with Santek raising its disposal prices to WMADS and Regional by about 16\% versus 88\% in the baseline.
+However, here we find that the RRC effects are much smaller, with Santek raising its disposal prices to WMADS and Regional by about 16\% versus 88\% in the baseline.
 This difference occurs because in the pure vertical case, the highly popular Santek-Santek and Republic-Republic products are not fully owned by the merging firms, which means that some of the benefits from customers induced to divert from withholding disposal would accrue to the independent Santek collection and Republic disposal firms.
@@
 
-When we compare the pure downstream case to the partial downstream merger between an integrated Santek and Republic's collection assets, we see that consumer harm increases.
-This occurs even though the move to the partial merger introduces a large EDM effect for Santek's sales of disposal to Republic.
+When we compare the pure downstream case to the partial downstream merger between an integrated Santek and Republic's collection assets (third row of Figure \ref{fig:TrashFakeRepSan}), we see that consumer harm increases.
+This occurs even though the move to the partial downstream merger introduces a large EDM effect for Santek's sales of disposal to Republic.
 That input price falls by over 70\%.
@@
 The preceding analysis demonstrates how various horizontal and vertical effects of mergers interact in the model.
-We find that in our empirical application, EDM only results in net benefits for consumers in the pure vertical case, whereas all other complex mergers result in consumer harm.
+We find that in our empirical application, EDM only results in net benefits for consumers in the pure vertical case, whereas all complex mergers result in consumer harm.
 Indeed, we see with the downstream and partial downstream Republic/Santek examples that moving from a pure horizontal merger to a complex merger that adds EDM effects can actually result in more harm to consumers.
-In the complex mergers we study, we find that EDM benefits are often outweighted by RRC effecs and the loss of horizontal competition.
+In the complex mergers we study, we find that EDM benefits are outweighed by RRC effects and the loss of horizontal competition.
 
@@
 \includegraphics[scale=.9]{output/trashelast.png}\\
-\footnotesize{Notes: The absolute value of the market elasticity corresponding to the change in outside good share is in parentheses. Baseline results assume 0 outside share.  }
+\footnotesize{Notes: The absolute value of the market elasticity corresponding to the outside good share is in parentheses. Baseline results assume 0 outside share.  }
 \label{fig:TrashElast}
@@
 \footnotesize{Notes: Simulations use nested logit demand.
-``Integrated'' assumes that all vertically integrated products are in the same nest, while ``Landfill'' assumes that all collectors that use the same landfill are in the same nest.
+``Integrated'' assumes that all vertically integrated products are in the same nest, while ``Landfill'' assumes that all collection firms that use the same landfill are in the same nest.
 Baseline results assume a nesting parameter of 0.
@@
 MSW disposal prices are collected from the 2019 Waste Business Journal's Directory of Waste Processing \& Disposal Sites.
-The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in dollars per ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.} Republic and Santek both operate multiple landfills in the market with different prices.
+The measure of price is the ``gate rate,'' which is the posted price at the landfill, measured in dollars per ton.\footnote{Disposal prices for large customers may be bilaterally negotiated instead of paying the gate rate.
+We assume this behavior is not significant in Chattanooga.} Republic and Santek both operate multiple landfills in the market with different prices.
 The price used in the analysis for each is the volume-weighted average for their landfills.

### 2024-10-07 code/TrashSims.R
minor edits to figures 7 and 9 in paper See Gloria's 10/2 email.

@@ -919,12 +919,12 @@ trashinterestingmergerbar <- ggplot(data=filter(mkt_mergersweep,!(Acquirer=="San
             #hjust=1.1,
             vjust=0.9,
             color="black",
-   position=position_dodge(width=.9),size =3) +# coord_flip()+
+   position=position_dodge(width=.9),size =2) +# coord_flip()+
   theme_bw() +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_y_discrete(limits=rev)
 
-ggsave(filename="./output/trashinterestingmergerbar.png",trashinterestingmergerbar,width = 6,height=4)
+ggsave(filename="./output/trashinterestingmergerbar.png",trashinterestingmergerbar,width = 6.5,height=4.5)
 
 
 
@@ -955,7 +955,7 @@ trashinterestingsantrep <- ggplot(data=filter(mkt_mergersweep,Acquirer %in% c("S
             position=position_dodge(width=.9),size =3) +# coord_flip()+
   theme_bw() +
   theme(legend.position="bottom",axis.text.x = element_text(angle = 45, hjust = 1)) +
-  scale_y_discrete(limits=rev) +ylab("Pre-merger Arrangement")
+  scale_y_discrete(limits=rev) +ylab("Merger Arrangement")
 
 ggsave(filename="./output/trashinterestingsantrep.png",trashinterestingsantrep,width = 7,height=4)
 

### 2024-10-07 doc/JIE/RefereeResponses.tex
minor nits before JIE submission

\documentclass[12pt]{article}
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs
\usepackage{parskip}

\newcounter{para}
\newcommand\mypara{\par\textbf{\stepcounter{para}\thepara)\space}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Preliminaries
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Detailed Responses to the Referees for \\ ``Beyond Horizontal and Vertical'' \\MS 2023138}
\author{Margaret Loudermilk, Gloria Sheu, and Charles Taragin}
\date{October 2024}
\begin{document}
\maketitle

We appreciate the opportunity to revise our paper, and we have made substantial changes in response to the suggestions from the editor and referees, which we found quite helpful.  The largest change is that we have removed the results for sampled mergers and now instead focus on the Republic/Santek merger example. In this document, the editor and referee comments are in italics, followed by our response.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Editor
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Response to the Editor}
\emph{My biggest concern is that I don’t understand why the range (much less distribution) of parameters you consider is interesting. Therefore, I propose you combine Sections 3 and 4 using the Republic/Santek merger as a ``central'' point of calibration and exploring perturbations of that point to establish how changes to the underlying model assumptions impact outcomes of interest.}

We understand the concern about the sampling distribution of our merger simulations in the previous draft of the paper.  In response, we took your suggestion of combining our simulation exercises into one section centered on the Republic/Santek example.  This is now Section 3 of the paper.

\emph{In particular, I will be looking for you to deliver on the following:}
\begin{enumerate}
\item  \emph{Be very clear on how the central point is parameterized/calibrated. Be clear about what data is used to pin down what parameters. Also note what parameters you will vary either because they are not well identified with available data or because you want to consider sensitivity.}

We have created a new Section 3.2 that explains our calibration.  We tie our explanation directly to the data inputs, which appear in Table 1, and to the set of equations we use for identification, which are the firm first order conditions in Section 2.  We have considered in Section 3.3 the sensitivity of our results to the share of the outside good/market elasticity, to adding a nesting parameter, and to the size of the bargaining parameter, which we discuss further below.

\item \emph{Be sure to compare simulations that allow for the complex nature of Republic/Santek versus the more vanilla simulations considered in ST (2021). This is important to capture this paper's contribution.}

This was an extremely useful comment in helping us focus our simulations on the most interesting cases.  We have added a new Section 3.5 that walks through examples of complex mergers alongside similar ``non-complex'' mergers like those in ST (2021).  For example, we show how our results change when moving from a pure vertical merger to a merger where both firms are already integrated and from a pure horizontal merger to a merger where one of the firms is already integrated.  This allows us to, for example, show how layering in additional vertical aspects to a horizontal merger can impact welfare.

\item \emph{Be sure to consider how the nature of bargaining and downstream diversion will impact the results. For the later, you would need to go beyond a multinational logit model, but a nested logit should be able to capture the intuition. I'm not asking you to estimate substitution in this context since it doesn't seem like you have the data to do so (although I'd love to be shown wrong on this point!). I do want to know whether varying substitution patterns which fit the data you do have are likely to alter the qualitative outcomes of the merger simulation.}

We have added three new sensitivity analyses, at the end of Section 3.3, in order to address this point.  We show how our simulated consumer welfare results are impacted by (1) the extent of substitution outside the market, (2) the use of the logit versus the nested logit, and (3) the bargaining power of retailers versus wholesalers.  In each of these exercises, we vary the relevant parameter in smooth increments and show how welfare changes relative to the baseline.  The other aspects of the simulations are held fixed, allowing for cleaner comparisons.  We find that our results change in an intuitive manner that we believe will help practitioners understand the impact of the key parameters on the model's conclusions.

As part of this change, we removed the logit demand function from the theory discussion in Section 2, to make clear that the model can accommodate multiple forms of demand.  We introduce our baseline logit assumption and the alternative nested logit assumption later in Section 3 when we come to our specific empirical application.

\end{enumerate}

\emph{When presenting your results, please consider whether the simulations should be presented as the result of a statistical sampling (as is done now) or more in the vein of counterfactual analysis where you vary one parameter at a time. The argument for the statistical sampling approach is that parameters may interact---although you haven't clearly established that in most of the
current results. The argument for the one parameter at a time approach is that it can be more closely tied to the institution presented in the theory section. If you are tempted to do the sampling approach because there are many parameters to vary, consider whether you should instead focus the reader's attention on the most important/illustrative results.}

As mentioned in our responses above, we have shifted away from the statistical sampling approach.  In the sensitivity analysis, we vary one parameter at a time to facilitate comparisons.  In comparing across different mergers, we have chosen a set of illustrative cases that give the reader concrete examples to focus on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% R1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\break
\section{Response to Referee 1}

\begin{enumerate}

    \item \emph{I hope that the authors provide the code so that researchers can (a) run their own simulations and (b) use the code for their own research. The authors should follow the AER/ReStud guidelines that are becoming standard in the profession (\footnotesize{\url{https://aeadataeditor.github.io/aea-de-guidance/preparing-for-data-deposit.html}}). Basically, make the code easily readable and post it to a code/data archive website like Zenodo (and link to it from the author's webpage and the paper).}

    \emph{Note: Footnote 16 states that the antitrust R package contains the code to do the simulations. I could not easily find the relevant parts of the package. It would be great if this machinery was also added to antitrust, but I would like to see a stand-alone replication package.}

     Much of the coding machinery we used has already been incorporated into the next version of the \texttt{antitrust} R package.  A new function, \texttt{vertical.barg}, has been added that uses observed shares, prices, and some margins to calibrate demand parameters and simulate the effect of complex mergers. We understand that having a replication package would also be helpful for readers, and we are planning to post a package once the paper has been finalized. Aside from containing all the code and data needed to run all the simulations presented in the paper, the replication package will contain code that allows calibration when downstream costs rather than prices are observed.

    \item \emph{I think the authors could ditch the Republic/Santek example. The authors admit that the example is for ``illustrative" purposes. Instead of a single example, the authors could choose a few representative mergers and look at specific fake data that mimics those cases. Eg. many upstream and few downstream or vise versa.}

    Based on feedback from the editor, we ended up removing the other simulations and focusing on the Republic/Santek example.  However, in the spirit of this comment, we chose a few representative mergers within the Republic/Santek context to provide the reader with additional intuition.

    \item \emph{There is one thing I really like about Section 4: the comparison between a naive merger analysis (eg. downstream/upstream horizontal effects only) and the more comprehensive analysis that includes all of the vertical effects. I would like to see this type of analysis in Section 3.}

    Thank you for your feedback.  We have kept that comparison with the downstream- and upstream-only models, and have moved it into Section 3 to go with the additional examples we have added for Republic/Santek.

    \item \emph{I think that the authors overclaim a bit. In the abstract (and this sentiment is repeated in the paper, for example on page 2) the authors: ``In our model, we find that mergers with both horizontal and vertical characteristics typically harm consumers."}

    \emph{First, I think the more important contribution is to point out that a naive analysis that ignores the vertical aspect of a merger can mis-quantify the effects of the merger (see point 3 above) or that the EDM benefits are typically small (a point already made).}

    \emph{Second, mergers arise in nature (and are prosecuted) endogenously. The median pre-merger HHI in the simulations is between 2800 and 4300 and the median Delta HHI is over 1000. Most of the mergers analyzed would be presumed harmful based on the classic Cournot intuition or the more recent Nocke/Whinston intuition. I think the authors should avoid statements like in their abstract that make it seem they are drawing generic conclusions about merger enforcement.}

    We understand this concern and have now removed those generic-sounding statements.  Any statements we make now are tied to the specific Republic/Santek example, and we have tried to make that clear in the text.

\textbf{Smaller Comments}

    \item \emph{Table 1: It seems a bit suspicious that the price quantiles are all integers. Maybe I am missing something, but the domain should be some subset of the real number line, making integers very unlikely.}

    This occurred because we rounded all those numbers to integers.  Sometimes we make choices in the name of clarity that backfire.  That table has now been removed, since those simulations no longer appear in the paper.

    \item \emph{Simulation details. I don't understand why shares are drawn from a Dirichlet distribution (or any distribution), Shares are endogenous functions of model primitives. Do you mean that you sampled $\delta$ from a distribution?}

    We no longer include the simulations based on sampling, so the Dirichlet distribution is not used in this draft.  To clarify on the use of market shares (which we do still use to calibrate the Republic/Santek exercise), we agree that they are endogenous to the model but they are also the type of data that practitioners and researchers would observe about a market.  Therefore, shares are commonly used for calibration.  It is helpful to draw the analogy from the structural estimation context: in recovering the parameters for a structural demand system, an econometrician usually uses variation in market shares and prices (often along with other data for instrumental variables) in order to estimate the model, even though the shares and prices are also functions of the model.  For antitrust practitioners calibrating a model, they also use shares and prices.  In both cases, we are trying to find parameters of the model that are consistent with the market outcomes that we observe in the data.  We have provided some of this background in our calibration discussion in Section 3.2.

    \item \emph{Simulation details. It might make more sense to calibrate to a market elasticity (for example -1) instead of the way it is done in the paper.}

    Since we have removed the simulations that were formerly in Section 3, this part of the calibration is no longer used.  In the Republic/Santek example, we now start by assuming no substitution outside of the waste management market (due to the strict government regulations mandating that commercial customers dispose of waste in proper landfills), and then vary the market elasticity in a sensitivity check that appears in Section 3.3.

    \item  \emph{It might make sense if a retailer and manufacturer had meaning in the model. One suggestion is to have $\delta$ be comprised of two components, a manufacturer effect, and a retailer effect, where the retailer effect is perfectly correlated across manufacturer products within a retailer, and the manufacturer effect is perfectly correlated across retailers for the same product. So something like  $\delta_{rw} = \delta_{r} + \delta_{w}$.}

    We tried something in this spirit in our nested logit sensitivity exercise for the Republic/Santek simulations.  We did one version where we grouped together all the products that use the same firm for collection and disposal (``integrated'' in Figure 4) and another version where we grouped together all products that use the same landfill (``landfill'' in Figure 4).  The discussion of these results is in Section 3.3.

    \item  \emph{In Section 2.2 (first paragraph), referencing a particular retailer with index r is confusing because r is the generic index for a retailer.}

    We have shifted to referring to the merging firms in Section 2.2 using index $t$ for the retailer and $v$ for the wholesaler.

    \item \emph{In Equation 4, I think there is an additional effect in the first term that Luco and Marshall (2020) would call the Edgeworth-Salinger effect.}

    This is a good point.  We have added a footnote to the discussion of that equation (now equation (8)) mentioning the Salinger (1991) result and the Luco and Marshall (2020) paper.

    \item \emph{The authors mention on multiple occasions that a retailer could cease to trade with a wholesaler (eg. page 8). I don't see how this is a possibility in the model.}

    We agree that this is confusing and have removed those statements.  We now say near the beginning of Section 2.1 that we are taking the observed set of retailer-wholesaler combinations offered as given.

    \item  \emph{I don't think ``Vertical" and ``Horizontal" need to be quoted in the title.}

    We have removed the quotation marks from the title.

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% R2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\break
\section{Response to Referee 2}
\emph{I recommend that the paper be shortened in two ways. First I think that it should report the results for the simulations of complex mergers but should omit the results for the simulations of mergers where the number of pre-existing rivals (that do not participate in the merger) changes. That is, I think the authors should only include simulation results for the case when
there are no pre-existing vertically integrated rivals that do not participate in the merger and report that the nature of the results does not change substantially when pre-existing vertically integrated rivals that do not participate in the merger are allowed to exist. In addition to shortening the paper, this will allow the authors to provide a simpler and much less confusing description of the set of examples they consider in their simulations. Second, I think that the material on generalizing the results to the case of increasing marginal costs should be omitted. I think this a somewhat separate issue that deserves separate treatment if substantial enough results can be derived.}

Thank you for this helpful feedback.  We have removed both sets of material and pivoted the draft to focusing on the Republic/Santek merger, which has greatly streamlined the paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% R3
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\break
\section{Response to Referee 3}
\begin{enumerate}
    \item \emph{I could not find a contribution that would merit publication at the Journal of Industrial Economics. The paper is very similar to Sheu and Taragin (2021). The model and simulation are essentially the same. At times, the paper reads as an appendix of the previous article.}

    We agree that we could have been clearer in our last draft about the distinction between this paper and our previous research.  We believe that removing the old Section 3 and changing the focus to the Republic/Santek merger helps address this issue.  In the new Section 3, we show how moving from pure vertical and pure horizontal mergers (as appear in Sheu and Taragin (2021)) to complex mergers where one or both of the firms are already integrated prior to the merger (new for this paper) impacts the results.

    \item \emph{The paper claims to simulate the welfare effects in a new set of mergers called ``complex mergers." Complex mergers are standard vertical and horizontal mergers where a firm is vertically integrated. They have been previously studied by Ho and Lee (2017) and Crawford et al. (2018), among others. (The related literature also needs to be improved.) So, the paper uses a stylized model to simulate standard mergers.}

    We agree that there has been past research on complex mergers, but we also believe that these mergers have been understudied relative to pure vertical and pure horizontal transactions.  Furthermore, we also believe that research showing how a more complicated structural model can be simplified for use by practitioners, keeping in mind the data and time constraints they face, is important.  We have personal experience with being told Crawford et al. (2018) was too complicated to be used as part of the court testimony we were helping to prepare in a merger litigation.

    We have added to the literature review in the Introduction to make clear that there are past papers that have studied complex mergers, including the two you mention above.  We also have added a discussion of earlier theory literature on vertical integration to provide more context for our paper.

    \item \emph{Some of the modeling assumptions are very restrictive. For instance, the logit demand system, simultaneous bargaining-Bertrand game, and the linear fees in the bargaining model. While these assumptions seem reasonable for a policy paper for practitioners, they are too strong for the JINDEC, as they hinder the credibility of the main exercises in the paper. The logit demand, in particular, is unnecessarily restrictive in terms of substitution patterns and could be driving the welfare results. For example, vertically and non-vertically integrated firms are clearly different from the consumers' perspective. One would expect consumers self-selecting into the firms' types, which would affect the compensating variation computation. To the very least, random coefficients on price and on whether the firm is vertically integrated are necessary to obtain more credible results.}

    We have removed the discussion of logit demand from the theory explanation in Section 2, to make clear that it is not a necessary part of the model.  In our merger simulations in Section 3.3, we present both logit and nested logit results, the latter with one option where all the integrated products are in the same nest separate from the non-integrated products (in a similar idea to what you suggest in your comment) and a second option where products that use the same landfill are in the same nest.  We agree that the logit assumption is restrictive, but we also note that it is commonly used in antitrust contexts, as the logit fits with the antitrust emphasis on market shares.  We have added a discussion of these issues in Section 3.2.

    We understand that simultaneous pricing and linear fees may not be appropriate for all applications.  We believe they are useful for some industries, like cable programming, where prices are fixed for periods of time and linear fee contracts are common.  We have added a discussion of these limitations in the Introduction.

    \item \emph{The RRC discussion was perplexing. What the paper calls ``RRC" differs from the classic RRC in the literature in Salop and Scheffman (1983). In the model, it arises due to the greater bargaining power of the integrated firms. But the paper's assumption that ``negotiation for a given input price treats other input prices and all downstream prices as given" (p. 6) rules out the possibility of a standard RRC effect. So the paper contradicts Crawford et al. (2018) footnote 62. Is this an error? It is paramount that this issue is clarified in the future, even at a different journal. Please include the profit functions and derivations in the Appendix (e.g. equation 5 and subsequent).}

    We understand the confusion around this point.  Our meaning of the term agrees with Crawford et al. (2018).  We use the term ``RRC'' in the way that they use it, to mean a shift in bargaining leverage that causes a vertically integrated firm to raise input prices to a downstream rival.  Crawford et al. (2018) refer to this as ``raising rivals' cost.'' See, for example, their discussion in Section 6.2.2, pages 934-936, of their paper.  Footnote 62 in their paper explains that this RRC effect is not motivated by the likelihood that downstream firms will raise their retail prices in response to higher wholesale prices.  The behavior is instead motivated by the possibility that consumers will switch who they purchase from, at fixed prices.  Additional discussion is provided in their footnote 36.  We and Crawford et al. (2018) do not use RRC to mean the incentive modeled by Salop and Scheffman (1983), where a dominant firm chooses an action in order to directly impact downstream retail prices.  We have added footnote 10 in the discussion around equation (9) to clarify this distinction.

    We have included the profit functions in Section 2.1 and added an explanation for how they appear in the firm first order conditions.

    \item \emph{The simulation exercise in Section 3 does not implement the simulation of the model in Section 2. What is the mapping between the two? How do you ensure market shares are consistent with equation 1 and firm pricing decisions? The paper states that: ``we calibrate the price coefficient $\alpha$ assuming that in the pre-merger world, there is a vertically integrated outside option that sets its Bertrand equilibrium price to \$5, has 15\% market share, and has zero marginal costs. All other goods are differenced." But how are prices consistent with the demand and profit maximization computed? Specifically, I could not understand how such a procedure fits the bargaining game upstream. They seem to be inconsistent. For instance, how do you ensure downstream (upstream) prices are consistent with equation 2 (3)? For these reasons, I could not follow the details of the simulation exercise.}

    We have removed those simulations from the paper.  All the simulations and calibration we have done rely on the first order conditions of the model in Section 2, to ensure that the resulting prices and market shares are consistent with the theory.  We have added a discussion clarifying this point in Section 3.2.

    \item \emph{Finally, I found the calibration exercise in Section 4 hard to follow. At times it reads as if the paper is collecting data. But it is closer to a calibration/simulation exercise motivated by the merger. But there is no estimation and the data is minimal. The theoretical model does not fit the setting in the case. First, the bargaining model is not present nor discussed in the DOJ lawsuit. Second, the RRC discussed in the DOJ lawsuit refers to foreclosure/exclusionary conduct concerns. This feature could typically have different welfare implications. See, e.g., Ordover et al. (1990) and the references therein. I found the analysis in this section not credible.}

    We have designed this exercise to closely mirror the types of data and calibration exercises that are frequently used by antitrust agencies and practitioners in merger review.  We have added an explanation of this purpose at the beginning of Section 3.

    The DOJ complaint in Republic/Santek at paragraph 46 explains that the DOJ was alleging harm due to increased upstream disposal prices to rival downstream collection firms, not just due to the possibility of total exclusion or foreclosure of these collection firms.  Specifically, the complaint states that the merged firm would have an ``incentive to raise the MSW disposal costs of independent haulers because Republic---no longer confronting competition from Santek in SCCW collection---would capture more of the business lost by independent haulers in the SCCW collection market.''  We agree that the complaint does not put this story in terms of bargaining, but the incentive discussed in this passage is very similar to the mechanism that leads to increased bargaining leverage and hence higher wholesale prices in our model.  The knowledge that downstream customers have fewer independent rivals from which to purchase collection is what allows the merged firm to negotiate for a higher upstream disposal rate.

\end{enumerate}

\end{document}

### 2024-11-05 presentations/FRB/BargainingSimsSlidesNov2024.tex
Added Gloria's 2024 presentation to the FTC as well as her 2024 presentation to the FRB.

\pdfminorversion=6
\documentclass{beamer}
\usefonttheme{serif}		% font theme
\usepackage{hyperref}   % displays URLs
\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{mathtools,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{setspace}   % controls line spacing
\usepackage{bibentry}   % cite without bibliography
\usepackage{anyfontsize}% font sizing
\usepackage{newcent}		% font

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Theme
\usetheme{Pittsburgh}
\definecolor{mk}{rgb}{.64,.1,.1}
\definecolor{gray}{rgb}{.6,.6,.6}
\usecolortheme{orchid}
\usenavigationsymbolstemplate{}
\useoutertheme{infolines}
\setbeamertemplate{footline}{
\hspace*{.5cm}
\scriptsize{\vspace*{2pt}
\insertshortauthor \hfill \insertshorttitle\hfill\insertframenumber\hspace*{.5cm}	
}
}

\setbeamercolor{footline}{fg=gray}

\setbeamertemplate{frametitle}{
\insertframetitle
\par
}

\setbeamersize{text margin left=.7cm,text margin right=.7cm}
\setbeamertemplate{itemize/enumerate body begin}{\footnotesize}
\setbeamertemplate{itemize/enumerate subbody begin}{\tiny}

\setbeamercolor{alerted text}{fg=mk}

\renewcommand\thempfootnote{\arabic{mpfootnote}} % return footnotes numbers to numerals

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Title Slide Information
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[]{Margaret Loudermilk\inst{1} \and Gloria Sheu\inst{2} \and Charles Taragin\inst{3}}
\institute[]{\inst{1} U.S. Department of Justice \and \inst{2}Federal Reserve Board of Governors \and \inst{3}Federal Reserve Board of Governors \bigbreak  \tiny{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors.  Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice.} }
\title[]{Beyond Horizontal and Vertical:\\ Measuring the Welfare Effects of Complex Integration}
\date{November 2024}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{frame}
\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Motivation}

``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically?'' \\
\scriptsize{(\textit{Request for Information on Merger Enforcement}, Question 1g, DOJ and FTC, 2022)}

\bigbreak \begin{itemize}
	\item Few producers operate at only one level of a supply chain
	\begin{itemize}
			\item Mergers frequently combine firms that are already vertically integrated
			\item Nevertheless, standard merger analysis often simplifies cases into purely horizontal and vertical categories
		\end{itemize}
	\item What happens when we incorporate some of these vertical relationships into a typical merger simulation model?
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Research Goals}

\begin{enumerate}
\item Incorporate additional vertical integration into the Sheu and Taragin (2021) merger simulation model
	\begin{itemize}
		\item Allow some merging parties to already be integrated, resulting in ``complex'' mergers
		\item Model welfare effects for consumers accounting for elimination of double marginalization (EDM) and raising rivals' cost (RRC)
	\end{itemize}
\item Apply model to Republic/Santek merger (2021) in solid waste management
	\begin{itemize}
		\item Compare results to partial models that do not account for vertical aspects
		\item Explore model implications for counterfactual mergers, comparing pure horizontal and pure vertical mergers to complex options
	\end{itemize}
\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Findings}

\begin{itemize}
	\item We simulate Republic/Santek using limited data inputs and generate realistic RRC and EDM effects.  We find net harm to consumers.  
	\item Simulating Republic/Santek as separate upstream and downstream horizontal mergers using a partial model can greatly misrepresent the extent of harm
	\item None of the counterfactual complex mergers we study result in net benefits to consumers
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Related Literature}

\begin{itemize}
	\item Merger simulation papers: Sheu and Taragin (2021); Werden and Froeb (1994); Werden and Froeb (2008)
	\item Bargaining in vertical supply chains: Draganska, Klapper, and Villas-Boas (2010); Crawford, Lee, Whinston, and Yurukoglu (2018); Ho and Lee (2019)
	\item Measuring impacts of vertical mergers: Moresi and Salop (2013); Rogerson (2014); DasVarma and De Stefano (2020); Rogerson (2020); Domnenko and Sibley (2023); De Stefano and Salinger (2024)
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Part I: Theory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
	\frametitle{Outline}
	\small

	\begin{enumerate}

		\item \textbf{Theory}
		\item Republic/Santek Application
		\item Conclusion

	\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Model Overview}

\begin{itemize}
  \item Each wholesaler $w$ in the set $\mathcal{W}$ offers a single product to retailers $r$ in the set $\mathcal{R}$,
  who in turn sell to consumers
  \begin{enumerate}
    \item Upstream: Nash bargaining between wholesalers and retailers
    \item Downstream: Bertrand differentiated products logit competition between retailers (nested logit for sensitivity analysis)
  \end{enumerate}
	\item All prices (upstream and downstream) assumed to be determined simultaneously
	\item Calibration follows Sheu and Taragin (2021)
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Downstream Bertrand Logit}

\begin{itemize}
  \item Assume firms set prices in Bertrand Nash equilibrium
	\item Market share of wholesaler $w$'s product purchased through retailer $r$ 
  \begin{equation*}
  s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{x \in \mathcal{R}} \sum_{y \in \mathcal{W}^x} \exp(\delta_{xy}   - \alpha p_{xy})}
  \end{equation*}
  \item Profit function for retailer $r$ with wholesale fee $p^W_{rx}$ and other costs $c^R_{rx}$
  \begin{equation*}
  \pi^r = \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx} M
  \end{equation*}
  \item First order conditions for downstream equilibrium prices
  \begin{equation*}
  \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} = 0
  \end{equation*}
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Upstream Nash Bargaining}

\begin{itemize}
  \item Assume a linear wholesale price set by Nash Bargaining with $\lambda_w$ bargaining power parameter
  \begin{equation*}
  \max_{p^W_{rw}} (\pi^r - d^r (\mathcal{W}^r \setminus \{w\}))^{\lambda_w} (\pi^w - d^w (\mathcal{R}^w \setminus \{r\}
  ))^{1 - \lambda_w}
  \end{equation*}
  \item Profit function for wholesaler $w$ with marginal cost $c^W_{xw}$ 
  \begin{equation*}
  \pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M
  \end{equation*}
	\item Profit for wholesaler $w$ if it does not sell to retailer $r$ 
  \begin{equation*}
  d^w (\mathcal{R}^w \setminus \{r\}) = \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] s_{xw}(\mathcal{W}^r   \setminus \{w\}) M
  \end{equation*}
  \item Profit for retailer $r$ if it does not buy from wholesaler $w$ 
  \begin{equation*}
  d^r (\mathcal{W}^r \setminus \{w\}) = \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}
  (\mathcal{W}^r \setminus \{w\}) M
  \end{equation*}
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Bargaining Equilibrium}

\begin{itemize}
	\item Given assumption of simultaneous pricing upstream and downstream, first order conditions have the form
	\begin{equation*}
	\begin{split}
	&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
	&\frac{1-\lambda_w}{\lambda_w} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right)
	\end{split}
	\end{equation*}
  \item A larger disagreement payoff increases bargaining leverage
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Integrated Firm Pricing: Retail}
Assume that retailer $t$ merges with wholesaler $v$. Then $t$'s price downstream for product $v$ is determined by

\begin{equation*}
\begin{split}
\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} &+ s_{tv}\\
\underbrace{[p_{tv}-c^R_{tv}-c^W_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}_\text{EDM effect} &+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_\text{upstream UPP effect} = 0.
\end{split}
\end{equation*}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Integrated Firm Pricing: Wholesale Price to Rival Retailer}
When wholesaler $v$ bargains with unaffiliated retailer $j$, the wholesale price is determined by

\tiny
\begin{equation*}
\begin{split}
&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
& - \overbrace{\overbrace{[p_{tv}-c^W_{tv}-c^R_{tv}] \Delta s_{tv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^j \setminus \{v\})}^{\text{RRC effect}} =\\
&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right)
\end{split}
\end{equation*}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Integrated Firm Pricing: Wholesale Price from Rival Wholesaler}
When retailer $t$ bargains with unaffiliated wholesaler $k$, the wholesale price is determined by

\tiny
\begin{equation*}
\begin{split}
&[p^W_{tk} - c^W_{tk}]s_{tk} - \sum_{x \in \mathcal{R}^k \setminus \{t\}} [p^W_{xk} - c^W_{xk}] \Delta s_{xk}(\mathcal{W}^t \setminus \{k\}) = \\
&\frac{1-\lambda_k}{\lambda_k} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
&\left.- \underbrace{[p_{tv} - c^W_{tv} - c^R_{tv}] \Delta s_{tv}(\mathcal{W}^t \setminus \{k\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{wholesale recapture leverage effect}}\right)
\end{split}
\end{equation*}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Part II: Republic/Santek
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
	\frametitle{Outline}
	\small

	\begin{enumerate}

		\item Theory
		\item \textbf{Republic/Santek Application}
		\item Conclusion

	\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Merger Background}

\begin{itemize}
\item Republic Services bought Santek Environmental in 2021
	\begin{itemize}
		\item Both operated in waste collection (downstream) and waste disposal (upstream)
	\end{itemize}
\item Many major competitors also vertically integrated
\item DOJ negotiated a settlement in both collection and disposal for the Chattanooga Area market
	\begin{itemize}
		\item Parties had 73 percent combined share in collection, 82 percent in disposal
		\item Waste Connections only other disposal competitor, also competed in collection along with Waste Management-ADS (WMADS) and a regional company (Regional)
	\end{itemize}
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Market Diagram}

\centering
\includegraphics[width=4in]{output/chattanooga_tree.png}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Data Inputs}

\centering{
\includegraphics[width=4in]{output/siminputs.pdf}}

\begin{itemize}
	\item Data are for 2019
	\item Data sources:
		\begin{itemize}
			\item Volumes/market shares from DOJ Competitive Impact Statement and Tennessee Department of Environmental Quality reports
			\item Prices from Waste Business Journal
			\item Margins from firm financial statements
		\end{itemize}
	\item Rely on public sources that may differ from information DOJ had access to
	\item Assume no substitution outside of market, due to environmental regulations
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Results}

\centering
\only<1>{\includegraphics[height=2.8in]{output/TrashSimsFirm.png}}
\only<2>{\includegraphics[height=2.8in]{output/TrashSimsFirm1.png}}
\only<3>{\includegraphics[height=2.8in]{output/TrashSimsFirm2.png}}
\only<4>{\includegraphics[height=2.8in]{output/TrashSimsFirm3.png}}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Results}

\begin{itemize}
	\item \$16.6 million annual harm to consumers
	\item Average upstream prices decrease about 3\% (EDM)
	\item Average downstream prices increase by 12.5\% (incomplete pass-through)
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Comparing to Partial Models}

\begin{itemize}
	\item We compare our results from the full vertical model to two alternatives:
		\begin{enumerate}
			\item Downstream-Only: hold prices of upstream products fixed, simulate merger in downstream logit model
			\item Upstream-Only: hold prices of downstream products fixed, simulate merger in upstream bargaining model
		\end{enumerate}
	\item Examine differences in prices and welfare
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Partial Model Results}
\centering
\only<1>{\includegraphics[height=2.8in]{output/TrashSimsCompareAll.png}}
\only<2>{\includegraphics[height=2.8in]{output/TrashSimsCompareAll1.png}}
\only<3>{\includegraphics[height=2.8in]{output/TrashSimsCompareAll2.png}}
\only<4>{\includegraphics[height=2.8in]{output/TrashSimsCompareAll3.png}}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Partial Model Results}

\begin{itemize}
	\item Downstream-Only 
		\begin{itemize}
			\item \$10 million annual harm to consumers
			\item Collection prices for Santek/WM-ADS, Santek/Regional too low (misses RRC)
			\item Collection price for Santek/Republic somewhat too high (misses EDM)
		\end{itemize}
	\item Upstream-Only
		\begin{itemize}
			\item Average upstream prices increase by about 11\% 
			\item Disposal prices for Santek/WM-ADS, Santek/Regional too high (overstates RRC)
		\end{itemize}
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Comparing to Counterfactual Mergers}

\begin{itemize}
	\item We compare welfare effects of Republic/Santek to three other merger options:
		\begin{enumerate}
			\item Santek/Waste Connections: two fully integrated firms that are not capacity constrained
			\item Santek/WMADS: integrated firm with a collection-only firm
			\item WMADS/Regional: horizontal merger between two collection-only firms
		\end{enumerate}
	\item Allows us to examine examples of different complex mergers
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Results for Other Mergers}

\centering
\includegraphics[height=2.8in]{output/trashinterestingmergerbar.png}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Results for Other Mergers}

\centering
\only<1>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar.png}}
\only<2>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar1.png}}
\only<3>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar2.png}}
\only<4>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar3.png}}
\only<5>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar4.png}}
\only<6>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar5.png}}
\only<7>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar6.png}}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Comparing to Mergers of Counterfactual Firms}

\begin{itemize}
	\item Suppose we split the upstream and downstream components of Republic and Santek into independent entities
	\item Simulate a new set of counterfactual mergers:
		\begin{enumerate}
			\item Non-complex: pure downstream horizontal, pure upstream horizontal, pure vertical
			\item Complex with one merging firm being integrated: integrated Santek with Republic collection, integrated Santek with Republic disposal
		\end{enumerate}
	\item Allows us to study complex mergers that involve the most popular products (Republic-Republic and Santek-Santek)
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Results for Other Republic/Santek Mergers}

\centering
\only<1>{\includegraphics[height=2.5in]{output/trashinterestingsantrep.png}}
\only<2>{\includegraphics[height=2.5in]{output/trashinterestingsantrep1.png}}
\only<3>{\includegraphics[height=2.5in]{output/trashinterestingsantrep2.png}}
\only<4>{\includegraphics[height=2.5in]{output/trashinterestingsantrep3.png}}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Results for Counterfactual Mergers}

\begin{itemize}
	\item Republic/Santek as proposed appears to have been one of the most harmful possiblities, due to the popularity of the Republic-Republic and Santek-Santek products
	\item Any gains from EDM are outweighed by RRC and loss of horizontal competition.  
		\begin{itemize}
			\item Despite EDM decreases in upstream prices of over 60\%
			\item Pure vertical merger between Santek disposal and Republic collection is the exception
		\end{itemize}
	\item The complex options tend to result in more consumer harm
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Part III: Conclusion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
	\frametitle{Outline}
	\small

	\begin{enumerate}

		\item Theory
		\item Republic/Santek Application
		\item \textbf{Conclusion}

	\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Summary}

\begin{itemize}
\item We study the impacts of additional vertical integration on the welfare effects of mergers 
\item Model shows the effects of EDM and RRC in Republic/Santek
	\begin{itemize}
		\item Partial model can understate or overstate effects
		\item Despite substantial input price decreases from EDM, hard to find instances where consumers gain on net
	\end{itemize}
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}

\begin{center} Thank You!  \end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Sensitivity Analysis}

We consider the way our results change in three sensitivity exercises
\begin{enumerate}
	\item Extent of substitution outside the market: vary share of outside good/market elasticity
	\item Assumed demand substitution: switch to nested logit and vary nesting parameter
		\begin{itemize}
			\item Option 1: nest products by integrated vs. not
			\item Option 2: nest products by disposal landfill provider
		\end{itemize}
	\item Bargaining power: vary $\lambda$
		\begin{itemize}
			\item Baseline calibration results in $\lambda_{Santek}=1$ and $\lambda_{WasteConn}=0.73$, so retailers have substantial advantage 
		\end{itemize}
\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Consumer Harm and Outside Good Share}

\centering
\includegraphics[height=2.8in]{output/trashelast.png}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Consumer Harm and Nested Logit}

\centering
\includegraphics[height=2.8in]{output/trashnest.png}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Consumer Harm and Bargaining Power}

\centering
\includegraphics[height=2.8in]{output/trashbarg.png}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Sensitivity Analysis Results}

\begin{enumerate}
	\item Consumer harm falls as substitution outside market increases.  Harm still above 5\% when 	market is unit elastic.
	\item Consumer harm rises as preferences for integrated products increases, falls as preferences for landfills increases.  Harm still above 5\% when landfill nesting parameter is as high as 0.6.
	\item Consumer harm falls as retailer bargaining power decreases.  Harm similar to baseline when bargaining power is as low as 0.4.  
\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-11-05 presentations/FTC/2024/BargainingSimsSlidesOct2023.tex
Added Gloria's 2024 presentation to the FTC as well as her 2024 presentation to the FRB.

--- 2024-11-05 presentations/FRB/BargainingSimsSlidesNov2024.tex
@@
 Department of Justice.} }
-\title[]{Beyond Horizontal and Vertical:\\ Measuring the Welfare Effects of Complex Integration}
-\date{November 2024}
+\title[]{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration}
+\date{October 2023}
 
@@
 	\begin{itemize}
-			\item Mergers frequently combine firms that are already vertically integrated
-			\item Nevertheless, standard merger analysis often simplifies cases into purely horizontal and vertical categories
+			\item Mergers frequently combine firms and impact rivals that are already vertically integrated
+			\item Nevertheless, standard merger analysis often simplifies a case into purely horizontal and vertical categories
 		\end{itemize}
@@
 	\begin{itemize}
-		\item Allow some merging parties to already be integrated, resulting in ``complex'' mergers
-		\item Model welfare effects for consumers accounting for elimination of double marginalization (EDM) and raising rivals' cost (RRC)
+		\item Allow some merging parties to already be integrated
+		\item Allow some rivals to be integrated 
+		\item Calculate welfare effects balancing UPP, EDM, RRC
 	\end{itemize}
+\item Study welfare effects of four types of mergers
+	\begin{enumerate}
+		\item Vertical: merger between unintegrated upstream and unintegrated downstream firms
+		\item Upstream: merger between an unintegrated upstream firm and an upstream firm that may already be integrated with a downstream firm 
+		\item Downstream: merger between an unintegrated downstream firm and a downstream firm that may already be integrated with a upstream firm 
+		\item Integrated: merger between two vertically integrated firms 
+	\end{enumerate}
 \item Apply model to Republic/Santek merger (2021) in solid waste management
-	\begin{itemize}
-		\item Compare results to partial models that do not account for vertical aspects
-		\item Explore model implications for counterfactual mergers, comparing pure horizontal and pure vertical mergers to complex options
-	\end{itemize}
 \end{enumerate}
@@
 \begin{itemize}
-	\item We simulate Republic/Santek using limited data inputs and generate realistic RRC and EDM effects.  We find net harm to consumers.  
-	\item Simulating Republic/Santek as separate upstream and downstream horizontal mergers using a partial model can greatly misrepresent the extent of harm
-	\item None of the counterfactual complex mergers we study result in net benefits to consumers
+	\item Vertical effects in ``horizontal'' mergers (both upstream and downstream) create net consumer benefits only in limited cases
+		\begin{itemize}
+			\item Mostly when upstream firms have more bargaining power (when EDM is expected to be larger)
+		\end{itemize}
+	\item Mergers between two integrated firms are typically net harmful to consumers 
+		\begin{itemize}
+			\item Only beneficial when upstream firms have a significant edge in bargaining power
+		\end{itemize}
+	\item The presence of integrated rivals does not have a large impact on realized harm 
+		\begin{itemize}
+			\item Can somewhat shift outcomes towards less harm, but differences not dramatic
+		\end{itemize}
+	\item We can model Republic/Santek using limited inputs and generate realistic RRC and EDM effects
 \end{itemize}
@@
 	\item Bargaining in vertical supply chains: Draganska, Klapper, and Villas-Boas (2010); Crawford, Lee, Whinston, and Yurukoglu (2018); Ho and Lee (2019)
-	\item Measuring impacts of vertical mergers: Moresi and Salop (2013); Rogerson (2014); DasVarma and De Stefano (2020); Rogerson (2020); Domnenko and Sibley (2023); De Stefano and Salinger (2024)
+	\item Measuring impacts of vertical mergers: Moresi and Salop (2013); Rogerson (2014); DasVarma and De Stefano (2020); Rogerson (2020); Domnenko and Sibley (2022)
 \end{itemize}
@@
 		\item \textbf{Theory}
+		\item Numerical Simulations
 		\item Republic/Santek Application
@@
 \begin{itemize}
-  \item Each wholesaler $w$ in the set $\mathcal{W}$ offers a single product to retailers $r$ in the set $\mathcal{R}$,
+  \item Each wholesaler $w$ in the set $\mathbb{W}$ offers a single product to retailers $r$ in the set $\mathbb{R}$,
   who in turn sell to consumers
@@
     \item Upstream: Nash bargaining between wholesalers and retailers
-    \item Downstream: Bertrand differentiated products logit competition between retailers (nested logit for sensitivity analysis)
+    \item Downstream: Bertrand differentiated products logit competition between retailers
   \end{enumerate}
@@
   \begin{equation*}
-  s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{x \in \mathcal{R}} \sum_{y \in \mathcal{W}^x} \exp(\delta_{xy}   - \alpha p_{xy})}
+  s_{rw} = \frac{\exp(\delta_{rw} - \alpha p_{rw})}{1 + \sum_{t \in \mathbb{R}} \sum_{x \in \mathbb{W}^t} \exp(\delta_{tx}   - \alpha p_{tx})}
   \end{equation*}
-  \item Profit function for retailer $r$ with wholesale fee $p^W_{rx}$ and other costs $c^R_{rx}$
+  \item Profit function for retailer $r$ with wholesale fee $p^W_{rw}$ and other costs $c^R_{rw}$
   \begin{equation*}
-  \pi^r = \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx} M
+  \pi^r = \sum_{w \in \mathbb{W}^r} [p_{rw} - p^W_{rw} - c^R_{rw}] s_{rw} M
   \end{equation*}
@@
   \begin{equation*}
-  \sum_{x \in \mathcal{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} = 0
+  \sum_{x \in \mathbb{W}^r} [p_{rx} - p^W_{rx} - c^R_{rx}] \frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} = 0
   \end{equation*}
@@
 \begin{itemize}
-  \item Assume a linear wholesale price set by Nash Bargaining with $\lambda_w$ bargaining power parameter
+  \item Assume a linear wholesale price set by Nash Bargaining with $\lambda$ bargaining power parameter
   \begin{equation*}
-  \max_{p^W_{rw}} (\pi^r - d^r (\mathcal{W}^r \setminus \{w\}))^{\lambda_w} (\pi^w - d^w (\mathcal{R}^w \setminus \{r\}
-  ))^{1 - \lambda_w}
+  \max_{p^W_{rw}} (\pi^r - d^r (\mathbb{W}^r \setminus \{w\}))^{\lambda} (\pi^w - d^w (\mathbb{R}^w \setminus \{r\}
+  ))^{1 - \lambda}
   \end{equation*}
-  \item Profit function for wholesaler $w$ with marginal cost $c^W_{xw}$ 
+  \item Profit function for wholesaler $w$ with marginal cost $c^W_{rw}$ 
   \begin{equation*}
-  \pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M
+  \pi^w = \sum_{r \in \mathbb{R}^w} [p^W_{rw} - c^W_{rw}] s_{rw} M
   \end{equation*}
@@
   \begin{equation*}
-  d^w (\mathcal{R}^w \setminus \{r\}) = \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] s_{xw}(\mathcal{W}^r   \setminus \{w\}) M
+  d^w (\mathbb{R}^w \setminus \{r\}) = \sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] s_{tw}(\mathbb{W}^r   \setminus \{w\}) M
   \end{equation*}
@@
   \begin{equation*}
-  d^r (\mathcal{W}^r \setminus \{w\}) = \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}
-  (\mathcal{W}^r \setminus \{w\}) M
+  d^r (\mathbb{W}^r \setminus \{w\}) = \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}
+  (\mathbb{W}^r \setminus \{w\}) M
   \end{equation*}
@@
 	\begin{split}
-	&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
-	&\frac{1-\lambda_w}{\lambda_w} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right)
+	&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
+	&\frac{1-\lambda}{\lambda} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right)
 	\end{split}
@@
 \frametitle{Integrated Firm Pricing: Retail}
-Assume that retailer $t$ merges with wholesaler $v$. Then $t$'s price downstream for product $v$ is determined by
+Assume that retailer $r$ merges with wholesaler $w$. Then $r$'s price downstream for product $w$ is determined by
 
@@
 \begin{split}
-\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} &+ s_{tv}\\
-\underbrace{[p_{tv}-c^R_{tv}-c^W_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}_\text{EDM effect} &+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_\text{upstream UPP effect} = 0.
+\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} &+ s_{rw}\\
+\underbrace{[p_{rw}-c^R_{rw}-c^W_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}_\text{EDM effect} &+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_\text{upstream UPP effect} = 0.
 \end{split}
@@
 \frametitle{Integrated Firm Pricing: Wholesale Price to Rival Retailer}
-When wholesaler $v$ bargains with unaffiliated retailer $j$, the wholesale price is determined by
+When wholesaler $w$ bargains with unaffiliated retailer $s$, the wholesale price is determined by
 
@@
 \begin{split}
-&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
-& - \overbrace{\overbrace{[p_{tv}-c^W_{tv}-c^R_{tv}] \Delta s_{tv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^j \setminus \{v\})}^{\text{RRC effect}} =\\
-&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right)
+&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
+& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}} =\\
+&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right)
 \end{split}
@@
 \frametitle{Integrated Firm Pricing: Wholesale Price from Rival Wholesaler}
-When retailer $t$ bargains with unaffiliated wholesaler $k$, the wholesale price is determined by
+When retailer $r$ bargains with unaffiliated wholesaler $v$, the wholesale price is determined by
 
@@
 \begin{split}
-&[p^W_{tk} - c^W_{tk}]s_{tk} - \sum_{x \in \mathcal{R}^k \setminus \{t\}} [p^W_{xk} - c^W_{xk}] \Delta s_{xk}(\mathcal{W}^t \setminus \{k\}) = \\
-&\frac{1-\lambda_k}{\lambda_k} \left([p_{tk} - p^W_{tk} - c^R_{tk}]s_{tk} - \sum_{x \in \mathcal{W}^t \setminus \{v,k\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^t \setminus \{k\})\right.\\
-&\left.- \underbrace{[p_{tv} - c^W_{tv} - c^R_{tv}] \Delta s_{tv}(\mathcal{W}^t \setminus \{k\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^t \setminus \{k\})}_{\text{wholesale recapture leverage effect}}\right)
+&[p^W_{rv} - c^W_{rv}]s_{rv} - \sum_{t \in \mathbb{R}^v \setminus \{r\}} [p^W_{tv} - c^W_{tv}] \Delta s_{tv}(\mathbb{W}^r \setminus \{v\}) = \\
+&\frac{1-\lambda}{\lambda} \left([p_{rv} - p^W_{rv} - c^R_{rv}]s_{rv} - \sum_{x \in \mathbb{W}^r \setminus \{w,v\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^r \setminus \{v\})\right.\\
+&\left.- \underbrace{[p_{rw} - c^W_{rw} - c^R_{rw}] \Delta s_{rw}(\mathbb{W}^r \setminus \{v\})}_{\text{EDM recapture effect}}-\underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^r \setminus \{v\})}_{\text{wholesale recapture leverage effect}}\right)
 \end{split}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% Part II: Republic/Santek
+%%% Part II: Numerical Simulations
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 		\item Theory
+		\item \textbf{Numerical Simulations}
+		\item Republic/Santek Application
+		\item Conclusion
+
+	\end{enumerate}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\frametitle{Data Generating Process}
+
+\begin{itemize}
+\item Simulate markets with 2, 3, 4, or 5 wholesalers/retailers
+\item Have 0 to 4 non-siloed integrated incumbents pre-merger
+	\begin{itemize}
+		\item 2 to 6 for integrated mergers
+	\end{itemize}
+\item Bargaining power parameter ranges from 0.1 to 0.9 in 0.1 increments
+\item Eliminate some instances
+	\begin{itemize}
+		\item Unprofitable mergers
+		\item Downstream markets that do not pass the Hypothetical Monopolist Test
+	\end{itemize}
+\item Results in approximately 1.6 million markets
+\end{itemize}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\frametitle{Include Four Types of Mergers}
+
+\begin{enumerate}
+	\item Vertical: merger between unintegrated upstream and unintegrated downstream firms
+	\item Upstream: merger between an unintegrated upstream firm and an upstream firm that may already be integrated with a downstream firm 
+	\item Downstream: merger between an unintegrated downstream firm and a downstream firm that may already be integrated with a upstream firm 
+	\item Integrated: merger between two vertically integrated firms 
+\end{enumerate}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\frametitle{Summary Statistics}
+
+\centering
+\includegraphics[width=4in,height=2.8in]{output/summarystats.pdf}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+
+\centering
+\only<1>{\includegraphics[width=4in]{output/surplussum.png}}
+\only<2>{\includegraphics[width=4in]{output/surplussum1.png}}
+\only<3>{\includegraphics[width=4in]{output/surplussum2.png}}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\centering
+\only<1>{\includegraphics[width=4in,height=3in]{output/CVvertincumbBW_consumer.png}}
+\only<2>{\includegraphics[width=4in,height=3in]{output/CVvertincumbBW_consumer1.png}}
+\only<3>{\includegraphics[width=4in,height=3in]{output/CVvertincumbBW_consumer2.png}}
+\only<4>{\includegraphics[width=4in,height=3in]{output/CVvertincumbBW_consumer3.png}}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\centering
+\only<1>{\includegraphics[width=4in,height=3in]{output/CVbargincumbent_consumer.png}}
+\only<2>{\includegraphics[width=4in,height=3in]{output/CVbargincumbent_consumer1.png}}
+\only<3>{\includegraphics[width=4in,height=3in]{output/CVbargincumbent_consumer2.png}}
+\only<4>{\includegraphics[width=4in,height=3in]{output/CVbargincumbent_consumer3.png}}
+\only<5>{\includegraphics[width=4in,height=3in]{output/CVbargincumbent_consumer4.png}}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% Part III: Republic/Santek
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+	\frametitle{Outline}
+	\small
+
+	\begin{enumerate}
+
+		\item Theory
+		\item Numerical Simulations
 		\item \textbf{Republic/Santek Application}
@@
 		\item Parties had 73 percent combined share in collection, 82 percent in disposal
-		\item Waste Connections only other disposal competitor, also competed in collection along with Waste Management-ADS (WMADS) and a regional company (Regional)
+		\item Waste Connections only other disposal competitor, also competed in collection along with Waste Management-ADS and a regional company
 	\end{itemize}
@@
 \begin{itemize}
-	\item Data are for 2019
-	\item Data sources:
+	\item Data Sources
 		\begin{itemize}
@@
 			\item Prices from Waste Business Journal
-			\item Margins from firm financial statements
+			\item Costs from firm financial statements
 		\end{itemize}
 	\item Rely on public sources that may differ from information DOJ had access to
-	\item Assume no substitution outside of market, due to environmental regulations
 \end{itemize}
@@
 \only<3>{\includegraphics[height=2.8in]{output/TrashSimsFirm2.png}}
-\only<4>{\includegraphics[height=2.8in]{output/TrashSimsFirm3.png}}
 
@@
 \begin{itemize}
-	\item \$16.6 million annual harm to consumers
+	\item \$16 million annual harm to consumers
 	\item Average upstream prices decrease about 3\% (EDM)
-	\item Average downstream prices increase by 12.5\% (incomplete pass-through)
+	\item Average downstream prices increase by about 12\% (incomplete pass-through)
 \end{itemize}
@@
 		\begin{itemize}
-			\item Average upstream prices increase by about 11\% 
+			\item Total welfare unchanged as effect is a transfer between firms
+			\item Average upstream prices increase by about 8\% 
 			\item Disposal prices for Santek/WM-ADS, Santek/Regional too high (overstates RRC)
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Comparing to Counterfactual Mergers}
-
-\begin{itemize}
-	\item We compare welfare effects of Republic/Santek to three other merger options:
-		\begin{enumerate}
-			\item Santek/Waste Connections: two fully integrated firms that are not capacity constrained
-			\item Santek/WMADS: integrated firm with a collection-only firm
-			\item WMADS/Regional: horizontal merger between two collection-only firms
-		\end{enumerate}
-	\item Allows us to examine examples of different complex mergers
-\end{itemize}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Results for Other Mergers}
-
-\centering
-\includegraphics[height=2.8in]{output/trashinterestingmergerbar.png}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Results for Other Mergers}
-
-\centering
-\only<1>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar.png}}
-\only<2>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar1.png}}
-\only<3>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar2.png}}
-\only<4>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar3.png}}
-\only<5>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar4.png}}
-\only<6>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar5.png}}
-\only<7>{\includegraphics[height=2.8in]{output/trashinterestingfirmbar6.png}}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Comparing to Mergers of Counterfactual Firms}
-
-\begin{itemize}
-	\item Suppose we split the upstream and downstream components of Republic and Santek into independent entities
-	\item Simulate a new set of counterfactual mergers:
-		\begin{enumerate}
-			\item Non-complex: pure downstream horizontal, pure upstream horizontal, pure vertical
-			\item Complex with one merging firm being integrated: integrated Santek with Republic collection, integrated Santek with Republic disposal
-		\end{enumerate}
-	\item Allows us to study complex mergers that involve the most popular products (Republic-Republic and Santek-Santek)
-\end{itemize}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Results for Other Republic/Santek Mergers}
-
-\centering
-\only<1>{\includegraphics[height=2.5in]{output/trashinterestingsantrep.png}}
-\only<2>{\includegraphics[height=2.5in]{output/trashinterestingsantrep1.png}}
-\only<3>{\includegraphics[height=2.5in]{output/trashinterestingsantrep2.png}}
-\only<4>{\includegraphics[height=2.5in]{output/trashinterestingsantrep3.png}}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Results for Counterfactual Mergers}
-
-\begin{itemize}
-	\item Republic/Santek as proposed appears to have been one of the most harmful possiblities, due to the popularity of the Republic-Republic and Santek-Santek products
-	\item Any gains from EDM are outweighed by RRC and loss of horizontal competition.  
-		\begin{itemize}
-			\item Despite EDM decreases in upstream prices of over 60\%
-			\item Pure vertical merger between Santek disposal and Republic collection is the exception
-		\end{itemize}
-	\item The complex options tend to result in more consumer harm
-\end{itemize}
-
-\end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% Part III: Conclusion
+%%% Part IV: Conclusion
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 		\item Theory
+		\item Numerical Simulations
 		\item Republic/Santek Application
@@
 \item We study the impacts of additional vertical integration on the welfare effects of mergers 
+	\begin{itemize}
+		\item Allow merging parties to already be integrated
+		\item Allow rivals to be integrated
+	\end{itemize}
+\item Most merger simulations show net negative impacts on consumers
+	\begin{itemize}
+		\item Typically only result in net gains when upstream bargaining power is high (more potential for EDM)
+		\item Presence of integrated rivals does not strongly protect consumers from harm 
+	\end{itemize}
 \item Model shows the effects of EDM and RRC in Republic/Santek
+\item Appendix to paper extends to case of increasing marginal costs
 	\begin{itemize}
-		\item Partial model can understate or overstate effects
-		\item Despite substantial input price decreases from EDM, hard to find instances where consumers gain on net
+		\item Allow marginal costs to increase linearly
+		\item Merger simulations tend to show more harm than under constant marginal costs
 	\end{itemize}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Sensitivity Analysis}
-
-We consider the way our results change in three sensitivity exercises
-\begin{enumerate}
-	\item Extent of substitution outside the market: vary share of outside good/market elasticity
-	\item Assumed demand substitution: switch to nested logit and vary nesting parameter
-		\begin{itemize}
-			\item Option 1: nest products by integrated vs. not
-			\item Option 2: nest products by disposal landfill provider
-		\end{itemize}
-	\item Bargaining power: vary $\lambda$
-		\begin{itemize}
-			\item Baseline calibration results in $\lambda_{Santek}=1$ and $\lambda_{WasteConn}=0.73$, so retailers have substantial advantage 
-		\end{itemize}
-\end{enumerate}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Consumer Harm and Outside Good Share}
-
-\centering
-\includegraphics[height=2.8in]{output/trashelast.png}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Consumer Harm and Nested Logit}
-
-\centering
-\includegraphics[height=2.8in]{output/trashnest.png}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Consumer Harm and Bargaining Power}
-
-\centering
-\includegraphics[height=2.8in]{output/trashbarg.png}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Sensitivity Analysis Results}
-
-\begin{enumerate}
-	\item Consumer harm falls as substitution outside market increases.  Harm still above 5\% when 	market is unit elastic.
-	\item Consumer harm rises as preferences for integrated products increases, falls as preferences for landfills increases.  Harm still above 5\% when landfill nesting parameter is as high as 0.6.
-	\item Consumer harm falls as retailer bargaining power decreases.  Harm similar to baseline when bargaining power is as low as 0.4.  
-\end{enumerate}
-
-\end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
